### 2020-06-15 article.Rmd
Add the main article

---
title: "A Closer Look at Fixed-Effects Regression in Structural Equation Modeling Using lavaan"
author: "Henrik Kenneth Andersen"
abstract: "This article provides an in-depth look at fixed-effects regression in the structural equation modeling (SEM) framework, specifically the application of fixed-effects in the `lavaan` package for `R`. It is meant as a applied guide for researchers, covering the underlying model specification, syntax, and summary output. It further discusses various common extentsions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of fixed-/random-effects hybrid model. \\par \\par \\textbf{Keywords:} Fixed-effects, structural equation modeling, lavaan, R, panel analysis" 
date: "`r Sys.Date()`"
output: 
  pdf_document:
    fig_caption: yes
    number_sections: yes
    includes: 
      in_header: header.tex
documentclass: article
classoption: a4paper
bibliography: references.bib   
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# ----- Packages

library( formatR)
library( knitr)

# ----- Script hook for printing only certain lines

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(if (abs(lines[1])>1) more else NULL, 
            x[lines], 
            if (length(x)>lines[abs(length(lines))]) more else NULL
           )
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
})

# From: https://community.rstudio.com/t/showing-only-the-first-few-lines-of-the-results-of-a-code-chunk/6963/2
# Retrieved on: 26.05.2020

# ----- Script hook to increase space between text and chunk output

hook_output_def = knitr::knit_hooks$get('output')
knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(options$vspaceout)) {
    end <- paste0("\\vspace{", options$vspaceout, "}")
    stringr::str_c(hook_output_def(x, options), end)
  } else {
    hook_output_def(x, options)
  }
})

hook_source_def = knitr::knit_hooks$get('source')
knitr::knit_hooks$set(source = function(x, options) {
  if (!is.null(options$vspaceecho)) {
    begin <- paste0("\\vspace{", options$vspaceecho, "}")
    stringr::str_c(begin, hook_source_def(x, options))
  } else {
    hook_source_def(x, options)
  }
})

# From: https://stackoverflow.com/questions/43379224/add-vertical-space-above-and-below-figures-code-chunks
# Retrieved on: 04.06.2020
```

\newpage

\pagenumbering{arabic} 
\setcounter{page}{1}

# Introduction {#intro}

Several years ago, @Curran2011 reflected positively on the growing use of panel studies in empirical social research. Some of the strengths of panel data are well-known, e.g., the ability to establish temporal precedence, increased statistical power and the reduction of potential alternative models. However, the perhaps greatest strength of panel data is that they allow for a more rigorous testing of substantive theories as a tool for \textit{causal analysis} [@Greene2012]. Panel data, i.e., repeated measures of the same observed units (people, schools, firms, countries, etc.), allow researchers to decompose the error term into a part that stays constant within units and the part that changes over time. This is the essence of fixed-effects (FE) regression. It drastically reduces the number of potential confounders of the relationship between variables of interest by allowing one to control for all time-invariant influences. Thus, by using each unit at an earlier point in time as its own control, and by examining solely \textit{within-unit changes}, we come closer to the \textit{counterfactual} or \textit{potential outcomes} ideal: what would the outcome have been, had the observational unit not been exposed to the treatment [@Angrist2009]?

Structural equation modeling (SEM) is a popular regression framework. One of its main strengths is its flexibility. Not only can complex causal structures with multiple dependent variables be tested simultaneously, but in longitudinal (and, more generally, hierarchical) studies both time-varying and invariant predictors can be included, and effects can easily be allowed to vary over time. Thus researchers can allow for and study effects that increase or fade over time, or that appear only in specific periods.  Beyond that, with the use of latent variables, SEM provides a way to deal with measurement error and get closer to the true underlying constructs of interest.

This article is intended as a guide for researchers looking for in-depth help with specifying FE models in SEM. It focuses on the `lavaan` [@R-lavaan] package for `R` [@R-base]. While `Mplus` [@Mplus] is unarguably the most robust SEM software currently available (in terms of features like alignment, latent variable interactions, for example) , the `lavaan` package has many benefits. First, it and `R` are open source and completely free. For researchers dipping their toes into SEM, there is no financial barrier to try, and no risk if they decide it is not for them. Second, the implementation of `lavaan` in the larger `R` environment is an enormous advantage. Instead of poring over reams of plain text, copying out coefficients by hand, every part of the `lavaan` output is available as an object. This means that all aspects of the fitted `lavaan` object, from fit indices, to coefficients and standard errors, to the model matrices can be accessed and easily integrated into tables and plots. Furthermore, `R` can be used for a great deal of applications. It can be used to manage and manipulate as well as simulate data, perform symbolic algebra, run more traditional analyses (e.g., multiple regression, logistic regression, principal component analysis), etc. Once one is comfortable using `R`, there is no longer any need to switch between different software for data preparation and analysis.

The following article outlines the basic idea of FE regression, the particularities of FE in SEM, and shows its implementation in `lavaan`. Using simulated data (found in the Appendix), it demonstrates and annotates the code for the most basic FE model and provides an overview of the summary output. After that, a number of potential extensions are discussed and demonstrated, from loosening assumptions, to dealing with measurement error, to specifying hybrid fixed- and random effects models in order to include time-invariant predictors.

# Fixed-effects {#fe}

The most basic fixed-effects model can be expressed as
\begin{align}
y_{it} & = \bm{x_{it}}\bm{\beta} + \alpha_{i} + \varepsilon_{it}, \label{fe}
\end{align}
where $y_{it}$ is the observed outcome variable for unit $i$ at time $t$, $\bm{x_{it}}$ is a $1 \times k$ vector of covariates linked to the outcome by the $k \times 1$ vector of coefficients $\bm{\beta}$. All stable, unit-specific characteristics such as date of birth, country of origin, sex, etc. are captured by the time-constant part of the error $\alpha_{i}$, whereas $\varepsilon_{it}$ is the idiosyncratic error term that varies between units and over time [@Bruederl2015].

The point of FE regression is to be able to drop assumptions regarding the relatedness of the model covariates and the unobserved time-constant effects, i.e., $\E[\bm{x_{it}}^{\intercal}\alpha_{i}]$. We can do so by manipulating Equation \eqref{fe} so that the individual effect, $\alpha_{i}$ is eliminated. In doing so, we no longer need to worry about potential bias due to unit-specific characteristics. This can be achieved in a number of ways (e.g., differencing, least squares dummy variable regression), but the most common method involves subtracting the over-time unit mean from each of the terms: 
\begin{align}
(y_{it} - \bar{y}_{i}) & = (\bm{x_{it}} - \bar{\bm{x}_{i}})\bm{\beta} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
\ddot{y}_{it} & = \bm{\ddot{x}_{it}}\bm{\beta} + \ddot{\varepsilon}_{it}.
\end{align}
Because the average of something that does not change is that thing itself, the individual effects get wiped out by the demeaning. This means that no assumptions about the relatedness of the model covariates and the unit-specific portion of the error are needed. The unbiasedness of the estimate is related solely to the strict exogeneity assumption imposed on the indiosyncratic errors, i.e., $\E[\ddot{\varepsilon}_{it}|\bm{\ddot{x}_{it}}] = 0$ which also implies $\E[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T$ [@Bruederl2015; @Wooldridge2002]. Strict exogeneity not only rules out contemporaneous correlations between the covariates and the error, but it also implies that for the estimates $\bm{\beta}$ to be unbiased, the covariates at all timepoints must not be correlated with the error at all timepoints.

## Fixed-effects in structural equation modeling

Analytically, Equation \eqref{fe} applies as well to FE models in SEM as we will see shortly. However, there are a number of differences we must address. First, SEM is a covariance-based approach. This means that the vectors of individual observations are only of interest insomuch as they can be converted into observed covariances. Second, the way in which we use a latent variable to represent the individual effects forces us to convert the data from stacked, long-format vectors of length $NT$ to $T$ wide-format vectors of length $N$. Lastly, the notation in SEM differs from traditional least squares-based models. There are a number of different model notations (see, for example @Bollen1989 for an overview), but the one that will serve us best is one that was proposed by @Graff1979:
\begin{align}
\bm{y}^{+} & = \bm{\Lambda_{y}}^{+} \bm{\eta}^{+}, \\
\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+}, 
\end{align}
where $\bm{\eta}^{+} = (\bm{y}, \bm{x}, \bm{\eta}, \bm{\xi})^{\intercal}$, $\bm{\zeta}^{+} = (\bm{\varepsilon}, \bm{\delta}, \bm{\zeta}, \bm{\xi})^{\intercal}$, $\bm{y}^{+} = (\bm{y}, \bm{x})^{\intercal}$. Furthermore, $\bm{y}$ is a vector of observed dependent variables and $\bm{x}$ is a vector of observed independent variables. $\bm{\eta}$ is a vector of the latent dependent variables and $\bm{\xi}$ is a vector of latent independent variables. $\bm{\varepsilon}$ and $\bm{\delta}$ are vectors of the errors of the observed dependent and independent variables, respectively, and $\bm{\zeta}$ is a vector of the errors, or disturbances, of the latent variables. Notice the $^{+}$ symbol is just meant to differentiate the vectors with them from those without them. That means, $\bm{\eta}^{+}$ is a vector that holds the observed and latent variables, both dependent (i.e., 'endogenous') and independent (i.e., 'exogeneous'), $\bm{\zeta}^{+}$ holds the errors for the observed variables and the disturbances of the latent variables. $\bm{y}^{+}$ holds just the observed variables, both dependent and independent, and $\bm{\Lambda_{y}}^{+}$ is a matrix of ones and zeros that selects the observed variables from $\bm{\eta}^{+}$. Lastly $\bm{B}$ is a matrix that holds the regression coefficients.

If we say that $p$ and $q$ stand for the number of observed dependent and independent variables, respectively, and $m$ and $n$ stand for the number of latent dependent and independent variables, respectively, then $\bm{\eta}^{+}$ and $\bm{\zeta}^{+}$ are $p + q + m + n$, $\bm{y}^{+}$ is $p + q$, $\bm{\Lambda_{y}}^{+}$ is $(p + q) \times (p + q + m + n)$ and $\bm{B}$ is $(p + q + m + n) \times (p + q + m + n)$ [@Bollen1989].

\begin{figure}
\begin{center}
\resizebox{0.75\textwidth}{!}{%
\begin{tikzpicture}
% Node styles ---
\tikzstyle{man} = [rectangle, thick, minimum size = 1cm, draw = black!80, fill = white!100, font = \sffamily] % Manifest variables
\tikzstyle{lat} = [circle, thick, minimum size = 1cm, draw = black!80, fill = white!100, font = \sffamily] % Latent variables
\tikzstyle{err} = [circle, draw = black!80, fill = white!100, font = \sffamily] % Errors 
% Edge styles
\tikzstyle{con} = [-latex, font = \sffamily] % Effects
\tikzstyle{cons} = [-latex, font = \sffamily\small] % Effects with smaller font 
\tikzstyle{cor} = [latex-latex, font = \sffamily] % Correlations
% Begin Figure ---
% Nodes 
\node at (+0.0,+0.0) [man] (x1) {$x_{1}$};
\node at (+2.0,+0.0) [man] (x2) {$x_{2}$};
\node at (+6.0,+0.0) [man] (x3) {$x_{3}$};
\node at (+4.0,+0.0) [lat] (eta) {$\alpha$};
\node at (+0.0,-2.0) [man] (y1) {$y_{1}$};
\node at (+2.0,-2.0) [man] (y2) {$y_{2}$};
\node at (+6.0,-2.0) [man] (y3) {$y_{3}$};
\node at (+0.0,-3.5) [err] (e1) {$\varepsilon_{1}$};
\node at (+2.0,-3.5) [err] (e2) {$\varepsilon_{2}$};
\node at (+6.0,-3.5) [err] (e3) {$\varepsilon_{3}$};
% Paths
\path (x1) edge [con, draw = black!100, left] node {$\beta$} (y1);
\path (x2) edge [con, draw = black!100, left, near start] node {$\beta$} (y2);
\path (x3) edge [con, draw = black!100, right] node {$\beta$} (y3);
\path (e1) edge [con, draw = black!100, left] node {1} (y1);
\path (e2) edge [con, draw = black!100, left] node {1} (y2);
\path (e3) edge [con, draw = black!100, right] node {1} (y3);
\path (eta) edge [con, draw = black!100, above, near start] node {1} (y1);
\path (eta) edge [con, draw = black!100, below] node {1} (y2);
\path (eta) edge [con, draw = black!100, above] node {1} (y3);
% correlations
\path (x1) edge [cor, bend left] node {} (x2);
\path (x1) edge [cor, bend left] node {} (eta);
\path (x1) edge [cor, bend left] node {} (x3);
\path (x2) edge [cor, bend left] node {} (eta);
\path (x2) edge [cor, bend left] node {} (x3);
\path (x3) edge [cor, bend right] node {} (eta);
\end{tikzpicture}
}
\caption{Typical three-wave FE-SEM model with contemporary effects\label{fig:fesem}}
\end{center}
\end{figure}

This notation may be confusing at first, but it has advantages. First, it allows us to be more flexible with our models. For example, it allows observed $\bm{x}$ to influence observed $\bm{y}$ directly (more common notation assumes that substantive effects occur only between latent variables, observed ones are only used as indicators). It also allows $\bm{\xi}$, i.e., any latent exogenous variables, to influence $\bm{y}$ directly. These two scenarios cover the traditional FE model with observed variables, and one in which latent variables are used to account for measurement error in the independent variables, which will be discussed later in this article. It is also consistent with the notation used in popular SEM software like `lavaan` and `Mplus`. That means the matrix representation of the model one sees if they type in `lavInspect( model, what = "est")` after specifying their model in `lavaan` will match up with the notation used here. Let us, however, make things more concrete and take a look at a simple, three-wave version of the typical FE model in SEM using this notation (see Figure \ref{fig:fesem}). For that, we have:
\begin{align}
\begin{split}
\bm{y}^{+} & = \bm{\Lambda_{y}}^{+}\bm{\eta}^{+} \\
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3}
\end{pmatrix} & = 
\begin{blockarray}{cccccccc}
 & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 y_{1} & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
 y_{2} & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
 y_{3} & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\ 
 x_{1} & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
 x_{2} & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
 x_{3} & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
 \end{block}
\end{blockarray}
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix},
\end{split}
\end{align}
\begin{align}
\begin{split}
\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+} \\
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix} & = 
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix}
\begin{blockarray}{cccccccc}
 & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
 y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
 y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
 x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \end{block}
\end{blockarray} + 
\begin{pmatrix}
\varepsilon_{1} \\
\varepsilon_{2} \\
\varepsilon_{3} \\
x_{1} = \delta_{1} \\
x_{2} = \delta_{2} \\
x_{3} = \delta_{3} \\
\alpha = \xi \\
\end{pmatrix}. \label{sem_matrix}
\end{split}
\end{align}
Notice in $\bm{\zeta}^{+}$, for the independent variables it does not matter whether we write, for example $x_{1}$ or $\delta_{1}$. By treating these variables as exogenous, we can say their entire variance is made up of 'error', i.e., it is the combined effect of all unobserved influences.

Admittedly, Equation \eqref{sem_matrix} may not look like much yet. We can remedy this by first putting the equation for $\bm{\eta}^{+}$ in reduced form, i.e., by getting rid of the dependent variable on the r.h.s.: 
\begin{align}
\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+} \\
\bm{\eta}^{+} - \bm{\eta}^{+}\bm{B} & = \bm{\zeta}^{+} \\
(\bm{I} - \bm{B})\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
\bm{\eta}^{+} & = (\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+},
\end{align}
where $\bm{I}$ is the identity matrix. By substituting this back into the equation for the observed variables we get $\bm{y}^{+} = \bm{\Lambda_{y}}^{+}[(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+}]$, which works out to:
\begin{align}
\begin{pmatrix} 
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\end{pmatrix} & = 
\begin{pmatrix}
\alpha + \beta x_{1} + \varepsilon_{1} \\
\alpha + \beta x_{2} + \varepsilon_{2} \\
\alpha + \beta x_{3} + \varepsilon_{3} \\
x_{1} \\
x_{2} \\
x_{3}
\end{pmatrix},
\end{align}
which is of course exactly what we should expect given Equation \eqref{fe}.

The traditional FE model in Equation \eqref{fe} works by eliminating the individual effects by demeaning, i.e., subtracting the over-time unit mean from each of the model variables. In SEM, we specify a latent variable, here we call it $\alpha$, to represent the combined effect of all time-invariant influences. Colloquially, we can say that this variable represents what the dependent variable has in common over time. For the sake of simplicity, let us ignore the independent variable for a moment, and rewrite Equation \eqref{fe} as
\begin{align}
y_{it} & = \alpha_{i} + \varepsilon_{it}.
\end{align}
If $\alpha$ and $\varepsilon$ are independent, then the variance of $y$ is $\Var(\alpha) + \Var(\varepsilon)$. Once we transform the long-format data into wide-format, then we can see obviously that the covariance between the $T$ columns of $y_{t}$ is just $\Var(\alpha)$
\begin{align}
y_{t} & = \alpha + \varepsilon_{t}, \\
\Cov(y_{t}, y_{s}) & = \E[(\alpha + \varepsilon_{t})(\alpha + \varepsilon_{s})] \\
 & = \alpha^{2} \\
 & = \Var(\alpha), 
\end{align}
assuming $\Cov(\alpha,\varepsilon_{t}) = 0, \ \forall \ t$ and $\Cov(\varepsilon_{t},\varepsilon_{s}) = 0, \ \forall \ s, t$. And in fact this is exactly how it works. For example, if we had three waves of observations, we could write the variances and covariances of the observed variables, in this case $\bm{y} = (y_{1},y_{2},y_{3})^{\intercal}$ as a system of six linear equations in the form of $\bm{A}\bm{x} = \bm{b}$
\begin{align}
\bm{A}\bm{x} & = \bm{b} \\
\begin{blockarray}{ccccc}
 & \alpha^{2} & \varepsilon_{1}^{2} & \varepsilon_{2}^{2} & \varepsilon_{3}^{2} \\
 \begin{block}{c(cccc)}
y_{1}^{2}  & 1 & 1 & 0 & 0 \\ 
y_{2}y_{1} & 1 & 0 & 0 & 0 \\
y_{3}y_{1} & 1 & 0 & 0 & 0 \\
y_{2}^{2}  & 1 & 0 & 1 & 0 \\
y_{3}y_{2} & 1 & 0 & 0 & 0 \\
y_{3}^{2}  & 1 & 0 & 0 & 1 \\
 \end{block}
\end{blockarray}
\begin{pmatrix}
\alpha^{2} \\
\varepsilon_{1}^{2} \\
\varepsilon_{2}^{2} \\
\varepsilon_{3}^{2}
\end{pmatrix} & = 
\begin{pmatrix}
y_{1}^{2}  \\ 
y_{2}y_{1}  \\
y_{3}y_{1} \\
y_{2}^{2}   \\
y_{3}y_{2}  \\
y_{3}^{2}
\end{pmatrix}
\end{align}
solve this equation: $(\bm{A}^{\intercal}\bm{A})^{-1} \bm{A}^{\intercal}\bm{b} = \bm{x}$ which works out to
\begin{align}
\bm{x} & = 
\begin{pmatrix}
\alpha^{2} \\
\varepsilon_{1}^{2} \\
\varepsilon_{2}^{2} \\
\varepsilon_{3}^{2}
\end{pmatrix}
= 
\begin{pmatrix}
.33(y_{2}y_{1}) + .33(y_{3}y_{1}) + .33(y_{3}y_{2}) \\
y_{1}^{2} - (.33(y_{2}y_{1}) + .33(y_{3}y_{1}) + .33(y_{3}y_{2})) \\
y_{2}^{2} - (.33(y_{2}y_{1}) + .33(y_{3}y_{1}) + .33(y_{3}y_{2})) \\
y_{3}^{2} - (.33(y_{2}y_{1}) + .33(y_{3}y_{1}) + .33(y_{3}y_{2}))
\end{pmatrix}
\end{align}
which means $\Var(\alpha) = \frac{\Cov(y_{2},y_{1}) + \Cov(y_{3},y_{1}) + \Cov(y_{3},y_{2})}{3}$ the individual effects is the average covariance between $y$ at the different timepoints. It is assumed, based on the long-format equation $y_{it} = \alpha_{i} + \varepsilon_{it}$, that $\Cov(y_{2},y_{1}) = \Cov(y_{3},y_{1}) = \Cov(y_{3},y_{2}) = \Var(\alpha)$ which is to say $3\frac{\Var(\alpha)}{3} = \Var(\alpha)$.

Remember, in Equation \eqref{fe}, $y_{it}$ is a stacked $NT \times 1$ vector which we converted into $T$ individual $N \times 1$ vectors for the SEM. For the long-format data

This explanation may not be satisfying to some, so we can get even more specific and show the intuition behind the use of latent variables to decompose the dependent variable into its between- and within-variance components. For the sake of simplicity, let us ignore the independent variable for a moment, and concentrate just on the dependent variable.

What essentially differentiates an FE from a random effects (RE) model is our assumption concerning the relationship between the unobserved individual effects and the model covariates [@Bollen2010]. The FE model assumes that $\E[\alpha x_{t}] \ne 0$. As such, if we fail to control for the correlation of the covariate and the time-invariant part of the error, then the coefficient of interest, here $\beta$ will be biased. The RE model assumes $\E[\alpha x_{t}] = 0$. The benefit of an RE model over a simple pooled ordinary least squares model (POLS) is that because it accounts for serial correlation in the composite error term, i.e., $\nu_{it} = \alpha_{i} + \varepsilon_{it}$, we achieve accurate standard errors. The coefficient of interest, however, is unaffected and will be unbiased.

Our assumption regarding whether the individual effects are correlated with the model covariates occurs in $\E[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] = \bm{\Psi}$, the covariance matrix of the errors
\begin{align}
\bm{y}^{+}\bm{y}^{+ \intercal} & = \E[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})^{\intercal}] \\
 & = \E[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\zeta}^{+ \intercal}(\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal})] \\
 & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \E[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal} \\
 & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \bm{\Psi} (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal}.
\end{align}
In the case of an FE model, $\bm{\Psi}$ will reflect our belief that the individual effects are correlated with the model covariates, here again for demonstration the three-wave model:
\begin{align}
\bm{\Psi} & =
\begin{blockarray}{cccccccc}
 & \varepsilon_{1} & \varepsilon_{2} & \varepsilon_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 \varepsilon_{1} & \varepsilon_{1}^{2} &                     &                     &              &              &              & \\
 \varepsilon_{2} & 0                   & \varepsilon_{2}^{2} &                     &              &              &              & \\
 \varepsilon_{3} & 0                   & 0                   & \varepsilon_{3}^{2} &              &              &              & \\
 x_{1}           & 0                   & 0                   & 0                   & x_{1}^{2}    &              &              & \\
 x_{2}           & 0                   & 0                   & 0                   & x_{2}x_{1}   & x_{2}^{2}    &              & \\
 x_{3}           & 0                   & 0                   & 0                   & x_{3}x_{1}   & x_{3}x_{2}   & x_{3}^{2}    & \\
 \alpha          & 0                   & 0                   & 0                   & \alpha x_{1} & \alpha x_{2} & \alpha x_{3} & \alpha^{2} \\
 \end{block}
\end{blockarray}.
\end{align}
Knowing this, we can work out the equation for the coefficient of interest, $\beta$. For the sake of simplicity, assume here and throughout mean-centered variables: 
\begin{align}
\Cov(y_{t},x_{t}) & = \E[y_{t}x_{t}] \\
 & = \E[(\alpha + \beta x_{t} + \varepsilon_{t})x_{t}] \\
 & = \E[\alpha x_{t} + \beta x_{t}^{2} + \varepsilon_{t}x_{t}] \\
 & = \Cov(\alpha, x_{t}) + \beta \Var(x_{t}) \\
\hat{\beta} & = \frac{\Cov(y_{t},x_{t}) - \Cov(\alpha, x_{t})}{\Var(x_{t})}. 
\end{align}
This should make intuitive sense. From the observed covariance between the dependent and the independent variable, we are partialling out the part that is due to the covariance between the independent variable and the individual effects per unit, and then dividing by the variance of the independent variable, as usual. For the RE model, we assume $\E[\alpha x_{t}] = 0$ and the equation reduces to $\hat{\beta} = \Cov(y_{t},x_{t})/\Var(x_{t})$. The rest of the model-implied covariance matrix results from $\bm{y}^{+}\bm{y}^{+ \intercal}$.

```{python, results='asis', echo=FALSE, eval=FALSE}
# Following sympy code double-checks the matrix algebra discussed in this section
from sympy import *

x1, x2, x3, y1, y2, y3, d1, d2, d3, e1, e2, e3, xi1, xi2, xi3, n1, n2, n3, a, z1, z2, z3, b = symbols('x1, x2, x3, y1, y2, y3, delta1, delta2, delta3, varepsilon1, varepsilon2, varepsilon3, xi1, xi2, xi3, eta1, eta2, eta3, alpha, zeta1, zeta2, zeta3, beta')

y = Matrix([[y1],[y2],[y3],[x1],[x2],[x3]])
Ly = Matrix([[1, 0, 0, 0, 0, 0, 0],[0, 1, 0, 0, 0, 0, 0],[0, 0, 1, 0, 0, 0, 0],[0, 0, 0, 1, 0, 0, 0],[0, 0, 0, 0, 1, 0, 0],[0, 0, 0, 0, 0, 1, 0]])
n = Matrix([[y1],[y2],[y3],[x1],[x2],[x3],[a]])
B = Matrix([[0, 0, 0, b, 0, 0, 1],[0, 0, 0, 0, b, 0, 1],[0, 0, 0, 0, 0, b, 1],[0, 0, 0, 0, 0, 0, 0],[0, 0, 0, 0, 0, 0, 0],[0, 0, 0, 0, 0, 0, 0],[0, 0, 0, 0, 0, 0, 0]])
z = Matrix([[e1],[e2],[e3],[x1],[x2],[x3],[a]])
I = Matrix([[1, 0, 0, 0, 0, 0, 0],[0, 1, 0, 0, 0, 0, 0],[0, 0, 1, 0, 0, 0, 0],[0, 0, 0, 1, 0, 0, 0],[0, 0, 0, 0, 1, 0, 0],[0, 0, 0, 0, 0, 1, 0],[0, 0, 0, 0, 0, 0, 1]])

print('$' + latex(Ly*Inverse(I - B)*z) + '$')

print('$' + latex((Ly*Inverse(I - B)*z)*(Ly*Inverse(I - B)*z).T) + '$')
```

# Fixed-effects in lavaan {#fesem}

The package `lavaan` needs to be installed once with `install.packages( "lavaan")`. To be able to use it, we need to load it for every new `R` session:

```{r, message=FALSE, warning=FALSE, error=FALSE}
library( lavaan)
```

For users unfamiliar with `R`, SEM analyses can be carried out with almost no knowledge of the language. Typically, someone unfamiliar with `R` would prepare their data using some other statistical software, and then save the intended dataset as a `.csv`, `.xlsx`, `.dta`, `.sav`, etc. file. The user must then import the data, preferably as a dataframe, and the rest occurs using the `lavaan` syntax.^[There are many online tutorials for importing data in various formats, see, for example some from  [datacamp](https://www.datacamp.com/community/tutorials/r-data-import-tutorial) or [Quick-R](https://www.statmethods.net/input/importingdata.html), or any of the literally thousands of posts on [stackoverflow](https://stackoverflow.com/search?q=r+import+data).]

Specifying the most basic fixed-effects model, like the one shown in @Bollen2010 involves four components. First, we define the latent individual effects variable using the `=~` 'measured by' operator at the same time constraining the factor loadings at each timepoint to one. I will call the latent variable `a` to stand for $\alpha$:

Constraining all of the factor loadings to one reflects our implicit assumption that the combined effect of the unit-specific unobserved factors is constant over time. This is the default behaviour of traditional POLS-based approaches to FE that use the stacked long-format data.

```{r, eval=FALSE}
a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
```

Second, we regress the dependent variable on the independent variable using the `~` regression operator. With stacked, long-format data, only one regression coefficient is estimated over all observed timepoints. To have our FE-SEM model mimic this behaviour, we need to constrain the the estimated coefficient to equal over time. We do so by adding the same label to the regression coefficient at every time point. We will use the label `b` (this label was chosen arbitrarily, we could have used any letter) and have it act as an equality constraint for the regression coefficient of interest $\beta$:

```{r eval=FALSE}
y1 ~ b*x1
y2 ~ b*x2 
y3 ~ b*x3
y4 ~ b*x4
y5 ~ b*x5
```

The key to a FE model, as opposed to an RE model are our assumptions about the relatedness of our covariate and the individual effects, i.e., $\E[x_{t}\alpha]$. For a FE model, we want to partial out any potential covariance between the independent variable and the individual effects. This accounts for any linear relationship between $x_{t}$ and the unit-specific characteristics influencing the dependent variable. Further, allowing unrestricted covariances between the independent variable itself over time will not affect how the coefficient $\beta$ is estimated, but will have an effect on the standard errors. To mimic the behaviour of a conventional FE model, we allow the independent variable to be correlated with the individual effects and itself over time. Covariances (including covariances between a variable and itself, i.e., variances) are specified using the `~~` operator:

```{r eval=FALSE}
a ~~ x1 + x2 + x3 + x4 + x5
x1 ~~ x2 + x3 + x4 + x5
x2 ~~ x3 + x4 + x5
x3 ~~ x4 + x5
x4 ~~ x5
```

The last component of our code involves the variances of the residuals. This component is optional, but we can constrain the residual variances to be equal over time to again mimic the behaviour of a conventional FE model using POLS on stacked data. Here, again, we use labels to make equality constraints. Because $y_{t}$ is endogenous, the `~~` operator specifies the variances of *residuals*, i.e. $\epsilon_{t}$.

```{r eval=FALSE}
y1 ~~ e*y1
y2 ~~ e*y2
y3 ~~ e*y3
y4 ~~ e*y4
y5 ~~ e*y5
```

# An example {#ex1}

```{r, code=readLines("simulation-code.R"), echo=FALSE, results=FALSE}
```

To demonstrate the application of FE models in SEM, a dataset can be simulated that embodies the FE assumptions. The code for data simulation can be found in Appendix A.

To show that the latent individual effects variables represent the *combined* effect of all time-invariant characteristics, the dependent variable will be influenced by two separate unit-specific variables, which we can call $\alpha_{1}$ and $\alpha_{2}$. We will construct the simulated data such that the independent variable is correlated with both of the time-invariant effects. This means that approaches that fail to account for this confounding influence, such as pooled ordinary least squares (POLS) or random effects (RE), will be biased.

The wide-format equations for the DGP can be described as:
\begin{align}
\bm{x_{t}} & = \bm{\alpha_{1}}\beta_{x_{t},\alpha_{1}} + \bm{\alpha_{2}}\beta_{x_{t},\alpha_{2}} + \bm{\delta_{t}}, \\
\bm{y_{t}} & = \bm{x_{t}}\beta_{y_{t},x_{t}} + \bm{\alpha_{1}}\beta_{y_{t},\alpha_{1}} + \bm{\alpha_{2}} \beta_{y_{t},\alpha_{2}} + \bm{\varepsilon_{t}} 
\end{align}
where both $\bm{\alpha_{1}}$ and $\bm{\alpha_{2}}$ are $\sim N(\mu_{\alpha_{j}}, \sigma_{\alpha_{j}})$, $j = 1, 2$. $\bm{\delta_{t}}$ is $\sim N(\mu_{\delta}, \sigma_{\delta})$ and $\bm{\varepsilon_{t}}$ is $\sim N(0,1)$.

For the following example, a sample size of 1,000, observed over five waves was chosen. The unique variance of $\bm{x}$, as well as both the individual-effect variables is $\sim N(0,1)$. The coefficient of interest, $\beta_{y,x}$ is set to be equal to $0.3$. A correlation between $\bm{x}$ and the individual effects is induced through $\beta_{x,\alpha_{1}} = 0.85$ and $\beta_{x,\alpha_{1}} = 0.50$. The dependent variable is also influenced by the individual effects variables with $\beta_{y_{t},\alpha_{1}} = 0.75$ and $\beta_{y_{t},\alpha_{2}} = 0.45$ These values were chosen arbitrarily.

Now, we run the FE-SEM in `lavaan`.

```{r}
fe_sem <- '
# Define individual effects variable 
a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
# Regressions, constrain coefficient to be equal over time
y1 ~ b*x1
y2 ~ b*x2 
y3 ~ b*x3
y4 ~ b*x4
y5 ~ b*x5
# Allow unrestricted correlation between eta and covariates
a ~~ x1 + x2 + x3 + x4 + x5
x1 ~~ x2 + x3 + x4 + x5
x2 ~~ x3 + x4 + x5
x3 ~~ x4 + x5
x4 ~~ x5
# Constrain residual variances to be equal over time
y1 ~~ e*y1
y2 ~~ e*y2
y3 ~~ e*y3
y4 ~~ e*y4
y5 ~~ e*y5
'
fe_sem.fit <- sem( model = fe_sem, 
                   data = dfw, 
                   estimator = "ML")
saveRDS(fe_sem, "fe_sem.Rda")
saveRDS(fe_sem.fit, "fe_sem.fit.Rda")
```

We can get a summary of the model with `summary()`. The first portion of the summary output gives an overview of some basic information and fit statistics. The maximum likelihood estimator is the default, so it did not have to be explicitly selected in the fitting function call. Other estimators are available, including generalized and unweighted least squares (`GLS` and `ULS`, respectively), robust standard errors maximum likelihood (`MLM`) and several others (see [the lavaan online tutorial for more](https://lavaan.ugent.be/tutorial/est.html)).

This part of the summary output also tells us that the analysis is based on 1,000 observations (missings would be shown here as well if there were any), and that the $\chi^{2}$ statistic is `r round( lavInspect( fe_sem.fit, "fit")[ "chisq"], 3)` based on `r lavInspect( fe_sem.fit, "fit")[ "df"]` degrees of freedom (55 observed covariances minus 1 error variance, 1 coefficient, 1 latent variable variance, 5 exogenous variable variances and 15 covariances for $55 - 23 = 32$ df). The p-value on the $\chi^{2}$ statistic is not significant with $p =$ `r round( lavInspect( fe_sem.fit, "fit")[ "pvalue"], 3)` which tells us the differences between the model-implied and observed covariance matrices are likely due to chance, and that the model fits the data well (given how the data was generated, it would be surprising if this were not the case). Other fit measures including typical comparative fit indices can be requested by either adding `fit.measures = TRUE` as a secondary argument to the `summary()` call, or by asking for a complete list of all available fit statistics using `lavInspect( model, "fit")` where `model` stands for the name of the fitted model, in this case `fe_sem.fit`.

```{r restate model, output.lines=1:20}
summary( fe_sem.fit)
```

Next the summary output shows the measurement models for the latent variables, if any. In this case the latent variable `a` for $\alpha$ is measured by each of the five observed dependent variables with factor loadings fixed to 1.0.

```{r measurement, echo=FALSE, output.lines=22:29}
summary( fe_sem.fit)
```

The regressions are shown next. Here, because we have constrained the regression coefficients to be equal over time ( the equality constraint label `(b)` is listed to the left of the estimates), the estimate of $\beta =$ `r round( lavInspect( fe_sem.fit, "list")[6, 14], 3)` (`r round( lavInspect( fe_sem.fit, "list")[6, 15], 3)`) is repeated five times. The corresponding z- and p-values show that the coefficient is, unsurprisingly, significant.

```{r regression, echo=FALSE, output.lines=31:42}
summary( fe_sem.fit)
```

Next, the covariance estimates are listed. First, the covariances between the latent individual effects variable and the independent variable over time are shown, and then the covariances between the independent variable with itself over time.

One should always take care to double-check that there are no unintended covariances listed here. Like `Mplus`, the `lavaan` package estimates some covariances per default, without the user explicitly having to add them to the model syntax. For example, covariances between latent variables are estimated per default. If one does not wish for them to covary, it must be explicitly stated, e.g., with `f1 ~~ 0*f2`, assuming the latent variables are called `f1` and `f2`, or by overriding the default behaviour for the entire model by adding `orthogonal = TRUE` (which sets the correltion between all latent variables to zero) to the fitting call.^[This is at least the current behaviour of both the `cfa` and `sem` wrappers. In fact, both wrappers seem to be identical in terms of the default settings, see @Rosseel2020.]

```{r covariance, echo=FALSE, output.lines=44:65}
summary( fe_sem.fit)
```

Finally, the variance estimates are listed. Here, we see that in order to mimic the behaviour of a traditional FE model, the error variances over time were specified to be equal using the equality constraint `(e)`. Notice the `.` beside `y1`, `y2`, etc.: this indicates that the listed variance refers to an endogenous variable, and that it is thus an error variance. In this case, these refer to the variances of $\varepsilon_{t}$. After that, the variances of the exogenous variables, both observed and unobserved are listed.

```{r variance, echo=FALSE, output.lines=67:79}
summary( fe_sem.fit)
```

# Conclusion

Fixed-effects regression in SEM has been outlined in well-known articles by [@Allison2011; @Bollen2010; @Teachman2001]. This article provides a focused look at the implementation of the basic model, as well as common extensions using the `lavaan` package in `R`.

The benefits of FE-SEM as opposed to traditional OLS-based FE-models are largely the same ones that apply to the SEM framework in general: for one, SEM allows for a great deal of flexibility. For example, it is easy to loosen model constraints as necessary. Measurement error in both the dependent and independent variables can be dealt with using latent variables to achieve unbiased and more efficient results. Researchers interested in time-invariant predictors can integrate them into a hybrid FE/RE model with ease. Further extensions, like measurement invariance testing [@Schoot2012; @Millsap2011; @Steenkamp1998] as well as lagged dependent variables [@Bollen2010; @Allison2017] for example, can also be implemented in a straightforward fashion.

The most basic FE-SEM is the basis for a variety of currently popular extended models, such as Latent Curve Models in general [@Curran2001; @Bollen2004], as well as special implementations like the Dynamic Panel Model [@Allison2017], the Random-Intercept Cross-Lagged Panel Model [@Hamaker2015] and the Latent Curve Model with Structured Residuals [@Curran2014]. For this reason it is all the more important for researchers to have a consolidated guide to a variety of common applications of fixed-effects in SEM.

\newpage

# References {-}

<div id="refs"></div>

### 2020-06-15 extensions.Rmd
Add the extensions supplementary materials

---
title: "Supplememtary examples for:"
#author: "Henrik Kenneth Andersen"
#date: "`r Sys.Date()`"
output: 
  pdf_document:
    fig_caption: yes
    number_sections: yes
    includes: 
      in_header: header.tex
documentclass: article
classoption: a4paper
bibliography: references.bib   
---

\pagenumbering{arabic} 
\setcounter{page}{1}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# ----- Packages

library( formatR)
library( knitr)
library( lavaan)

# ----- Script hook for printing only certain lines

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(if (abs(lines[1])>1) more else NULL, 
            x[lines], 
            if (length(x)>lines[abs(length(lines))]) more else NULL
           )
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
})

# From: https://community.rstudio.com/t/showing-only-the-first-few-lines-of-the-results-of-a-code-chunk/6963/2
# Retrieved on: 26.05.2020

# ----- Script hook to increase space between text and chunk output

hook_output_def = knitr::knit_hooks$get('output')
knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(options$vspaceout)) {
    end <- paste0("\\vspace{", options$vspaceout, "}")
    stringr::str_c(hook_output_def(x, options), end)
  } else {
    hook_output_def(x, options)
  }
})

hook_source_def = knitr::knit_hooks$get('source')
knitr::knit_hooks$set(source = function(x, options) {
  if (!is.null(options$vspaceecho)) {
    begin <- paste0("\\vspace{", options$vspaceecho, "}")
    stringr::str_c(begin, hook_source_def(x, options))
  } else {
    hook_source_def(x, options)
  }
})

# From: https://stackoverflow.com/questions/43379224/add-vertical-space-above-and-below-figures-code-chunks
# Retrieved on: 04.06.2020
```

Andersen, Henrik K. (2020). A closer look at fixed-effects regression in structural equation modeling using lavaan.

```{r echo=FALSE}
df <- readRDS("longData.Rda")
dfw <- readRDS("wideData.Rda")

n <- 1000
fe_sem <- readRDS("fe_sem.Rda")
fe_sem.fit <- readRDS("fe_sem.fit.Rda")
```

# Extensions {#exten}

The following discusses several opportunities to extend the basic FE model outlined in the main article, and provides concrete guidance on the implementation in `lavaan`. First, I go over a number of possibilities to relax assumptions associated with the traditional FE model. Then, I discuss the issue of measurement error and show how we can use latent variables to deal with it and allow us to properly estimate the coefficients of interest. Then, I show display a type of hybrid FE/RE model that allows us to control for time-invariant unobserved heterogeneity that allows us to include time-invariant predictors in the model.

## Relaxing assumptions meant to mimic traditional FE models {#relax}

There are a number of implicit assumptions attached to the typical FE model that can be relaxed in SEM. Some of these assumptions have been discussed already, and a full list of assumptions can be found in @Bollen2010. Here, I will go over just a few, concentrating on the implementation in `lavaan` and the opportunity to empirically test whether the adjustments are justified or not.

The assumptions we will discuss here pertain to the time-invariance of the effects of both the latent individual effects and the observed covariates, as well as a time-invariant error variance. We can also empirically test the correlation between the individual effects and the covariates to see whether a RE model is preferable to the FE model.

For example, we can rewrite the original FE equation as
\begin{align}
y_{it} & = \beta_{t}x_{it} + \lambda_{t}\alpha_{i} + \varepsilon_{it}
\end{align}
where $\beta$ becomes $\beta_{t}$ and the implicit regression weight of one turns to $\lambda_{t}$ to highlight the fact that the effect of $x$ as well as $\alpha$ on $y$ may vary over time. We can furthermore easily relax the assumption of time-constant error variance, i.e., $\sigma^{2}_{\varepsilon_{t}}$. As noted above, the assumption regarding $\E[\alpha x_{t}]$ in $\bm{\Psi}$ determines whether we have an FE or RE model. We can set these to zero and test whether the RE model would be preferable to the FE model. In general, if the individual effects are truly uncorrelated with the model covariates, it is advisable to switch to an RE model since because it uses up less degrees of freedom, it will have smaller asymptotic standard errors [@Bollen2010].

In the following `lavaan` code, we simply remove the factor loadings of one for the latent individual effect variable which allows them to be estimated freely at each timepoint. For the effect of the covariate, we can either delete the constraints `b` in `yt ~ b*xt` or give each regression a different label, e.g., `b1`, `b2`, `b3`, etc. Similarly, to allow the error variance to vary over time, we turn the constraints `e` into simple labels, i.e., `e1`, `e2`, `e3`, etc., or again just delete them. In fact, regarding the error variances, they will be estimated necessarily, and do not need to be explicitly mentioned in the model syntax at all. Finally, to move from an FE to an RE model, we could simply constrain the correlations between the individual effects and the covariates to zero, i.e., `a ~~ 0*x1 + 0*x2 + 0*x3 + 0*x4 + 0*x5`.

```{r eval=FALSE}
fe_sem_fullyrelaxed <- '
# Define individual effects variable 
a =~ y1 + y2 + y3 + y4 + y5
# Regressions, constrain coefficient to be equal over time
y1 ~ b1*x1
y2 ~ b2*x2 
y3 ~ b3*x3
y4 ~ b4*x4
y5 ~ b5*x5
# Allow unrestricted correlation between eta and covariates
a ~~ x1 + x2 + x3 + x4 + x5
# Alternatively: constrain all to 0 for RE model, or
# just individual correlations
# a ~~ 0*x1 + 0*x2 + 0*x3 + 0*x4 + 0*x5
x1 ~~ x2 + x3 + x4 + x5
x2 ~~ x3 + x4 + x5
x3 ~~ x4 + x5
x4 ~~ x5
# Constrain residual variances to be equal over time
y1 ~~ e1*y1
y2 ~~ e2*y2
y3 ~~ e3*y3
y4 ~~ e4*y4
y5 ~~ e5*y5
'
fe_sem_fullyrelaxed.fit <- sem( model = fe_sem_fullyrelaxed, 
                                data = dfw, 
                                estimator = "ML")
```

```{r echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
re_sem <- '
# Define individual effects variable 
a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
# Regressions, constrain coefficient to be equal over time
y1 ~ b*x1
y2 ~ b*x2 
y3 ~ b*x3
y4 ~ b*x4
y5 ~ b*x5
# Allow unrestricted correlation between eta and covariates
a ~~ 0*x1 + 0*x2 + 0*x3 + 0*x4 + 0*x5
x1 ~~ x2 + x3 + x4 + x5
x2 ~~ x3 + x4 + x5
x3 ~~ x4 + x5
x4 ~~ x5
# Constrain residual variances to be equal over time
y1 ~~ e*y1
y2 ~~ e*y2
y3 ~~ e*y3
y4 ~~ e*y4
y5 ~~ e*y5
'
re_sem.fit <- sem( model = re_sem, 
                   data = dfw, 
                   estimator = "ML")
```

As outlined in the @Bollen2010 text, the researcher has the opportunity to test each of the assumptions empirically and decide whether a more parsimonious, i.e., restrictive model is justifiable. For each assumption, a likelihood ratio test can be carried out to determine whether the improvement to model fit resulting from the relaxation of various assumptions is significant or whether the more parsimonious model is preferable after all.

If we use the original model `fe_sem.fit` as a starting point, the best strategy for testing these assumptions is to work in a stepwise fashion, relaxing one assumption at a time. We can begin by first constraining the correlation between $\alpha$ and $x_{t}$ to zero (`re_sem`) for an RE model. If turning from an FE to an RE model does not significantly worsen model fit, we can go forward with the rest of the steps with the RE model. If, however, the fit does worsen significantly, it is likely better to stick with the FE model; moving forward then with it to see if a less restrictive FE model is preferable. We can perform a likelihood ratio test in `R` using the `anova()` function:

\small
```{r}
anova( fe_sem.fit, re_sem.fit)
```
\normalsize

The table that is generated shows a comparison of the nested models, in decending order according to degrees of freedom. The RE model does not estimate the correlations between the individual effects and the covariates, so it is more parsimonious and thus listed at the bottom. The `Chisq` column shows the $\chi^{2}$ statistic for both models and the `Chisq diff` column calculates the difference between the two. Obviously, according to the DGP, the correlation between the individual effects and $x_{t}$ is not zero, so fixing these to zero leads to a substantial amount of misfit. The last column puts the $\chi^{2}$ difference in relation to the difference in degrees of freedom and gives a p-value for the probability that the difference is solely due to chance. Here, the change in $\chi^{2}$ is highly significant, so the FE model should be retained.

After now having established once and for all that FE is our preferred model, we can begin relaxing the rest of the assumptions. I show the following merely as a demonstration of the procedure, we know already from the DGP that the parsimonious model as specified in `fe_sem.fit` is appropriate. We can next allow the error variances (`fe_semb.fit`), the effect of $x$ on $y$ (`fe_semc.fit`) and finally the factor loadings of the individual effects (`fe_semd.fit`) all to vary over time.

```{r echo=FALSE}
fe_semb <- '
# Define individual effects variable 
a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
# Regressions, constrain coefficient to be equal over time
y1 ~ b*x1
y2 ~ b*x2 
y3 ~ b*x3
y4 ~ b*x4
y5 ~ b*x5
# Allow unrestricted correlation between eta and covariates
a ~~ x1 + x2 + x3 + x4 + x5
x1 ~~ x2 + x3 + x4 + x5
x2 ~~ x3 + x4 + x5
x3 ~~ x4 + x5
x4 ~~ x5
# Constrain residual variances to be equal over time
y1 ~~ e1*y1
y2 ~~ e2*y2
y3 ~~ e3*y3
y4 ~~ e4*y4
y5 ~~ e5*y5
'
fe_semb.fit <- sem( model = fe_semb, 
                   data = dfw, 
                   estimator = "ML")

fe_semc <- '
# Define individual effects variable 
a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
# Regressions, constrain coefficient to be equal over time
y1 ~ b1*x1
y2 ~ b2*x2 
y3 ~ b3*x3
y4 ~ b4*x4
y5 ~ b5*x5
# Allow unrestricted correlation between eta and covariates
a ~~ x1 + x2 + x3 + x4 + x5
x1 ~~ x2 + x3 + x4 + x5
x2 ~~ x3 + x4 + x5
x3 ~~ x4 + x5
x4 ~~ x5
# Constrain residual variances to be equal over time
y1 ~~ e1*y1
y2 ~~ e2*y2
y3 ~~ e3*y3
y4 ~~ e4*y4
y5 ~~ e5*y5
'
fe_semc.fit <- sem( model = fe_semc, 
                   data = dfw, 
                   estimator = "ML")

fe_semd <- '
# Define individual effects variable 
a =~ y1 + y2 + y3 + y4 + y5
# Regressions, constrain coefficient to be equal over time
y1 ~ b1*x1
y2 ~ b2*x2 
y3 ~ b3*x3
y4 ~ b4*x4
y5 ~ b5*x5
# Allow unrestricted correlation between eta and covariates
a ~~ x1 + x2 + x3 + x4 + x5
x1 ~~ x2 + x3 + x4 + x5
x2 ~~ x3 + x4 + x5
x3 ~~ x4 + x5
x4 ~~ x5
# Constrain residual variances to be equal over time
y1 ~~ e1*y1
y2 ~~ e2*y2
y3 ~~ e3*y3
y4 ~~ e4*y4
y5 ~~ e5*y5
'
fe_semd.fit <- sem( model = fe_semd, 
                   data = dfw, 
                   estimator = "ML")
```

\small
```{r}
anova( fe_sem.fit, fe_semb.fit, fe_semc.fit, fe_semd.fit)
```
\normalsize

Keep in mind that a less parsimonious model (fewer degrees of freedom) can never fit worse than a more parsimonious one (more degrees of freedom). I.e., chance variations due to sampling error mean that adding constraints to a model will tend to always worsen fit, at least minimally. The question here is whether the improvement to fit by loosening constraints is meaningful or not. In the table above, we should not expect any meaningful improvements moving from `fe_sem.fit` to `fe_semd.fit`. Here, using simulated data, we have the luxury of knowing that any significant differences in $\chi^{2}$ are due to chance. With real data, it is up to the researcher to apply their best judgment and decide whether the results are plausible or not.

## Measurement error {#measerr}

What if the observed variables are not measured perfectly? Then what we observe, call them $\tilde{x}_{t}$ and $\tilde{y}_{t}$ are composites of the true score we are after, i.e., $x_{t}$ and $y_{t}$, plus an additive measurement error portion:
\begin{align}
\tilde{x}_{t} & = x_{t} + \upsilon_{t}, \\
\tilde{y}_{t} & = y_{t} + \nu_{t}.
\end{align}
How does this affect our model? Well, first notice that measurement error in the dependent variable is typically less of a serious problem than measurement error in the independent variables. Let us assume again mean-centered variables so that we can ignore the intercept, and consider the following simple bivariate equation:
\begin{align}
y & = \beta x + \varepsilon
\end{align}
if $y$ is measured imperfectly and what we observe is $\tilde{y} = y + \nu$, then we can rewrite the equation as: 
\begin{align}
(\tilde{y} - \nu) & = \beta x + \varepsilon \\
\tilde{y} & = \beta x + \varepsilon + \nu.
\end{align}
The measurement error in $y$ just gets added to the regression error. As long as $\nu$ is uncorrelated with $x$, then the regression coefficient will be unbiased [@Pischke2007; @Wooldridge2009]. However, this will increase the error variance and thus make the estimates less precise.

We will look at the effect of measurement error in the dependent variable using an example shortly. For now though, let us be safe in the knowledge that the coefficient of interest is likely unbiased, and concentrate on the more serious problem of error in the independent variable.

The intuition behind the problem of measurement error in the independent variable(s) can be explained as follows. Take $\tilde{x} = x + \upsilon$ and substitute this into the equation for $y$: 
\begin{align}
y & = \beta x + \varepsilon \\
 & = \beta(\tilde{x} - \upsilon) + \varepsilon \\
 & = \beta\tilde{x} + (\varepsilon - \beta\upsilon).
\end{align}
Since $\tilde{x}$ is obviously correlated with $\upsilon$ (unless the variance of $\upsilon$ is so small so that the correlation is essentially negligible), then the composite error in this regression is also correlated with the independent variable and thus the estimated coefficient of $\beta$ will be biased.

### The consequences of measurement error

To demonstrate the effect of measurement error on the FE-SEM model, and then provide a strategy for dealing with measurement error in SEM, let us first modify the simulated dataset by generating multiple *indicators* of the independent and dependent variables that all measure the intended variable imprecisely. Returning to our panel data, we have three indicators of each the independent and dependent variable, per timepoint:  
\begin{align}
\tilde{x}_{kt} & = x_{t} + \upsilon_{kt}, \\
\tilde{y}_{kt} & = y_{t} + \nu_{kt}
\end{align}
where $k = 1, 2, 3$ and $t = 1, ..., T$. This is like repeatedly presenting a respondent with a multi-item scale designed to measure things like stress, depression, xenophobia, etc. over the course of a panel study. To create the observed indicators, a random amount of measurement error (ranging from $\{\sigma^{2}_{\upsilon_{k}}, \sigma^{2}_{\nu_{k}}\}\in \{1.0, 1.1, 1.2, 1.3, 1.4, 1.5\}$) was added to the true variables. The code for generating impercise indicators can be found in Appendix B.

Let us first focus on the issue of imprecise measurements of the independent variable of interest and run the same FE-SEM model above, but this time we will use one of the measurement error sullied indicators, here $\tilde{x}_{1t}$, instead of the true independent variable, $x_{t}$.

```{r}
fe_sem2 <- '
# Define individual effects variable 
a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
# Regressions, constrain coefficient to be equal over time
# Now the imprecisely measured indicator tilde{x}_kt
# instead of the true variable x_t
y1 ~ b*x11 
y2 ~ b*x12 
y3 ~ b*x13
y4 ~ b*x14
y5 ~ b*x15
# Allow unrestricted correlation between eta and covariates
a ~~ x11 + x12 + x13 + x14 + x15
x11 ~~ x12 + x13 + x14 + x15
x12 ~~ x13 + x14 + x15
x13 ~~ x14 + x15
x14 ~~ x15
# Constrain residual variances to be equal over time
y1 ~~ e*y1
y2 ~~ e*y2
y3 ~~ e*y3
y4 ~~ e*y4
y5 ~~ e*y5
'
fe_sem2.fit <- sem( model = fe_sem2, 
                    data = dfw, 
                    estimator = "ML")
```

Now, for the sake of brevity, let us look just at the estimated coefficients for $\beta$.

```{r, echo=TRUE, output.lines=32:43}
summary( fe_sem2.fit)
```

Obviously, the estimated coefficient $\hat{\beta} =$ `r round( lavInspect( fe_sem2.fit, "list")[ 6, 14], 3)` is substantially smaller than the true population coefficient of $\beta = 0.3$. And the discrepancy is not just due to sampling error. In fact, we can derive the bias we are observing here.

For a simple bivariate regression model, it is straightforward to quantify the bias due to measurement error. It will be 
\begin{align}
\Cov(y, \tilde{x}) & = \E[y \tilde{x}] \\
 & = \E[(\beta\tilde{x} + \varepsilon)\tilde{x}] \\
 & = \E[\beta\tilde{x}^{2} + \varepsilon \tilde{x}] \\
 & = \beta \Var(\tilde{x}) \\
\hat{\beta} & = \frac{\Cov(y, \tilde{x})}{\Var(\tilde{x})} \\
 & = \frac{\E[(\beta x + \varepsilon)(x + \upsilon)]}{\E[(x + \upsilon)^{2}]} \\
 & = \frac{\E[\beta x^{2} + \beta x \upsilon + \varepsilon x + \varepsilon \upsilon]}{\E[x^{2} + 2 x \upsilon + \upsilon^{2}]} \\
 & = \beta \frac{\Var(x)}{\Var(x) + \Var(\upsilon)}.
\end{align}
which results if we assume that $\E[x \upsilon] = 0$, $\E[x \varepsilon] = 0$, $\E[\tilde{x} \varepsilon] = 0$ and $\E[\varepsilon \upsilon] = 0$ [@Wooldridge2009]. However, the model we are interested is not a bivariate model, so what was the point of showing the this? For one, it points out that the bias will always move the estimated coefficient closer to 0, since $\Var(x) \le \Var(x) + \Var(\upsilon)$. This means positive effects will be biased downwards and negative effects biased upwards, always towards zero. This is why it is referred to as *attenuation bias*. Second, it will help to familiarize ourselves with this equation to better understand the one for the multivariate case.

Indeed, the magnitude of the bias in a multivariate model is somewhat more complex to derive, but it will be 
\begin{align}
\hat{\beta} & = \beta \frac{\Var(\theta)}{\Var(\theta) + \Var(\upsilon)}
\end{align}
where $\theta$ is just the residual of a regression in which the true underlying variable is regressed on all other covariates. In this case, we need to regress $x_{t}$ on $\alpha_{1}$ and $\alpha_{2}$ for: $x_{t} = \gamma_{1}\alpha_{1} + \gamma_{2}\alpha_{2} + \theta_{t}$ [@Wooldridge2009, p. 318--320].

Normally it is not possible to reconstruct the bias since in cases where we have to rely on indicators, we would not know the true underlying variable. Furthermore, in the case of a fixed-effects model, the covariates are the unobserved time-invariant characteristics. However, because we are working with simulated data, we have everything we need. Going back to the results above, we can get the residuals of $x_{t}$ by either running a regression and saving the residuals, or we could skip a step and get them directly using the 'residual maker' matrix [@Ruettenauer2020] which is $\bm{M} = \bm{I} - \bm{A}(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}$ and $\bm{A} = \begin{pmatrix}\bm{\alpha_{1}}, \bm{\alpha_{2}}\end{pmatrix}$ is the $n \times 2$ matrix of the covariates.

```{r}
# Make the n x n identity matrix
Id <- diag( n)

# n x 2 matrix of covariates a1 and a2
A <- matrix( c( rep( 1, n), dfw$a1, dfw$a2), 
             nrow = n, ncol = 3)

# The residual maker matrix M = I - A(A'A)^-1 A'
M <- Id - A %*% solve( t( A) %*% A) %*% t( A)

# Save the residuals, t for 'theta'
t <- M %*% dfw$x1

# Re-run the FE model from above with the 'true' 
# independent variable for the correct estimate for beta 
fe_sem.fit <- sem( model = fe_sem, data = dfw, estimator = "ML")

# The equation for the biased beta 
lavInspect( fe_sem.fit, "list")[ 6, 14]*
  (( var( t))/( var( t) + var( dfw$x11 - dfw$x1)))
```

```{r, echo=FALSE, results=FALSE}
b <- lavInspect( fe_sem.fit, "list")[ 6, 14]
vart <- var( t)
varu <- var( dfw$x11 - dfw$x1)
```

From this we can see that the biased estimate above of $\hat{\beta} =$ `r round( lavInspect( fe_sem2.fit, "list")[ 6, 14], 3)` roughly comes from $\beta \frac{\Var(\theta_{t})}{\Var(\theta_{t}) + \Var(\upsilon_{t})} = `r round( b, 3)` \frac{`r  round( vart, 3)`}{`r  round( vart + varu, 3)`}$ = `r round( b*(( vart)/( vart + varu)), 3)`; 'roughly' because the equation here is the population equation. Due to sampling error, the estimates will tend vary slightly.

### Using latent variables to deal with measurement error

The way we deal with measurement error in SEM is surprisingly similar to the logic of fixed-effects regression. Namely, if we have multiple cross-sectional observations of the underlying construct of interest, then we can define a latent variable that represents the common variance across those multiple variables. Contrast this with the use of longitudinal repeated measures to isolate the common variance across time.

So, if we do in fact have multiple cross-sectional indicators for the underlying variables of interest, then we can partition them into an explained and unexplained portion:  
\begin{align}
x_{kt} & = \lambda_{kt}^{x}\xi_{t} + \delta_{kt}, \\
y_{kt} & = \lambda_{kt}^{y}\eta_{t} + \varepsilon_{kt},
\end{align}
where $x_{kt}$ and $y_{kt}$ are the $k^{th}$ indicators, $\xi_{t}$ and $\eta_{t}$ are latent factors representing the common variance across the cross-sectional repeated measures, and $\delta_{kt}$ and $\varepsilon_{kt}$ are the unexplained portions of $x_{t}$ and $y_{t}$, respectively. The latent factors are linked to the observed indicators through the factor loadings $\lambda_{kt}$.

Thus, our FE regression equation changes from $y_{t} = \beta x_{t} + \alpha + \varepsilon_{t}$ to:
\begin{align}
\eta_{t} & = \beta \xi_{t} + \alpha + \zeta_{t}
\end{align}
where $\zeta_{t}$ represents the disturbance, in other words the residual of the latent dependent variable $\eta_{t}$.

```{python, results='asis', eval=FALSE, echo=FALSE}
# Following sympy code double-checks how to specify this model 
# using matrix algebra. Turn eval=TRUE to see the vector y.

from sympy import *

x1, x2, x3, y1, y2, y3, d1, d2, d3, e1, e2, e3, xi1, xi2, xi3, n1, n2, n3, a, z1, z2, z3, b, x11, x21, x31, x12, x22, x32, x13, x23, x33, xi1, xi2, xi3, d11, d21, d31, d12, d22, d32, d13, d23, d33, l21, l31, l22, l32, l23, l33 = symbols('x1, x2, x3, y1, y2, y3, delta1, delta2, delta3, varepsilon1, varepsilon2, varepsilon3, xi1, xi2, xi3, eta1, eta2, eta3, alpha, zeta1, zeta2, zeta3, beta, x11, x21, x31, x12, x22, x32, x13, x23, x33, xi1, xi2, xi3, delta11, delta21, delta31, delta12, delta22, delta32, delta13, delta23, delta33, lambda21, lambda31, lambda22, lambda32, lambda23, lambda33')

y = Matrix([[y1],[y2],[y3],[x11],[x21],[x31],[x12],[x22],[x32],[x13],[x23],[x33]])
Ly = Matrix([[1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0],
             [0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0],
             [0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0],
             [0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0],
             [0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0],
             [0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0],
             [0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0],
             [0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0],
             [0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0],
             [0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0],
             [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0],
             [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0]])
n = Matrix([[y1],[y2],[y3],[x11],[x21],[x31],[x12],[x22],[x32],[x13],[x23],[x33],[a],[xi1],[xi2],[xi3]])
B = Matrix([[0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, b, 0, 0],
            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, b, 0],
            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, b],
            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0],
            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, l21, 0, 0],
            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, l31, 0, 0],
            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0],
            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, l22, 0],
            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, l32, 0],
            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1],
            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, l23],
            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, l33],
            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0],
            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0],
            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0],
            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0]])
z = Matrix([[e1],[e2],[e3],[d11],[d21],[d31],[d12],[d22],[d32],[d13],[d23],[d33],[a],[xi1],[xi2],[xi3]])
I = eye(16)

print('$' + latex(Ly*Inverse(I - B)*z) + '$')
```

First, however, let us double-check that measurement error in the dependent variable only increases the error variance (thus also increasing standard errors and reducing $R^{2}$), but does not systematically bias the coefficients of interest. The next model uses the indicators of $x$ and specifies latent variables to represent the valid cross-sectional variance. The dependent variable in the model is one of the impercisely measured indicators of $y$.

```{r}
fe_sem3 <- '
# Define individual effects variable 
a =~ 1*y11 + 1*y12 + 1*y13 + 1*y14 + 1*y15
# Measurement model for independent variables, xi 
xi1 =~ 1*x11 + x21 + x31 
xi2 =~ 1*x12 + x22 + x32
xi3 =~ 1*x13 + x23 + x33
xi4 =~ 1*x14 + x24 + x34
xi5 =~ 1*x15 + x25 + x35
# Regressions, constrain coefficient to be equal over time
y11 ~ b*xi1
y12 ~ b*xi2 
y13 ~ b*xi3
y14 ~ b*xi4
y15 ~ b*xi5
# Allow unrestricted correlation between eta and covariates
a ~~ xi1 + xi2 + xi3 + xi4 + xi5
xi1 ~~ xi2 + xi3 + xi4 + xi5
xi2 ~~ xi3 + xi4 + xi5
xi3 ~~ xi4 + xi5
xi4 ~~ xi5
# Constrain residual variances to be equal over time
y11 ~~ e*y11
y12 ~~ e*y12
y13 ~~ e*y13
y14 ~~ e*y14
y15 ~~ e*y15
'
fe_sem3.fit <- sem( model = fe_sem3, 
                    data = dfw, 
                    estimator = "ML")
```

```{r, output.lines=51:62}
summary( fe_sem3.fit)
```

The estimated coefficient here in model `fe_sem3.fit` is $\hat{\beta}_{y_{1t},\xi_{t}} =$ `r round( lavInspect( fe_sem3.fit, "list")[ 21, 14], 3)` which is very close to the estimated coefficient in the first, correctly specified model `fe_sem.fit`, where $\hat{\beta}_{y_{t},x_{t}} =$ `r round( lavInspect( fe_sem.fit, "list")[ 6, 14], 3)`. Notice, however, that the standard error of the estimate is substantially larger, with `r round( lavInspect( fe_sem3.fit, "list")[ 21, 15], 3)` in `fe_sem3.fit` vs. `r round( lavInspect( fe_sem.fit, "list")[ 6, 15], 3)` in `fe_sem.fit` in which $y$ was measured without error. The explained variance ($R^{2}$) in the dependent variable was also much higher in the first model:

```{r}
lavInspect( fe_sem.fit, "r2")[ 1:5]
```

compared to the current model:

```{r}
lavInspect( fe_sem3.fit, "r2")[ 1:5]
```

Finally, to see the benefits of removing measurement error from the dependent variable in terms of standard errors and $R^{2}$ statistics, we can specify a model with latent variables representing the valid cross-sectional variance in $y$.

```{r}
fe_sem4 <- '
# Measurement model for dependent variable, n for eta
n1 =~ 1*y11 + y21 + y31
n2 =~ 1*y12 + y22 + y32
n3 =~ 1*y13 + y23 + y33
n4 =~ 1*y14 + y24 + y34
n5 =~ 1*y15 + y25 + y35
# Define individual effects variable 
a =~ 1*n1 + 1*n2 + 1*n3 + 1*n4 + 1*n5
# Measurement model for independent variables, xi 
xi1 =~ 1*x11 + x21 + x31 
xi2 =~ 1*x12 + x22 + x32
xi3 =~ 1*x13 + x23 + x33
xi4 =~ 1*x14 + x24 + x34
xi5 =~ 1*x15 + x25 + x35
# Regressions, constrain coefficient to be equal over time
n1 ~ b*xi1
n2 ~ b*xi2 
n3 ~ b*xi3
n4 ~ b*xi4
n5 ~ b*xi5
# Allow unrestricted correlation between eta and covariates
a ~~ xi1 + xi2 + xi3 + xi4 + xi5
xi1 ~~ xi2 + xi3 + xi4 + xi5
xi2 ~~ xi3 + xi4 + xi5
xi3 ~~ xi4 + xi5
xi4 ~~ xi5
# Constrain residual variances to be equal over time
n1 ~~ e*n1
n2 ~~ e*n2
n3 ~~ e*n3
n4 ~~ e*n4
n5 ~~ e*n5
'
fe_sem4.fit <- sem( model = fe_sem4, 
                    data = dfw, 
                    estimator = "ML")
```

```{r, output.lines=71:82}
summary( fe_sem4.fit)
```

Generally, the use of a latent variable to account for measurement error in the dependent variable will help smooth out chance variation in the estimated coefficient and increase model efficiency, i.e., make the estimated coefficient closer to the true parameter more often. The standard errors will tend to be smaller compared to a model in which the dependent variable still contains measurement error (e.g., `fe_sem4.fit`: `r round( lavInspect( fe_sem4.fit, "list")[ 36, 15], 3)` vs. `fe_sem3.fit`: `r round( lavInspect( fe_sem3.fit, "list")[ 21, 15], 3)`) and the model fit will be better.

## Time-invariant predictors

What if we do not just want to just control for the effects of all time-invariant variables, but investigate some of them in detail? Many time-invariant variables, like sex, birth cohort, nationality, education, etc. can be interesting on their own. And typically, many of these variables are readily available in a given dataset. The traditional OLS-based FE model does not allow for this, as it wipes out the effect of *all* time-invariant variables, whether observed or not.

In SEM, we can easily specify a type of *hybrid* FE/RE model [@Bollen2010] that allows us to control for time-invariant unobserved heterogeneity while also investigating the effects of specific observed time-invariant predictors.^[These types of models have become well known outside of SEM as well, see for example @Allison2011; @Schunck2013; @Bell2018.]

In the next example, we continue with the most complex model we have specified so far, `fe_sem4.fit` in which measurement error in both the independent and dependent variables is accounted for using latent variables. Now, we would like as well to specifically investigate the effect of $\alpha_{2}$ on the dependent variable. The equation for this model changes to: $\eta_{t} = \beta \xi_{t} + \alpha + \gamma \alpha_{2} + \zeta_{t}$.

```{r eval=TRUE}
fe_sem5 <- '
# Measurement model for dependent variable, n for eta
n1 =~ 1*y11 + y21 + y31
n2 =~ 1*y12 + y22 + y32
n3 =~ 1*y13 + y23 + y33
n4 =~ 1*y14 + y24 + y34
n5 =~ 1*y15 + y25 + y35
# Define individual effects variable 
a =~ 1*n1 + 1*n2 + 1*n3 + 1*n4 + 1*n5
# Measurement model for independent variables, xi 
xi1 =~ 1*x11 + x21 + x31 
xi2 =~ 1*x12 + x22 + x32
xi3 =~ 1*x13 + x23 + x33
xi4 =~ 1*x14 + x24 + x34
xi5 =~ 1*x15 + x25 + x35
# Regressions, constrain coefficient to be equal over time
n1 ~ b*xi1 + g*a2
n2 ~ b*xi2 + g*a2
n3 ~ b*xi3 + g*a2
n4 ~ b*xi4 + g*a2
n5 ~ b*xi5 + g*a2
# Allow unrestricted correlation between eta and covariates
a ~~ xi1 + xi2 + xi3 + xi4 + xi5 + 0*a2
a2 ~~ xi1 + xi2 + xi3 + xi4 + xi5
xi1 ~~ xi2 + xi3 + xi4 + xi5
xi2 ~~ xi3 + xi4 + xi5
xi3 ~~ xi4 + xi5
xi4 ~~ xi5
# Constrain residual variances to be equal over time
n1 ~~ e*n1
n2 ~~ e*n2
n3 ~~ e*n3
n4 ~~ e*n4
n5 ~~ e*n5
'
fe_sem5.fit <- sem( model = fe_sem5, 
                    data = dfw, 
                    estimator = "ML")
```

Keep in mind, based on the DGP, the true parameters are $\beta = 0.3$ and $\gamma = 0.45$.

```{r, output.lines=71:87, eval=TRUE}
summary( fe_sem5.fit)
```

From this we can see that such a hybrid model is does a good job of estimating the coefficients of interest, with $\hat{\beta} =$ `r round( lavInspect( fe_sem5.fit, "list")[ 36, 14], 3)` (`r round( lavInspect( fe_sem5.fit, "list")[ 36, 15], 3)`) and $\hat{\gamma} =$ `r round( lavInspect( fe_sem5.fit, "list")[ 37, 14], 3)` (`r round( lavInspect( fe_sem5.fit, "list")[ 37, 15], 3)`).

It is important, however, to realize that the unbiasedness of $\hat{\gamma}$ in this model is dependent on the assumption that $\E[\zeta | \bm{\xi_{t}}, \alpha_{2}] = 0$. In other words, the idiosyncratic error is mean independent of $\bm{\xi_{t}} = \xi_{1}, \xi_{2}, ..., \xi_{T}$ as well as $\alpha_{2}$. The first part is easier to accept because we are controlling for all potential time-invariant confounders that could induce a relationship between the independent variable and the error. The unbiasedness of $\hat{\gamma}$, on the other hand rests on the assumption that the time-invariant predictor is independent of the error. If $\alpha_{2}$ represented the respondent's intelligence and $\eta_{t}$ represented the respondent's income, for example, then $\hat{\gamma}$ would be biased if both were dependent on a third time-invariant variable, say level of schooling. For this reason, we need to treat the regression on a time-invariant predictor like any other regular multivariate regression model and look to include all plausible potential confounders as controls in the model, or turn to other methods, e.g., instrumental variables.

\newpage

# References {-}

<div id="refs"></div>

### 2020-06-15 scripthooks.R
Add the script hooks for outputs

@@ -1,0 +1,56 @@
+# ----- Script hooks 
+
+# Script hook for printing only certain lines
+
+hook_output <- knit_hooks$get("output")
+knit_hooks$set(output = function(x, options) {
+  lines <- options$output.lines
+  if (is.null(lines)) {
+    return(hook_output(x, options))  # pass to default hook
+  }
+  x <- unlist(strsplit(x, "\n"))
+  more <- "..."
+  if (length(lines)==1) {        # first n lines
+    if (length(x) > lines) {
+      # truncate the output, but add ....
+      x <- c(head(x, lines), more)
+    }
+  } else {
+    x <- c(if (abs(lines[1])>1) more else NULL, 
+           x[lines], 
+           if (length(x)>lines[abs(length(lines))]) more else NULL
+    )
+  }
+  # paste these lines together
+  x <- paste(c(x, ""), collapse = "\n")
+  hook_output(x, options)
+})
+
+# From: https://community.rstudio.com/t/showing-only-the-first-few-lines-of-the-results-of-a-code-chunk/6963/2
+# Retrieved on: 26.05.2020
+
+
+# Script hook to increase space between text and chunk output 
+
+hook_output_def = knitr::knit_hooks$get('output')
+knitr::knit_hooks$set(output = function(x, options) {
+  if (!is.null(options$vspaceout)) {
+    end <- paste0("\\vspace{", options$vspaceout, "}")
+    stringr::str_c(hook_output_def(x, options), end)
+  } else {
+    hook_output_def(x, options)
+  }
+})
+
+hook_source_def = knitr::knit_hooks$get('source')
+knitr::knit_hooks$set(source = function(x, options) {
+  if (!is.null(options$vspaceecho)) {
+    begin <- paste0("\\vspace{", options$vspaceecho, "}")
+    stringr::str_c(begin, hook_source_def(x, options))
+  } else {
+    hook_source_def(x, options)
+  }
+})
+
+# From: https://stackoverflow.com/questions/43379224/add-vertical-space-above-and-below-figures-code-chunks
+# Retrieved on: 04.06.2020

### 2020-06-15 article.Rmd
Removed the script hooks from the main article .Rmd

@@
 library( knitr)
+```
 
-# ----- Script hook for printing only certain lines
-
-hook_output <- knit_hooks$get("output")
-knit_hooks$set(output = function(x, options) {
-   lines <- options$output.lines
-   if (is.null(lines)) {
-     return(hook_output(x, options))  # pass to default hook
-   }
-   x <- unlist(strsplit(x, "\n"))
-   more <- "..."
-   if (length(lines)==1) {        # first n lines
-     if (length(x) > lines) {
-       # truncate the output, but add ....
-       x <- c(head(x, lines), more)
-     }
-   } else {
-     x <- c(if (abs(lines[1])>1) more else NULL, 
-            x[lines], 
-            if (length(x)>lines[abs(length(lines))]) more else NULL
-           )
-   }
-   # paste these lines together
-   x <- paste(c(x, ""), collapse = "\n")
-   hook_output(x, options)
-})
-
-# From: https://community.rstudio.com/t/showing-only-the-first-few-lines-of-the-results-of-a-code-chunk/6963/2
-# Retrieved on: 26.05.2020
-
-# ----- Script hook to increase space between text and chunk output
-
-hook_output_def = knitr::knit_hooks$get('output')
-knitr::knit_hooks$set(output = function(x, options) {
-  if (!is.null(options$vspaceout)) {
-    end <- paste0("\\vspace{", options$vspaceout, "}")
-    stringr::str_c(hook_output_def(x, options), end)
-  } else {
-    hook_output_def(x, options)
-  }
-})
-
-hook_source_def = knitr::knit_hooks$get('source')
-knitr::knit_hooks$set(source = function(x, options) {
-  if (!is.null(options$vspaceecho)) {
-    begin <- paste0("\\vspace{", options$vspaceecho, "}")
-    stringr::str_c(begin, hook_source_def(x, options))
-  } else {
-    hook_source_def(x, options)
-  }
-})
-
-# From: https://stackoverflow.com/questions/43379224/add-vertical-space-above-and-below-figures-code-chunks
-# Retrieved on: 04.06.2020
+```{r, code=readLines("scripthooks.R"), echo=FALSE, results=FALSE}
 ```

### 2020-06-15 article.Rmd
Remove the python code for the symbolic algebra

@@
 
-```{python, results='asis', echo=FALSE, eval=FALSE}
-# Following sympy code double-checks the matrix algebra discussed in this section
-from sympy import *
-
-x1, x2, x3, y1, y2, y3, d1, d2, d3, e1, e2, e3, xi1, xi2, xi3, n1, n2, n3, a, z1, z2, z3, b = symbols('x1, x2, x3, y1, y2, y3, delta1, delta2, delta3, varepsilon1, varepsilon2, varepsilon3, xi1, xi2, xi3, eta1, eta2, eta3, alpha, zeta1, zeta2, zeta3, beta')
-
-y = Matrix([[y1],[y2],[y3],[x1],[x2],[x3]])
-Ly = Matrix([[1, 0, 0, 0, 0, 0, 0],[0, 1, 0, 0, 0, 0, 0],[0, 0, 1, 0, 0, 0, 0],[0, 0, 0, 1, 0, 0, 0],[0, 0, 0, 0, 1, 0, 0],[0, 0, 0, 0, 0, 1, 0]])
-n = Matrix([[y1],[y2],[y3],[x1],[x2],[x3],[a]])
-B = Matrix([[0, 0, 0, b, 0, 0, 1],[0, 0, 0, 0, b, 0, 1],[0, 0, 0, 0, 0, b, 1],[0, 0, 0, 0, 0, 0, 0],[0, 0, 0, 0, 0, 0, 0],[0, 0, 0, 0, 0, 0, 0],[0, 0, 0, 0, 0, 0, 0]])
-z = Matrix([[e1],[e2],[e3],[x1],[x2],[x3],[a]])
-I = Matrix([[1, 0, 0, 0, 0, 0, 0],[0, 1, 0, 0, 0, 0, 0],[0, 0, 1, 0, 0, 0, 0],[0, 0, 0, 1, 0, 0, 0],[0, 0, 0, 0, 1, 0, 0],[0, 0, 0, 0, 0, 1, 0],[0, 0, 0, 0, 0, 0, 1]])
-
-print('$' + latex(Ly*Inverse(I - B)*z) + '$')
-
-print('$' + latex((Ly*Inverse(I - B)*z)*(Ly*Inverse(I - B)*z).T) + '$')
-```
-
 # Fixed-effects in lavaan {#fesem}

### 2020-06-15 extensions.Rmd
Remove the scripthooks and the python code for symbolic algebra

@@
 library( lavaan)
+```
 
-# ----- Script hook for printing only certain lines
-
-hook_output <- knit_hooks$get("output")
-knit_hooks$set(output = function(x, options) {
-   lines <- options$output.lines
-   if (is.null(lines)) {
-     return(hook_output(x, options))  # pass to default hook
-   }
-   x <- unlist(strsplit(x, "\n"))
-   more <- "..."
-   if (length(lines)==1) {        # first n lines
-     if (length(x) > lines) {
-       # truncate the output, but add ....
-       x <- c(head(x, lines), more)
-     }
-   } else {
-     x <- c(if (abs(lines[1])>1) more else NULL, 
-            x[lines], 
-            if (length(x)>lines[abs(length(lines))]) more else NULL
-           )
-   }
-   # paste these lines together
-   x <- paste(c(x, ""), collapse = "\n")
-   hook_output(x, options)
-})
-
-# From: https://community.rstudio.com/t/showing-only-the-first-few-lines-of-the-results-of-a-code-chunk/6963/2
-# Retrieved on: 26.05.2020
-
-# ----- Script hook to increase space between text and chunk output
-
-hook_output_def = knitr::knit_hooks$get('output')
-knitr::knit_hooks$set(output = function(x, options) {
-  if (!is.null(options$vspaceout)) {
-    end <- paste0("\\vspace{", options$vspaceout, "}")
-    stringr::str_c(hook_output_def(x, options), end)
-  } else {
-    hook_output_def(x, options)
-  }
-})
-
-hook_source_def = knitr::knit_hooks$get('source')
-knitr::knit_hooks$set(source = function(x, options) {
-  if (!is.null(options$vspaceecho)) {
-    begin <- paste0("\\vspace{", options$vspaceecho, "}")
-    stringr::str_c(begin, hook_source_def(x, options))
-  } else {
-    hook_source_def(x, options)
-  }
-})
-
-# From: https://stackoverflow.com/questions/43379224/add-vertical-space-above-and-below-figures-code-chunks
-# Retrieved on: 04.06.2020
+```{r, code=readLines("scripthooks.R"), echo=FALSE, results=FALSE}
 ```
@@
 
-```{python, results='asis', eval=FALSE, echo=FALSE}
-# Following sympy code double-checks how to specify this model 
-# using matrix algebra. Turn eval=TRUE to see the vector y.
-
-from sympy import *
-
-x1, x2, x3, y1, y2, y3, d1, d2, d3, e1, e2, e3, xi1, xi2, xi3, n1, n2, n3, a, z1, z2, z3, b, x11, x21, x31, x12, x22, x32, x13, x23, x33, xi1, xi2, xi3, d11, d21, d31, d12, d22, d32, d13, d23, d33, l21, l31, l22, l32, l23, l33 = symbols('x1, x2, x3, y1, y2, y3, delta1, delta2, delta3, varepsilon1, varepsilon2, varepsilon3, xi1, xi2, xi3, eta1, eta2, eta3, alpha, zeta1, zeta2, zeta3, beta, x11, x21, x31, x12, x22, x32, x13, x23, x33, xi1, xi2, xi3, delta11, delta21, delta31, delta12, delta22, delta32, delta13, delta23, delta33, lambda21, lambda31, lambda22, lambda32, lambda23, lambda33')
-
-y = Matrix([[y1],[y2],[y3],[x11],[x21],[x31],[x12],[x22],[x32],[x13],[x23],[x33]])
-Ly = Matrix([[1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0],
-             [0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0],
-             [0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0],
-             [0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0],
-             [0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0],
-             [0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0],
-             [0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0],
-             [0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0],
-             [0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0],
-             [0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0],
-             [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0],
-             [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0]])
-n = Matrix([[y1],[y2],[y3],[x11],[x21],[x31],[x12],[x22],[x32],[x13],[x23],[x33],[a],[xi1],[xi2],[xi3]])
-B = Matrix([[0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, b, 0, 0],
-            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, b, 0],
-            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, b],
-            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0],
-            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, l21, 0, 0],
-            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, l31, 0, 0],
-            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0],
-            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, l22, 0],
-            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, l32, 0],
-            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1],
-            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, l23],
-            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, l33],
-            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0],
-            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0],
-            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0],
-            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0]])
-z = Matrix([[e1],[e2],[e3],[d11],[d21],[d31],[d12],[d22],[d32],[d13],[d23],[d33],[a],[xi1],[xi2],[xi3]])
-I = eye(16)
-
-print('$' + latex(Ly*Inverse(I - B)*z) + '$')
-```
-
 First, however, let us double-check that measurement error in the dependent variable only increases the error variance (thus also increasing standard errors and reducing $R^{2}$), but does not systematically bias the coefficients of interest.

### 2020-06-15 sympy-doublecheck-matrixnotation.py
Add the python code for double-checking the matrix notation and results using symbolic algebra (sympy)

@@ -1,0 +1,215 @@
+# --------------------------------------------- #
+# Sympy code to double-check the model matrices 
+# 
+# Henrik Kenneth Andersen 
+# 
+# 15.06.2020 
+# --------------------------------------------- #
+
+# The following code just double-checks the matrix notation used in the article. 
+
+
+from sympy import *
+
+
+# FE-SEM ------------------------------------------------------------------
+
+# ----- Symbols 
+
+x1, x2, x3, y1, y2, y3, d1, d2, d3, e1, e2, e3, xi1, xi2, xi3, n1, n2, n3, a, z1, z2, z3, b = symbols('x1, x2, x3, y1, y2, y3, delta1, delta2, delta3, varepsilon1, varepsilon2, varepsilon3, xi1, xi2, xi3, eta1, eta2, eta3, alpha, zeta1, zeta2, zeta3, beta')
+
+
+# ----- Matrices 
+
+# y
+y = Matrix([[y1],
+            [y2],
+            [y3],
+            [x1],
+            [x2],
+            [x3]])
+
+# Lambda_y            
+Ly = Matrix([[1, 0, 0, 0, 0, 0, 0],
+             [0, 1, 0, 0, 0, 0, 0],
+             [0, 0, 1, 0, 0, 0, 0],
+             [0, 0, 0, 1, 0, 0, 0],
+             [0, 0, 0, 0, 1, 0, 0],
+             [0, 0, 0, 0, 0, 1, 0]])
+
+# eta
+n = Matrix([[y1],
+            [y2],
+            [y3],
+            [x1],
+            [x2],
+            [x3],
+            [a]])
+
+# Beta            
+B = Matrix([[0, 0, 0, b, 0, 0, 1],
+            [0, 0, 0, 0, b, 0, 1],
+            [0, 0, 0, 0, 0, b, 1],
+            [0, 0, 0, 0, 0, 0, 0],
+            [0, 0, 0, 0, 0, 0, 0],
+            [0, 0, 0, 0, 0, 0, 0],
+            [0, 0, 0, 0, 0, 0, 0]])
+
+# zeta            
+z = Matrix([[e1],
+            [e2],
+            [e3],
[…]

### 2020-06-15 scripthooks.R
Make some small edits

@@ -1,5 +1,7 @@
 # ----- Script hooks 
 
+library( knitr)
+
 # Script hook for printing only certain lines
 
 hook_output <- knit_hooks$get("output")

### 2020-06-15 extensions.Rmd
Add the extensions

@@
 library( lavaan)
-```
 
-```{r, code=readLines("scripthooks.R"), echo=FALSE, results=FALSE}
+source("scripthooks.R")
 ```
@@
 
-n <- 1000
+n <- readRDS("n.Rda")
+waves <- readRDS("waves.Rda")
+
 fe_sem <- readRDS("fe_sem.Rda")

### 2020-06-16 article.Rmd
Rerun and generate the main article
Major revisions to text

@@
 ---
-title: "A Closer Look at Fixed-Effects Regression in Structural Equation Modeling Using lavaan"
-author: "Henrik Kenneth Andersen"
+title: \singlespacing A closer look at fixed-effects regression in structural equation modeling using \texttt{lavaan}
+author: Henrik Kenneth Andersen
 abstract: "This article provides an in-depth look at fixed-effects regression in the structural equation modeling (SEM) framework, specifically the application of fixed-effects in the `lavaan` package for `R`.
 It is meant as a applied guide for researchers, covering the underlying model specification, syntax, and summary output.
-It further discusses various common extentsions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of fixed-/random-effects hybrid model.
-\\par \\par \\textbf{Keywords:} Fixed-effects, structural equation modeling, lavaan, R, panel analysis"
+Online supplementary materials further discuss various common extentsions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of hybrid fixed-/random-effects model.
+\\par  \\textbf{Keywords:} Fixed-effects, structural equation modeling, lavaan, R, panel analysis"
 date: "`r Sys.Date()`"
@@
 classoption: a4paper
+urlcolor: blue
+link-citations: yes
 bibliography: references.bib   
@@
 ```{r setup, include=FALSE}
+# Knitr options 
 knitr::opts_chunk$set(echo = TRUE)
 
-# ----- Packages
+# Packages (some are repeated below, add them here as well just to be sure)
+library(formatR)
+library(knitr)
+library(lavaan)
 
-library( formatR)
-library( knitr)
-```
+hook_chunk = knitr::knit_hooks$get('chunk')
 
-```{r, code=readLines("scripthooks.R"), echo=FALSE, results=FALSE}
+knitr::knit_hooks$set(chunk = function(x, options) {
+  regular_output = hook_chunk(x, options)
+  # add latex commands if chunk option singlespacing is TRUE
+  if (isTRUE(options$singlespacing)) 
+    sprintf("\\singlespacing\n %s \n\\doublespacing", regular_output)
+  else
+    regular_output
+})
+
+knitr::opts_chunk$set(echo = TRUE, singlespacing = TRUE)
+
+# Script hooks for chunk outputs
+source("scripthooks.R")
 ```
@@
 Some of the strengths of panel data are well-known, e.g., the ability to establish temporal precedence, increased statistical power and the reduction of potential alternative models.
-However, the perhaps greatest strength of panel data is that they allow for a more rigorous testing of substantive theories as a tool for \textit{causal analysis} [@Greene2012].
+However, perhaps the greatest strength of panel data is that they allow for a more rigorous testing of substantive theories.
 Panel data, i.e., repeated measures of the same observed units (people, schools, firms, countries, etc.), allow researchers to decompose the error term into a part that stays constant within units and the part that changes over time.
-This is the essence of fixed-effects (FE) regression.
-It drastically reduces the number of potential confounders of the relationship between variables of interest by allowing one to control for all time-invariant influences.
-Thus, by using each unit at an earlier point in time as its own control, and by examining solely \textit{within-unit changes}, we come closer to the \textit{counterfactual} or \textit{potential outcomes} ideal: what would the outcome have been, had the observational unit not been exposed to the treatment [@Angrist2009]?
+The part that does not change over time can be seen as the combined effect of all time-invariant influences (e.g., sex, date of birth, nationality) on the dependent variable.
+Fixed-effects (FE) regression involves controlling for these time-invariant influences via a number of various methods.
+It thus drastically reduces the number of potential confounders of the relationship between variables of interest.
 
@@
 It focuses on the `lavaan` [@R-lavaan] package for `R` [@R-base].
-While `Mplus` [@Mplus] is unarguably the most robust SEM software currently available (in terms of features like alignment, latent variable interactions, for example) , the `lavaan` package has many benefits.
+While `Mplus` [@Mplus] is arguably the most robust SEM software currently available (in terms of features like alignment, latent variable interactions, for example) , the `lavaan` package has many benefits.
 First, it and `R` are open source and completely free.
@@
 Instead of poring over reams of plain text, copying out coefficients by hand, every part of the `lavaan` output is available as an object.
-This means that all aspects of the fitted `lavaan` object, from fit indices, to coefficients and standard errors, to the model matrices can be accessed and easily integrated into tables and plots.
+This means that all aspects of the model, from fit indices, to coefficients and standard errors, to the model matrices, can be accessed and easily integrated into tables and plots.
 Furthermore, `R` can be used for a great deal of applications.
@@
 The following article outlines the basic idea of FE regression, the particularities of FE in SEM, and shows its implementation in `lavaan`.
-Using simulated data (found in the Appendix), it demonstrates and annotates the code for the most basic FE model and provides an overview of the summary output.
-After that, a number of potential extensions are discussed and demonstrated, from loosening assumptions, to dealing with measurement error, to specifying hybrid fixed- and random effects models in order to include time-invariant predictors.
+Using [simulated data](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R), it demonstrates and annotates the code for the most basic FE model and provides an overview of the summary output.
+A number of potential extensions to the basic model, including relaxing various assumptions, dealing with measurement error in both the independent and dependent variables, as well as the inclusion of time-invariant predictors in the form of a hybrid fixed-/random-effects model, are shown in detail in the form of [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf).
 
@@
 
-The most basic fixed-effects model can be expressed as
+To begin, let us review the conventional, least squares-based FE model. It is normally written as
 \begin{align}
-y_{it} & = \bm{x_{it}}\bm{\beta} + \alpha_{i} + \varepsilon_{it}, \label{fe}
+y_{it} & = \bm{x_{it}}\bm{\beta} + \alpha_{i} + \varepsilon_{it} \label{eq:fe}, 
 \end{align}
@@
 
-The point of FE regression is to be able to drop assumptions regarding the relatedness of the model covariates and the unobserved time-constant effects, i.e., $\E[\bm{x_{it}}^{\intercal}\alpha_{i}]$.
-We can do so by manipulating Equation \eqref{fe} so that the individual effect, $\alpha_{i}$ is eliminated.
+At least a subset of all the possible time-invariant, unit-specific characteristics affecting the dependent variable are unobservable.
+Some, like sex and nationality, will be available in a typical dataset, but others, like (for example) ability or motivation will remain unobserved.
+If the conditional mean of any of these unit-specific characteristics depends on the model covariates, i.e., $\E[\alpha_{i}|\bm{x_{it}}] \ne \E[\alpha_{i}]$ (which implies $\Cov(\bm{x_{it}}^{\intercal},\alpha_{i}) \ne \bm{0}$), then the coefficient of interest, $\beta$, will be biased.
+
+The point of FE regression is to be able to drop assumptions regarding the relatedness of the model covariates and the unobserved time-constant effects.
+We can do so by manipulating Equation \eqref{eq:fe} so that the individual effect, $\alpha_{i}$ is eliminated.
 In doing so, we no longer need to worry about potential bias due to unit-specific characteristics.
@@
 \begin{align}
-(y_{it} - \bar{y}_{i}) & = (\bm{x_{it}} - \bar{\bm{x}_{i}})\bm{\beta} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
+(y_{it} - \bar{y}_{i}) & = (\bm{x_{it}} - \bar{\bm{x}}_{i})\bm{\beta} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
 \ddot{y}_{it} & = \bm{\ddot{x}_{it}}\bm{\beta} + \ddot{\varepsilon}_{it}.
@@
 This means that no assumptions about the relatedness of the model covariates and the unit-specific portion of the error are needed.
-The unbiasedness of the estimate is related solely to the strict exogeneity assumption imposed on the indiosyncratic errors, i.e., $\E[\ddot{\varepsilon}_{it}|\bm{\ddot{x}_{it}}] = 0$ which also implies $\E[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T$ [@Bruederl2015; @Wooldridge2002].
-Strict exogeneity not only rules out contemporaneous correlations between the covariates and the error, but it also implies that for the estimates $\bm{\beta}$ to be unbiased, the covariates at all timepoints must not be correlated with the error at all timepoints.
+The unbiasedness of the estimate is related solely to the strict exogeneity assumption imposed on the idiosyncratic errors, i.e., $\E[\ddot{\varepsilon}_{it}|\bm{\ddot{x}_{it}}] = \E[\ddot{\varepsilon}_{it}] = 0$ which also implies $\E[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T$ [@Bruederl2015; @Wooldridge2002].
+The assumption of strict exogeneity comes from the demeaning, where $\E[\bm{\ddot{x}}_{it}^{\intercal}\ddot{\varepsilon}_{it}] = \E[(\bm{x_{it}} - \bm{\bar{x}_{i}})^{\intercal}(\varepsilon_{it} - \bar{\varepsilon}_{i})] = \E[(\bm{x_{it}} - T^{-1}(\bm{x_{i1}} + \bm{x_{i2}} + ...
++ \bm{x_{iT}}))(\varepsilon_{it} - T^{-1}(\varepsilon_{i1} + \varepsilon_{i2} + ...
++ \varepsilon_{iT}))]$ which implies that for $\E[\bm{\ddot{x}_{it}}^{\intercal}\ddot{\varepsilon}_{it}]$ to hold, the covariates must be uncorrelated not just with the contemporaneous errors, but the errors at all timepoints.
 
@@
 
-Analytically, Equation \eqref{fe} applies as well to FE models in SEM as we will see shortly.
+Analytically, Equation \eqref{eq:fe} applies as well to FE models in SEM as we will see shortly.
 However, there are a number of differences we must address.
@@
 This means that the vectors of individual observations are only of interest insomuch as they can be converted into observed covariances.
-Second, the way in which we use a latent variable to represent the individual effects forces us to convert the data from stacked, long-format vectors of length $NT$ to $T$ wide-format vectors of length $N$.
+Second, the way in which we use a latent variable to represent the individual effects forces us to convert the data from stacked, long-format vectors of length $NT$ to $T$ individual vectors of length $N$.
 Lastly, the notation in SEM differs from traditional least squares-based models.
@@
 where $\bm{\eta}^{+} = (\bm{y}, \bm{x}, \bm{\eta}, \bm{\xi})^{\intercal}$, $\bm{\zeta}^{+} = (\bm{\varepsilon}, \bm{\delta}, \bm{\zeta}, \bm{\xi})^{\intercal}$, $\bm{y}^{+} = (\bm{y}, \bm{x})^{\intercal}$.
-Furthermore, $\bm{y}$ is a vector of observed dependent variables and $\bm{x}$ is a vector of observed independent variables.
+$\bm{y}$ is a vector of observed dependent variables and $\bm{x}$ is a vector of observed independent variables.
 $\bm{\eta}$ is a vector of the latent dependent variables and $\bm{\xi}$ is a vector of latent independent variables.
@@
 Notice the $^{+}$ symbol is just meant to differentiate the vectors with them from those without them.
-That means, $\bm{\eta}^{+}$ is a vector that holds the observed and latent variables, both dependent (i.e., 'endogenous') and independent (i.e., 'exogeneous'), $\bm{\zeta}^{+}$ holds the errors for the observed variables and the disturbances of the latent variables.
+That means, $\bm{\eta}^{+}$ is a vector that holds the observed and latent variables, both dependent (in SEM they are referred to as 'endogenous') and independent (i.e., 'exogeneous')^[For our purposes, the terms endogenous and dependent, on the one hand, and exogenous and independent, on the other, can be used interchangeably.], $\bm{\zeta}^{+}$ holds the errors for the observed variables and the disturbances of the latent variables.
 $\bm{y}^{+}$ holds just the observed variables, both dependent and independent, and $\bm{\Lambda_{y}}^{+}$ is a matrix of ones and zeros that selects the observed variables from $\bm{\eta}^{+}$.
@@
 If we say that $p$ and $q$ stand for the number of observed dependent and independent variables, respectively, and $m$ and $n$ stand for the number of latent dependent and independent variables, respectively, then $\bm{\eta}^{+}$ and $\bm{\zeta}^{+}$ are $p + q + m + n$, $\bm{y}^{+}$ is $p + q$, $\bm{\Lambda_{y}}^{+}$ is $(p + q) \times (p + q + m + n)$ and $\bm{B}$ is $(p + q + m + n) \times (p + q + m + n)$ [@Bollen1989].
+
+This notation may be confusing at first, but it has advantages.
+First, it allows us to be more flexible with our models.
+For example, it allows observed $\bm{x}$ to directly influence observed $\bm{y}$ (more common notation assumes that substantive effects occur only between latent variables, observed ones are only used as indicators, see for example @Bollen1989, @Kline2016).
+It also allows $\bm{\xi}$, i.e., any latent exogenous variables, to influence $\bm{y}$ directly.
+These two scenarios cover the traditional FE model with observed variables, and one in which latent variables are used to account for measurement error in the independent variables.
+It is also consistent with the notation used in popular SEM software like `lavaan` and `Mplus`.
+That means the matrix representation of the model one sees if they type in `lavInspect(model, what = "est")` after specifying their model in `lavaan` will match up with the notation used here.^[In `Mplus`, the model matrices can be requested by including `OUTPUT: TECH1` in the input file.] Let us, however, make things more concrete and take a look at a simple, three-wave version of the typical FE-SEM using this notation (shown graphically in Figure \ref{fig:fesem}).
+For that, we have:
+\begin{align}
+\begin{split}
+\bm{y}^{+} & = \bm{\Lambda_{y}}^{+}\bm{\eta}^{+} \\
+\begin{pmatrix}
+y_{1} \\
+y_{2} \\
+y_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3}
+\end{pmatrix} & = 
+\begin{blockarray}{cccccccc}
+ & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
+ \begin{block}{c(ccccccc)}
+ y_{1} & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
+ y_{2} & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
+ y_{3} & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\ 
+ x_{1} & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
+ x_{2} & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
+ x_{3} & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
+ \end{block}
+\end{blockarray}
+\begin{pmatrix}
+y_{1} \\
+y_{2} \\
+y_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3} \\
+\alpha
+\end{pmatrix},
+\end{split} \label{eq:y}
+\end{align}
+
+\begin{align}
+\begin{split}
+\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+} \\
+\begin{pmatrix}
+y_{1} \\
+y_{2} \\
+y_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3} \\
+\alpha
+\end{pmatrix} & = 
+\begin{pmatrix}
+y_{1} \\
+y_{2} \\
+y_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3} \\
+\alpha
+\end{pmatrix}
+\begin{blockarray}{cccccccc}
+ & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
+ \begin{block}{c(ccccccc)}
+ y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
+ y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
+ y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
+ x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+ x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+ x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+ \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+ \end{block}
+\end{blockarray} + 
+\begin{pmatrix}
+\varepsilon_{1} \\
+\varepsilon_{2} \\
+\varepsilon_{3} \\
+x_{1} = \delta_{1} \\
+x_{2} = \delta_{2} \\
+x_{3} = \delta_{3} \\
+\alpha = \xi \\
+\end{pmatrix}. 
+\end{split} \label{eq:eta}
+\end{align} 
+Notice in $\bm{\zeta}^{+}$, for the independent variables it does not matter whether we write, for example $x_{1}$ or $\delta_{1}$.
+By treating these variables as exogenous, we can say their entire variance is made up of 'error', i.e., it is the combined effect of all unobserved influences.
+
+Admittedly, Equations \eqref{eq:y} and \eqref{eq:eta} may not look like much yet.
+We can remedy this by first putting the equation for $\bm{\eta}^{+}$ in reduced form, i.e., by getting rid of the dependent variable on the r.h.s.:
+\begin{align}
+\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+} \\
+\bm{\eta}^{+} - \bm{\eta}^{+}\bm{B} & = \bm{\zeta}^{+} \\
+(\bm{I} - \bm{B})\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
+\bm{\eta}^{+} & = (\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+},
+\end{align}
+where $\bm{I}$ is the identity matrix.
+By substituting this back into the equation for the observed variables we get $\bm{y}^{+} = \bm{\Lambda_{y}}^{+}[(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+}]$, which works out to:
+\begin{align}
+\begin{pmatrix} 
+y_{1} \\
+y_{2} \\
+y_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3} \\
+\end{pmatrix} & = 
+\begin{pmatrix}
+\alpha + \beta x_{1} + \varepsilon_{1} \\
+\alpha + \beta x_{2} + \varepsilon_{2} \\
+\alpha + \beta x_{3} + \varepsilon_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3}
+\end{pmatrix},
+\end{align}
+which is of course exactly what we should expect given Equation \eqref{eq:fe}^[We can use the `sympy` package in `python` to verify and show the steps for this and other examples, see the [supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/sympy-doublecheck-matrixnotation.py) for the code.].
 
@@
 
-This notation may be confusing at first, but it has advantages.
-First, it allows us to be more flexible with our models.
-For example, it allows observed $\bm{x}$ to influence observed $\bm{y}$ directly (more common notation assumes that substantive effects occur only between latent variables, observed ones are only used as indicators).
-It also allows $\bm{\xi}$, i.e., any latent exogenous variables, to influence $\bm{y}$ directly.
-These two scenarios cover the traditional FE model with observed variables, and one in which latent variables are used to account for measurement error in the independent variables, which will be discussed later in this article.
-It is also consistent with the notation used in popular SEM software like `lavaan` and `Mplus`.
-That means the matrix representation of the model one sees if they type in `lavInspect( model, what = "est")` after specifying their model in `lavaan` will match up with the notation used here.
-Let us, however, make things more concrete and take a look at a simple, three-wave version of the typical FE model in SEM using this notation (see Figure \ref{fig:fesem}).
-For that, we have:
-\begin{align}
-\begin{split}
-\bm{y}^{+} & = \bm{\Lambda_{y}}^{+}\bm{\eta}^{+} \\
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3}
-\end{pmatrix} & = 
-\begin{blockarray}{cccccccc}
- & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
- \begin{block}{c(ccccccc)}
- y_{1} & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
- y_{2} & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
- y_{3} & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\ 
- x_{1} & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
- x_{2} & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
- x_{3} & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
- \end{block}
-\end{blockarray}
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix},
-\end{split}
-\end{align}
-\begin{align}
-\begin{split}
-\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+} \\
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix} & = 
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix}
-\begin{blockarray}{cccccccc}
- & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
- \begin{block}{c(ccccccc)}
- y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
- y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
- y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
- x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- \end{block}
-\end{blockarray} + 
-\begin{pmatrix}
-\varepsilon_{1} \\
-\varepsilon_{2} \\
-\varepsilon_{3} \\
-x_{1} = \delta_{1} \\
-x_{2} = \delta_{2} \\
-x_{3} = \delta_{3} \\
-\alpha = \xi \\
-\end{pmatrix}. \label{sem_matrix}
-\end{split}
-\end{align}
-Notice in $\bm{\zeta}^{+}$, for the independent variables it does not matter whether we write, for example $x_{1}$ or $\delta_{1}$.
-By treating these variables as exogenous, we can say their entire variance is made up of 'error', i.e., it is the combined effect of all unobserved influences.
-
-Admittedly, Equation \eqref{sem_matrix} may not look like much yet.
-We can remedy this by first putting the equation for $\bm{\eta}^{+}$ in reduced form, i.e., by getting rid of the dependent variable on the r.h.s.:
-\begin{align}
-\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+} \\
-\bm{\eta}^{+} - \bm{\eta}^{+}\bm{B} & = \bm{\zeta}^{+} \\
-(\bm{I} - \bm{B})\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
-\bm{\eta}^{+} & = (\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+},
-\end{align}
-where $\bm{I}$ is the identity matrix.
-By substituting this back into the equation for the observed variables we get $\bm{y}^{+} = \bm{\Lambda_{y}}^{+}[(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+}]$, which works out to:
-\begin{align}
-\begin{pmatrix} 
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\end{pmatrix} & = 
-\begin{pmatrix}
-\alpha + \beta x_{1} + \varepsilon_{1} \\
-\alpha + \beta x_{2} + \varepsilon_{2} \\
-\alpha + \beta x_{3} + \varepsilon_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3}
-\end{pmatrix},
-\end{align}
-which is of course exactly what we should expect given Equation \eqref{fe}.
-
-The traditional FE model in Equation \eqref{fe} works by eliminating the individual effects by demeaning, i.e., subtracting the over-time unit mean from each of the model variables.
-In SEM, we specify a latent variable, here we call it $\alpha$, to represent the combined effect of all time-invariant influences.
-Colloquially, we can say that this variable represents what the dependent variable has in common over time.
-For the sake of simplicity, let us ignore the independent variable for a moment, and rewrite Equation \eqref{fe} as
-\begin{align}
-y_{it} & = \alpha_{i} + \varepsilon_{it}.
-\end{align}
-If $\alpha$ and $\varepsilon$ are independent, then the variance of $y$ is $\Var(\alpha) + \Var(\varepsilon)$.
-Once we transform the long-format data into wide-format, then we can see obviously that the covariance between the $T$ columns of $y_{t}$ is just $\Var(\alpha)$
-\begin{align}
-y_{t} & = \alpha + \varepsilon_{t}, \\
-\Cov(y_{t}, y_{s}) & = \E[(\alpha + \varepsilon_{t})(\alpha + \varepsilon_{s})] \\
- & = \alpha^{2} \\
- & = \Var(\alpha), 
-\end{align}
-assuming $\Cov(\alpha,\varepsilon_{t}) = 0, \ \forall \ t$ and $\Cov(\varepsilon_{t},\varepsilon_{s}) = 0, \ \forall \ s, t$.
-And in fact this is exactly how it works.
-For example, if we had three waves of observations, we could write the variances and covariances of the observed variables, in this case $\bm{y} = (y_{1},y_{2},y_{3})^{\intercal}$ as a system of six linear equations in the form of $\bm{A}\bm{x} = \bm{b}$
-\begin{align}
-\bm{A}\bm{x} & = \bm{b} \\
-\begin{blockarray}{ccccc}
- & \alpha^{2} & \varepsilon_{1}^{2} & \varepsilon_{2}^{2} & \varepsilon_{3}^{2} \\
- \begin{block}{c(cccc)}
-y_{1}^{2}  & 1 & 1 & 0 & 0 \\ 
-y_{2}y_{1} & 1 & 0 & 0 & 0 \\
-y_{3}y_{1} & 1 & 0 & 0 & 0 \\
-y_{2}^{2}  & 1 & 0 & 1 & 0 \\
-y_{3}y_{2} & 1 & 0 & 0 & 0 \\
-y_{3}^{2}  & 1 & 0 & 0 & 1 \\
- \end{block}
-\end{blockarray}
-\begin{pmatrix}
-\alpha^{2} \\
-\varepsilon_{1}^{2} \\
-\varepsilon_{2}^{2} \\
-\varepsilon_{3}^{2}
-\end{pmatrix} & = 
-\begin{pmatrix}
-y_{1}^{2}  \\ 
-y_{2}y_{1}  \\
-y_{3}y_{1} \\
-y_{2}^{2}   \\
-y_{3}y_{2}  \\
-y_{3}^{2}
-\end{pmatrix}
-\end{align}
-solve this equation: $(\bm{A}^{\intercal}\bm{A})^{-1} \bm{A}^{\intercal}\bm{b} = \bm{x}$ which works out to
-\begin{align}
-\bm{x} & = 
-\begin{pmatrix}
-\alpha^{2} \\
-\varepsilon_{1}^{2} \\
-\varepsilon_{2}^{2} \\
-\varepsilon_{3}^{2}
-\end{pmatrix}
-= 
-\begin{pmatrix}
-.33(y_{2}y_{1}) + .33(y_{3}y_{1}) + .33(y_{3}y_{2}) \\
-y_{1}^{2} - (.33(y_{2}y_{1}) + .33(y_{3}y_{1}) + .33(y_{3}y_{2})) \\
-y_{2}^{2} - (.33(y_{2}y_{1}) + .33(y_{3}y_{1}) + .33(y_{3}y_{2})) \\
-y_{3}^{2} - (.33(y_{2}y_{1}) + .33(y_{3}y_{1}) + .33(y_{3}y_{2}))
-\end{pmatrix}
-\end{align}
-which means $\Var(\alpha) = \frac{\Cov(y_{2},y_{1}) + \Cov(y_{3},y_{1}) + \Cov(y_{3},y_{2})}{3}$ the individual effects is the average covariance between $y$ at the different timepoints.
-It is assumed, based on the long-format equation $y_{it} = \alpha_{i} + \varepsilon_{it}$, that $\Cov(y_{2},y_{1}) = \Cov(y_{3},y_{1}) = \Cov(y_{3},y_{2}) = \Var(\alpha)$ which is to say $3\frac{\Var(\alpha)}{3} = \Var(\alpha)$.
-
-Remember, in Equation \eqref{fe}, $y_{it}$ is a stacked $NT \times 1$ vector which we converted into $T$ individual $N \times 1$ vectors for the SEM. For the long-format data
-
-This explanation may not be satisfying to some, so we can get even more specific and show the intuition behind the use of latent variables to decompose the dependent variable into its between- and within-variance components.
-For the sake of simplicity, let us ignore the independent variable for a moment, and concentrate just on the dependent variable.
-
 What essentially differentiates an FE from a random effects (RE) model is our assumption concerning the relationship between the unobserved individual effects and the model covariates [@Bollen2010].
 The FE model assumes that $\E[\alpha x_{t}] \ne 0$.
-As such, if we fail to control for the correlation of the covariate and the time-invariant part of the error, then the coefficient of interest, here $\beta$ will be biased.
-The RE model assumes $\E[\alpha x_{t}] = 0$.
-The benefit of an RE model over a simple pooled ordinary least squares model (POLS) is that because it accounts for serial correlation in the composite error term, i.e., $\nu_{it} = \alpha_{i} + \varepsilon_{it}$, we achieve accurate standard errors.
-The coefficient of interest, however, is unaffected and will be unbiased.
-
+As such, if we fail to control for the correlation of the covariate and the time-invariant part of the error, then the coefficient of interest, here $\beta$, will be biased.
 Our assumption regarding whether the individual effects are correlated with the model covariates occurs in $\E[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] = \bm{\Psi}$, the covariance matrix of the errors
@@
 
-The package `lavaan` needs to be installed once with `install.packages( "lavaan")`. To be able to use it, we need to load it for every new `R` session:
+The package `lavaan` needs to be installed once with `install.packages("lavaan")`. To be able to use it, we need to load it for every new `R` session:
 
@@
 file.
-The user must then import the data, preferably as a dataframe, and the rest occurs using the `lavaan` syntax.^[There are many online tutorials for importing data in various formats, see, for example some from  [datacamp](https://www.datacamp.com/community/tutorials/r-data-import-tutorial) or [Quick-R](https://www.statmethods.net/input/importingdata.html), or any of the literally thousands of posts on [stackoverflow](https://stackoverflow.com/search?q=r+import+data).]
+The user must then import the data, preferably as a dataframe, and the rest occurs using the `lavaan` syntax.^[There are many online tutorials for importing data in various formats, see, for example some from  [datacamp](https://www.datacamp.com/community/tutorials/r-data-import-tutorial) or [Quick-R](https://www.statmethods.net/input/importingdata.html), or any of the many posts on [stackoverflow](https://stackoverflow.com/search?q=r+import+data).]
 
-Specifying the most basic fixed-effects model, like the one shown in @Bollen2010 involves four components.
-First, we define the latent individual effects variable using the `=~` 'measured by' operator at the same time constraining the factor loadings at each timepoint to one.
+Specifying the most basic fixed-effects model, like the one shown in @Bollen2010 (the same model as Equation \eqref{eq:fe} but with just one time-varying predictor) involves four components.
+First, we define the latent individual effects variable using the `=~` 'measured by' or 'manifested by' [@R-lavaan] operator at the same time constraining the factor loadings at each timepoint to one.
 I will call the latent variable `a` to stand for $\alpha$:
@@
 
-```{r, code=readLines("simulation-code.R"), echo=FALSE, results=FALSE}
+```{r echo=FALSE}
+df <- readRDS("longData.Rda")
+dfw <- readRDS("wideData.Rda")
 ```
 
-To demonstrate the application of FE models in SEM, a dataset can be simulated that embodies the FE assumptions. The code for data simulation can be found in Appendix A.
+To demonstrate the application of FE models in SEM, a dataset can be simulated that embodies the FE assumptions.
+Again, the code for data simulation can be found in the [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R).
 
 To show that the latent individual effects variables represent the *combined* effect of all time-invariant characteristics, the dependent variable will be influenced by two separate unit-specific variables, which we can call $\alpha_{1}$ and $\alpha_{2}$.
-We will construct the simulated data such that the independent variable is correlated with both of the time-invariant effects.
-This means that approaches that fail to account for this confounding influence, such as pooled ordinary least squares (POLS) or random effects (RE), will be biased.
+We will construct the simulated data such that the independent variable is correlated with both of the time-invariant variables.
+This means that approaches that fail to account for this confounding influence, such as pooled ordinary least squares (POLS) or RE, will be biased.
 
-The wide-format equations for the DGP can be described as:
+The wide-format equations for the data generating process can be described as:
 \begin{align}
@@
 \end{align}
-where both $\bm{\alpha_{1}}$ and $\bm{\alpha_{2}}$ are $\sim N(\mu_{\alpha_{j}}, \sigma_{\alpha_{j}})$, $j = 1, 2$.
-$\bm{\delta_{t}}$ is $\sim N(\mu_{\delta}, \sigma_{\delta})$ and $\bm{\varepsilon_{t}}$ is $\sim N(0,1)$.
+where, for the sake of simplicity, $\bm{\alpha_{1}}$, $\bm{\alpha_{2}}$, $\bm{\delta_{t}}$, $\sim N(\mu_{\delta}, \sigma_{\delta})$ and $\bm{\varepsilon_{t}}$ are $\sim N(0,1)$.
 
-For the following example, a sample size of 1,000, observed over five waves was chosen.
-The unique variance of $\bm{x}$, as well as both the individual-effect variables is $\sim N(0,1)$.
+For the following example, a sample size of 1,000, observed over five waves, was chosen.
+The unique variance of $\bm{x}$, as well as both the individual-effect variables is also $\sim N(0,1)$.
 The coefficient of interest, $\beta_{y,x}$ is set to be equal to $0.3$.
 A correlation between $\bm{x}$ and the individual effects is induced through $\beta_{x,\alpha_{1}} = 0.85$ and $\beta_{x,\alpha_{1}} = 0.50$.
+With the variances above set to one, the covariances will be roughly $\Cov(x_{t},\alpha_{1}) = 0.85$ and $\Cov(x_{t},\alpha_{2}) = 0.5$.
 The dependent variable is also influenced by the individual effects variables with $\beta_{y_{t},\alpha_{1}} = 0.75$ and $\beta_{y_{t},\alpha_{2}} = 0.45$ These values were chosen arbitrarily.
@@
 '
-fe_sem.fit <- sem( model = fe_sem, 
+fe_sem.fit <- sem(model = fe_sem, 
                   data = dfw, 
                   estimator = "ML")
+```
+
+```{r, echo=FALSE, message=FALSE, error=FALSE, warning=FALSE}
 saveRDS(fe_sem, "fe_sem.Rda")
@@
 The p-value on the $\chi^{2}$ statistic is not significant with $p =$ `r round( lavInspect( fe_sem.fit, "fit")[ "pvalue"], 3)` which tells us the differences between the model-implied and observed covariance matrices are likely due to chance, and that the model fits the data well (given how the data was generated, it would be surprising if this were not the case).
-Other fit measures including typical comparative fit indices can be requested by either adding `fit.measures = TRUE` as a secondary argument to the `summary()` call, or by asking for a complete list of all available fit statistics using `lavInspect( model, "fit")` where `model` stands for the name of the fitted model, in this case `fe_sem.fit`.
+Other fit measures including typical comparative fit indices can be requested by either adding `fit.measures = TRUE` as a secondary argument to the `summary()` call, or by asking for a complete list of all available fit statistics using `lavInspect(model, "fit")` where `model` stands for the name of the fitted model, in this case `fe_sem.fit`.
 
 ```{r restate model, output.lines=1:20}
-summary( fe_sem.fit)
+summary(fe_sem.fit)
 ```
@@
 ```{r measurement, echo=FALSE, output.lines=22:29}
-summary( fe_sem.fit)
+summary(fe_sem.fit)
 ```
@@
 The regressions are shown next.
-Here, because we have constrained the regression coefficients to be equal over time ( the equality constraint label `(b)` is listed to the left of the estimates), the estimate of $\beta =$ `r round( lavInspect( fe_sem.fit, "list")[6, 14], 3)` (`r round( lavInspect( fe_sem.fit, "list")[6, 15], 3)`) is repeated five times.
+Here, because we have constrained the regression coefficients to be equal over time (the equality constraint label `(b)` is listed to the left of the estimates), the estimate of $\beta =$ `r round( lavInspect( fe_sem.fit, "list")[6, 14], 3)` (`r round( lavInspect( fe_sem.fit, "list")[6, 15], 3)`) is repeated five times.
 The corresponding z- and p-values show that the coefficient is, unsurprisingly, significant.
@@
 ```{r regression, echo=FALSE, output.lines=31:42}
-summary( fe_sem.fit)
+summary(fe_sem.fit)
 ```
@@
 For example, covariances between latent variables are estimated per default.
-If one does not wish for them to covary, it must be explicitly stated, e.g., with `f1 ~~ 0*f2`, assuming the latent variables are called `f1` and `f2`, or by overriding the default behaviour for the entire model by adding `orthogonal = TRUE` (which sets the correltion between all latent variables to zero) to the fitting call.^[This is at least the current behaviour of both the `cfa` and `sem` wrappers.
+If one does not wish for them to covary, it must be explicitly stated, e.g., with `f1 ~~ 0*f2`, assuming the latent variables are called `f1` and `f2`, or by overriding the default behaviour for the entire model by adding `orthogonal = TRUE` (which sets the correlation between all latent variables to zero) to the fitting call.^[This is at least the current behaviour of both the `cfa` and `sem` wrappers.
 In fact, both wrappers seem to be identical in terms of the default settings, see @Rosseel2020.]
@@
 ```{r covariance, echo=FALSE, output.lines=44:65}
-summary( fe_sem.fit)
+summary(fe_sem.fit)
 ```
@@
 ```{r variance, echo=FALSE, output.lines=67:79}
-summary( fe_sem.fit)
+summary(fe_sem.fit)
 ```

### 2020-06-16 article.Rmd
Fixed the spacing

@@
 library(lavaan)
-
-hook_chunk = knitr::knit_hooks$get('chunk')
-
-knitr::knit_hooks$set(chunk = function(x, options) {
-  regular_output = hook_chunk(x, options)
-  # add latex commands if chunk option singlespacing is TRUE
-  if (isTRUE(options$singlespacing)) 
-    sprintf("\\singlespacing\n %s \n\\doublespacing", regular_output)
-  else
-    regular_output
-})
-
-knitr::opts_chunk$set(echo = TRUE, singlespacing = TRUE)
 

### 2020-06-16 scripthooks.R

@@ -2,6 +2,24 @@
 
 library( knitr)
 
+# Make code singlespacing
+
+hook_chunk = knitr::knit_hooks$get('chunk')
+
+knitr::knit_hooks$set(chunk = function(x, options) {
+  regular_output = hook_chunk(x, options)
+  # add latex commands if chunk option singlespacing is TRUE
+  if (isTRUE(options$singlespacing)) 
+    sprintf("\\singlespacing\n %s \n\\doublespacing", regular_output)
+  else
+    regular_output
+})
+
+knitr::opts_chunk$set(echo = TRUE, singlespacing = TRUE)
+
+# From: https://stackoverflow.com/questions/56704022/how-can-i-automate-different-spacing-between-text-and-code-blocks-in-r-markdown
+# Retrieved on: 16.06.2020
+
 # Script hook for printing only certain lines
 
 hook_output <- knit_hooks$get("output")

### 2020-06-19 article.Rmd
Fixed a mistake in the matrix algebra
Fixed another mistake in the algebra
Made major revisions

@@
 ---
-title: \singlespacing A closer look at fixed-effects regression in structural equation modeling using \texttt{lavaan}
+title: \singlespacing A closer look at Fixed effects regression in structural equation modeling using \texttt{lavaan}
 author: Henrik Kenneth Andersen
-abstract: "This article provides an in-depth look at fixed-effects regression in the structural equation modeling (SEM) framework, specifically the application of fixed-effects in the `lavaan` package for `R`.
+abstract: "This article provides an in-depth look at Fixed effects regression in the structural equation modeling (SEM) framework, specifically the application of Fixed effects in the `lavaan` package for `R`.
 It is meant as a applied guide for researchers, covering the underlying model specification, syntax, and summary output.
 Online supplementary materials further discuss various common extentsions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of hybrid fixed-/random-effects model.
-\\par  \\textbf{Keywords:} Fixed-effects, structural equation modeling, lavaan, R, panel analysis"
+\\par  \\textbf{Keywords:} Fixed effects, structural equation modeling, lavaan, R, panel analysis"
 date: "`r Sys.Date()`"
@@
 The part that does not change over time can be seen as the combined effect of all time-invariant influences (e.g., sex, date of birth, nationality) on the dependent variable.
-Fixed-effects (FE) regression involves controlling for these time-invariant influences via a number of various methods.
+Fixed effects (FE) regression involves controlling for these time-invariant influences via a number of various methods.
 It thus drastically reduces the number of potential confounders of the relationship between variables of interest.
@@
 
-# Fixed-effects {#fe}
+# Panel models {#panel}
 
-To begin, let us review the conventional, least squares-based FE model. It is normally written as
+To begin, let us start by reviewing a general panel model [@Bollen2010], also referred to as the 'unobserved effects model' [@Wooldridge2012]
 \begin{align}
-y_{it} & = \bm{x_{it}}\bm{\beta} + \alpha_{i} + \varepsilon_{it} \label{eq:fe}, 
+y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \alpha_{i} + \varepsilon_{it} \label{eq:gpm}
 \end{align}
-where $y_{it}$ is the observed outcome variable for unit $i$ at time $t$, $\bm{x_{it}}$ is a $1 \times k$ vector of covariates linked to the outcome by the $k \times 1$ vector of coefficients $\bm{\beta}$.
-All stable, unit-specific characteristics such as date of birth, country of origin, sex, etc.
-are captured by the time-constant part of the error $\alpha_{i}$, whereas $\varepsilon_{it}$ is the idiosyncratic error term that varies between units and over time [@Bruederl2015].
+where $y_{it}$ is the dependent variable for unit $i$ at time $t$, $\bm{x}_{it}$ is a $1 \times K$ vector of time-varying covariates (which could include a constant) linked to the dependent variable by the $K \times 1$ vector of coefficients $\bm{\beta}$, $\bm{z}_{i}$ is a $M \times 1$ vector of time-invariant covariates linked to the dependent variable by the $M \times 1$ vector of coefficients in $\bm{\gamma}$, $\alpha_{i}$ is the unit-specific effect that represents the combined effect of all unobserved unit-specific variables affecting the dependent variable and $\varepsilon_{it}$ is the idiosyncratic error.
 
-At least a subset of all the possible time-invariant, unit-specific characteristics affecting the dependent variable are unobservable.
-Some, like sex and nationality, will be available in a typical dataset, but others, like (for example) ability or motivation will remain unobserved.
-If the conditional mean of any of these unit-specific characteristics depends on the model covariates, i.e., $\E[\alpha_{i}|\bm{x_{it}}] \ne \E[\alpha_{i}]$ (which implies $\Cov(\bm{x_{it}}^{\intercal},\alpha_{i}) \ne \bm{0}$), then the coefficient of interest, $\beta$, will be biased.
+We can make stating some of the model assumptions easier by rewriting the model in matrix notation
+\begin{align}
+\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\iota}_{T}\alpha_{i} + \bm{\varepsilon}_{i}
+\end{align}
+where $\bm{y}_{i}$ and $\bm{\varepsilon}_{i}$ are $T \times 1$ vectors, $\bm{X}_{i}$ and $\bm{Z}_{i}$ are $T \times K$ and $T \times M$ matrices, respectively, $\bm{\iota}_{T}$ is a $T \times 1$ vector of ones and $\alpha_{i}$ is a scalar.
+$\bm{\beta}$ and $\bm{\gamma}$ are unchanged from Equation \eqref{eq:gpm}.
 
-The point of FE regression is to be able to drop assumptions regarding the relatedness of the model covariates and the unobserved time-constant effects.
-We can do so by manipulating Equation \eqref{eq:fe} so that the individual effect, $\alpha_{i}$ is eliminated.
-In doing so, we no longer need to worry about potential bias due to unit-specific characteristics.
-This can be achieved in a number of ways (e.g., differencing, least squares dummy variable regression), but the most common method involves subtracting the over-time unit mean from each of the terms:
+We typically make the following assumptions [see, e.g., @Wooldridge2002; @Schmidheiny2019]:
+
+- Linearity: the model is linear in its parameters and $\E[\varepsilon_{it}] = 0$ and $\E[\alpha_{i}] = 0$. 
+- Independence: the observations are independent across individuals, but not necessarily across time. 
+- Strict exogeneity: $\E[\varepsilon_{it} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] = 0$, i.e., the idiosyncratic error at each point in time is mean independent of $\bm{X}_{i}$ at \textit{all} points in time, $t = 1, ..., T$, as well as $\bm{z}_{i}$ and $\alpha_{i}$ .
+- Error variance: homoscedastic and non-serially correlated errors, i.e., $\Var(\varepsilon_{i}|\bm{X}_{i},\bm{z}_{i}, \alpha_{i}) = \sigma^{2}_{\varepsilon} \bm{I}$.
+
+## Random effects
+
+If we are willing to make some assumptions about the distribution of the individual effects and their relationship with the observed time-varying covariates, we are left with the random effects model [@Wooldridge2002; @Wooldridge2012; @Schmidheiny2019; @Skrondal2004].
+Namely, we assume the individual effects are normally distributed, with a constant (homoscedastic) variance conditional on the model covariates, i.e., $\alpha_{i}|\bm{X}_{i}, \bm{z}_{i} \sim N(0, \sigma^{2}_{\alpha})$.^[As long as $\bm{X}_{i}$ contains a constant, the assumption that $\alpha_{i}$ has a mean of zero is unproblematic [@Wooldridge2012].] Further, we assume that the expectation of the individual effects conditional on the model covariates is zero, i.e., $\E[\alpha_{i}|\bm{X}_{i},\bm{z}_{i}] = 0$, which implies they are also uncorrelated.
+
+The random effects model works by defining a composite error term: $\nu_{it} = \alpha_{i} + \varepsilon_{it}$ and rewriting the model in Equation \eqref{eq:gpm} as 
 \begin{align}
-(y_{it} - \bar{y}_{i}) & = (\bm{x_{it}} - \bar{\bm{x}}_{i})\bm{\beta} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
-\ddot{y}_{it} & = \bm{\ddot{x}_{it}}\bm{\beta} + \ddot{\varepsilon}_{it}.
+y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \nu_{it}.
 \end{align}
-Because the average of something that does not change is that thing itself, the individual effects get wiped out by the demeaning.
+By adding the individual effects to the composite error in each time period, the composite error becomes serially correlated over time, to give the following conditional covariance matrix for the errors,
+\begin{align}
+\Var(\nu | \bm{X}, \bm{z}) & = \bm{\Omega}_{\nu} = 
+\begin{pmatrix}
+\bm{\Omega}_{\nu 1} & \hdots & 0 & \hdots & 0 \\
+\vdots & \ddots & \vdots & \ddots & \vdots \\
+0 & \hdots & \bm{\Omega}_{\nu i} & \hdots & 0 \\
+\vdots & \ddots & \vdots & \ddots & \vdots \\
+0 & \hdots & 0 & \hdots & \bm{\Omega}_{\nu N}
+\end{pmatrix}
+\end{align}
+with typical elements 
+\begin{align}
+\Var(\nu_{i} | \bm{X}_{i}, \bm{z}_{i}) & = \bm{\Omega}_{\nu i} = 
+\begin{pmatrix}
+\sigma^{2}_{\nu} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\alpha} \\
+\sigma^{2}_{\alpha} & \sigma^{2}_{\nu} & \hdots & \sigma^{2}_{\alpha} \\
+\vdots & \vdots & \ddots & \vdots \\
+\sigma^{2}_{\alpha} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\nu}
+\end{pmatrix}
+\end{align}
+where $\sigma^{2}_{\nu} = \sigma^{2}_{\alpha} + \sigma^{2}_{\varepsilon}$.
+This means that in the conditional covariance matrix of the errors, given the time-varying and -invariant covariates, units over time will be correlated due to the individual effects.
+Keep the covariance structure of the errors in mind as it will help make sense of the use of latent variables to decompose the dependent variable into between- and within-variance components.
+
+## Fixed effects
+
+What if the individual effects are independent of the observed time-varying covariates, i.e., $\E[\alpha_{i}|\bm{X}_{i}] \ne 0$?
+Then grouping them in with the error will cause $\bm{\beta}$ to be biased [@Wooldridge2002; @Wooldridge2012].
+In order to drop assumptions involving the individual effects, a number of methods are available (e.g., differencing, least squares dummy variable regression), but the most common approach involves \textit{demeaning} the equation [@Bruederl2015].
+Demeaning involves subtracting the per-unit, over-time average from each of the model terms, i.e.,
+\begin{align}
+(y_{it} - \bar{y}_{i}) & = (\bm{x_{it}} - \bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \bar{\bm{z}}_{i})\bm{\gamma} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
+\ddot{y}_{it} & = \bm{\ddot{x}_{it}}\bm{\beta} + \ddot{\varepsilon}_{it}
+\end{align}
+where, for example, $\bar{y}_{i} = T^{-1}\sum_{i = 1}^{T}y_{it}$ (and the other over-time means are calculated analogously), and the variables with the dots above them represent the demeaned versions.
+Because the average of something that does not change is that thing itself, the individual effects, along with any time-invariant predictors, get wiped out by the demeaning.
 This means that no assumptions about the relatedness of the model covariates and the unit-specific portion of the error are needed.
 The unbiasedness of the estimate is related solely to the strict exogeneity assumption imposed on the idiosyncratic errors, i.e., $\E[\ddot{\varepsilon}_{it}|\bm{\ddot{x}_{it}}] = \E[\ddot{\varepsilon}_{it}] = 0$ which also implies $\E[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T$ [@Bruederl2015; @Wooldridge2002].
-The assumption of strict exogeneity comes from the demeaning, where $\E[\bm{\ddot{x}}_{it}^{\intercal}\ddot{\varepsilon}_{it}] = \E[(\bm{x_{it}} - \bm{\bar{x}_{i}})^{\intercal}(\varepsilon_{it} - \bar{\varepsilon}_{i})] = \E[(\bm{x_{it}} - T^{-1}(\bm{x_{i1}} + \bm{x_{i2}} + ...
-+ \bm{x_{iT}}))(\varepsilon_{it} - T^{-1}(\varepsilon_{i1} + \varepsilon_{i2} + ...
-+ \varepsilon_{iT}))]$ which implies that for $\E[\bm{\ddot{x}_{it}}^{\intercal}\ddot{\varepsilon}_{it}]$ to hold, the covariates must be uncorrelated not just with the contemporaneous errors, but the errors at all timepoints.
+The downside to this approach is that no time-invariant predictors can be included in the model.
 
-## Fixed-effects in structural equation modeling
+# Fixed effects in structural equation modeling
 
-Analytically, Equation \eqref{eq:fe} applies as well to FE models in SEM as we will see shortly.
-However, there are a number of differences we must address.
-First, SEM is a covariance-based approach.
-This means that the vectors of individual observations are only of interest insomuch as they can be converted into observed covariances.
-Second, the way in which we use a latent variable to represent the individual effects forces us to convert the data from stacked, long-format vectors of length $NT$ to $T$ individual vectors of length $N$.
-Lastly, the notation in SEM differs from traditional least squares-based models.
+Moving from the conventional methods outlined above to SEM, we must state the FE model in a different way.
+We turn to latent variables to account for time-invariant unobserved heterogeneity.
+In fact, besides accounting for measurement error and the representation of abstract hypothetical concepts, unobserved heterogeneity is one of the main uses of latent variables in SEM [@Skrondal2004].
+
+We first need to convert the data from stacked, long-format vectors of length $NT$ into $T$ individual vectors of length $N$.
+To see why this is necessary, consider what effect this has on the vector of responses $y_{it}$.
+Let us, for a minute ignore any covariates and focus just on the dependent variable (a so-called 'intercept-only' model) so that we have $y_{it} = \alpha_{i} + \varepsilon_{it}$.
+When we convert the data to wide-format, we get $T$ individual equations,
+\begin{align}
+y_{1i} & = \alpha_{i} + \varepsilon_{1i} \\
+y_{2i} & = \alpha_{i} + \varepsilon_{2i} \\
+ & \vdots \\
+y_{Ti} & = \alpha_{i} + \varepsilon_{Ti}. \label{eq:wide}
+\end{align}
+Because the idiosyncratic errors are assumed to be uncorrelated across units and across time, the covariance between any two of the new wide vectors $\Cov(y_{ti},y_{si}) = \Var(\alpha_{i}), \ t \ne s$.
+Otherwise, when $t = s$, the variance is $\Var(\alpha_{i}) + \Var(\varepsilon_{it})$.
+This is exactly the structure we saw above in a typical element of $\bm{\Omega}_{\nu}$.
+
+And in fact this is exactly how a latent variable is used to account for time-invariant unobserved heterogeneity.
+The dependent variable at each timepoint is regressed onto the latent variable, see Figure \ref{fig:fesem}.
+Here, the regression weights or 'factor loadings' are fixed to one to represent our assumption that the effect of the time-invariant unobserved heterogeneity is constant over time.
+It also means that the estimated variance of the latent variable is equal to the average covariance between the dependent variable over time.
+If $y_{it} = \alpha_{i} + \varepsilon_{it}$ is the true data generating process, then the covariance between all timepoints is constant and thus equal to the average covariance between all timepoints.
+
+To show this, consider the following matrix equation of the nonredundant covariances in a three-wave intercept-only model that just restates Equation \eqref{eq:wide}, and which we can solve easily with least squares:
+\begin{align}
+\bm{b} & = \bm{A}\bm{x} \\
+\begin{pmatrix}
+\Var(y_{1}) \\
+\Cov(y_{2},y_{1}) \\
+\Cov(y_{3},y_{1}) \\
+\Var(y_{2}) \\
+\Cov(y_{3},y_{2}) \\
+\Var(y_{3})
+\end{pmatrix} & = 
+\begin{blockarray}{ccccc}
+ & \psi & \phi_{1} & \phi_{2} & \phi_{3}  \\
+ \begin{block}{c(cccc)}
+ y_{1}^{2}  & 1 & 1 & 0 & 0  \\
+ y_{2}y_{1} & 1 & 0 & 0 & 0  \\
+ y_{3}y_{1} & 1 & 0 & 0 & 0  \\ 
+ y_{2}^{2}  & 1 & 0 & 1 & 0  \\
+ y_{3}y_{2} & 1 & 0 & 0 & 0  \\
+ y_{3}^{2}  & 1 & 0 & 0 & 1  \\
+ \end{block}
+\end{blockarray}
+\begin{pmatrix}
+\psi \\
+\phi_{1} \\
+\phi_{2} \\
+\phi_{3}
+\end{pmatrix}
+\end{align}
+where $\psi = \Var(\alpha)$, $\phi_{t} = \Var(\varepsilon_{t})$. We can solve this equation like so
+\begin{align}
+(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} = 
+\begin{pmatrix}
+\psi \\
+\phi_{1} \\
+\phi_{2} \\
+\phi_{3}
+\end{pmatrix} = 
+\begin{pmatrix}
+.33\Cov(y_{2},y_{1}) + .33\Cov(y_{3},y_{1}) + .33\Cov(y_{3},y_{2}) \\
+\Var(y_{1}) - \psi \\
+\Var(y_{2}) - \psi \\
+\Var(y_{3}) - \psi
+\end{pmatrix}
+\end{align}
+which shows that if our assumption of constant covariances between the dependent variable across time holds, the estimated variance of $\alpha$ is just what it should be: the average covariance between units of $y$ over time.
+Once we add in observed covariates, the estimated covariance of $\alpha$ then become the \textit{conditional} covariance of $y$ over time, given those covariates.
+
+Having explained how a latent variable is used to estimate the individual effects, we need to state the FE model using SEM-compatible matrix notation.
 There are a number of different model notations (see, for example @Bollen1989 for an overview), but the one that will serve us best is one that was proposed by @Graff1979:
@@
 \bm{y}^{+} & = \bm{\Lambda_{y}}^{+} \bm{\eta}^{+}, \\
-\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+}, 
+\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+}, 
 \end{align}
@@
 \begin{split}
-\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+} \\
+\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+} \\
 \begin{pmatrix}
@@
 \end{pmatrix} & = 
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix}
 \begin{blockarray}{cccccccc}
@@
  \end{block}
-\end{blockarray} + 
+\end{blockarray}
+\begin{pmatrix}
+y_{1} \\
+y_{2} \\
+y_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3} \\
+\alpha
+\end{pmatrix} + 
 \begin{pmatrix}
@@
 \begin{align}
-\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+} \\
-\bm{\eta}^{+} - \bm{\eta}^{+}\bm{B} & = \bm{\zeta}^{+} \\
+\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+} \\
+\bm{\eta}^{+} - \bm{B}\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
 (\bm{I} - \bm{B})\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
@@
 
-# Fixed-effects in lavaan {#fesem}
+# Fixed effects in lavaan {#fesem}
 
@@
 
-Specifying the most basic fixed-effects model, like the one shown in @Bollen2010 (the same model as Equation \eqref{eq:fe} but with just one time-varying predictor) involves four components.
+Specifying the most basic Fixed effects model, like the one shown in @Bollen2010 (the same model as Equation \eqref{eq:fe} but with just one time-varying predictor) involves four components.
 First, we define the latent individual effects variable using the `=~` 'measured by' or 'manifested by' [@R-lavaan] operator at the same time constraining the factor loadings at each timepoint to one.
@@
 
-# An example {#ex1}
+# A simulated example {#ex1}
 
@@
 
-Fixed-effects regression in SEM has been outlined in well-known articles by [@Allison2011; @Bollen2010; @Teachman2001].
+Fixed effects regression in SEM has been outlined in well-known articles by [@Allison2011; @Bollen2010; @Teachman2001].
 This article provides a focused look at the implementation of the basic model, as well as common extensions using the `lavaan` package in `R`.
@@
 The most basic FE-SEM is the basis for a variety of currently popular extended models, such as Latent Curve Models in general [@Curran2001; @Bollen2004], as well as special implementations like the Dynamic Panel Model [@Allison2017], the Random-Intercept Cross-Lagged Panel Model [@Hamaker2015] and the Latent Curve Model with Structured Residuals [@Curran2014].
-For this reason it is all the more important for researchers to have a consolidated guide to a variety of common applications of fixed-effects in SEM.
+For this reason it is all the more important for researchers to have a consolidated guide to a variety of common applications of Fixed effects in SEM.
 

### 2020-06-20 article.Rmd
More minor revisions

@@
 ---
-title: \singlespacing A closer look at Fixed effects regression in structural equation modeling using \texttt{lavaan}
+title: \singlespacing A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}
 author: Henrik Kenneth Andersen
-abstract: "This article provides an in-depth look at Fixed effects regression in the structural equation modeling (SEM) framework, specifically the application of Fixed effects in the `lavaan` package for `R`.
+abstract: "This article provides an in-depth look at fixed effects regression in the structural equation modeling (SEM) framework, specifically the application of fixed effects in the `lavaan` package for `R`.
 It is meant as a applied guide for researchers, covering the underlying model specification, syntax, and summary output.
-Online supplementary materials further discuss various common extentsions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of hybrid fixed-/random-effects model.
+Online supplementary materials further discuss various common extentsions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of hybrid fixed-/ random effects model.
 \\par  \\textbf{Keywords:} Fixed effects, structural equation modeling, lavaan, R, panel analysis"
@@
 
-To begin, let us start by reviewing a general panel model [@Bollen2010], also referred to as the 'unobserved effects model' [@Wooldridge2012]
+To begin, let us start by reviewing a general panel model [@Bollen2010], also referred to as the 'unobserved effects model' [@Wooldridge2012] (we will return to this model in the [online section](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf) of the article when we discuss loosening assumptions)
 \begin{align}
@@
 \end{align}
-where $y_{it}$ is the dependent variable for unit $i$ at time $t$, $\bm{x}_{it}$ is a $1 \times K$ vector of time-varying covariates (which could include a constant) linked to the dependent variable by the $K \times 1$ vector of coefficients $\bm{\beta}$, $\bm{z}_{i}$ is a $M \times 1$ vector of time-invariant covariates linked to the dependent variable by the $M \times 1$ vector of coefficients in $\bm{\gamma}$, $\alpha_{i}$ is the unit-specific effect that represents the combined effect of all unobserved unit-specific variables affecting the dependent variable and $\varepsilon_{it}$ is the idiosyncratic error.
+where $y_{it}$ is the dependent variable for unit $i, \ i = 1, ..., N$ at time $t, \ t = 1, ..., T$, $\bm{x}_{it}$ is a $1 \times K$ vector of time-varying covariates (which could include a constant) linked to the dependent variable by the $K \times 1$ vector of coefficients $\bm{\beta}$, $\bm{z}_{i}$ is a $1 \times M$ vector of time-invariant covariates linked to the dependent variable by the $M \times 1$ vector of coefficients in $\bm{\gamma}$, $\alpha_{i}$ represents the combined effect of all unobserved unit-specific variables affecting the dependent variable and $\varepsilon_{it}$ is the idiosyncratic error.
 
-We can make stating some of the model assumptions easier by rewriting the model in matrix notation
+We can make stating some of the model assumptions easier by rewriting it in matrix notation
 \begin{align}
@@
 
-We typically make the following assumptions [see, e.g., @Wooldridge2002; @Schmidheiny2019]:
+We typically make the following assumptions [see, e.g., @Wooldridge2002; @Schmidheiny2019] about this model:
 
@@
 
-## Random effects
+## Random effects {#re}
 
@@
 This means that in the conditional covariance matrix of the errors, given the time-varying and -invariant covariates, units over time will be correlated due to the individual effects.
-Keep the covariance structure of the errors in mind as it will help make sense of the use of latent variables to decompose the dependent variable into between- and within-variance components.
+We should keep the covariance structure of the errors in mind as it will help make sense of the use of latent variables to decompose the dependent variable into between- and within-variance components, discussed below in Section \ref{fe-sem}.
 
-## Fixed effects
+## Fixed effects {#fe}
 
-What if the individual effects are independent of the observed time-varying covariates, i.e., $\E[\alpha_{i}|\bm{X}_{i}] \ne 0$?
+What if the individual effects are \textit{note} independent of the observed time-varying covariates, i.e., $\E[\alpha_{i}|\bm{X}_{i}] \ne 0$?
 Then grouping them in with the error will cause $\bm{\beta}$ to be biased [@Wooldridge2002; @Wooldridge2012].
-In order to drop assumptions involving the individual effects, a number of methods are available (e.g., differencing, least squares dummy variable regression), but the most common approach involves \textit{demeaning} the equation [@Bruederl2015].
+In order to drop assumptions involving the individual effects, a number of methods are available (e.g., differencing, least squares dummy variable regression), but the most common approach is to \textit{demean} the equation [@Bruederl2015].
 Demeaning involves subtracting the per-unit, over-time average from each of the model terms, i.e.,
 \begin{align}
-(y_{it} - \bar{y}_{i}) & = (\bm{x_{it}} - \bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \bar{\bm{z}}_{i})\bm{\gamma} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
-\ddot{y}_{it} & = \bm{\ddot{x}_{it}}\bm{\beta} + \ddot{\varepsilon}_{it}
+(y_{it} - \bar{y}_{i}) & = (\bm{x}_{it} - \bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \bar{\bm{z}}_{i})\bm{\gamma} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
+\ddot{y}_{it} & = \bm{\ddot{x}}_{it}\bm{\beta} + \ddot{\varepsilon}_{it} \label{eq:fe}
 \end{align}
@@
 This means that no assumptions about the relatedness of the model covariates and the unit-specific portion of the error are needed.
-The unbiasedness of the estimate is related solely to the strict exogeneity assumption imposed on the idiosyncratic errors, i.e., $\E[\ddot{\varepsilon}_{it}|\bm{\ddot{x}_{it}}] = \E[\ddot{\varepsilon}_{it}] = 0$ which also implies $\E[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T$ [@Bruederl2015; @Wooldridge2002].
+The unbiasedness of the estimate is related solely to the strict exogeneity assumption imposed on the idiosyncratic errors, i.e., $\E[\ddot{\varepsilon}_{it}|\bm{\ddot{x}}_{it}] = \E[\ddot{\varepsilon}_{it}] = 0$ which also implies $\E[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T$ [@Bruederl2015; @Wooldridge2002].
 The downside to this approach is that no time-invariant predictors can be included in the model.
+In the [online section](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf), it will be discussed how to get around this restriction using SEM.
 
-# Fixed effects in structural equation modeling
+# Fixed effects in structural equation modeling {#fe-sem}
 
@@
 We turn to latent variables to account for time-invariant unobserved heterogeneity.
-In fact, besides accounting for measurement error and the representation of abstract hypothetical concepts, unobserved heterogeneity is one of the main uses of latent variables in SEM [@Skrondal2004].
+In fact, besides accounting for measurement error and the representation of abstract hypothetical concepts, unobserved heterogeneity has historically been one of the main uses of latent variables in SEM [@Skrondal2004].
 
@@
 It also means that the estimated variance of the latent variable is equal to the average covariance between the dependent variable over time.
-If $y_{it} = \alpha_{i} + \varepsilon_{it}$ is the true data generating process, then the covariance between all timepoints is constant and thus equal to the average covariance between all timepoints.
+If $y_{it} = \alpha_{i} + \varepsilon_{it}$ is the true data generating process, then the expected covariances between all timepoints are constant and thus equal to the average covariance between all timepoints.
 
@@
 \Var(y_{3}) - \psi
-\end{pmatrix}
+\end{pmatrix}.
 \end{align}
-which shows that if our assumption of constant covariances between the dependent variable across time holds, the estimated variance of $\alpha$ is just what it should be: the average covariance between units of $y$ over time.
+In other words, $\psi = .33\Cov(y_{2},y_{1}) + .33\Cov(y_{3},y_{1}) + .33\Cov(y_{3},y_{2})$ is the same as $\psi = \frac{1}{3} 3\Cov(y_{t},y_{s}) = \Cov(y_{t},y_{s})$ where $\Cov(y_{t},y_{s})$ is the covariance between any two columns of the wide-format dependent variable.
+This shows that if our assumption of constant covariances between the dependent variable across time holds, the estimated variance of $\alpha$ is just what it should be: the average covariance between units of $y$ over time.
 Once we add in observed covariates, the estimated covariance of $\alpha$ then become the \textit{conditional} covariance of $y$ over time, given those covariates.
@@
 This notation may be confusing at first, but it has advantages.
-First, it allows us to be more flexible with our models.
+First, it allows us the flexibility we need for the models.
 For example, it allows observed $\bm{x}$ to directly influence observed $\bm{y}$ (more common notation assumes that substantive effects occur only between latent variables, observed ones are only used as indicators, see for example @Bollen1989, @Kline2016).
@@
 These two scenarios cover the traditional FE model with observed variables, and one in which latent variables are used to account for measurement error in the independent variables.
-It is also consistent with the notation used in popular SEM software like `lavaan` and `Mplus`.
-That means the matrix representation of the model one sees if they type in `lavInspect(model, what = "est")` after specifying their model in `lavaan` will match up with the notation used here.^[In `Mplus`, the model matrices can be requested by including `OUTPUT: TECH1` in the input file.] Let us, however, make things more concrete and take a look at a simple, three-wave version of the typical FE-SEM using this notation (shown graphically in Figure \ref{fig:fesem}).
-For that, we have:
+It is also consistent with the notation used for these models in `lavaan`.
+In fact, `lavaan` switches automatically between matrix notations depending on the specified model.
+That means the matrix representation of the model one sees if they type in `lavInspect(model, what = "est")` after specifying their model in `lavaan` will match up with the notation used here.^[In `Mplus`, the model matrices can be requested by including `OUTPUT: TECH1` in the input file.] It does have a potential disadvantage however.
+Besides being less intuitive than the typical $\bm{\eta} = \bm{B}\bm{\eta} + \bm{\Gamma}\bm{\xi} + \bm{\zeta}$ notation, it means that by including the observed covariates in the stacked long vector $\bm{y}^{+}$, they are treated as another response variable with variances and covariances to be estimated by the model (instead of just using the sample statistics).
+This means that the assumption of multivariate normality (otherwise just imposed on the dependent variables) also applies to the independent ones [@Skrondal2004, p.
+75].
+This can be problematic for noncontinuous independent variables like sex, nationality dummies, marriage status (married/unmarried), etc.
+See @Skrondal2004 for more on this topic.
+
+Let us, however, make things more concrete and take a look at a simple, three-wave version of the typical FE-SEM using this notation (shown graphically in Figure \ref{fig:fesem}). For that, we have:
 \begin{align}
@@
 \end{align} 
-Notice in $\bm{\zeta}^{+}$, for the independent variables it does not matter whether we write, for example $x_{1}$ or $\delta_{1}$.
-By treating these variables as exogenous, we can say their entire variance is made up of 'error', i.e., it is the combined effect of all unobserved influences.
+Notice in $\bm{\zeta}^{+}$, for the independent variables we could either write, for example $x_{1}$ or $\delta_{1}$.
+As mentioned above, this is due to the model notation treating the independent variables like dependent variables with variances/covariances to be estimated.
+For the sake of simplicity, we will ignore this subtlety and refer to the observed variable from now on, keeping in mind that if the multivariate normality assumption holds, the estimated statistics will likely be sufficiently close to the sample ones for it to not make much of a difference.
 
@@
 
-Specifying the most basic Fixed effects model, like the one shown in @Bollen2010 (the same model as Equation \eqref{eq:fe} but with just one time-varying predictor) involves four components.
+Specifying the most basic fixed effects model, like the one shown in @Bollen2010 (the same model as Equation \eqref{eq:fe} but with just one time-varying predictor) involves four components.
 First, we define the latent individual effects variable using the `=~` 'measured by' or 'manifested by' [@R-lavaan] operator at the same time constraining the factor loadings at each timepoint to one.
@@
 The most basic FE-SEM is the basis for a variety of currently popular extended models, such as Latent Curve Models in general [@Curran2001; @Bollen2004], as well as special implementations like the Dynamic Panel Model [@Allison2017], the Random-Intercept Cross-Lagged Panel Model [@Hamaker2015] and the Latent Curve Model with Structured Residuals [@Curran2014].
-For this reason it is all the more important for researchers to have a consolidated guide to a variety of common applications of Fixed effects in SEM.
+For this reason it is all the more important for researchers to have a consolidated guide to a variety of common applications of fixed effects in SEM.
 

### 2020-06-22 article.Rmd
Minor revisions

@@
 
-This article is intended as a guide for researchers looking for in-depth help with specifying FE models in SEM.
+There are a number of articles describing basic concept of panel model regression, and FE regression in SEM [e.g., @Allison2011; @Bollen2010; @Teachman2001].
+This article is intended as a \textit{practical guide} for researchers looking for in-depth help with specifying FE models in SEM.
 It focuses on the `lavaan` [@R-lavaan] package for `R` [@R-base].
@@
 Using [simulated data](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R), it demonstrates and annotates the code for the most basic FE model and provides an overview of the summary output.
-A number of potential extensions to the basic model, including relaxing various assumptions, dealing with measurement error in both the independent and dependent variables, as well as the inclusion of time-invariant predictors in the form of a hybrid fixed-/random-effects model, are shown in detail in the form of [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf).
+A number of potential extensions to the basic model, including relaxing various assumptions, dealing with measurement error in both the independent and dependent variables, as well as the inclusion of time-invariant predictors in the form of a hybrid fixed-/ random effects model, are shown in detail in the form of [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf).
 
@@
 \end{align}
-where $y_{it}$ is the dependent variable for unit $i, \ i = 1, ..., N$ at time $t, \ t = 1, ..., T$, $\bm{x}_{it}$ is a $1 \times K$ vector of time-varying covariates (which could include a constant) linked to the dependent variable by the $K \times 1$ vector of coefficients $\bm{\beta}$, $\bm{z}_{i}$ is a $1 \times M$ vector of time-invariant covariates linked to the dependent variable by the $M \times 1$ vector of coefficients in $\bm{\gamma}$, $\alpha_{i}$ represents the combined effect of all unobserved unit-specific variables affecting the dependent variable and $\varepsilon_{it}$ is the idiosyncratic error.
+where $y_{it}$ is the dependent variable for unit $i, \ i = 1, ..., N$ at time $t, \ t = 1, ..., T$, $\bm{x}_{it}$ is a $1 \times K$ vector of time-varying covariates (which could include a constant) linked to the dependent variable by the $K \times 1$ vector of coefficients $\bm{\beta}$.
+$\bm{z}_{i}$ is a $1 \times M$ vector of time-invariant covariates linked to the dependent variable by the $M \times 1$ vector of coefficients in $\bm{\gamma}$, $\alpha_{i}$ represents the combined effect of all unobserved unit-specific variables affecting the dependent variable and $\varepsilon_{it}$ is the idiosyncratic error.
 
@@
 - Linearity: the model is linear in its parameters and $\E[\varepsilon_{it}] = 0$ and $\E[\alpha_{i}] = 0$. 
-- Independence: the observations are independent across individuals, but not necessarily across time. 
-- Strict exogeneity: $\E[\varepsilon_{it} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] = 0$, i.e., the idiosyncratic error at each point in time is mean independent of $\bm{X}_{i}$ at \textit{all} points in time, $t = 1, ..., T$, as well as $\bm{z}_{i}$ and $\alpha_{i}$ .
+- Independence: the observations are independent across individuals (assured by random sampling in the cross-section), but not necessarily across time. 
+- Strict exogeneity: $\E[\varepsilon_{it} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] = 0$, i.e., the idiosyncratic error at each point in time is mean independent of $\bm{X}_{i}$ at \textit{all} points in time, $t = 1, ..., T$, as well as $\bm{z}_{i}$ and $\alpha_{i}$.
 - Error variance: homoscedastic and non-serially correlated errors, i.e., $\Var(\varepsilon_{i}|\bm{X}_{i},\bm{z}_{i}, \alpha_{i}) = \sigma^{2}_{\varepsilon} \bm{I}$.
@@
 
-If we are willing to make some assumptions about the distribution of the individual effects and their relationship with the observed time-varying covariates, we are left with the random effects model [@Wooldridge2002; @Wooldridge2012; @Schmidheiny2019; @Skrondal2004].
+If we are willing to make some assumptions about the distribution of the individual effects and, more importantly, their lack of a relationship with the observed time-varying covariates, we are left with the random effects model [@Wooldridge2002; @Wooldridge2012; @Schmidheiny2019; @Skrondal2004].
 Namely, we assume the individual effects are normally distributed, with a constant (homoscedastic) variance conditional on the model covariates, i.e., $\alpha_{i}|\bm{X}_{i}, \bm{z}_{i} \sim N(0, \sigma^{2}_{\alpha})$.^[As long as $\bm{X}_{i}$ contains a constant, the assumption that $\alpha_{i}$ has a mean of zero is unproblematic [@Wooldridge2012].] Further, we assume that the expectation of the individual effects conditional on the model covariates is zero, i.e., $\E[\alpha_{i}|\bm{X}_{i},\bm{z}_{i}] = 0$, which implies they are also uncorrelated.
@@
 \begin{align}
-y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \nu_{it}.
+y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \nu_{it}, \ \text{or} \\
+\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\nu}_{i}.
 \end{align}
@@
 \begin{align}
-\Var(\nu | \bm{X}, \bm{z}) & = \bm{\Omega}_{\nu} = 
+\Var(\bm{\nu} | \bm{X}, \bm{z}) & = \bm{\Omega}_{\bm{\nu}} = 
 \begin{pmatrix}
-\bm{\Omega}_{\nu 1} & \hdots & 0 & \hdots & 0 \\
+\bm{\Omega}_{\bm{\nu}_{1}} & \hdots & 0 & \hdots & 0 \\
 \vdots & \ddots & \vdots & \ddots & \vdots \\
-0 & \hdots & \bm{\Omega}_{\nu i} & \hdots & 0 \\
+0 & \hdots & \bm{\Omega}_{\bm{\nu}_{i}} & \hdots & 0 \\
 \vdots & \ddots & \vdots & \ddots & \vdots \\
-0 & \hdots & 0 & \hdots & \bm{\Omega}_{\nu N}
+0 & \hdots & 0 & \hdots & \bm{\Omega}_{\bm{\nu}_{N}}
 \end{pmatrix}
@@
 \begin{align}
-\Var(\nu_{i} | \bm{X}_{i}, \bm{z}_{i}) & = \bm{\Omega}_{\nu i} = 
+\Var(\bm{\nu}_{i} | \bm{X}_{i}, \bm{z}_{i}) & = \bm{\Omega}_{\bm{\nu}_{i}} = 
 \begin{pmatrix}
@@
 \sigma^{2}_{\alpha} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\nu}
-\end{pmatrix}
+\end{pmatrix} \label{eq:omeganui}
 \end{align}
@@
 
-What if the individual effects are \textit{note} independent of the observed time-varying covariates, i.e., $\E[\alpha_{i}|\bm{X}_{i}] \ne 0$?
+What if the individual effects are \textit{not} independent of the observed time-varying covariates, i.e., $\E[\alpha_{i}|\bm{X}_{i}] \ne 0$?
 Then grouping them in with the error will cause $\bm{\beta}$ to be biased [@Wooldridge2002; @Wooldridge2012].
@@
 (y_{it} - \bar{y}_{i}) & = (\bm{x}_{it} - \bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \bar{\bm{z}}_{i})\bm{\gamma} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
-\ddot{y}_{it} & = \bm{\ddot{x}}_{it}\bm{\beta} + \ddot{\varepsilon}_{it} \label{eq:fe}
+\ddot{y}_{it} & = \bm{\ddot{x}}_{it}\bm{\beta} + \ddot{\varepsilon}_{it}, \ \text{or} \\
+\bm{\ddot{y}}_{i} & = \bm{\ddot{X}}_{i}\bm{\beta} + \bm{\ddot{\varepsilon}}_{i} \label{eq:fe}
 \end{align}
@@
 
+## Modeling time-invariant unobserved heterogeneity as a latent variable
+
 We first need to convert the data from stacked, long-format vectors of length $NT$ into $T$ individual vectors of length $N$.
@@
 Because the idiosyncratic errors are assumed to be uncorrelated across units and across time, the covariance between any two of the new wide vectors $\Cov(y_{ti},y_{si}) = \Var(\alpha_{i}), \ t \ne s$.
-Otherwise, when $t = s$, the variance is $\Var(\alpha_{i}) + \Var(\varepsilon_{it})$.
-This is exactly the structure we saw above in a typical element of $\bm{\Omega}_{\nu}$.
+Otherwise, when $t = s$, the covariance $\Cov(y_{ti},y_{ti}) = \Var(\alpha_{i}) + \Var(\varepsilon_{ti})$.
+This is the structure we saw above in a typical element of $\bm{\Omega}_{\nu}$.
 
@@
 Here, the regression weights or 'factor loadings' are fixed to one to represent our assumption that the effect of the time-invariant unobserved heterogeneity is constant over time.
-It also means that the estimated variance of the latent variable is equal to the average covariance between the dependent variable over time.
-If $y_{it} = \alpha_{i} + \varepsilon_{it}$ is the true data generating process, then the expected covariances between all timepoints are constant and thus equal to the average covariance between all timepoints.
+It also means that the estimated variance of the latent variable is equal to the \textit{average covariance between the wide-format columns of the dependent variable over time}.
+If $y_{it} = \alpha_{i} + \varepsilon_{it}$ is the true data generating process, then the relationship between two units over time is just $\alpha$, regardless of the time distance.
+Referring back to the random effects structure of $\bm{\Omega}_{\bm{\nu}_{i}}$ in Equality \eqref{eq:omeganui} for a generic unit $i$, we see the covariance on all of the off-diagonals is $\sigma^{2}_{\alpha}$.
+And, as we know, the average of something that does not change is that thing itself.
+I.e., if $T(T-1)/2 = h$ is the number of elements on either the upper- or lower triangle of $\bm{\Omega}_{\bm{\nu}_{i}}$, then we have $h^{-1}\sum_{i=1}^{h}\sigma^{2}_{\alpha} = \frac{h \sigma^{2}_{\alpha}}{h} = \sigma^{2}_{\alpha}$.
 
-To show this, consider the following matrix equation of the nonredundant covariances in a three-wave intercept-only model that just restates Equation \eqref{eq:wide}, and which we can solve easily with least squares:
+To show this, consider the following matrix equation of the variances and the nonredundant covariances in a three-wave intercept-only model that follows directly from Equation \eqref{eq:wide} (assuming $\Cov(\varepsilon_{t1},\varepsilon_{si}) = 0, \ t \ne s$), and which we can solve easily with least squares:
 \begin{align}
@@
 \end{align}
-where $\psi = \Var(\alpha)$, $\phi_{t} = \Var(\varepsilon_{t})$. We can solve this equation like so
+where $\psi = \Var(\alpha)$, $\phi_{t} = \Var(\varepsilon_{t})$. We can solve this equation to show
 \begin{align}
-(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} = 
+\bm{b} & = \bm{A}\bm{x} \\
+(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{A}\bm{x} \\
+(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} & = \bm{x} \\
+\begin{pmatrix}
+.33\Cov(y_{2},y_{1}) + .33\Cov(y_{3},y_{1}) + .33\Cov(y_{3},y_{2}) \\
+\Var(y_{1}) - \psi \\
+\Var(y_{2}) - \psi \\
+\Var(y_{3}) - \psi
+\end{pmatrix} & = 
 \begin{pmatrix}
@@
 \phi_{3}
-\end{pmatrix} = 
-\begin{pmatrix}
-.33\Cov(y_{2},y_{1}) + .33\Cov(y_{3},y_{1}) + .33\Cov(y_{3},y_{2}) \\
-\Var(y_{1}) - \psi \\
-\Var(y_{2}) - \psi \\
-\Var(y_{3}) - \psi
 \end{pmatrix}.
 \end{align}
-In other words, $\psi = .33\Cov(y_{2},y_{1}) + .33\Cov(y_{3},y_{1}) + .33\Cov(y_{3},y_{2})$ is the same as $\psi = \frac{1}{3} 3\Cov(y_{t},y_{s}) = \Cov(y_{t},y_{s})$ where $\Cov(y_{t},y_{s})$ is the covariance between any two columns of the wide-format dependent variable.
 This shows that if our assumption of constant covariances between the dependent variable across time holds, the estimated variance of $\alpha$ is just what it should be: the average covariance between units of $y$ over time.
 Once we add in observed covariates, the estimated covariance of $\alpha$ then become the \textit{conditional} covariance of $y$ over time, given those covariates.
+
+## Model notation
 
@@
 
+## Assumptions
+
 What essentially differentiates an FE from a random effects (RE) model is our assumption concerning the relationship between the unobserved individual effects and the model covariates [@Bollen2010].
@@
 First, we define the latent individual effects variable using the `=~` 'measured by' or 'manifested by' [@R-lavaan] operator at the same time constraining the factor loadings at each timepoint to one.
-I will call the latent variable `a` to stand for $\alpha$:
-
+I will call the latent variable `a` to stand for $\alpha$.
 Constraining all of the factor loadings to one reflects our implicit assumption that the combined effect of the unit-specific unobserved factors is constant over time.
@@
 We do so by adding the same label to the regression coefficient at every time point.
-We will use the label `b` (this label was chosen arbitrarily, we could have used any letter) and have it act as an equality constraint for the regression coefficient of interest $\beta$:
+We will use the label `b` (this label was chosen arbitrarily, we could have used any letter or string of characters) and have it act as an equality constraint for the regression coefficient of interest $\beta$:
 
@@
 \end{align}
-where, for the sake of simplicity, $\bm{\alpha_{1}}$, $\bm{\alpha_{2}}$, $\bm{\delta_{t}}$, $\sim N(\mu_{\delta}, \sigma_{\delta})$ and $\bm{\varepsilon_{t}}$ are $\sim N(0,1)$.
+where, for the sake of simplicity, $\bm{\alpha_{1}}$, $\bm{\alpha_{2}}$, $\bm{\delta_{t}}$ and $\bm{\varepsilon_{t}}$ are $\sim N(0,1)$.
 
@@
 
-The most basic FE-SEM is the basis for a variety of currently popular extended models, such as Latent Curve Models in general [@Curran2001; @Bollen2004], as well as special implementations like the Dynamic Panel Model [@Allison2017], the Random-Intercept Cross-Lagged Panel Model [@Hamaker2015] and the Latent Curve Model with Structured Residuals [@Curran2014].
-For this reason it is all the more important for researchers to have a consolidated guide to a variety of common applications of fixed effects in SEM.
+The most basic FE-SEM is furthermore the basis for a variety of currently popular extended models, such as Latent Curve Models in general [@Curran2001; @Bollen2004], as well as special implementations like the Dynamic Panel Model [@Allison2017], the Random-Intercept Cross-Lagged Panel Model [@Hamaker2015] and the Latent Curve Model with Structured Residuals [@Curran2014].
+For this reason it is all the more important for researchers to have a good grasp on the method of applying FE-SEM.
+This article is meant to serve as a consolidated resource for researchers looking for concrete advise on specifying FE and more general panel models in SEM.
 

### 2020-07-09 extensions.Rmd
Made minor edits to the text

@@
 First, I go over a number of possibilities to relax assumptions associated with the traditional FE model.
-Then, I discuss the issue of measurement error and show how we can use latent variables to deal with it and allow us to properly estimate the coefficients of interest.
-Then, I show display a type of hybrid FE/RE model that allows us to control for time-invariant unobserved heterogeneity that allows us to include time-invariant predictors in the model.
+Then, I discuss the issue of measurement error and show how we can use latent variables to deal with it and properly estimate the coefficients of interest.
+Then, I show a type of hybrid FE/RE model that allows us to control for time-invariant unobserved heterogeneity while including time-invariant predictors in the model.
 
@@
 There are a number of implicit assumptions attached to the typical FE model that can be relaxed in SEM.
-Some of these assumptions have been discussed already, and a full list of assumptions can be found in @Bollen2010.
+Some of these assumptions have been discussed already, and a fairly comprehenisve list of assumptions can be found in @Bollen2010.
 Here, I will go over just a few, concentrating on the implementation in `lavaan` and the opportunity to empirically test whether the adjustments are justified or not.
@@
 We can furthermore easily relax the assumption of time-constant error variance, i.e., $\sigma^{2}_{\varepsilon_{t}}$.
-As noted above, the assumption regarding $\E[\alpha x_{t}]$ in $\bm{\Psi}$ determines whether we have an FE or RE model.
+As noted in the [main article](https://github.com/henrik-andersen/FE-SEM/blob/master/article.pdf), the assumption regarding $\E[\alpha x_{t}]$ in $\bm{\Psi}$ determines whether we have an FE or RE model.
 We can set these to zero and test whether the RE model would be preferable to the FE model.
-In general, if the individual effects are truly uncorrelated with the model covariates, it is advisable to switch to an RE model since because it uses up less degrees of freedom, it will have smaller asymptotic standard errors [@Bollen2010].
+In general, if the individual effects are truly uncorrelated with the model covariates, it is advisable to switch to an RE model since because it uses up less degrees of freedom, it will have smaller standard errors [@Bollen2010].
 
@@
 
-As outlined in the @Bollen2010 text, the researcher has the opportunity to test each of the assumptions empirically and decide whether a more parsimonious, i.e., restrictive model is justifiable.
+As outlined in @Bollen2010, the researcher has the opportunity to test each of the assumptions empirically and decide whether a more parsimonious, i.e., restrictive model is justifiable.
 For each assumption, a likelihood ratio test can be carried out to determine whether the improvement to model fit resulting from the relaxation of various assumptions is significant or whether the more parsimonious model is preferable after all.
 
-If we use the original model `fe_sem.fit` as a starting point, the best strategy for testing these assumptions is to work in a stepwise fashion, relaxing one assumption at a time.
+If we use the original model `fe_sem.fit` (from the [main article](https://github.com/henrik-andersen/FE-SEM/blob/master/article.pdf)) as a starting point, the best strategy for testing these assumptions is to work in a stepwise fashion, relaxing one assumption at a time.
 We can begin by first constraining the correlation between $\alpha$ and $x_{t}$ to zero (`re_sem`) for an RE model.
@@
 
-To demonstrate the effect of measurement error on the FE-SEM model, and then provide a strategy for dealing with measurement error in SEM, let us first modify the simulated dataset by generating multiple *indicators* of the independent and dependent variables that all measure the intended variable imprecisely.
+To demonstrate the effect of measurement error on the FE-SEM model, and then provide a strategy for dealing with measurement error in SEM, the [simulated dataset](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R) generates multiple *indicators* of the independent and dependent variables that all measure the intended variable imprecisely.
 Returning to our panel data, we have three indicators of each the independent and dependent variable, per timepoint:
@@
 over the course of a panel study.
-To create the observed indicators, a random amount of measurement error (ranging from $\{\sigma^{2}_{\upsilon_{k}}, \sigma^{2}_{\nu_{k}}\}\in \{1.0, 1.1, 1.2, 1.3, 1.4, 1.5\}$) was added to the true variables.
-The code for generating impercise indicators can be found in Appendix B.
+To create the observed indicators, a random amount of measurement error (ranging from $\{\sigma^{2}_{\upsilon_{k}}, \sigma^{2}_{\nu_{k}}\}\in \{1.0, 1.1, 1.2, 1.3, 1.4, 1.5\}$) was added to the true variables, again see [the simulation code](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R).
 
 Let us first focus on the issue of imprecise measurements of the independent variable of interest and run the same FE-SEM model above, but this time we will use one of the measurement error sullied indicators, here $\tilde{x}_{1t}$, instead of the true independent variable, $x_{t}$.
+As for the naming conventions in the `R` code, `x11` stands for the first indicator ($k = 1$) at the first point in time ($t = 1$), whereas for example `x35` stands for the third indicator ($k = 3$) at the fifth point in time ($t = 5$).
 
@@
 where $\zeta_{t}$ represents the disturbance, in other words the residual of the latent dependent variable $\eta_{t}$. 
-
 First, however, let us double-check that measurement error in the dependent variable only increases the error variance (thus also increasing standard errors and reducing $R^{2}$), but does not systematically bias the coefficients of interest.
-The next model uses the indicators of $x$ and specifies latent variables to represent the valid cross-sectional variance.
+The next model uses the indicators of $x$ and specifies latent variables ($\xi_{t}$, `xi` in the code) to represent the valid cross-sectional variance.
 The dependent variable in the model is one of the impercisely measured indicators of $y$.
@@
 
-Finally, to see the benefits of removing measurement error from the dependent variable in terms of standard errors and $R^{2}$ statistics, we can specify a model with latent variables representing the valid cross-sectional variance in $y$.
+Finally, to see the benefits of removing measurement error from the dependent variable in terms of standard errors and $R^{2}$ statistics, we can specify a model with latent variables representing the valid cross-sectional variance in $y$ (`n` for $\eta$ in the code).
 
@@
 
-Generally, the use of a latent variable to account for measurement error in the dependent variable will help smooth out chance variation in the estimated coefficient and increase model efficiency, i.e., make the estimated coefficient closer to the true parameter more often.
-The standard errors will tend to be smaller compared to a model in which the dependent variable still contains measurement error (e.g., `fe_sem4.fit`: `r round( lavInspect( fe_sem4.fit, "list")[ 36, 15], 3)` vs.
-`fe_sem3.fit`: `r round( lavInspect( fe_sem3.fit, "list")[ 21, 15], 3)`) and the model fit will be better.
+Here, the effect $\hat{\beta}_{\eta_{t},\xi_{t}}$ is somewhat further off of the true effect of 0.3 than the preceding models.
+This will depend on how the latent variables are estimated, which themselves will depend on the underlying correlations between the indicators.
+Again, if the main goal of the model is to avoid bias, it may be advisable to just leave the manifest dependent variable as it is, and worry about measurement error in the independent variables.
 
@@
 It is important, however, to realize that the unbiasedness of $\hat{\gamma}$ in this model is dependent on the assumption that $\E[\zeta | \bm{\xi_{t}}, \alpha_{2}] = 0$.
-In other words, the idiosyncratic error is mean independent of $\bm{\xi_{t}} = \xi_{1}, \xi_{2}, ..., \xi_{T}$ as well as $\alpha_{2}$.
+In other words, the idiosyncratic error is mean independent of $\bm{\xi_{t}} = (\xi_{1}, \xi_{2}, ..., \xi_{T})$ as well as $\alpha_{2}$.
 The first part is easier to accept because we are controlling for all potential time-invariant confounders that could induce a relationship between the independent variable and the error.
 The unbiasedness of $\hat{\gamma}$, on the other hand rests on the assumption that the time-invariant predictor is independent of the error.
-If $\alpha_{2}$ represented the respondent's intelligence and $\eta_{t}$ represented the respondent's income, for example, then $\hat{\gamma}$ would be biased if both were dependent on a third time-invariant variable, say level of schooling.
+If $\alpha_{2}$ represented the respondent's intelligence and $\eta_{t}$, the dependent variable, represented the respondent's income, for example, then $\hat{\gamma}$ would be biased if both were dependent on a third time-invariant variable, say level of schooling, if it is not controlled for.
 For this reason, we need to treat the regression on a time-invariant predictor like any other regular multivariate regression model and look to include all plausible potential confounders as controls in the model, or turn to other methods, e.g., instrumental variables.

### 2020-07-09 article.Rmd
Made some minor edits

@@
 
-To show this, consider the following matrix equation of the variances and the nonredundant covariances in a three-wave intercept-only model that follows directly from Equation \eqref{eq:wide} (assuming $\Cov(\varepsilon_{t1},\varepsilon_{si}) = 0, \ t \ne s$), and which we can solve easily with least squares:
+To show this, consider the following matrix equation of the variances and the nonredundant covariances in a three-wave intercept-only model that follows directly from Equation \eqref{eq:wide} (assuming $\Cov(\varepsilon_{ti},\varepsilon_{si}) = 0, \ t \ne s$), and which we can solve easily with least squares:
 \begin{align}
@@
  \begin{block}{c(cccc)}
- y_{1}^{2}  & 1 & 1 & 0 & 0  \\
- y_{2}y_{1} & 1 & 0 & 0 & 0  \\
- y_{3}y_{1} & 1 & 0 & 0 & 0  \\ 
- y_{2}^{2}  & 1 & 0 & 1 & 0  \\
- y_{3}y_{2} & 1 & 0 & 0 & 0  \\
- y_{3}^{2}  & 1 & 0 & 0 & 1  \\
+ \Var(y_{1})  & 1 & 1 & 0 & 0  \\
+ \Cov(y_{2},y_{1}) & 1 & 0 & 0 & 0  \\
+ \Cov(y_{3},y_{1}) & 1 & 0 & 0 & 0  \\ 
+ \Var(y_{2})  & 1 & 0 & 1 & 0  \\
+ \Cov(y_{3},y_{2}) & 1 & 0 & 0 & 0  \\
+ \Var(y_{3})  & 1 & 0 & 0 & 1  \\
  \end{block}
@@
 \end{align} 
-Notice in $\bm{\zeta}^{+}$, for the independent variables we could either write, for example $x_{1}$ or $\delta_{1}$.
+Notice in $\bm{\zeta}^{+}$, for the independent variables we could either write, for example $x_{t}$ or $\delta_{t}$.
 As mentioned above, this is due to the model notation treating the independent variables like dependent variables with variances/covariances to be estimated.
@@
 The key to a FE model, as opposed to an RE model are our assumptions about the relatedness of our covariate and the individual effects, i.e., $\E[x_{t}\alpha]$.
-For a FE model, we want to partial out any potential covariance between the independent variable and the individual effects.
+For an FE model, we want to partial out any potential covariance between the independent variable and the individual effects.
 This accounts for any linear relationship between $x_{t}$ and the unit-specific characteristics influencing the dependent variable.

### 2020-07-22 article.Rmd
Made some major revisions to the text

@@
 \\par  \\textbf{Keywords:} Fixed effects, structural equation modeling, lavaan, R, panel analysis"
+note: "Preprint version, please do not cite without permission."
 date: "`r Sys.Date()`"
@@
 Fixed effects (FE) regression involves controlling for these time-invariant influences via a number of various methods.
-It thus drastically reduces the number of potential confounders of the relationship between variables of interest.
+It thus accounts for a likely and common source of bias.
 
@@
 While `Mplus` [@Mplus] is arguably the most robust SEM software currently available (in terms of features like alignment, latent variable interactions, for example) , the `lavaan` package has many benefits.
-First, it and `R` are open source and completely free.
+First, like `R` it is open source and completely free.
 For researchers dipping their toes into SEM, there is no financial barrier to try, and no risk if they decide it is not for them.
@@
 Using [simulated data](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R), it demonstrates and annotates the code for the most basic FE model and provides an overview of the summary output.
-A number of potential extensions to the basic model, including relaxing various assumptions, dealing with measurement error in both the independent and dependent variables, as well as the inclusion of time-invariant predictors in the form of a hybrid fixed-/ random effects model, are shown in detail in the form of [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf).
+One of the main strengths of panel SEM compared to the more traditional methods of panel analysis is its flexibility.
+Therefore, a number of potential extensions to the basic model, including relaxing various assumptions, dealing with measurement error in both the independent and dependent variables, as well as the inclusion of time-invariant predictors in the form of a hybrid fixed-/ random effects model, are shown in detail in the form of [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf).
 
@@
 
-To begin, let us start by reviewing a general panel model [@Bollen2010], also referred to as the 'unobserved effects model' [@Wooldridge2012] (we will return to this model in the [online section](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf) of the article when we discuss loosening assumptions)
+To begin, let us start by reviewing a general panel model [@Bollen2010], also referred to as the 'unobserved effects model' [@Wooldridge2012; @R-plm_a] (we will return to this model in the [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf) of the article when we discuss loosening assumptions)
 \begin{align}
@@
 where $y_{it}$ is the dependent variable for unit $i, \ i = 1, ..., N$ at time $t, \ t = 1, ..., T$, $\bm{x}_{it}$ is a $1 \times K$ vector of time-varying covariates (which could include a constant) linked to the dependent variable by the $K \times 1$ vector of coefficients $\bm{\beta}$.
-$\bm{z}_{i}$ is a $1 \times M$ vector of time-invariant covariates linked to the dependent variable by the $M \times 1$ vector of coefficients in $\bm{\gamma}$, $\alpha_{i}$ represents the combined effect of all unobserved unit-specific variables affecting the dependent variable and $\varepsilon_{it}$ is the idiosyncratic error.
+$\bm{z}_{i}$ is a $1 \times M$ vector of time-invariant covariates linked to the dependent variable by the $M \times 1$ vector of coefficients in $\bm{\gamma}$, $\alpha_{i}$ represents the combined effect of all unobserved time-constant variables affecting the dependent variable and $\varepsilon_{it}$ is the idiosyncratic error.
 
@@
 
-We typically make the following assumptions [see, e.g., @Wooldridge2002; @Schmidheiny2019] about this model:
+Consistency of the following models requires the assumption of strict exogeneity, although what constitutes strict exogeneity differs between the random and fixed effects setups.
+Each assumption will be discussed shortly.
+Apart from that, we typically make the following assumptions [see, e.g., @Wooldridge2002; @Schmidheiny2019] about this model:
 
-- Linearity: the model is linear in its parameters and $\E[\varepsilon_{it}] = 0$ and $\E[\alpha_{i}] = 0$. 
+- Linearity: the model is linear in its parameters. 
 - Independence: the observations are independent across individuals (assured by random sampling in the cross-section), but not necessarily across time. 
-- Strict exogeneity: $\E[\varepsilon_{it} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] = 0$, i.e., the idiosyncratic error at each point in time is mean independent of $\bm{X}_{i}$ at \textit{all} points in time, $t = 1, ..., T$, as well as $\bm{z}_{i}$ and $\alpha_{i}$.
-- Error variance: homoscedastic and non-serially correlated errors, i.e., $\Var(\varepsilon_{i}|\bm{X}_{i},\bm{z}_{i}, \alpha_{i}) = \sigma^{2}_{\varepsilon} \bm{I}$.
+- The usual rank condition: we have more observations than independent variables and there is no perfect collinearity between any of the independent variables.
 
@@
 
-If we are willing to make some assumptions about the distribution of the individual effects and, more importantly, their lack of a relationship with the observed time-varying covariates, we are left with the random effects model [@Wooldridge2002; @Wooldridge2012; @Schmidheiny2019; @Skrondal2004].
-Namely, we assume the individual effects are normally distributed, with a constant (homoscedastic) variance conditional on the model covariates, i.e., $\alpha_{i}|\bm{X}_{i}, \bm{z}_{i} \sim N(0, \sigma^{2}_{\alpha})$.^[As long as $\bm{X}_{i}$ contains a constant, the assumption that $\alpha_{i}$ has a mean of zero is unproblematic [@Wooldridge2012].] Further, we assume that the expectation of the individual effects conditional on the model covariates is zero, i.e., $\E[\alpha_{i}|\bm{X}_{i},\bm{z}_{i}] = 0$, which implies they are also uncorrelated.
-
-The random effects model works by defining a composite error term: $\nu_{it} = \alpha_{i} + \varepsilon_{it}$ and rewriting the model in Equation \eqref{eq:gpm} as 
+For the random effects (RE) model, we define a composite error term: $\nu_{it} = \alpha_{i} + \varepsilon_{it}$ and rewrite the model in Equation \eqref{eq:gpm} as 
 \begin{align}
 y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \nu_{it}, \ \text{or} \\
-\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\nu}_{i}.
+\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\nu}_{i},
 \end{align}
-By adding the individual effects to the composite error in each time period, the composite error becomes serially correlated over time, to give the following conditional covariance matrix for the errors,
+where $\bm{\nu}_{i} = \alpha_{i} \bm{\iota}_{T} + \bm{\varepsilon}_{i}$ and $\bm{\iota}_{T}$ is a $T \times 1$ vector of ones. The strict exogeneity assumption in the RE model implies
 \begin{align}
-\Var(\bm{\nu} | \bm{X}, \bm{z}) & = \bm{\Omega}_{\bm{\nu}} = 
+\E[\varepsilon_{it} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = 0, \\
+\E[\alpha_{i} |\bm{X}_{i}, \bm{z}_{i}] = \E[\alpha_{i}] & = 0,
+\end{align}
+where $\bm{X}_{i} = \bm{x}_{i1}, ..., \bm{x}_{iT}$.
+For both parts, the assumption that the unconditional expectations are 0 is unproblematic as long as a constant is included in the regression.
+The first part says the idiosyncratic errors at each timepoint are assumed to be independent of the explanatory variables at \textit{all} timepoints which is stronger than just assuming that they are \textit{contemporaneously} independent.
+This implies that they are also uncorrelated, i.e., $\E[\bm{x}_{is}^{\intercal}\varepsilon_{it}] = \bm{0}$ and $\E[\bm{z}_{i}^{\intercal}\varepsilon_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T$ [@Wooldridge2002; @Bruederl2015].
+We assume the idiosyncratic errors are further independent of the individual effects, which implies $\E[\alpha_{i}\varepsilon_{it}] = 0$.
+
+The second part is the potentially controversial assumption: it states that the individual effects are uncorrelated with the independent variables.
+We can use an example to show why this is often controversial: If we are interested in the question of whether married men earn more than unmarried men, then the second part of the strict exogeneity assumption means that a man's marriage status would have to be uncorrelated with all the time-invariant characteristics that could potentially make that man an attractive marriage candidate in the first place; e.g., looks, personality, family's status, profession, etc.
+[@Bruederl2015].^[Assuming, for the sake of argument, that these characteristics are constant over time.]
+
+From what we have discussed so far, the $T \times T$ covariance matrix of the errors $\bm{\Omega}_{i} = \E[\bm{\nu}_{i}\bm{\nu}_{i}^{\intercal}]$ can be constructed.
+However, the standard random effects model adds the additional assumptions
+\begin{alignat}{3}
+\E[\varepsilon_{it}^{2} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \E[\varepsilon_{it}^{2}] && = \sigma_{\varepsilon}^{2}, && \ t = 1, ..., T, \\
+\E[\varepsilon_{it}\varepsilon_{is}| \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \E[\varepsilon_{it}\varepsilon_{is}] && = 0, && \ \forall \ t \ne s 
+\end{alignat}
+i.e., the idiosyncratic errors are conditionally homoscedastic and serially uncorrelated and 
+\begin{alignat}{2}
+\E[\alpha_{i}^{2}| \bm{X}_{i},\bm{z}_{i}] & = \E[\alpha_{i}^{2}] && = \sigma^{2}_{\alpha}
+\end{alignat}
+i.e., the individual effects are conditionally homoscedastic (they are necessarily serially correlated as long as $\sigma_{\alpha}^{2} > 0$) we arrive at the typical random effects structure of the $NT \times NT$ matrix $\bm{\Omega}$:
+
+\begin{align}
+\bm{\Omega} & = 
 \begin{pmatrix}
-\bm{\Omega}_{\bm{\nu}_{1}} & \hdots & 0 & \hdots & 0 \\
+\bm{\Omega}_{1} & \hdots & 0 & \hdots & 0 \\
 \vdots & \ddots & \vdots & \ddots & \vdots \\
-0 & \hdots & \bm{\Omega}_{\bm{\nu}_{i}} & \hdots & 0 \\
+0 & \hdots & \bm{\Omega}_{i} & \hdots & 0 \\
 \vdots & \ddots & \vdots & \ddots & \vdots \\
-0 & \hdots & 0 & \hdots & \bm{\Omega}_{\bm{\nu}_{N}}
+0 & \hdots & 0 & \hdots & \bm{\Omega}_{N}
 \end{pmatrix}
 \end{align}
-with typical elements 
+with $T \times T$ typical elements 
 \begin{align}
-\Var(\bm{\nu}_{i} | \bm{X}_{i}, \bm{z}_{i}) & = \bm{\Omega}_{\bm{\nu}_{i}} = 
+\bm{\Omega}_{i} & = 
 \begin{pmatrix}
@@
 \end{align}
-where $\sigma^{2}_{\nu} = \sigma^{2}_{\alpha} + \sigma^{2}_{\varepsilon}$.
+where $\sigma^{2}_{\nu} = \sigma^{2}_{\alpha} + \sigma^{2}_{\varepsilon}$ [@Wooldridge2002; @Schmidheiny2019].
 This means that in the conditional covariance matrix of the errors, given the time-varying and -invariant covariates, units over time will be correlated due to the individual effects.
@@
 
+Estimation of the RE model can be done using feasible generalized least squares (GLS) in which the two unknowns in $\bm{\Omega}$, $\sigma_{\alpha}^{2}$ and $\sigma_{\nu}^{2}$, are first estimated using pooled ordinary least squares (pooled OLS or POLS), where^[Normally a degrees-of-freedom correction is applied by subtracting off the number of independent variables that is negligible in large N samples.
+It is ignored here for the sake of simplicity.]
+\begin{align}
+\hat{\sigma}_{\nu}^{2} & = \frac{1}{NT}\sum_{i = 1}^{N}\sum_{t = 1}^{T}\hat{\nu}_{it}^{2}, \\
+\hat{\sigma}_{\varepsilon}^{2} & = \frac{1}{NT - N}\sum_{i = 1}^{N}\sum_{t = 1}^{T}(\hat{\nu}_{it} - \bar{\hat{\nu}}_{i})^{2}
+\end{align}
+and $\hat{\nu}_{it} = y_{it} - \bm{x}_{it}\hat{\bm{\beta}}_{\tiny{POLS}} - \bm{z}_{i}\hat{\bm{\gamma}}_{\tiny{POLS}}$, $\bar{\hat{\nu}}_{i} = T^{-1}\sum_{t = 1}^{T}\hat{\nu}_{it}$ and $\hat{\sigma}^{2}_{\alpha} = \hat{\sigma}^{2}_{\nu} - \hat{\sigma}^{2}_{\varepsilon}$.
+Then, the coefficients are estimated using those estimates in the variance matrix $\hat{\bm{\Omega}}$:
+\begin{align}
+\begin{pmatrix}
+\hat{\bm{\beta}}_{RE} \\
+\hat{\bm{\gamma}}_{RE}
+\end{pmatrix} & = 
+(\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{W})^{-1}\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{y}
+\end{align}
+where $\bm{W} = \begin{pmatrix}\bm{X} & \bm{Z}\end{pmatrix}$ and $\bm{X}$ is $NT \times K$ and $\bm{Z}$ is $NT \times M$ and $\bm{y}$ is $NT \times 1$.
+
+In practice, however, computational problems can arise with large cross-sectional samples, where it can become difficult to invert the $\hat{\bm{\Omega}}$ matrix.
+One solution is to use 'partial-demeaning' to transform the data before performing simple OLS:
+\begin{align}
+(y_{it} - \theta\bar{y}_{i}) & = (\bm{x}_{it} - \theta\bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \theta \bar{\bm{z}}_{i})\bm{\gamma} + (\varepsilon_{it} - \theta \bar{\varepsilon}_{i}) \label{eq:partialdemean}
+\end{align} 
+where $\theta = 1 - [\sigma_{\alpha}^{2}/(\sigma_{\alpha}^{2} + T \sigma_{\varepsilon}^{2})]^{1/2}$, and $\bar{y}_{i} = T^{-1}\sum_{t = 1}^{T}y_{it}$, $\bar{\bm{x}}_{i} = T^{-1}\sum_{t = 1}^{T}\bm{x}_{it}$, $\bar{\bm{z}}_{i} = T^{-1}\sum_{t = 1}^{T}\bm{z}_{i}$ and $\bar{\varepsilon}_{i} = T^{-1}\sum_{t = 1}^{T}\varepsilon_{it}$ [@R-plm_a].
+
+The RE model can also be estimated in the maximum likelihood framework, where in the associated literature panel models are generally referred to as either mixed models, hierarchical models or longitudinal models.
+The typical RE model discussed here is the equivalent to a mixed model with random intercepts and fixed slopes [@R-plm_a].
+Under the assumption of normality, along with homoscedasticity and serially uncorrelated errors, the maximum likelihood estimator is the same as the OLS estimator.
+For more on the topic of mixed models, see for example @R-lme4, @Bates2010.
+
 ## Fixed effects {#fe}
 
-What if the individual effects are \textit{not} independent of the observed time-varying covariates, i.e., $\E[\alpha_{i}|\bm{X}_{i}] \ne 0$?
-Then grouping them in with the error will cause $\bm{\beta}$ to be biased [@Wooldridge2002; @Wooldridge2012].
+For fixed effects, we assume that the individual effects are \textit{not independent} of the model covariates, i.e., $\E[\alpha_{i}|\bm{X}_{i},\bm{z}_{i}] \ne \E[\alpha_{i}] \ne 0$.
+Under this assumption, grouping the individual effects in with the composite error will cause the coefficients of interest, here specifically $\bm{\beta}$ to be inconsistent [@Wooldridge2002; @Wooldridge2012].
+We write the model therefore again in terms of the general panel model in Equation \eqref{eq:gpm} with separate individual effects and idiosyncratic error terms.
 In order to drop assumptions involving the individual effects, a number of methods are available (e.g., differencing, least squares dummy variable regression), but the most common approach is to \textit{demean} the equation [@Bruederl2015].
-Demeaning involves subtracting the per-unit, over-time average from each of the model terms, i.e.,
+Demeaning is the same as the transformation applied in Equation \eqref{eq:partialdemean} in the special case where $\theta = 1$.
+I.e., demeaning involves subtracting the per-unit, over-time average from each of the model terms, i.e.,
 \begin{align}
@@
 \end{align}
-where, for example, $\bar{y}_{i} = T^{-1}\sum_{i = 1}^{T}y_{it}$ (and the other over-time means are calculated analogously), and the variables with the dots above them represent the demeaned versions.
+where the over-time averages are calculated the same as above, and the variables with the dots above them represent the demeaned versions. 
 Because the average of something that does not change is that thing itself, the individual effects, along with any time-invariant predictors, get wiped out by the demeaning.
 This means that no assumptions about the relatedness of the model covariates and the unit-specific portion of the error are needed.
-The unbiasedness of the estimate is related solely to the strict exogeneity assumption imposed on the idiosyncratic errors, i.e., $\E[\ddot{\varepsilon}_{it}|\bm{\ddot{x}}_{it}] = \E[\ddot{\varepsilon}_{it}] = 0$ which also implies $\E[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T$ [@Bruederl2015; @Wooldridge2002].
+Consistency of the estimates is related solely to the strict exogeneity assumption imposed on the idiosyncratic errors, i.e., $\E[\ddot{\varepsilon}_{it}|\bm{\ddot{x}}_{it}] = \E[\ddot{\varepsilon}_{it}] = 0$ which also implies $\E[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T$ [@Bruederl2015; @Wooldridge2002].
+
+Having demeaned the data, the typical FE estimator is POLS on the transformed data
+\begin{align}
+\bm{\beta}_{FE} & = (\ddot{\bm{X}}^{\intercal}\ddot{\bm{X}})^{-1}\ddot{\bm{X}}^{\intercal}\ddot{\bm{y}}
+\end{align}
+[@Bruederl2015].
 The downside to this approach is that no time-invariant predictors can be included in the model.
-In the [online section](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf), it will be discussed how to get around this restriction using SEM.
+However, there are alternative approaches in the random effects and mixed model frameworks that allow them to be included.
+These models are sometimes referred to as 'within-between' or 'hybrid' models, often based on the Chamberlain -@Chamberlain1980 and Mundlak -@Mundlak1978 approaches, see for example @Bell2018; @Allison2011; @Schunck2013; @Enders2007.
+In the [online appendix](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf), it will be discussed how to also get around this restriction using SEM.
 
@@
 To see why this is necessary, consider what effect this has on the vector of responses $y_{it}$.
-Let us, for a minute ignore any covariates and focus just on the dependent variable (a so-called 'intercept-only' model) so that we have $y_{it} = \alpha_{i} + \varepsilon_{it}$.
+Let us, for a minute ignore any covariates and focus just on the dependent variable (a so-called 'intercept-only' or 'null' model) so that we have $y_{it} = \alpha_{i} + \varepsilon_{it}$.
 When we convert the data to wide-format, we get $T$ individual equations,
 \begin{align}
-y_{1i} & = \alpha_{i} + \varepsilon_{1i} \\
-y_{2i} & = \alpha_{i} + \varepsilon_{2i} \\
- & \vdots \\
-y_{Ti} & = \alpha_{i} + \varepsilon_{Ti}. \label{eq:wide}
+\bm{y}_{t} & = \bm{\alpha} + \bm{\varepsilon}_{t} \\
+\begin{bmatrix}
+y_{1t} \\
+y_{2t} \\
+\vdots \\
+y_{Nt}\end{bmatrix} & = 
+\begin{bmatrix}
+\alpha_{1} \\
+\alpha_{2} \\
+\vdots \\
+\alpha_{N}
+\end{bmatrix} + 
+\begin{bmatrix}
+\varepsilon_{1t} \\
+\varepsilon_{2t} \\
+\vdots \\
+\varepsilon_{Nt}
+\end{bmatrix} \label{eq:wide}
 \end{align}
+for each $t = 1, 2, ..., T$.
 Because the idiosyncratic errors are assumed to be uncorrelated across units and across time, the covariance between any two of the new wide vectors $\Cov(y_{ti},y_{si}) = \Var(\alpha_{i}), \ t \ne s$.
 Otherwise, when $t = s$, the covariance $\Cov(y_{ti},y_{ti}) = \Var(\alpha_{i}) + \Var(\varepsilon_{ti})$.
-This is the structure we saw above in a typical element of $\bm{\Omega}_{\nu}$.
+This is the structure we saw above in a typical element of $\bm{\Omega}$.
 
@@
 If $y_{it} = \alpha_{i} + \varepsilon_{it}$ is the true data generating process, then the relationship between two units over time is just $\alpha$, regardless of the time distance.
-Referring back to the random effects structure of $\bm{\Omega}_{\bm{\nu}_{i}}$ in Equality \eqref{eq:omeganui} for a generic unit $i$, we see the covariance on all of the off-diagonals is $\sigma^{2}_{\alpha}$.
+Referring back to the random effects structure of $\bm{\Omega}_{i}$ in Equality \eqref{eq:omeganui} for a generic unit $i$, we see the covariance on all of the off-diagonals is $\sigma^{2}_{\alpha}$.
 And, as we know, the average of something that does not change is that thing itself.
-I.e., if $T(T-1)/2 = h$ is the number of elements on either the upper- or lower triangle of $\bm{\Omega}_{\bm{\nu}_{i}}$, then we have $h^{-1}\sum_{i=1}^{h}\sigma^{2}_{\alpha} = \frac{h \sigma^{2}_{\alpha}}{h} = \sigma^{2}_{\alpha}$.
+I.e., if $T(T-1)/2 = h$ is the number of elements on either the upper- or lower triangle of $\bm{\Omega}_{i}$, then we have $h^{-1}\sum_{i=1}^{h}\sigma^{2}_{\alpha} = \frac{h \sigma^{2}_{\alpha}}{h} = \sigma^{2}_{\alpha}$.
 
@@
 \begin{align}
-\bm{b} & = \bm{A}\bm{x} \\
+\bm{A}\bm{x} & = \bm{b} \\
+\begin{pmatrix}
+1 & 1 & 0 & 0 \\
+1 & 0 & 0 & 0 \\
+1 & 0 & 0 & 0 \\
+1 & 0 & 1 & 0 \\
+1 & 0 & 0 & 0 \\
+1 & 0 & 0 & 1
+\end{pmatrix}
+\begin{pmatrix}
+\psi \\
+\phi_{1} \\
+\phi_{2} \\
+\phi_{3}
+\end{pmatrix} & = 
 \begin{pmatrix}
@@
 \Var(y_{3})
-\end{pmatrix} & = 
-\begin{blockarray}{ccccc}
- & \psi & \phi_{1} & \phi_{2} & \phi_{3}  \\
- \begin{block}{c(cccc)}
- \Var(y_{1})  & 1 & 1 & 0 & 0  \\
- \Cov(y_{2},y_{1}) & 1 & 0 & 0 & 0  \\
- \Cov(y_{3},y_{1}) & 1 & 0 & 0 & 0  \\ 
- \Var(y_{2})  & 1 & 0 & 1 & 0  \\
- \Cov(y_{3},y_{2}) & 1 & 0 & 0 & 0  \\
- \Var(y_{3})  & 1 & 0 & 0 & 1  \\
- \end{block}
-\end{blockarray}
-\begin{pmatrix}
-\psi \\
-\phi_{1} \\
-\phi_{2} \\
-\phi_{3}
 \end{pmatrix}
@@
 \begin{align}
-\bm{b} & = \bm{A}\bm{x} \\
-(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{A}\bm{x} \\
-(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} & = \bm{x} \\
+\bm{A}\bm{x} & = \bm{b} \\
+(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{A}\bm{x} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} \\
+\hat{\bm{x}} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} \\
 \begin{pmatrix}
-.33\Cov(y_{2},y_{1}) + .33\Cov(y_{3},y_{1}) + .33\Cov(y_{3},y_{2}) \\
-\Var(y_{1}) - \psi \\
-\Var(y_{2}) - \psi \\
-\Var(y_{3}) - \psi
+\hat{\psi} \\
+\hat{\phi}_{1} \\
+\hat{\phi}_{2} \\
+\hat{\phi}_{3}
 \end{pmatrix} & = 
 \begin{pmatrix}
-\psi \\
-\phi_{1} \\
-\phi_{2} \\
-\phi_{3}
+\frac{1}{3}\Cov(y_{2},y_{1}) + \frac{1}{3}\Cov(y_{3},y_{1}) + \frac{1}{3}\Cov(y_{3},y_{2}) \\
+\Var(y_{1}) - \hat{\psi} \\
+\Var(y_{2}) - \hat{\psi} \\
+\Var(y_{3}) - \hat{\psi}
 \end{pmatrix}.
@@
 $\bm{y}^{+}$ holds just the observed variables, both dependent and independent, and $\bm{\Lambda_{y}}^{+}$ is a matrix of ones and zeros that selects the observed variables from $\bm{\eta}^{+}$.
-Lastly $\bm{B}$ is a matrix that holds the regression coefficients.
+Lastly, $\bm{B}$ is a matrix that holds the regression coefficients.
 
@@
 
-Let us, however, make things more concrete and take a look at a simple, three-wave version of the typical FE-SEM using this notation (shown graphically in Figure \ref{fig:fesem}). For that, we have:
+\begin{figure}
+\begin{center}
+\resizebox{0.75\textwidth}{!}{%
+\begin{tikzpicture}
+% Node styles ---
+\tikzstyle{man} = [rectangle, thick, minimum size = 1cm, draw = black!80, fill = white!100, font = \sffamily] % Manifest variables
+\tikzstyle{lat} = [circle, thick, minimum size = 1cm, draw = black!80, fill = white!100, font = \sffamily] % Latent variables
+\tikzstyle{err} = [circle, draw = black!80, fill = white!100, font = \sffamily] % Errors 
+% Edge styles
+\tikzstyle{con} = [-latex, font = \sffamily] % Effects
+\tikzstyle{cons} = [-latex, font = \sffamily\small] % Effects with smaller font 
+\tikzstyle{cor} = [latex-latex, font = \sffamily] % Correlations
+% Begin Figure ---
+% Nodes 
+\node at (+0.0,+0.0) [man] (x1) {$x_{1}$};
+\node at (+2.0,+0.0) [man] (x2) {$x_{2}$};
+\node at (+6.0,+0.0) [man] (x3) {$x_{3}$};
+\node at (+4.0,+0.0) [lat] (eta) {$\alpha$};
+\node at (+0.0,-2.0) [man] (y1) {$y_{1}$};
+\node at (+2.0,-2.0) [man] (y2) {$y_{2}$};
+\node at (+6.0,-2.0) [man] (y3) {$y_{3}$};
+\node at (+0.0,-3.5) [err] (e1) {$\varepsilon_{1}$};
+\node at (+2.0,-3.5) [err] (e2) {$\varepsilon_{2}$};
+\node at (+6.0,-3.5) [err] (e3) {$\varepsilon_{3}$};
+% Paths
+\path (x1) edge [con, draw = black!100, left] node {$\beta$} (y1);
+\path (x2) edge [con, draw = black!100, left, near start] node {$\beta$} (y2);
+\path (x3) edge [con, draw = black!100, right] node {$\beta$} (y3);
+\path (e1) edge [con, draw = black!100, left] node {1} (y1);
+\path (e2) edge [con, draw = black!100, left] node {1} (y2);
+\path (e3) edge [con, draw = black!100, right] node {1} (y3);
+\path (eta) edge [con, draw = black!100, above, near start] node {1} (y1);
+\path (eta) edge [con, draw = black!100, below] node {1} (y2);
+\path (eta) edge [con, draw = black!100, above] node {1} (y3);
+% correlations
+\path (x1) edge [cor, bend left] node {} (x2);
+\path (x1) edge [cor, bend left] node {} (eta);
+\path (x1) edge [cor, bend left] node {} (x3);
+\path (x2) edge [cor, bend left] node {} (eta);
+\path (x2) edge [cor, bend left] node {} (x3);
+\path (x3) edge [cor, bend right] node {} (eta);
+\end{tikzpicture}
+}
+\caption{Typical three-wave FE-SEM model with contemporary effects\label{fig:fesem}}
+\end{center}
+\end{figure}
+
+Let us, however, make things more concrete and take a look at a simple, three-wave version of the typical FE-SEM using this notation (shown graphically in Figure \ref{fig:fesem}).
+For that, we have the following matrix notation (with labels on the outside of the matrices):
 \begin{align}
@@
 \end{align}
-
 \begin{align}
@@
 
-\begin{figure}
-\begin{center}
-\resizebox{0.75\textwidth}{!}{%
-\begin{tikzpicture}
-% Node styles ---
-\tikzstyle{man} = [rectangle, thick, minimum size = 1cm, draw = black!80, fill = white!100, font = \sffamily] % Manifest variables
-\tikzstyle{lat} = [circle, thick, minimum size = 1cm, draw = black!80, fill = white!100, font = \sffamily] % Latent variables
-\tikzstyle{err} = [circle, draw = black!80, fill = white!100, font = \sffamily] % Errors 
-% Edge styles
-\tikzstyle{con} = [-latex, font = \sffamily] % Effects
-\tikzstyle{cons} = [-latex, font = \sffamily\small] % Effects with smaller font 
-\tikzstyle{cor} = [latex-latex, font = \sffamily] % Correlations
-% Begin Figure ---
-% Nodes 
-\node at (+0.0,+0.0) [man] (x1) {$x_{1}$};
-\node at (+2.0,+0.0) [man] (x2) {$x_{2}$};
-\node at (+6.0,+0.0) [man] (x3) {$x_{3}$};
-\node at (+4.0,+0.0) [lat] (eta) {$\alpha$};
-\node at (+0.0,-2.0) [man] (y1) {$y_{1}$};
-\node at (+2.0,-2.0) [man] (y2) {$y_{2}$};
-\node at (+6.0,-2.0) [man] (y3) {$y_{3}$};
-\node at (+0.0,-3.5) [err] (e1) {$\varepsilon_{1}$};
-\node at (+2.0,-3.5) [err] (e2) {$\varepsilon_{2}$};
-\node at (+6.0,-3.5) [err] (e3) {$\varepsilon_{3}$};
-% Paths
-\path (x1) edge [con, draw = black!100, left] node {$\beta$} (y1);
-\path (x2) edge [con, draw = black!100, left, near start] node {$\beta$} (y2);
-\path (x3) edge [con, draw = black!100, right] node {$\beta$} (y3);
-\path (e1) edge [con, draw = black!100, left] node {1} (y1);
-\path (e2) edge [con, draw = black!100, left] node {1} (y2);
-\path (e3) edge [con, draw = black!100, right] node {1} (y3);
-\path (eta) edge [con, draw = black!100, above, near start] node {1} (y1);
-\path (eta) edge [con, draw = black!100, below] node {1} (y2);
-\path (eta) edge [con, draw = black!100, above] node {1} (y3);
-% correlations
-\path (x1) edge [cor, bend left] node {} (x2);
-\path (x1) edge [cor, bend left] node {} (eta);
-\path (x1) edge [cor, bend left] node {} (x3);
-\path (x2) edge [cor, bend left] node {} (eta);
-\path (x2) edge [cor, bend left] node {} (x3);
-\path (x3) edge [cor, bend right] node {} (eta);
-\end{tikzpicture}
-}
-\caption{Typical three-wave FE-SEM model with contemporary effects\label{fig:fesem}}
-\end{center}
-\end{figure}
-
 ## Assumptions
@@
 \begin{align}
-\bm{\Psi} & =
+\bm{\Psi} & = \E
 \begin{blockarray}{cccccccc}
@@
  & = \Cov(\alpha, x_{t}) + \beta \Var(x_{t}) \\
-\hat{\beta} & = \frac{\Cov(y_{t},x_{t}) - \Cov(\alpha, x_{t})}{\Var(x_{t})}. 
+\hat{\beta}_{FE-SEM} & = \frac{\Cov(y_{t},x_{t}) - \Cov(\alpha, x_{t})}{\Var(x_{t})}. 
 \end{align}
@@
 From the observed covariance between the dependent and the independent variable, we are partialling out the part that is due to the covariance between the independent variable and the individual effects per unit, and then dividing by the variance of the independent variable, as usual.
-For the RE model, we assume $\E[\alpha x_{t}] = 0$ and the equation reduces to $\hat{\beta} = \Cov(y_{t},x_{t})/\Var(x_{t})$.
+For the RE model, we assume $\E[\alpha x_{t}] = 0$ and the equation reduces to $\hat{\beta}_{RE-SEM} = \Cov(y_{t},x_{t})/\Var(x_{t})$.
 The rest of the model-implied covariance matrix results from $\bm{y}^{+}\bm{y}^{+ \intercal}$.
@@
 Here, again, we use labels to make equality constraints.
-Because $y_{t}$ is endogenous, the `~~` operator specifies the variances of *residuals*, i.e.
-$\epsilon_{t}$.
+Because $y_{t}$ is endogenous, the `~~` operator specifies the variances of *residuals*, i.e., $\varepsilon_{t}$.
 
@@
 We will construct the simulated data such that the independent variable is correlated with both of the time-invariant variables.
-This means that approaches that fail to account for this confounding influence, such as pooled ordinary least squares (POLS) or RE, will be biased.
+This means that approaches that fail to account for this confounding influence, such as POLS or RE, will be biased.
 

### 2020-07-22 extensions.Rmd
Made some major revisions to the text

@@
 
-The following discusses several opportunities to extend the basic FE model outlined in the main article, and provides concrete guidance on the implementation in `lavaan`.
-First, I go over a number of possibilities to relax assumptions associated with the traditional FE model.
+The following goes into some more detail on the comparison of traditional FE and FE-SEM models and discusses several opportunities to extend the basic FE model outlined in the main article, and provides concrete guidance on the implementation in `lavaan`.
+First, I verify that the FE-SEM model does, in fact, return essentially identical results compared to the more traditional methods.
+Then, I go over a number of possibilities to relax assumptions associated with the traditional FE model.
 Then, I discuss the issue of measurement error and show how we can use latent variables to deal with it and properly estimate the coefficients of interest.
 Then, I show a type of hybrid FE/RE model that allows us to control for time-invariant unobserved heterogeneity while including time-invariant predictors in the model.
+
+## A comparison with non-SEM methods using the simulated data
+
+Just to be sure that the FE-SEM results do, in fact, line up with the more traditional methods outlined in Section 2 of the main article, we can use the long-format data (see [the supplementary materials at](https://github.com/henrik-andersen/FE-SEM)) to run the typical FE model using the `plm` package [@R-plm_a].
+By default, the `plm` function assumes the dataframe is structured so that the first two columns correspond to the individual and time indices, see the [documentation](https://cran.r-project.org/web/packages/plm/plm.pdf) or @R-plm_a.
+
+```{r message=FALSE, warning=FALSE, error=FALSE}
+library(plm)
+
+# Run the FE model in plm 
+fe1 <- plm(y ~ x, 
+           effect = "individual", model = "within", 
+           data = df)
+summary(fe1)
+```
+
+From this, we see that the results are, indeed, essentially identical, with $\hat{\beta}_{FE-SEM} =`r round(lavInspect(fe_sem.fit, "list")[6, 14], 3)` \ (`r round(lavInspect(fe_sem.fit, "list")[6, 15], 3)`)$ and $\hat{\beta}_{FE} = `r round(summary(fe1)$coefficients[1, 1], 3)` \ (`r round(summary(fe1)$coefficients[1, 2], 3)`)$.
+
+Other methods of estimating FE models work in the random or mixed effects model framework.
+For example, we can include the cluster means per individual of the time-varying independent variables, here $x$, in the equation to achieve within estimates [@Mundlak1978; @Chamberlain1980; @Wooldridge2002].
+
+```{r}
+# Generate the cluster means for x per id 
+clusterMeanx <- aggregate(df$x, by = list(df$id), FUN = mean)
+# Rename the columns 
+names(clusterMeanx) <- c("id", "xbar")
+
+# Add the cluster means back into df  
+df <- merge(df, clusterMeanx, by = "id")
+```
+
+Here using the `plm` function in the `random` setup:
+
+```{r}
+fe2 <- plm(y ~ x + xbar, 
+           effect = "individual", model = "random",
+           data = df)
+summary(fe2)
+```
+
+And here using the `lmer` function of the `lme4` package [@Bates2015] to estimate a mixed model:
+
+```{r message=FALSE, error=FALSE, warning=FALSE}
+library(lme4)
+
+# Run the mixed model in lmer with the cluster means for x 
+mixed1 <- lmer(y ~ x + xbar + (1 | id), data = df)
+summary(mixed1)
+```
+
+In both cases, the models return the same estimates as the FE and FE-SEM models.
+Also, in both the `random` setup using the `plm` function, and the mixed model using the `lmer` function, we get estimates of the variance components, $\hat{\sigma}^{2}_{\alpha}$ and $\hat{\sigma}^{2}_{\varepsilon}$:
+
+```{r}
+# Print the variance components for the plm model  
+print(ercomp(fe2), 
+      digits = 3)
+
+# Print the variance components for the lmer model 
+print(VarCorr(mixed1), 
+      comp = c("Variance", "Std.Dev"), 
+      digits = 3)
+```
+
+From this we see both models report the same estimated variance components, $\hat{\sigma}^{2}_{\alpha} = `r round(ercomp(fe2)[[1]][2], 3)`$ and $\hat{\sigma}^{2}_{\varepsilon} = `r round(ercomp(fe2)[[1]][1], 3)`$, telling us that about `r round(ercomp(fe2)[[1]][2]/(ercomp(fe2)[[1]][1] + ercomp(fe2)[[1]][2]), 3)`% of the residual variance is due to the differences between individuals (shown in the `share` column of the `ercomp()` output).
+This is what is referred to as the intraclass correlation coefficient, or ICC [@Hox2010].
 

### 2020-07-23 article.Rmd
Made some minor changes to the article and extensions, mostly format

@@
 date: "`r Sys.Date()`"
+geometry: margin=1in
 output: 
@@
 bibliography: references.bib   
+fontsize: 12pt
 ---
@@
 It focuses on the `lavaan` [@R-lavaan] package for `R` [@R-base].
-While `Mplus` [@Mplus] is arguably the most robust SEM software currently available (in terms of features like alignment, latent variable interactions, for example) , the `lavaan` package has many benefits.
+While `Mplus` [@Mplus] is arguably the most robust SEM software currently available (in terms of features like alignment, latent variable interactions, for example), the `lavaan` package has many benefits.
 First, like `R` it is open source and completely free.

### 2020-07-23 extensions.Rmd
Made some minor changes to the article and extensions, mostly format

@@
 ---
-title: "Supplememtary examples for:"
+title: "Supplememtary materials for:"
 #author: "Henrik Kenneth Andersen"
 #date: "`r Sys.Date()`"
+geometry: margin=1in
 output: 
@@
 bibliography: references.bib   
+fontsize: 12pt
 ---
+
+\vspace{-75pt}
 
@@
 
-Andersen, Henrik K. (2020). A closer look at fixed-effects regression in structural equation modeling using lavaan.
+[Andersen, Henrik K. 2020. "A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}"](https://github.com/henrik-andersen/FE-SEM/blob/master/article.pdf).
 
@@
 
-# Extensions {#exten}
+---
 
@@
 
-## A comparison with non-SEM methods using the simulated data
+# A comparison with non-SEM methods
 
@@
 This is what is referred to as the intraclass correlation coefficient, or ICC [@Hox2010].
+
+# Extensions {#exten}
 

### 2020-07-28 article.tex
Changed the formating, made some minor edits

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[
  12pt,
  a4paper]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={A closer look at fixed effects regression in structural equation modeling using},
  pdfauthor={Henrik Kenneth Andersen},
  colorlinks=true,
  linkcolor=Maroon,
  filecolor=Maroon,
  citecolor=Blue,
  urlcolor=blue,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% ----- Packages

\usepackage{booktabs}
\usepackage{titling}
\usepackage{bm}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{subfig}
\usepackage{tikz}
\usetikzlibrary{arrows}
\usepackage[nofiglist]{endfloat}
%\floatsetup[figure]{capposition=top}
%\usepackage{kbordermatrix}
\usepackage{blkarray}
\usepackage{setspace}
\usepackage{etoolbox}
\doublespacing

% ----- Math operators

\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\mathrm{Var}}
\DeclareMathOperator{\Cov}{\mathrm{Cov}}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

% ----- Graphics

%\graphicspath{{./images/}}

% ----- Equation labels

\mathtoolsset{showonlyrefs}

% ----- Page numbering

\pagenumbering{gobble}

% ----- Spacing

%\renewenvironment{knitrout}{\vspace{10em}}{}

\BeforeBeginEnvironment{equation}{\begin{singlespace}\vspace*{-\baselineskip}}
\AfterEndEnvironment{equation}{\end{singlespace}\noindent\ignorespaces}
\BeforeBeginEnvironment{align}{\begin{singlespace}\vspace*{-\baselineskip}}
\AfterEndEnvironment{align}{\end{singlespace}\noindent\ignorespaces}

\title{\singlespacing A closer look at fixed effects regression in structural
equation modeling using \texttt{lavaan}}
\author{Henrik Kenneth Andersen}
\date{2020-07-27}

\begin{document}
\maketitle
\begin{abstract}
This article provides an in-depth look at fixed effects regression in
the structural equation modeling (SEM) framework, specifically the
application of fixed effects in the \texttt{lavaan} package for
\texttt{R}. It is meant as a applied guide for researchers, covering the
underlying model specification, syntax, and summary output. Online
supplementary materials further discuss various common extentsions to
the basic fixed-effect model, demonstrating how to relax model
assumptions, deal with measurement error in both the dependent and
independent variables, and include time-invariant predictors in a type
of hybrid fixed-/ random effects model. extbf\{Keywords: Fixed effects,
structural equation modeling, lavaan, R, panel analysis\}
\end{abstract}

\newpage

\pagenumbering{arabic} 
\setcounter{page}{1}

\hypertarget{intro}{%
\section{Introduction}\label{intro}}

Several years ago, Curran and Bauer
(\protect\hyperlink{ref-Curran2011}{2011}) reflected positively on the
growing use of panel studies in empirical social research. Some of the
strengths of panel data are well-known, e.g., the ability to establish
temporal precedence, increased statistical power and the reduction of
potential alternative models. However, perhaps the greatest strength of
panel data is that they allow for a more rigorous testing of substantive
theories. Panel data, i.e., repeated measures of the same observed units
(people, schools, firms, countries, etc.), allow researchers to
decompose the error term into a part that stays constant within units
and the part that changes over time. The part that does not change over
time can be seen as the combined effect of all time-invariant influences
(e.g., sex, date of birth, nationality) on the dependent variable. Fixed
effects (FE) regression involves controlling for these time-invariant
influences via a number of various methods. It thus accounts for a
likely and common source of bias.

Structural equation modeling (SEM) is a popular regression framework.
One of its main strengths is its flexibility. Not only can complex
causal structures with multiple dependent variables be tested
simultaneously, but in longitudinal (and, more generally, hierarchical)
studies both time-varying and invariant predictors can be included, and
effects can easily be allowed to vary over time. Thus researchers can
allow for and study effects that increase or fade over time, or that
appear only in specific periods. Beyond that, with the use of latent
variables, SEM provides a way to deal with measurement error and get
closer to the true underlying constructs of interest.

There are a number of articles describing basic concept of panel model
regression, and FE regression in SEM (e.g., Allison
\protect\hyperlink{ref-Allison2011}{2011}; Bollen and Brand
\protect\hyperlink{ref-Bollen2010}{2010}; Teachman et al.
\protect\hyperlink{ref-Teachman2001}{2001}). This article is intended as
a \textit{practical guide} for researchers looking for in-depth help
with specifying FE models in SEM. It focuses on the \texttt{lavaan}
(Rosseel \protect\hyperlink{ref-R-lavaan}{2012}) package for \texttt{R}
(R Core Team \protect\hyperlink{ref-R-base}{2017}). While \texttt{Mplus}
(Muthén and Muthén, \protect\hyperlink{ref-Mplus}{n.d.}) is arguably the
most robust SEM software currently available (in terms of features like
alignment, latent variable interactions, for example), the
\texttt{lavaan} package has many benefits. First, like \texttt{R} it is
open source and completely free. For researchers dipping their toes into
SEM, there is no financial barrier to try, and no risk if they decide it
is not for them. Second, the implementation of \texttt{lavaan} in the
larger \texttt{R} environment is an enormous advantage. Instead of
poring over reams of plain text, copying out coefficients by hand, every
part of the \texttt{lavaan} output is available as an object. This means
that all aspects of the model, from fit indices, to coefficients and
standard errors, to the model matrices, can be accessed and easily
integrated into tables and plots. Furthermore, \texttt{R} can be used
for a great deal of applications. It can be used to manage and
manipulate as well as simulate data, perform symbolic algebra, run more
traditional analyses (e.g., multiple regression, logistic regression,
principal component analysis), etc. Once one is comfortable using
\texttt{R}, there is no longer any need to switch between different
software for data preparation and analysis.

The following article outlines the basic idea of FE regression, the
particularities of FE in SEM, and shows its implementation in
\texttt{lavaan}. Using
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R}{simulated
data}, it demonstrates and annotates the code for the most basic FE
model and provides an overview of the summary output. One of the main
strengths of panel SEM compared to the more traditional methods of panel
analysis is its flexibility. Therefore, a number of potential extensions
to the basic model, including relaxing various assumptions, dealing with
measurement error in both the independent and dependent variables, as
well as the inclusion of time-invariant predictors in the form of a
hybrid fixed-/ random effects model, are shown in detail in the form of
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
supplementary materials}.

\hypertarget{panel}{%
\section{Panel models}\label{panel}}

To begin, let us start by reviewing a general panel model (Bollen and
Brand \protect\hyperlink{ref-Bollen2010}{2010}), also referred to as the
`unobserved effects model' (Wooldridge
\protect\hyperlink{ref-Wooldridge2012}{2012}; Croissant and Millo
\protect\hyperlink{ref-R-plm_a}{2008}) (we will return to this model in
the
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
supplementary materials} of the article when we discuss loosening
assumptions) \begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \alpha_{i} + \varepsilon_{it} \label{eq:gpm}
\end{align} where \(y_{it}\) is the dependent variable for unit
\(i, \ i = 1, ..., N\) at time \(t, \ t = 1, ..., T\), \(\bm{x}_{it}\)
is a \(1 \times K\) vector of time-varying covariates (which could
include a constant) linked to the dependent variable by the
\(K \times 1\) vector of coefficients \(\bm{\beta}\). \(\bm{z}_{i}\) is
a \(1 \times M\) vector of time-invariant covariates linked to the
dependent variable by the \(M \times 1\) vector of coefficients in
\(\bm{\gamma}\), \(\alpha_{i}\) represents the combined effect of all
unobserved time-constant variables affecting the dependent variable and
\(\varepsilon_{it}\) is the idiosyncratic error.

We can make stating some of the model assumptions easier by rewriting it
in matrix notation \begin{align}
\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\iota}_{T}\alpha_{i} + \bm{\varepsilon}_{i}
\end{align} where \(\bm{y}_{i}\) and \(\bm{\varepsilon}_{i}\) are
\(T \times 1\) vectors, \(\bm{X}_{i}\) and \(\bm{Z}_{i}\) are
\(T \times K\) and \(T \times M\) matrices, respectively,
\(\bm{\iota}_{T}\) is a \(T \times 1\) vector of ones and \(\alpha_{i}\)
is a scalar. \(\bm{\beta}\) and \(\bm{\gamma}\) are unchanged from
Equation \eqref{eq:gpm}.

Consistency of the following models requires the assumption of strict
exogeneity, although what constitutes strict exogeneity differs between
the random and fixed effects setups. Each assumption will be discussed
shortly. Apart from that, we typically make the following assumptions
(see, e.g., Wooldridge \protect\hyperlink{ref-Wooldridge2002}{2002};
Schmidheiny \protect\hyperlink{ref-Schmidheiny2019}{2019}) about this
model:

\begin{itemize}
\tightlist
\item
  Linearity: the model is linear in its parameters.
\item
  Independence: the observations are independent across individuals
  (assured by random sampling in the cross-section), but not necessarily
  across time.
\item
  The usual rank condition: we have more observations than independent
  variables and there is no perfect collinearity between any of the
  independent variables.
\end{itemize}

\hypertarget{re}{%
\subsection{Random effects}\label{re}}

For the random effects (RE) model, we define a composite error term:
\(\nu_{it} = \alpha_{i} + \varepsilon_{it}\) and rewrite the model in
Equation \eqref{eq:gpm} as \begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \nu_{it}, \ \text{or} \\
\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\nu}_{i},
\end{align} where
\(\bm{\nu}_{i} = \alpha_{i} \bm{\iota}_{T} + \bm{\varepsilon}_{i}\) and
\(\bm{\iota}_{T}\) is a \(T \times 1\) vector of ones. The strict
exogeneity assumption in the RE model implies \begin{align}
\E[\varepsilon_{it} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = 0, \\
\E[\alpha_{i} |\bm{X}_{i}, \bm{z}_{i}] = \E[\alpha_{i}] & = 0,
\end{align} where \(\bm{X}_{i} = \bm{x}_{i1}, ..., \bm{x}_{iT}\). For
both parts, the assumption that the unconditional expectations are 0 is
unproblematic as long as a constant is included in the regression. The
first part says the idiosyncratic errors at each timepoint are assumed
to be independent of the explanatory variables at \textit{all}
timepoints which is stronger than just assuming that they are
\textit{contemporaneously} independent. This implies that they are also
uncorrelated, i.e.,
\(\E[\bm{x}_{is}^{\intercal}\varepsilon_{it}] = \bm{0}\) and
\(\E[\bm{z}_{i}^{\intercal}\varepsilon_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T\)
(Wooldridge \protect\hyperlink{ref-Wooldridge2002}{2002}; Brüderl and
Ludwig \protect\hyperlink{ref-Bruederl2015}{2015}). We assume the
idiosyncratic errors are further independent of the individual effects,
which implies \(\E[\alpha_{i}\varepsilon_{it}] = 0\).

The second part is the potentially controversial assumption: it states
that the individual effects are uncorrelated with the independent
variables. We can use an example to show why this is often
controversial: If we are interested in the question of whether married
men earn more than unmarried men, then the second part of the strict
exogeneity assumption means that a man's marriage status would have to
be uncorrelated with all the time-invariant characteristics that could
potentially make that man an attractive marriage candidate in the first
place; e.g., looks, personality, family's status, profession, etc.
(Brüderl and Ludwig
\protect\hyperlink{ref-Bruederl2015}{2015}).\footnote{Assuming, for the
  sake of argument, that these characteristics are constant over time.}

From what we have discussed so far, the \(T \times T\) covariance matrix
of the errors
\(\bm{\Omega}_{i} = \E[\bm{\nu}_{i}\bm{\nu}_{i}^{\intercal}]\) can be
constructed. However, the standard random effects model adds the
additional assumptions \begin{alignat}{3}
\E[\varepsilon_{it}^{2} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \E[\varepsilon_{it}^{2}] && = \sigma_{\varepsilon}^{2}, && \ t = 1, ..., T, \\
\E[\varepsilon_{it}\varepsilon_{is}| \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \E[\varepsilon_{it}\varepsilon_{is}] && = 0, && \ \forall \ t \ne s 
\end{alignat} i.e., the idiosyncratic errors are conditionally
homoscedastic and serially uncorrelated and \begin{alignat}{2}
\E[\alpha_{i}^{2}| \bm{X}_{i},\bm{z}_{i}] & = \E[\alpha_{i}^{2}] && = \sigma^{2}_{\alpha}
\end{alignat} i.e., the individual effects are conditionally
homoscedastic (they are necessarily serially correlated as long as
\(\sigma_{\alpha}^{2} > 0\)) we arrive at the typical random effects
structure of the \(NT \times NT\) matrix \(\bm{\Omega}\):

\begin{align}
\bm{\Omega} & = 
\begin{pmatrix}
\bm{\Omega}_{1} & \hdots & 0 & \hdots & 0 \\
\vdots & \ddots & \vdots & \ddots & \vdots \\
0 & \hdots & \bm{\Omega}_{i} & \hdots & 0 \\
\vdots & \ddots & \vdots & \ddots & \vdots \\
0 & \hdots & 0 & \hdots & \bm{\Omega}_{N}
\end{pmatrix}
\end{align} with \(T \times T\) typical elements \begin{align}
\bm{\Omega}_{i} & = 
\begin{pmatrix}
\sigma^{2}_{\nu} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\alpha} \\
\sigma^{2}_{\alpha} & \sigma^{2}_{\nu} & \hdots & \sigma^{2}_{\alpha} \\
\vdots & \vdots & \ddots & \vdots \\
\sigma^{2}_{\alpha} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\nu}
\end{pmatrix} \label{eq:omeganui}
\end{align} where
\(\sigma^{2}_{\nu} = \sigma^{2}_{\alpha} + \sigma^{2}_{\varepsilon}\)
(Wooldridge \protect\hyperlink{ref-Wooldridge2002}{2002}; Schmidheiny
\protect\hyperlink{ref-Schmidheiny2019}{2019}). This means that in the
conditional covariance matrix of the errors, given the time-varying and
-invariant covariates, units over time will be correlated due to the
individual effects. We should keep the covariance structure of the
errors in mind as it will help make sense of the use of latent variables
to decompose the dependent variable into between- and within-variance
components, discussed below in Section \ref{fe-sem}.

Estimation of the RE model can be done using feasible generalized least
squares (GLS) in which the two unknowns in \(\bm{\Omega}\),
\(\sigma_{\alpha}^{2}\) and \(\sigma_{\nu}^{2}\), are first estimated
using pooled ordinary least squares (pooled OLS or POLS),
where\footnote{Normally a degrees-of-freedom correction is applied by
  subtracting off the number of independent variables that is negligible
  in large N samples. It is ignored here for the sake of simplicity.}
\begin{align}
\hat{\sigma}_{\nu}^{2} & = \frac{1}{NT}\sum_{i = 1}^{N}\sum_{t = 1}^{T}\hat{\nu}_{it}^{2}, \\
\hat{\sigma}_{\varepsilon}^{2} & = \frac{1}{NT - N}\sum_{i = 1}^{N}\sum_{t = 1}^{T}(\hat{\nu}_{it} - \bar{\hat{\nu}}_{i})^{2}
\end{align} and
\(\hat{\nu}_{it} = y_{it} - \bm{x}_{it}\hat{\bm{\beta}}_{\tiny{POLS}} - \bm{z}_{i}\hat{\bm{\gamma}}_{\tiny{POLS}}\),
\(\bar{\hat{\nu}}_{i} = T^{-1}\sum_{t = 1}^{T}\hat{\nu}_{it}\) and
\(\hat{\sigma}^{2}_{\alpha} = \hat{\sigma}^{2}_{\nu} - \hat{\sigma}^{2}_{\varepsilon}\).
Then, the coefficients are estimated using those estimates in the
variance matrix \(\hat{\bm{\Omega}}\): \begin{align}
\begin{pmatrix}
\hat{\bm{\beta}}_{RE} \\
\hat{\bm{\gamma}}_{RE}
\end{pmatrix} & = 
(\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{W})^{-1}\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{y}
\end{align} where
\(\bm{W} = \begin{pmatrix}\bm{X} & \bm{Z}\end{pmatrix}\) and \(\bm{X}\)
is \(NT \times K\) and \(\bm{Z}\) is \(NT \times M\) and \(\bm{y}\) is
\(NT \times 1\).

In practice, however, computational problems can arise with large
cross-sectional samples, where it can become difficult to invert the
\(\hat{\bm{\Omega}}\) matrix. One solution is to use `partial-demeaning'
to transform the data before performing simple OLS: \begin{align}
(y_{it} - \theta\bar{y}_{i}) & = (\bm{x}_{it} - \theta\bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \theta \bar{\bm{z}}_{i})\bm{\gamma} + (\varepsilon_{it} - \theta \bar{\varepsilon}_{i}) \label{eq:partialdemean}
\end{align} where
\(\theta = 1 - [\sigma_{\alpha}^{2}/(\sigma_{\alpha}^{2} + T \sigma_{\varepsilon}^{2})]^{1/2}\),
and \(\bar{y}_{i} = T^{-1}\sum_{t = 1}^{T}y_{it}\),
\(\bar{\bm{x}}_{i} = T^{-1}\sum_{t = 1}^{T}\bm{x}_{it}\),
\(\bar{\bm{z}}_{i} = T^{-1}\sum_{t = 1}^{T}\bm{z}_{i}\) and
\(\bar{\varepsilon}_{i} = T^{-1}\sum_{t = 1}^{T}\varepsilon_{it}\)
(Croissant and Millo \protect\hyperlink{ref-R-plm_a}{2008}).

The RE model can also be estimated in the maximum likelihood framework,
where in the associated literature panel models are generally referred
to as either mixed models, hierarchical models or longitudinal models.
The typical RE model discussed here is the equivalent to a mixed model
with random intercepts and fixed slopes (Croissant and Millo
\protect\hyperlink{ref-R-plm_a}{2008}). Under the assumption of
normality, along with homoscedasticity and serially uncorrelated errors,
the maximum likelihood estimator is the same as the OLS estimator. For
more on the topic of mixed models, see for example Bates et al.
(\protect\hyperlink{ref-R-lme4}{2015}), Bates
(\protect\hyperlink{ref-Bates2010}{2010}).

\hypertarget{fe}{%
\subsection{Fixed effects}\label{fe}}

For fixed effects, we assume that the individual effects are
\textit{not independent} of the model covariates, i.e.,
\(\E[\alpha_{i}|\bm{X}_{i},\bm{z}_{i}] \ne \E[\alpha_{i}] \ne 0\). Under
this assumption, grouping the individual effects in with the composite
error will cause the coefficients of interest, here specifically
\(\bm{\beta}\) to be inconsistent (Wooldridge
\protect\hyperlink{ref-Wooldridge2002}{2002},
\protect\hyperlink{ref-Wooldridge2012}{2012}). We write the model
therefore again in terms of the general panel model in Equation
\eqref{eq:gpm} with separate individual effects and idiosyncratic error
terms. In order to drop assumptions involving the individual effects, a
number of methods are available (e.g., differencing, least squares dummy
variable regression), but the most common approach is to \textit{demean}
the equation (Brüderl and Ludwig
\protect\hyperlink{ref-Bruederl2015}{2015}). Demeaning is the same as
the transformation applied in Equation \eqref{eq:partialdemean} in the
special case where \(\theta = 1\). I.e., demeaning involves subtracting
the per-unit, over-time average from each of the model terms, i.e.,
\begin{align}
(y_{it} - \bar{y}_{i}) & = (\bm{x}_{it} - \bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \bar{\bm{z}}_{i})\bm{\gamma} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
\ddot{y}_{it} & = \bm{\ddot{x}}_{it}\bm{\beta} + \ddot{\varepsilon}_{it}, \ \text{or} \\
\bm{\ddot{y}}_{i} & = \bm{\ddot{X}}_{i}\bm{\beta} + \bm{\ddot{\varepsilon}}_{i} \label{eq:fe}
\end{align} where the over-time averages are calculated the same as
above, and the variables with the dots above them represent the demeaned
versions. Because the average of something that does not change is that
thing itself, the individual effects, along with any time-invariant
predictors, get wiped out by the demeaning. This means that no
assumptions about the relatedness of the model covariates and the
unit-specific portion of the error are needed. Consistency of the
estimates is related solely to the strict exogeneity assumption imposed
on the idiosyncratic errors, i.e.,
\(\E[\ddot{\varepsilon}_{it}|\bm{\ddot{x}}_{it}] = \E[\ddot{\varepsilon}_{it}] = 0\)
which also implies
\(\E[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T\)
(Brüderl and Ludwig \protect\hyperlink{ref-Bruederl2015}{2015};
Wooldridge \protect\hyperlink{ref-Wooldridge2002}{2002}).

Having demeaned the data, the typical FE estimator is POLS on the
transformed data \begin{align}
\bm{\beta}_{FE} & = (\ddot{\bm{X}}^{\intercal}\ddot{\bm{X}})^{-1}\ddot{\bm{X}}^{\intercal}\ddot{\bm{y}}
\end{align} (Brüderl and Ludwig
\protect\hyperlink{ref-Bruederl2015}{2015}). The downside to this
approach is that no time-invariant predictors can be included in the
model. However, there are alternative approaches in the random effects
and mixed model frameworks that allow them to be included. These models
are sometimes referred to as `within-between' or `hybrid' models, often
based on the Chamberlain (\protect\hyperlink{ref-Chamberlain1980}{1980})
and Mundlak (\protect\hyperlink{ref-Mundlak1978}{1978}) approaches, see
for example Bell, Fairbrother, and Jones
(\protect\hyperlink{ref-Bell2018}{2018}); Allison
(\protect\hyperlink{ref-Allison2011}{2011}); Schunck
(\protect\hyperlink{ref-Schunck2013}{2013}); Enders and Tofighi
(\protect\hyperlink{ref-Enders2007}{2007}). In the
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
appendix}, it will be discussed how to also get around this restriction
using SEM.

\hypertarget{fe-sem}{%
\section{Fixed effects in structural equation modeling}\label{fe-sem}}

Moving from the conventional methods outlined above to SEM, we must
state the FE model in a different way. We turn to latent variables to
account for time-invariant unobserved heterogeneity. In fact, besides
accounting for measurement error and the representation of abstract
hypothetical concepts, unobserved heterogeneity has historically been
one of the main uses of latent variables in SEM (Skrondal and
Rabe-Hesketh \protect\hyperlink{ref-Skrondal2004}{2004}).

\hypertarget{modeling-time-invariant-unobserved-heterogeneity-as-a-latent-variable}{%
\subsection{Modeling time-invariant unobserved heterogeneity as a latent
variable}\label{modeling-time-invariant-unobserved-heterogeneity-as-a-latent-variable}}

We first need to convert the data from stacked, long-format vectors of
length \(NT\) into \(T\) individual vectors of length \(N\). To see why
this is necessary, consider what effect this has on the vector of
responses \(y_{it}\). Let us, for a minute ignore any covariates and
focus just on the dependent variable (a so-called `intercept-only' or
`null' model) so that we have
\(y_{it} = \alpha_{i} + \varepsilon_{it}\). When we convert the data to
wide-format, we get \(T\) individual equations, \begin{align}
\bm{y}_{t} & = \bm{\alpha} + \bm{\varepsilon}_{t} \\
\begin{bmatrix}
y_{1t} \\
y_{2t} \\
\vdots \\
y_{Nt}\end{bmatrix} & = 
\begin{bmatrix}
\alpha_{1} \\
\alpha_{2} \\
\vdots \\
\alpha_{N}
\end{bmatrix} + 
\begin{bmatrix}
\varepsilon_{1t} \\
\varepsilon_{2t} \\
\vdots \\
\varepsilon_{Nt}
\end{bmatrix} \label{eq:wide}
\end{align} for each \(t = 1, 2, ..., T\). Because the idiosyncratic
errors are assumed to be uncorrelated across units and across time, the
covariance between any two of the new wide vectors
\(\Cov(y_{ti},y_{si}) = \Var(\alpha_{i}), \ t \ne s\). Otherwise, when
\(t = s\), the covariance
\(\Cov(y_{ti},y_{ti}) = \Var(\alpha_{i}) + \Var(\varepsilon_{ti})\).
This is the structure we saw above in a typical element of
\(\bm{\Omega}\).

And in fact this is exactly how a latent variable is used to account for
time-invariant unobserved heterogeneity. The dependent variable at each
timepoint is regressed onto the latent variable, see Figure
\ref{fig:fesem}. Here, the regression weights or `factor loadings' are
fixed to one to represent our assumption that the effect of the
time-invariant unobserved heterogeneity is constant over time. It also
means that the estimated variance of the latent variable is equal to the
\textit{average covariance between the wide-format columns of the dependent variable over time}.
If \(y_{it} = \alpha_{i} + \varepsilon_{it}\) is the true data
generating process, then the relationship between two units over time is
just \(\alpha\), regardless of the time distance. Referring back to the
random effects structure of \(\bm{\Omega}_{i}\) in Equality
\eqref{eq:omeganui} for a generic unit \(i\), we see the covariance on
all of the off-diagonals is \(\sigma^{2}_{\alpha}\). And, as we know,
the average of something that does not change is that thing itself.
I.e., if \(T(T-1)/2 = h\) is the number of elements on either the upper-
or lower triangle of \(\bm{\Omega}_{i}\), then we have
\(h^{-1}\sum_{i=1}^{h}\sigma^{2}_{\alpha} = \frac{h \sigma^{2}_{\alpha}}{h} = \sigma^{2}_{\alpha}\).

To show this, consider the following matrix equation of the variances
and the nonredundant covariances in a three-wave intercept-only model
that follows directly from Equation \eqref{eq:wide} (assuming
\(\Cov(\varepsilon_{ti},\varepsilon_{si}) = 0, \ t \ne s\)), and which
we can solve easily with least squares: \begin{align}
\bm{A}\bm{x} & = \bm{b} \\
\begin{pmatrix}
1 & 1 & 0 & 0 \\
1 & 0 & 0 & 0 \\
1 & 0 & 0 & 0 \\
1 & 0 & 1 & 0 \\
1 & 0 & 0 & 0 \\
1 & 0 & 0 & 1
\end{pmatrix}
\begin{pmatrix}
\psi \\
\phi_{1} \\
\phi_{2} \\
\phi_{3}
\end{pmatrix} & = 
\begin{pmatrix}
\Var(y_{1}) \\
\Cov(y_{2},y_{1}) \\
\Cov(y_{3},y_{1}) \\
\Var(y_{2}) \\
\Cov(y_{3},y_{2}) \\
\Var(y_{3})
\end{pmatrix}
\end{align} where \(\psi = \Var(\alpha)\),
\(\phi_{t} = \Var(\varepsilon_{t})\). We can solve this equation to show
\begin{align}
\bm{A}\bm{x} & = \bm{b} \\
(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{A}\bm{x} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} \\
\hat{\bm{x}} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} \\
\begin{pmatrix}
\hat{\psi} \\
\hat{\phi}_{1} \\
\hat{\phi}_{2} \\
\hat{\phi}_{3}
\end{pmatrix} & = 
\begin{pmatrix}
\frac{1}{3}\Cov(y_{2},y_{1}) + \frac{1}{3}\Cov(y_{3},y_{1}) + \frac{1}{3}\Cov(y_{3},y_{2}) \\
\Var(y_{1}) - \hat{\psi} \\
\Var(y_{2}) - \hat{\psi} \\
\Var(y_{3}) - \hat{\psi}
\end{pmatrix}.
\end{align} This shows that if our assumption of constant covariances
between the dependent variable across time holds, the estimated variance
of \(\alpha\) is just what it should be: the average covariance between
units of \(y\) over time. Once we add in observed covariates, the
estimated covariance of \(\alpha\) then become the \textit{conditional}
covariance of \(y\) over time, given those covariates.

\hypertarget{model-notation}{%
\subsection{Model notation}\label{model-notation}}

Having explained how a latent variable is used to estimate the
individual effects, we need to state the FE model using SEM-compatible
matrix notation. There are a number of different model notations (see,
for example Bollen (\protect\hyperlink{ref-Bollen1989}{1989}) for an
overview), but the one that will serve us best is one that was proposed
by Graff (\protect\hyperlink{ref-Graff1979}{1979}): \begin{align}
\bm{y}^{+} & = \bm{\Lambda_{y}}^{+} \bm{\eta}^{+}, \\
\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+}, 
\end{align} where
\(\bm{\eta}^{+} = (\bm{y}, \bm{x}, \bm{\eta}, \bm{\xi})^{\intercal}\),
\(\bm{\zeta}^{+} = (\bm{\varepsilon}, \bm{\delta}, \bm{\zeta}, \bm{\xi})^{\intercal}\),
\(\bm{y}^{+} = (\bm{y}, \bm{x})^{\intercal}\). \(\bm{y}\) is a vector of
observed dependent variables and \(\bm{x}\) is a vector of observed
independent variables. \(\bm{\eta}\) is a vector of the latent dependent
variables and \(\bm{\xi}\) is a vector of latent independent variables.
\(\bm{\varepsilon}\) and \(\bm{\delta}\) are vectors of the errors of
the observed dependent and independent variables, respectively, and
\(\bm{\zeta}\) is a vector of the errors, or disturbances, of the latent
variables. Notice the \(^{+}\) symbol is just meant to differentiate the
vectors with them from those without them. That means, \(\bm{\eta}^{+}\)
is a vector that holds the observed and latent variables, both dependent
(in SEM they are referred to as `endogenous') and independent (i.e.,
`exogeneous')\footnote{For our purposes, the terms endogenous and
  dependent, on the one hand, and exogenous and independent, on the
  other, can be used interchangeably.}, \(\bm{\zeta}^{+}\) holds the
errors for the observed variables and the disturbances of the latent
variables. \(\bm{y}^{+}\) holds just the observed variables, both
dependent and independent, and \(\bm{\Lambda_{y}}^{+}\) is a matrix of
ones and zeros that selects the observed variables from
\(\bm{\eta}^{+}\). Lastly, \(\bm{B}\) is a matrix that holds the
regression coefficients.

If we say that \(p\) and \(q\) stand for the number of observed
dependent and independent variables, respectively, and \(m\) and \(n\)
stand for the number of latent dependent and independent variables,
respectively, then \(\bm{\eta}^{+}\) and \(\bm{\zeta}^{+}\) are
\(p + q + m + n\), \(\bm{y}^{+}\) is \(p + q\), \(\bm{\Lambda_{y}}^{+}\)
is \((p + q) \times (p + q + m + n)\) and \(\bm{B}\) is
\((p + q + m + n) \times (p + q + m + n)\) (Bollen
\protect\hyperlink{ref-Bollen1989}{1989}).

This notation may be confusing at first, but it has advantages. First,
it allows us the flexibility we need for the models. For example, it
allows observed \(\bm{x}\) to directly influence observed \(\bm{y}\)
(more common notation assumes that substantive effects occur only
between latent variables, observed ones are only used as indicators, see
for example Bollen (\protect\hyperlink{ref-Bollen1989}{1989}), Kline
(\protect\hyperlink{ref-Kline2016}{2016})). It also allows \(\bm{\xi}\),
i.e., any latent exogenous variables, to influence \(\bm{y}\) directly.
These two scenarios cover the traditional FE model with observed
variables, and one in which latent variables are used to account for
measurement error in the independent variables. It is also consistent
with the notation used for these models in \texttt{lavaan}. In fact,
\texttt{lavaan} switches automatically between matrix notations
depending on the specified model. That means the matrix representation
of the model one sees if they type in
\texttt{lavInspect(model,\ what\ =\ "est")} after specifying their model
in \texttt{lavaan} will match up with the notation used here.\footnote{In
  \texttt{Mplus}, the model matrices can be requested by including
  \texttt{OUTPUT:\ TECH1} in the input file.} It does have a potential
disadvantage however. Besides being less intuitive than the typical
\(\bm{\eta} = \bm{B}\bm{\eta} + \bm{\Gamma}\bm{\xi} + \bm{\zeta}\)
notation, it means that by including the observed covariates in the
stacked long vector \(\bm{y}^{+}\), they are treated as another response
variable with variances and covariances to be estimated by the model
(instead of just using the sample statistics). This means that the
assumption of multivariate normality (otherwise just imposed on the
dependent variables) also applies to the independent ones (Skrondal and
Rabe-Hesketh \protect\hyperlink{ref-Skrondal2004}{2004}, 75). This can
be problematic for noncontinuous independent variables like sex,
nationality dummies, marriage status (married/unmarried), etc. See
Skrondal and Rabe-Hesketh (\protect\hyperlink{ref-Skrondal2004}{2004})
for more on this topic.

\begin{figure}
\begin{center}
\resizebox{0.75\textwidth}{!}{%
\begin{tikzpicture}
% Node styles ---
\tikzstyle{man} = [rectangle, thick, minimum size = 1cm, draw = black!80, fill = white!100, font = \sffamily] % Manifest variables
\tikzstyle{lat} = [circle, thick, minimum size = 1cm, draw = black!80, fill = white!100, font = \sffamily] % Latent variables
\tikzstyle{err} = [circle, draw = black!80, fill = white!100, font = \sffamily] % Errors 
% Edge styles
\tikzstyle{con} = [-latex, font = \sffamily] % Effects
\tikzstyle{cons} = [-latex, font = \sffamily\small] % Effects with smaller font 
\tikzstyle{cor} = [latex-latex, font = \sffamily] % Correlations
% Begin Figure ---
% Nodes 
\node at (+0.0,+0.0) [man] (x1) {$x_{1}$};
\node at (+2.0,+0.0) [man] (x2) {$x_{2}$};
\node at (+6.0,+0.0) [man] (x3) {$x_{3}$};
\node at (+4.0,+0.0) [lat] (eta) {$\alpha$};
\node at (+0.0,-2.0) [man] (y1) {$y_{1}$};
\node at (+2.0,-2.0) [man] (y2) {$y_{2}$};
\node at (+6.0,-2.0) [man] (y3) {$y_{3}$};
\node at (+0.0,-3.5) [err] (e1) {$\varepsilon_{1}$};
\node at (+2.0,-3.5) [err] (e2) {$\varepsilon_{2}$};
\node at (+6.0,-3.5) [err] (e3) {$\varepsilon_{3}$};
% Paths
\path (x1) edge [con, draw = black!100, left] node {$\beta$} (y1);
\path (x2) edge [con, draw = black!100, left, near start] node {$\beta$} (y2);
\path (x3) edge [con, draw = black!100, right] node {$\beta$} (y3);
\path (e1) edge [con, draw = black!100, left] node {1} (y1);
\path (e2) edge [con, draw = black!100, left] node {1} (y2);
\path (e3) edge [con, draw = black!100, right] node {1} (y3);
\path (eta) edge [con, draw = black!100, above, near start] node {1} (y1);
\path (eta) edge [con, draw = black!100, below] node {1} (y2);
\path (eta) edge [con, draw = black!100, above] node {1} (y3);
% correlations
\path (x1) edge [cor, bend left] node {} (x2);
\path (x1) edge [cor, bend left] node {} (eta);
\path (x1) edge [cor, bend left] node {} (x3);
\path (x2) edge [cor, bend left] node {} (eta);
\path (x2) edge [cor, bend left] node {} (x3);
\path (x3) edge [cor, bend right] node {} (eta);
\end{tikzpicture}
}
\caption{Typical three-wave FE-SEM model with contemporary effects\label{fig:fesem}}
\end{center}
\end{figure}

Let us, however, make things more concrete and take a look at a simple,
three-wave version of the typical FE-SEM using this notation (shown
graphically in Figure \ref{fig:fesem}). For that, we have the following
matrix notation (with labels on the outside of the matrices):
\begin{align}
\begin{split}
\bm{y}^{+} & = \bm{\Lambda_{y}}^{+}\bm{\eta}^{+} \\
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3}
\end{pmatrix} & = 
\begin{blockarray}{cccccccc}
 & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 y_{1} & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
 y_{2} & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
 y_{3} & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\ 
 x_{1} & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
 x_{2} & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
 x_{3} & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
 \end{block}
\end{blockarray}
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix},
\end{split} \label{eq:y}
\end{align} \begin{align}
\begin{split}
\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+} \\
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix} & = 
\begin{blockarray}{cccccccc}
 & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
 y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
 y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
 x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \end{block}
\end{blockarray}
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix} + 
\begin{pmatrix}
\varepsilon_{1} \\
\varepsilon_{2} \\
\varepsilon_{3} \\
x_{1} = \delta_{1} \\
x_{2} = \delta_{2} \\
x_{3} = \delta_{3} \\
\alpha = \xi \\
\end{pmatrix}. 
\end{split} \label{eq:eta}
\end{align} Notice in \(\bm{\zeta}^{+}\), for the independent variables
we could either write, for example \(x_{t}\) or \(\delta_{t}\). As
mentioned above, this is due to the model notation treating the
independent variables like dependent variables with
variances/covariances to be estimated. For the sake of simplicity, we
will ignore this subtlety and refer to the observed variable from now
on, keeping in mind that if the multivariate normality assumption holds,
the estimated statistics will likely be sufficiently close to the sample
ones for it to not make much of a difference.

Admittedly, Equations \eqref{eq:y} and \eqref{eq:eta} may not look like
much yet. We can remedy this by first putting the equation for
\(\bm{\eta}^{+}\) in reduced form, i.e., by getting rid of the dependent
variable on the r.h.s.: \begin{align}
\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+} \\
\bm{\eta}^{+} - \bm{B}\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
(\bm{I} - \bm{B})\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
\bm{\eta}^{+} & = (\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+},
\end{align} where \(\bm{I}\) is the identity matrix. By substituting
this back into the equation for the observed variables we get
\(\bm{y}^{+} = \bm{\Lambda_{y}}^{+}[(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+}]\),
which works out to: \begin{align}
\begin{pmatrix} 
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\end{pmatrix} & = 
\begin{pmatrix}
\alpha + \beta x_{1} + \varepsilon_{1} \\
\alpha + \beta x_{2} + \varepsilon_{2} \\
\alpha + \beta x_{3} + \varepsilon_{3} \\
x_{1} \\
x_{2} \\
x_{3}
\end{pmatrix},
\end{align} which is of course exactly what we should expect given
Equation \eqref{eq:fe}\footnote{We can use the \texttt{sympy} package in
  \texttt{python} to verify and show the steps for this and other
  examples, see the
  \href{https://github.com/henrik-andersen/FE-SEM/blob/master/sympy-doublecheck-matrixnotation.py}{supplementary
  materials} for the code.}.

\hypertarget{assumptions}{%
\subsection{Assumptions}\label{assumptions}}

What essentially differentiates an FE from a random effects (RE) model
is our assumption concerning the relationship between the unobserved
individual effects and the model covariates (Bollen and Brand
\protect\hyperlink{ref-Bollen2010}{2010}). The FE model assumes that
\(\E[\alpha x_{t}] \ne 0\). As such, if we fail to control for the
correlation of the covariate and the time-invariant part of the error,
then the coefficient of interest, here \(\beta\), will be biased. Our
assumption regarding whether the individual effects are correlated with
the model covariates occurs in
\(\E[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] = \bm{\Psi}\), the
covariance matrix of the errors \begin{align}
\bm{y}^{+}\bm{y}^{+ \intercal} & = \E[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})^{\intercal}] \\
 & = \E[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\zeta}^{+ \intercal}(\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal})] \\
 & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \E[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal} \\
 & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \bm{\Psi} (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal}.
\end{align} In the case of an FE model, \(\bm{\Psi}\) will reflect our
belief that the individual effects are correlated with the model
covariates, here again for demonstration the three-wave model:
\begin{align}
\bm{\Psi} & = \E
\begin{blockarray}{cccccccc}
 & \varepsilon_{1} & \varepsilon_{2} & \varepsilon_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 \varepsilon_{1} & \varepsilon_{1}^{2} &                     &                     &              &              &              & \\
 \varepsilon_{2} & 0                   & \varepsilon_{2}^{2} &                     &              &              &              & \\
 \varepsilon_{3} & 0                   & 0                   & \varepsilon_{3}^{2} &              &              &              & \\
 x_{1}           & 0                   & 0                   & 0                   & x_{1}^{2}    &              &              & \\
 x_{2}           & 0                   & 0                   & 0                   & x_{2}x_{1}   & x_{2}^{2}    &              & \\
 x_{3}           & 0                   & 0                   & 0                   & x_{3}x_{1}   & x_{3}x_{2}   & x_{3}^{2}    & \\
 \alpha          & 0                   & 0                   & 0                   & \alpha x_{1} & \alpha x_{2} & \alpha x_{3} & \alpha^{2} \\
 \end{block}
\end{blockarray}.
\end{align} Knowing this, we can work out the equation for the
coefficient of interest, \(\beta\). For the sake of simplicity, assume
here and throughout mean-centered variables: \begin{align}
\Cov(y_{t},x_{t}) & = \E[y_{t}x_{t}] \\
 & = \E[(\alpha + \beta x_{t} + \varepsilon_{t})x_{t}] \\
 & = \E[\alpha x_{t} + \beta x_{t}^{2} + \varepsilon_{t}x_{t}] \\
 & = \Cov(\alpha, x_{t}) + \beta \Var(x_{t}) \\
\hat{\beta}_{FE-SEM} & = \frac{\Cov(y_{t},x_{t}) - \Cov(\alpha, x_{t})}{\Var(x_{t})}. 
\end{align} This should make intuitive sense. From the observed
covariance between the dependent and the independent variable, we are
partialling out the part that is due to the covariance between the
independent variable and the individual effects per unit, and then
dividing by the variance of the independent variable, as usual. For the
RE model, we assume \(\E[\alpha x_{t}] = 0\) and the equation reduces to
\(\hat{\beta}_{RE-SEM} = \Cov(y_{t},x_{t})/\Var(x_{t})\). The rest of
the model-implied covariance matrix results from
\(\bm{y}^{+}\bm{y}^{+ \intercal}\).

\hypertarget{fesem}{%
\section{Fixed effects in lavaan}\label{fesem}}

The package \texttt{lavaan} needs to be installed once with
\texttt{install.packages("lavaan")}. To be able to use it, we need to
load it for every new \texttt{R} session:

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{( lavaan)}
\end{Highlighting}
\end{Shaded}

\doublespacing

For users unfamiliar with \texttt{R}, SEM analyses can be carried out
with almost no knowledge of the language. Typically, someone unfamiliar
with \texttt{R} would prepare their data using some other statistical
software, and then save the intended dataset as a \texttt{.csv},
\texttt{.xlsx}, \texttt{.dta}, \texttt{.sav}, etc. file. The user must
then import the data, preferably as a dataframe, and the rest occurs
using the \texttt{lavaan} syntax.\footnote{There are many online
  tutorials for importing data in various formats, see, for example some
  from
  \href{https://www.datacamp.com/community/tutorials/r-data-import-tutorial}{datacamp}
  or
  \href{https://www.statmethods.net/input/importingdata.html}{Quick-R},
  or any of the many posts on
  \href{https://stackoverflow.com/search?q=r+import+data}{stackoverflow}.}

Specifying the most basic fixed effects model, like the one shown in
Bollen and Brand (\protect\hyperlink{ref-Bollen2010}{2010}) (the same
model as Equation \eqref{eq:fe} but with just one time-varying
predictor) involves four components. First, we define the latent
individual effects variable using the \texttt{=\textasciitilde{}}
`measured by' or `manifested by' (Rosseel
\protect\hyperlink{ref-R-lavaan}{2012}) operator at the same time
constraining the factor loadings at each timepoint to one. I will call
the latent variable \texttt{a} to stand for \(\alpha\). Constraining all
of the factor loadings to one reflects our implicit assumption that the
combined effect of the unit-specific unobserved factors is constant over
time. This is the default behaviour of traditional POLS-based approaches
to FE that use the stacked long-format data.

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{a =}\ErrorTok{~}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y1 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y2 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y3 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y4 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y5}
\end{Highlighting}
\end{Shaded}

\doublespacing

Second, we regress the dependent variable on the independent variable
using the \texttt{\textasciitilde{}} regression operator. With stacked,
long-format data, only one regression coefficient is estimated over all
observed timepoints. To have our FE-SEM model mimic this behaviour, we
need to constrain the the estimated coefficient to equal over time. We
do so by adding the same label to the regression coefficient at every
time point. We will use the label \texttt{b} (this label was chosen
arbitrarily, we could have used any letter or string of characters) and
have it act as an equality constraint for the regression coefficient of
interest \(\beta\):

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{y1 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x1}
\NormalTok{y2 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x2 }
\NormalTok{y3 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x3}
\NormalTok{y4 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x4}
\NormalTok{y5 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x5}
\end{Highlighting}
\end{Shaded}

\doublespacing

The key to a FE model, as opposed to an RE model are our assumptions
about the relatedness of our covariate and the individual effects, i.e.,
\(\E[x_{t}\alpha]\). For an FE model, we want to partial out any
potential covariance between the independent variable and the individual
effects. This accounts for any linear relationship between \(x_{t}\) and
the unit-specific characteristics influencing the dependent variable.
Further, allowing unrestricted covariances between the independent
variable itself over time will not affect how the coefficient \(\beta\)
is estimated, but will have an effect on the standard errors. To mimic
the behaviour of a conventional FE model, we allow the independent
variable to be correlated with the individual effects and itself over
time. Covariances (including covariances between a variable and itself,
i.e., variances) are specified using the
\texttt{\textasciitilde{}\textasciitilde{}} operator:

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{a }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x1 }\OperatorTok{+}\StringTok{ }\NormalTok{x2 }\OperatorTok{+}\StringTok{ }\NormalTok{x3 }\OperatorTok{+}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
\NormalTok{x1 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x2 }\OperatorTok{+}\StringTok{ }\NormalTok{x3 }\OperatorTok{+}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
\NormalTok{x2 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x3 }\OperatorTok{+}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
\NormalTok{x3 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
\NormalTok{x4 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x5}
\end{Highlighting}
\end{Shaded}

\doublespacing

The last component of our code involves the variances of the residuals.
This component is optional, but we can constrain the residual variances
to be equal over time to again mimic the behaviour of a conventional FE
model using POLS on stacked data. Here, again, we use labels to make
equality constraints. Because \(y_{t}\) is endogenous, the
\texttt{\textasciitilde{}\textasciitilde{}} operator specifies the
variances of \emph{residuals}, i.e., \(\varepsilon_{t}\).

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{y1 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y1}
\NormalTok{y2 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y2}
\NormalTok{y3 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y3}
\NormalTok{y4 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y4}
\NormalTok{y5 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y5}
\end{Highlighting}
\end{Shaded}

\doublespacing

\hypertarget{ex1}{%
\section{A simulated example}\label{ex1}}

\singlespacing

\doublespacing

To demonstrate the application of FE models in SEM, a dataset can be
simulated that embodies the FE assumptions. Again, the code for data
simulation can be found in the
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R}{online
supplementary materials}.

To show that the latent individual effects variables represent the
\emph{combined} effect of all time-invariant characteristics, the
dependent variable will be influenced by two separate unit-specific
variables, which we can call \(\alpha_{1}\) and \(\alpha_{2}\). We will
construct the simulated data such that the independent variable is
correlated with both of the time-invariant variables. This means that
approaches that fail to account for this confounding influence, such as
POLS or RE, will be biased.

The wide-format equations for the data generating process can be
described as: \begin{align}
\bm{x_{t}} & = \bm{\alpha_{1}}\beta_{x_{t},\alpha_{1}} + \bm{\alpha_{2}}\beta_{x_{t},\alpha_{2}} + \bm{\delta_{t}}, \\
\bm{y_{t}} & = \bm{x_{t}}\beta_{y_{t},x_{t}} + \bm{\alpha_{1}}\beta_{y_{t},\alpha_{1}} + \bm{\alpha_{2}} \beta_{y_{t},\alpha_{2}} + \bm{\varepsilon_{t}} 
\end{align} where, for the sake of simplicity, \(\bm{\alpha_{1}}\),
\(\bm{\alpha_{2}}\), \(\bm{\delta_{t}}\) and \(\bm{\varepsilon_{t}}\)
are \(\sim N(0,1)\).

For the following example, a sample size of 1,000, observed over five
waves, was chosen. The unique variance of \(\bm{x}\), as well as both
the individual-effect variables is also \(\sim N(0,1)\). The coefficient
of interest, \(\beta_{y,x}\) is set to be equal to \(0.3\). A
correlation between \(\bm{x}\) and the individual effects is induced
through \(\beta_{x,\alpha_{1}} = 0.85\) and
\(\beta_{x,\alpha_{1}} = 0.50\). With the variances above set to one,
the covariances will be roughly \(\Cov(x_{t},\alpha_{1}) = 0.85\) and
\(\Cov(x_{t},\alpha_{2}) = 0.5\). The dependent variable is also
influenced by the individual effects variables with
\(\beta_{y_{t},\alpha_{1}} = 0.75\) and
\(\beta_{y_{t},\alpha_{2}} = 0.45\) These values were chosen
arbitrarily.

Now, we run the FE-SEM in \texttt{lavaan}.

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fe_sem <-}\StringTok{ '}
\StringTok{# Define individual effects variable }
\StringTok{a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5}
\StringTok{# Regressions, constrain coefficient to be equal over time}
\StringTok{y1 ~ b*x1}
\StringTok{y2 ~ b*x2 }
\StringTok{y3 ~ b*x3}
\StringTok{y4 ~ b*x4}
\StringTok{y5 ~ b*x5}
\StringTok{# Allow unrestricted correlation between eta and covariates}
\StringTok{a ~~ x1 + x2 + x3 + x4 + x5}
\StringTok{x1 ~~ x2 + x3 + x4 + x5}
\StringTok{x2 ~~ x3 + x4 + x5}
\StringTok{x3 ~~ x4 + x5}
\StringTok{x4 ~~ x5}
\StringTok{# Constrain residual variances to be equal over time}
\StringTok{y1 ~~ e*y1}
\StringTok{y2 ~~ e*y2}
\StringTok{y3 ~~ e*y3}
\StringTok{y4 ~~ e*y4}
\StringTok{y5 ~~ e*y5}
\StringTok{'}
\NormalTok{fe_sem.fit <-}\StringTok{ }\KeywordTok{sem}\NormalTok{(}\DataTypeTok{model =}\NormalTok{ fe_sem, }
                  \DataTypeTok{data =}\NormalTok{ dfw, }
                  \DataTypeTok{estimator =} \StringTok{"ML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\doublespacing

\singlespacing

\doublespacing

We can get a summary of the model with \texttt{summary()}. The first
portion of the summary output gives an overview of some basic
information and fit statistics. The maximum likelihood estimator is the
default, so it did not have to be explicitly selected in the fitting
function call. Other estimators are available, including generalized and
unweighted least squares (\texttt{GLS} and \texttt{ULS}, respectively),
robust standard errors maximum likelihood (\texttt{MLM}) and several
others (see \href{https://lavaan.ugent.be/tutorial/est.html}{the lavaan
online tutorial for more}).

This part of the summary output also tells us that the analysis is based
on 1,000 observations (missings would be shown here as well if there
were any), and that the \(\chi^{2}\) statistic is 30.138 based on 32
degrees of freedom (55 observed covariances minus 1 error variance, 1
coefficient, 1 latent variable variance, 5 exogenous variable variances
and 15 covariances for \(55 - 23 = 32\) df). The p-value on the
\(\chi^{2}\) statistic is not significant with \(p =\) 0.561 which tells
us the differences between the model-implied and observed covariance
matrices are likely due to chance, and that the model fits the data well
(given how the data was generated, it would be surprising if this were
not the case). Other fit measures including typical comparative fit
indices can be requested by either adding \texttt{fit.measures\ =\ TRUE}
as a secondary argument to the \texttt{summary()} call, or by asking for
a complete list of all available fit statistics using
\texttt{lavInspect(model,\ "fit")} where \texttt{model} stands for the
name of the fitted model, in this case \texttt{fe\_sem.fit}.

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{(fe_sem.fit)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## lavaan 0.6-6 ended normally after 37 iterations
## 
##   Estimator                                         ML
##   Optimization method                           NLMINB
##   Number of free parameters                         31
##   Number of equality constraints                     8
##                                                       
##   Number of observations                          1000
##                                                       
## Model Test User Model:
##                                                       
##   Test statistic                                30.138
##   Degrees of freedom                                32
##   P-value (Chi-square)                           0.561
## 
## Parameter Estimates:
## 
##   Standard errors                             Standard
##   Information                                 Expected
##   Information saturated (h1) model          Structured
...
\end{verbatim}

\doublespacing

Next the summary output shows the measurement models for the latent
variables, if any. In this case the latent variable \texttt{a} for
\(\alpha\) is measured by each of the five observed dependent variables
with factor loadings fixed to 1.0.

\singlespacing

\begin{verbatim}
...
## Latent Variables:
##                    Estimate  Std.Err  z-value  P(>|z|)
##   a =~                                                
##     y1                1.000                           
##     y2                1.000                           
##     y3                1.000                           
##     y4                1.000                           
##     y5                1.000                           
...
\end{verbatim}

\doublespacing

The regressions are shown next. Here, because we have constrained the
regression coefficients to be equal over time (the equality constraint
label \texttt{(b)} is listed to the left of the estimates), the estimate
of \(\beta =\) 0.294 (0.016) is repeated five times. The corresponding
z- and p-values show that the coefficient is, unsurprisingly,
significant.

\singlespacing

\begin{verbatim}
...
## Regressions:
##                    Estimate  Std.Err  z-value  P(>|z|)
##   y1 ~                                                
##     x1         (b)    0.294    0.016   18.809    0.000
##   y2 ~                                                
##     x2         (b)    0.294    0.016   18.809    0.000
##   y3 ~                                                
##     x3         (b)    0.294    0.016   18.809    0.000
##   y4 ~                                                
##     x4         (b)    0.294    0.016   18.809    0.000
##   y5 ~                                                
##     x5         (b)    0.294    0.016   18.809    0.000
...
\end{verbatim}

\doublespacing

Next, the covariance estimates are listed. First, the covariances
between the latent individual effects variable and the independent
variable over time are shown, and then the covariances between the
independent variable with itself over time.

One should always take care to double-check that there are no unintended
covariances listed here. Like \texttt{Mplus}, the \texttt{lavaan}
package estimates some covariances per default, without the user
explicitly having to add them to the model syntax. For example,
covariances between latent variables are estimated per default. If one
does not wish for them to covary, it must be explicitly stated, e.g.,
with \texttt{f1\ \textasciitilde{}\textasciitilde{}\ 0*f2}, assuming the
latent variables are called \texttt{f1} and \texttt{f2}, or by
overriding the default behaviour for the entire model by adding
\texttt{orthogonal\ =\ TRUE} (which sets the correlation between all
latent variables to zero) to the fitting call.\footnote{This is at least
  the current behaviour of both the \texttt{cfa} and \texttt{sem}
  wrappers. In fact, both wrappers seem to be identical in terms of the
  default settings, see Rosseel et al.
  (\protect\hyperlink{ref-Rosseel2020}{2020}).}

\singlespacing

\begin{verbatim}
...
## Covariances:
##                    Estimate  Std.Err  z-value  P(>|z|)
##   a ~~                                                
##     x1                0.844    0.055   15.355    0.000
##     x2                0.867    0.056   15.441    0.000
##     x3                0.845    0.055   15.400    0.000
##     x4                0.822    0.053   15.455    0.000
##     x5                0.820    0.053   15.572    0.000
##   x1 ~~                                               
##     x2                0.908    0.070   12.900    0.000
##     x3                0.935    0.069   13.466    0.000
##     x4                0.921    0.067   13.661    0.000
##     x5                0.914    0.067   13.716    0.000
##   x2 ~~                                               
##     x3                0.889    0.070   12.675    0.000
##     x4                0.922    0.069   13.423    0.000
##     x5                0.889    0.068   13.165    0.000
##   x3 ~~                                               
##     x4                0.865    0.067   12.976    0.000
##     x5                0.901    0.066   13.554    0.000
##   x4 ~~                                               
##     x5                0.850    0.064   13.285    0.000
...
\end{verbatim}

\doublespacing

Finally, the variance estimates are listed. Here, we see that in order
to mimic the behaviour of a traditional FE model, the error variances
over time were specified to be equal using the equality constraint
\texttt{(e)}. Notice the \texttt{.} beside \texttt{y1}, \texttt{y2},
etc.: this indicates that the listed variance refers to an endogenous
variable, and that it is thus an error variance. In this case, these
refer to the variances of \(\varepsilon_{t}\). After that, the variances
of the exogenous variables, both observed and unobserved are listed.

\singlespacing

\begin{verbatim}
...
## Variances:
##                    Estimate  Std.Err  z-value  P(>|z|)
##    .y1         (e)    1.022    0.023   44.721    0.000
##    .y2         (e)    1.022    0.023   44.721    0.000
##    .y3         (e)    1.022    0.023   44.721    0.000
##    .y4         (e)    1.022    0.023   44.721    0.000
##    .y5         (e)    1.022    0.023   44.721    0.000
##     x1                1.986    0.089   22.361    0.000
##     x2                2.079    0.093   22.361    0.000
##     x3                1.987    0.089   22.361    0.000
##     x4                1.860    0.083   22.361    0.000
##     x5                1.814    0.081   22.361    0.000
##     a                 0.799    0.052   15.310    0.000
\end{verbatim}

\doublespacing

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

Fixed effects regression in SEM has been outlined in well-known articles
by (Allison \protect\hyperlink{ref-Allison2011}{2011}; Bollen and Brand
\protect\hyperlink{ref-Bollen2010}{2010}; Teachman et al.
\protect\hyperlink{ref-Teachman2001}{2001}). This article provides a
focused look at the implementation of the basic model, as well as common
extensions using the \texttt{lavaan} package in \texttt{R}.

The benefits of FE-SEM as opposed to traditional OLS-based FE-models are
largely the same ones that apply to the SEM framework in general: for
one, SEM allows for a great deal of flexibility. For example, it is easy
to loosen model constraints as necessary. Measurement error in both the
dependent and independent variables can be dealt with using latent
variables to achieve unbiased and more efficient results. Researchers
interested in time-invariant predictors can integrate them into a hybrid
FE/RE model with ease. Further extensions, like measurement invariance
testing (Schoot, Lugtig, and Hox
\protect\hyperlink{ref-Schoot2012}{2012}; Millsap
\protect\hyperlink{ref-Millsap2011}{2011}; Steenkamp and Baumgartner
\protect\hyperlink{ref-Steenkamp1998}{1998}) as well as lagged dependent
variables (Bollen and Brand \protect\hyperlink{ref-Bollen2010}{2010};
Allison, Williams, and Moral-Benito
\protect\hyperlink{ref-Allison2017}{2017}) for example, can also be
implemented in a straightforward fashion.

The most basic FE-SEM is furthermore the basis for a variety of
currently popular extended models, such as Latent Curve Models in
general (Curran and Bollen \protect\hyperlink{ref-Curran2001}{2001};
Bollen and Curran \protect\hyperlink{ref-Bollen2004}{2004}), as well as
special implementations like the Dynamic Panel Model (Allison, Williams,
and Moral-Benito \protect\hyperlink{ref-Allison2017}{2017}), the
Random-Intercept Cross-Lagged Panel Model (Hamaker, Kuiper, and Grasman
\protect\hyperlink{ref-Hamaker2015}{2015}) and the Latent Curve Model
with Structured Residuals (Curran et al.
\protect\hyperlink{ref-Curran2014}{2014}). For this reason it is all the
more important for researchers to have a good grasp on the method of
applying FE-SEM. This article is meant to serve as a consolidated
resource for researchers looking for concrete advise on specifying FE
and more general panel models in SEM.

\newpage

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Allison2011}{}%
Allison, Paul. 2011. \emph{Fixed Effects Regression Models}. Thousand
Oaks: Sage Publications.

\leavevmode\hypertarget{ref-Allison2017}{}%
Allison, Paul, Richard Williams, and Enrique Moral-Benito. 2017.
``Maximum Likelihood for Cross-Lagged Panel Models with Fixed Effects.''
\emph{Socius} 3: 1--17. \url{https://doi.org/10.1177/2378023117710578}.

\leavevmode\hypertarget{ref-Bates2010}{}%
Bates, Douglas M. 2010. \emph{Lme4: Mixed-Effects Modeling with R}.
Springer.

\leavevmode\hypertarget{ref-R-lme4}{}%
Bates, Douglas, Martin Mächler, Ben Bolker, and Steve Walker. 2015.
``Fitting Linear Mixed-Effects Models Using lme4.'' \emph{Journal of
Statistical Software} 67 (1): 1--48.
\url{https://doi.org/10.18637/jss.v067.i01}.

\leavevmode\hypertarget{ref-Bell2018}{}%
Bell, Andrew, Malcolm Fairbrother, and Kelvyn Jones. 2018. ``Fixed and
Random Effects Models: Making an Informed Choice.'' \emph{Quality and
Quantity} 53: 1051--74.

\leavevmode\hypertarget{ref-Bollen1989}{}%
Bollen, Kenneth. 1989. \emph{Structural Equations with Latent
Variables}. New York, Chichester: Wiley.

\leavevmode\hypertarget{ref-Bollen2010}{}%
Bollen, Kenneth, and Jennie Brand. 2010. ``A General Panel Model with
Random and Fixed Effects: A Structural Equations Approach.''
\emph{Social Forces} 89(1): 1--34.

\leavevmode\hypertarget{ref-Bollen2004}{}%
Bollen, Kenneth, and Patrick Curran. 2004. ``Autoregressive Latent
Trajectory (ALT) Models: A Synthesis of Two Traditions.''
\emph{Sociological Methods and Research} 32(3): 336--83.
\url{https://doi.org/10.1177/0049124103260222}.

\leavevmode\hypertarget{ref-Bruederl2015}{}%
Brüderl, Josef, and Volker Ludwig. 2015. ``Fixed-Effects Panel
Regression.'' In \emph{The Sage Handbook of Regression Analysis and
Causal Inference}, edited by Henning Best and Christof Wolf, 327--57.
London, Thousand Oaks: Sage Publications.

\leavevmode\hypertarget{ref-Chamberlain1980}{}%
Chamberlain, Gary. 1980. ``Analysis of Covariance with Qualitative
Data.'' \emph{Review of Economic Studies} 47(1): 225--38.

\leavevmode\hypertarget{ref-R-plm_a}{}%
Croissant, Yves, and Giovanni Millo. 2008. ``Panel Data Econometrics in
R: The plm Package.'' \emph{Journal of Statistical Software} 27 (2):
1--43. \url{https://doi.org/10.18637/jss.v027.i02}.

\leavevmode\hypertarget{ref-Curran2011}{}%
Curran, Patrick, and Daniel Bauer. 2011. ``The Disaggregation of
Within-Person and Between-Person Effects in Longitudinal Models of
Change.'' \emph{Annual Review of Psychology} 62: 583--619.

\leavevmode\hypertarget{ref-Curran2001}{}%
Curran, Patrick, and Kenneth Bollen. 2001. ``The Best of Both Worlds:
Combining Autoregressive and Latent Curve Models.'' In \emph{New Methods
for the Analysis of Change}, edited by L. Collins and A. Sayer, 107--35.
Washington, DC: American Psychological Press.
\url{https://doi.org/\%20doi:10.1037/10409-004}.

\leavevmode\hypertarget{ref-Curran2014}{}%
Curran, Patrick, Andrea Howard, Sierra Bainter, Stephanie Lane, and
James McGinley. 2014. ``The Separation of Between-Person and
Within-Person Components of Individual Change over Time: A Latent Growth
Curve Model with Structured Residuals.'' \emph{Journal of Consulting and
Clinical Psychology} 82(5): 879--94.
\url{https://doi.org/10.1037/a0035297}.

\leavevmode\hypertarget{ref-Enders2007}{}%
Enders, Craig, and Davood Tofighi. 2007. ``Centering Predictor Variables
in Cross-Sectional Multilevel Models: A New Look at an Old Issue.''
\emph{Psychological Methods} 12(2): 121--38.

\leavevmode\hypertarget{ref-Graff1979}{}%
Graff, J. 1979. ``Verallgemeinertes LISREL-Modell.'' Mannheim, Germany.

\leavevmode\hypertarget{ref-Hamaker2015}{}%
Hamaker, Ellen, Rebecca Kuiper, and Raoul Grasman. 2015. ``A Critique of
the Cross-Lagged Panel Model.'' \emph{Psychological Methods} 20(1):
102--16. \url{https://doi.org/10.1037/a0038889}.

\leavevmode\hypertarget{ref-Kline2016}{}%
Kline, Rex. 2016. \emph{Principles and Practice of Structural Equation
Modeling. Fourth Edition}. New York: The Guilford Press.

\leavevmode\hypertarget{ref-Millsap2011}{}%
Millsap, Roger. 2011. \emph{Statistical Approaches to Measurement
Invariance}. New York, London: Routledge.

\leavevmode\hypertarget{ref-Mundlak1978}{}%
Mundlak, Yair. 1978. ``On the Pooling of Time Series and Cross Section
Data.'' \emph{Econometrica} 46(1): 69--85.

\leavevmode\hypertarget{ref-Mplus}{}%
Muthén, Linda, and Bengt Muthén. n.d. \emph{Mplus User's Guide. Eighth
Edition}. Los Angeles, California: Muthén \& Muthén.

\leavevmode\hypertarget{ref-R-base}{}%
R Core Team. 2017. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{https://www.R-project.org/}.

\leavevmode\hypertarget{ref-R-lavaan}{}%
Rosseel, Yves. 2012. ``lavaan: An R Package for Structural Equation
Modeling.'' \emph{Journal of Statistical Software} 48 (2): 1--36.
\url{http://www.jstatsoft.org/v48/i02/}.

\leavevmode\hypertarget{ref-Rosseel2020}{}%
Rosseel, Yves, Terrence D. Jorgensen, Daniel Oberski, Jarrett Byrnes,
Leonard Vanbrabant, Victoria Savalei, Ed Merkle, et al. 2020.
\emph{Lavaan: Latent Variable Analysis}.
\url{https://CRAN.R-project.org/package=lavaan}.

\leavevmode\hypertarget{ref-Schmidheiny2019}{}%
Schmidheiny, Kurt. 2019. ``Panel Data: Fixed and Random Effects.''
Unversität Basel.

\leavevmode\hypertarget{ref-Schoot2012}{}%
Schoot, Rens van de, Peter Lugtig, and Joop Hox. 2012. ``A Checklist for
Testing Measurement Invariance.'' \emph{European Journal of
Developmental Psychology} 9(4): 486--92.

\leavevmode\hypertarget{ref-Schunck2013}{}%
Schunck, Reinhard. 2013. ``Within and Between Estimates in
Random-Effects Models: Advantages and Drawbacks of Random Effects and
Hybrid Models.'' \emph{The Stata Journal} 13(1): 65--76.

\leavevmode\hypertarget{ref-Skrondal2004}{}%
Skrondal, Anders, and Sophia Rabe-Hesketh. 2004. \emph{Generalized
Latent Variable Modeling}. Boca Raton, London, New York, Washington,
D.C.: Chapman \& Hall/CRC.

\leavevmode\hypertarget{ref-Steenkamp1998}{}%
Steenkamp, Jan-Benedict, and Hans Baumgartner. 1998. ``Assessing
Measurement Invariance in Cross-National Consumer Research.''
\emph{Journal of Consumer Research} 25(1): 78--90.

\leavevmode\hypertarget{ref-Teachman2001}{}%
Teachman, Jay, Greg Duncan, Jean Yeung, and Dan Levy. 2001. ``Covariance
Structure Models for Fixed and Random Effects.'' \emph{Sociological
Methods and Research} 30(2): 271--88.

\leavevmode\hypertarget{ref-Wooldridge2002}{}%
Wooldridge, Jeffery. 2002. \emph{Econometric Analysis of Cross Sectional
and Panel Data}. Cambridge, Massachusetts: The MIT Press.

\leavevmode\hypertarget{ref-Wooldridge2012}{}%
---------. 2012. \emph{Introductory Econometrics: A Modern Approach,
5\(^{th}\) Edition}. Mason, Ohio: Thomson South-Western.

\end{document}

### 2020-07-28 article_format/article_format.Rmd
Changed the formating, made some minor edits

--- 2020-07-23 article.Rmd
@@
 ---
-title: \singlespacing A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}
-author: Henrik Kenneth Andersen
-abstract: "This article provides an in-depth look at fixed effects regression in the structural equation modeling (SEM) framework, specifically the application of fixed effects in the `lavaan` package for `R`.
+title: |
+  A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}
+#type: ARTICLE TEMPLATE
+author:
+  - name: Henrik Kenneth Andersen
+    affil: a
+    email: henrik.andersen@soziologie.tu-chemnitz.de
+affiliation:
+  - num: a
+    address: |
+      Chemnitz University of Technology, Institute of Sociology, Chair for Empirical Social Research, Thüringer Weg 9, 09126 Chemnitz, Germany
+bibliography: references2.bib
+# appendix: appendix.tex
+abstract: |
+  This article provides an in-depth look at fixed effects regression in the structural equation modeling (SEM) framework, specifically the application of fixed effects in the `lavaan` package for `R`.
 It is meant as a applied guide for researchers, covering the underlying model specification, syntax, and summary output.
-Online supplementary materials further discuss various common extentsions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of hybrid fixed-/ random effects model.
-\\par  \\textbf{Keywords:} Fixed effects, structural equation modeling, lavaan, R, panel analysis"
-note: "Preprint version, please do not cite without permission."
-date: "`r Sys.Date()`"
-geometry: margin=1in
-output: 
-  pdf_document:
-    fig_caption: yes
-    number_sections: yes
-    includes: 
-      in_header: header.tex
-documentclass: article
-classoption: a4paper
-urlcolor: blue
-link-citations: yes
-bibliography: references.bib   
-fontsize: 12pt
+Online supplementary materials further discuss various common extensions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of hybrid fixed-/ random effects model.
+keywords: |
+  Fixed effects, structural equation modeling, lavaan, R, panel analysis
+header-includes: |
+  \usepackage{hyperref}
+  \usepackage[utf8]{inputenc}
+  \usepackage{booktabs}
+  \usepackage{bm}
+  \usepackage{mathtools}
+  \usepackage{amssymb}
+  \usepackage{amsmath}
+  \usepackage{tikz}
+  \usetikzlibrary{arrows}
+  \usepackage[nofiglist]{endfloat}
+  \usepackage{blkarray}
+  \usepackage{setspace}
+  \usepackage{etoolbox}
+  \DeclareMathOperator{\E}{\mathbb{E}}
+  \DeclareMathOperator{\Var}{\mathrm{Var}}
+  \DeclareMathOperator{\Cov}{\mathrm{Cov}}
+  \DeclareMathOperator*{\argmax}{arg\,max}
+  \DeclareMathOperator*{\argmin}{arg\,min}
+  \mathtoolsset{showonlyrefs}
+  \BeforeBeginEnvironment{equation}{\begin{singlespace}\vspace*{-\baselineskip}}
+  \AfterEndEnvironment{equation}{\end{singlespace}\noindent\ignorespaces}
+  \BeforeBeginEnvironment{align}{\begin{singlespace}\vspace*{-\baselineskip}}
+  \AfterEndEnvironment{align}{\end{singlespace}\noindent\ignorespaces}
+  \def\tightlist{}
+  \pagenumbering{gobble}
+output: rticles::tf_article
 ---
@@
 # Script hooks for chunk outputs
-source("scripthooks.R")
+source("../scripthooks.R")
 ```
@@
 \setcounter{page}{1}
+
+\doublespacing
 
@@
 
-The following article outlines the basic idea of FE regression, the particularities of FE in SEM, and shows its implementation in `lavaan`.
+The following article outlines the basic idea of panel regression, the particularities of panel regression SEM, and shows its implementation in `lavaan`.
+The focus on FE panel regression stems from the fact that the assumptions needed to justify other common models, e.g., random effects, are often implausible.
+As we will see, the practical differences between the two are, in any event, very small.
 Using [simulated data](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R), it demonstrates and annotates the code for the most basic FE model and provides an overview of the summary output.
@@
 
-To begin, let us start by reviewing a general panel model [@Bollen2010], also referred to as the 'unobserved effects model' [@Wooldridge2012; @R-plm_a] (we will return to this model in the [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf) of the article when we discuss loosening assumptions)
+To begin, let us start by reviewing a general panel model [@Bollen2010], also referred to as the 'unobserved effects model' [@Wooldridge2012; @R-plm_a] (we will return to this model in the [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf) when we discuss loosening assumptions)
 \begin{align}
@@
 Each assumption will be discussed shortly.
-Apart from that, we typically make the following assumptions [see, e.g., @Wooldridge2002; @Schmidheiny2019] about this model:
+Apart from that, we typically make the following assumptions about this model [see, e.g., @Wooldridge2002; @Schmidheiny2019]:
 
@@
 ## Random effects {#re}
+
+Somewhat unintuitively, the modern approach to FE regression treats the unobserved effect $\alpha_{i}$ as a \textit{random} variable, rather than a fixed parameter to be estimated for each of the cross sectional units [@Wooldridge2002].
+And in fact, this view carries over to the way in which FE regression is applied in SEM.
+So, in order to help facilitate the discussion on the application of FE-SEM later on, it makes sense to spend some time discussing the basic idea of random effects regression (again, as we will see, the practical difference between the two models is very small).
 
@@
 The second part is the potentially controversial assumption: it states that the individual effects are uncorrelated with the independent variables.
-We can use an example to show why this is often controversial: If we are interested in the question of whether married men earn more than unmarried men, then the second part of the strict exogeneity assumption means that a man's marriage status would have to be uncorrelated with all the time-invariant characteristics that could potentially make that man an attractive marriage candidate in the first place; e.g., looks, personality, family's status, profession, etc.
+We can use an intuitive concrete example to show why this is often controversial: If we are interested in the question of whether married men earn more than unmarried men, then the second part of the strict exogeneity assumption means that a man's marriage status would have to be uncorrelated with all the time-invariant characteristics that could potentially make that man an attractive marriage candidate in the first place; e.g., looks, personality, family's status, profession, etc.
 [@Bruederl2015].^[Assuming, for the sake of argument, that these characteristics are constant over time.]
@@
 \end{alignat}
-i.e., the individual effects are conditionally homoscedastic (they are necessarily serially correlated as long as $\sigma_{\alpha}^{2} > 0$) we arrive at the typical random effects structure of the $NT \times NT$ matrix $\bm{\Omega}$:
+i.e., the individual effects are conditionally homoscedastic (they are necessarily serially correlated as long as $\sigma_{\alpha}^{2} > 0$).
+From that, we arrive at the typical random effects structure of the $NT \times NT$ matrix $\bm{\Omega}$:
 
@@
 In practice, however, computational problems can arise with large cross-sectional samples, where it can become difficult to invert the $\hat{\bm{\Omega}}$ matrix.
-One solution is to use 'partial-demeaning' to transform the data before performing simple OLS:
+One solution is to use 'partial-demeaning' to transform the data before performing simple POLS:
 \begin{align}
@@
 
-The RE model can also be estimated in the maximum likelihood framework, where in the associated literature panel models are generally referred to as either mixed models, hierarchical models or longitudinal models.
+The RE model can also be estimated in the maximum likelihood framework, where in the associated literature on panel models are generally referred to as either mixed models, hierarchical models or longitudinal models.
 The typical RE model discussed here is the equivalent to a mixed model with random intercepts and fixed slopes [@R-plm_a].
@@
 The dependent variable at each timepoint is regressed onto the latent variable, see Figure \ref{fig:fesem}.
-Here, the regression weights or 'factor loadings' are fixed to one to represent our assumption that the effect of the time-invariant unobserved heterogeneity is constant over time.
-It also means that the estimated variance of the latent variable is equal to the \textit{average covariance between the wide-format columns of the dependent variable over time}.
-If $y_{it} = \alpha_{i} + \varepsilon_{it}$ is the true data generating process, then the relationship between two units over time is just $\alpha$, regardless of the time distance.
+Here, the regression weights or 'factor loadings' are fixed to one to represent our assumption that the effect of the time-invariant unobserved heterogeneity is constant over time.^[In fact, the initial FE-SEM setup shown in the main article mimics the POLS methods described above in that it assumes constant effects and error variances over time.
+These assumptions can be loosened and tested, as will be shown in the [supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf).
+For now, for the sake of simplicity and comparability, we retain the assumptions associated with the 'pooled' models for the most part.] It also means that the estimated variance of the latent variable is equal to the \textit{average covariance between the wide-format columns of the dependent variable over time}.
+If $y_{it} = \alpha_{i} + \varepsilon_{it}$ is the true data generating process, then the relationship between two units over time is just $\Var(\alpha)$, regardless of the time distance.
 Referring back to the random effects structure of $\bm{\Omega}_{i}$ in Equality \eqref{eq:omeganui} for a generic unit $i$, we see the covariance on all of the off-diagonals is $\sigma^{2}_{\alpha}$.
@@
 
-To show this, consider the following matrix equation of the variances and the nonredundant covariances in a three-wave intercept-only model that follows directly from Equation \eqref{eq:wide} (assuming $\Cov(\varepsilon_{ti},\varepsilon_{si}) = 0, \ t \ne s$), and which we can solve easily with least squares:
+To elaborate on this concept some more, consider the following matrix equation of the variances and the nonredundant covariances in a three-wave intercept-only model that follows directly from Equation \eqref{eq:wide} (assuming $\Cov(\varepsilon_{ti},\varepsilon_{si}) = 0, \ t \ne s$), and which we can solve easily with least squares:
 \begin{align}
@@
 \end{align}
-This shows that if our assumption of constant covariances between the dependent variable across time holds, the estimated variance of $\alpha$ is just what it should be: the average covariance between units of $y$ over time.
+So if, in fact the covariance between any two wide-format columns of $y$ is $\Cov(y_{t},y_{s}) = \Var(\alpha), \ \forall \ s \ne t$, then $\hat{\psi} = \frac{1}{3}\Var(\alpha) + \frac{1}{3}\Var(\alpha) + \frac{1}{3}\Var(\alpha) = \frac{3 \Var(\alpha)}{3} = \Var(\alpha)$.
+This shows that if our assumption about the underlying DGP is correct, i.e., $y_{it} = \alpha_{i} + \varepsilon_{it}$ and $\Cov(\varepsilon_{t},\varepsilon_{s}) = 0, \ \forall \ t \ne s$, then the estimated variance of $\alpha$ is just what it should be: the average covariance between units of $y$ over time.
 Once we add in observed covariates, the estimated covariance of $\alpha$ then become the \textit{conditional} covariance of $y$ over time, given those covariates.
@@
 
-What essentially differentiates an FE from a random effects (RE) model is our assumption concerning the relationship between the unobserved individual effects and the model covariates [@Bollen2010].
+What essentially differentiates an FE from an RE model is our assumption concerning the relationship between the unobserved individual effects and the model covariates [@Bollen2010].
 The FE model assumes that $\E[\alpha x_{t}] \ne 0$.
@@
 ```{r echo=FALSE}
-df <- readRDS("longData.Rda")
-dfw <- readRDS("wideData.Rda")
+df <- readRDS("../longData.Rda")
+dfw <- readRDS("../wideData.Rda")
 ```
@@
 ```{r, echo=FALSE, message=FALSE, error=FALSE, warning=FALSE}
-saveRDS(fe_sem, "fe_sem.Rda")
-saveRDS(fe_sem.fit, "fe_sem.fit.Rda")
+saveRDS(fe_sem, "../fe_sem.Rda")
+saveRDS(fe_sem.fit, "../fe_sem.fit.Rda")
 ```
@@
 Fixed effects regression in SEM has been outlined in well-known articles by [@Allison2011; @Bollen2010; @Teachman2001].
-This article provides a focused look at the implementation of the basic model, as well as common extensions using the `lavaan` package in `R`.
+This article provides a focused look at the implementation of the basic model using the `lavaan` package in `R`.
+The [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf) further discuss common extensions and some tools for evaluating and loosening model assumptions.
 
@@
 The most basic FE-SEM is furthermore the basis for a variety of currently popular extended models, such as Latent Curve Models in general [@Curran2001; @Bollen2004], as well as special implementations like the Dynamic Panel Model [@Allison2017], the Random-Intercept Cross-Lagged Panel Model [@Hamaker2015] and the Latent Curve Model with Structured Residuals [@Curran2014].
-For this reason it is all the more important for researchers to have a good grasp on the method of applying FE-SEM.
+For this reason it is all the more important for researchers to have a good grasp on the method of applying panel regression in SEM, and understanding the intuition of controlling for time-invariant confounders.
 This article is meant to serve as a consolidated resource for researchers looking for concrete advise on specifying FE and more general panel models in SEM.
 
-\newpage
-
-# References {-}
-
-<div id="refs"></div>
+# References

### 2020-07-28 article_format/article_format.tex
Changed the formating, made some minor edits

--- 2020-07-28 article.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
-%
-\documentclass[
-  12pt,
-  a4paper]{article}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={A closer look at fixed effects regression in structural equation modeling using},
-  pdfauthor={Henrik Kenneth Andersen},
-  colorlinks=true,
-  linkcolor=Maroon,
-  filecolor=Maroon,
-  citecolor=Blue,
-  urlcolor=blue,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[margin=1in]{geometry}
+% interactcadsample.tex
+% v1.03 - April 2017
+
+\documentclass[]{interact}
+
+\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
+\usepackage{subfigure}% Support for small, `sub' figures and tables
+%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required
+
+\usepackage{natbib}% Citation support using natbib.sty
+\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
+\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty
+
+\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{corollary}[theorem]{Corollary}
+\newtheorem{proposition}[theorem]{Proposition}
+
+\theoremstyle{definition}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{example}[theorem]{Example}
+
+\theoremstyle{remark}
+\newtheorem{remark}{Remark}
+\newtheorem{notation}{Notation}
+
+% see https://stackoverflow.com/a/47122900
 \usepackage{color}
@@
 \newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-% ----- Packages
-
+
+\usepackage{hyperref}
+\usepackage[utf8]{inputenc}
 \usepackage{booktabs}
-\usepackage{titling}
 \usepackage{bm}
@@
 \usepackage{amsmath}
-\usepackage{subfig}
 \usepackage{tikz}
@@
 \usepackage[nofiglist]{endfloat}
-%\floatsetup[figure]{capposition=top}
-%\usepackage{kbordermatrix}
 \usepackage{blkarray}
@@
 \usepackage{etoolbox}
-\doublespacing
-
-% ----- Math operators
-
 \DeclareMathOperator{\E}{\mathbb{E}}
@@
 \DeclareMathOperator*{\argmin}{arg\,min}
-
-% ----- Graphics
-
-%\graphicspath{{./images/}}
-
-% ----- Equation labels
-
 \mathtoolsset{showonlyrefs}
-
-% ----- Page numbering
-
-\pagenumbering{gobble}
-
-% ----- Spacing
-
-%\renewenvironment{knitrout}{\vspace{10em}}{}
-
 \BeforeBeginEnvironment{equation}{\begin{singlespace}\vspace*{-\baselineskip}}
@@
 \AfterEndEnvironment{align}{\end{singlespace}\noindent\ignorespaces}
-
-\title{\singlespacing A closer look at fixed effects regression in structural
-equation modeling using \texttt{lavaan}}
-\author{Henrik Kenneth Andersen}
-\date{2020-07-27}
+\def\tightlist{}
+\pagenumbering{gobble}
 
 \begin{document}
+
+\articletype{}
+
+\title{A closer look at fixed effects regression in structural equation
+modeling using \texttt{lavaan}}
+
+\author{\name{Henrik Kenneth Andersen$^{a}$}
+\affil{$^{a}$Chemnitz University of Technology, Institute of Sociology, Chair for
+Empirical Social Research, Thüringer Weg 9, 09126 Chemnitz, Germany}
+}
+
+\thanks{CONTACT Henrik Kenneth Andersen. Email: \href{mailto:henrik.andersen@soziologie.tu-chemnitz.de}{\nolinkurl{henrik.andersen@soziologie.tu-chemnitz.de}}}
+
 \maketitle
+
 \begin{abstract}
@@
 underlying model specification, syntax, and summary output. Online
-supplementary materials further discuss various common extentsions to
-the basic fixed-effect model, demonstrating how to relax model
-assumptions, deal with measurement error in both the dependent and
-independent variables, and include time-invariant predictors in a type
-of hybrid fixed-/ random effects model. extbf\{Keywords: Fixed effects,
-structural equation modeling, lavaan, R, panel analysis\}
+supplementary materials further discuss various common extensions to the
+basic fixed-effect model, demonstrating how to relax model assumptions,
+deal with measurement error in both the dependent and independent
+variables, and include time-invariant predictors in a type of hybrid
+fixed-/ random effects model.
 \end{abstract}
+
+\begin{keywords}
+Fixed effects, structural equation modeling, lavaan, R, panel analysis
+\end{keywords}
 
@@
 
+\doublespacing
+
 \hypertarget{intro}{%
@@
 
-Several years ago, Curran and Bauer
-(\protect\hyperlink{ref-Curran2011}{2011}) reflected positively on the
+Several years ago, \citet{Curran2011} reflected positively on the
 growing use of panel studies in empirical social research. Some of the
@@
 There are a number of articles describing basic concept of panel model
-regression, and FE regression in SEM (e.g., Allison
-\protect\hyperlink{ref-Allison2011}{2011}; Bollen and Brand
-\protect\hyperlink{ref-Bollen2010}{2010}; Teachman et al.
-\protect\hyperlink{ref-Teachman2001}{2001}). This article is intended as
-a \textit{practical guide} for researchers looking for in-depth help
-with specifying FE models in SEM. It focuses on the \texttt{lavaan}
-(Rosseel \protect\hyperlink{ref-R-lavaan}{2012}) package for \texttt{R}
-(R Core Team \protect\hyperlink{ref-R-base}{2017}). While \texttt{Mplus}
-(Muthén and Muthén, \protect\hyperlink{ref-Mplus}{n.d.}) is arguably the
-most robust SEM software currently available (in terms of features like
-alignment, latent variable interactions, for example), the
-\texttt{lavaan} package has many benefits. First, like \texttt{R} it is
-open source and completely free. For researchers dipping their toes into
-SEM, there is no financial barrier to try, and no risk if they decide it
-is not for them. Second, the implementation of \texttt{lavaan} in the
-larger \texttt{R} environment is an enormous advantage. Instead of
-poring over reams of plain text, copying out coefficients by hand, every
-part of the \texttt{lavaan} output is available as an object. This means
-that all aspects of the model, from fit indices, to coefficients and
-standard errors, to the model matrices, can be accessed and easily
-integrated into tables and plots. Furthermore, \texttt{R} can be used
-for a great deal of applications. It can be used to manage and
-manipulate as well as simulate data, perform symbolic algebra, run more
-traditional analyses (e.g., multiple regression, logistic regression,
-principal component analysis), etc. Once one is comfortable using
-\texttt{R}, there is no longer any need to switch between different
-software for data preparation and analysis.
-
-The following article outlines the basic idea of FE regression, the
-particularities of FE in SEM, and shows its implementation in
-\texttt{lavaan}. Using
+regression, and FE regression in SEM
+\citep[e.g.,][]{Allison2011, Bollen2010, Teachman2001}. This article is
+intended as a \textit{practical guide} for researchers looking for
+in-depth help with specifying FE models in SEM. It focuses on the
+\texttt{lavaan} \citep{R-lavaan} package for \texttt{R} \citep{R-base}.
+While \texttt{Mplus} \citep{Mplus} is arguably the most robust SEM
+software currently available (in terms of features like alignment,
+latent variable interactions, for example), the \texttt{lavaan} package
+has many benefits. First, like \texttt{R} it is open source and
+completely free. For researchers dipping their toes into SEM, there is
+no financial barrier to try, and no risk if they decide it is not for
+them. Second, the implementation of \texttt{lavaan} in the larger
+\texttt{R} environment is an enormous advantage. Instead of poring over
+reams of plain text, copying out coefficients by hand, every part of the
+\texttt{lavaan} output is available as an object. This means that all
+aspects of the model, from fit indices, to coefficients and standard
+errors, to the model matrices, can be accessed and easily integrated
+into tables and plots. Furthermore, \texttt{R} can be used for a great
+deal of applications. It can be used to manage and manipulate as well as
+simulate data, perform symbolic algebra, run more traditional analyses
+(e.g., multiple regression, logistic regression, principal component
+analysis), etc. Once one is comfortable using \texttt{R}, there is no
+longer any need to switch between different software for data
+preparation and analysis.
+
+The following article outlines the basic idea of panel regression, the
+particularities of panel regression SEM, and shows its implementation in
+\texttt{lavaan}. The focus on FE panel regression stems from the fact
+that the assumptions needed to justify other common models, e.g., random
+effects, are often implausible. As we will see, the practical
+differences between the two are, in any event, very small. Using
 \href{https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R}{simulated
@@
 
-To begin, let us start by reviewing a general panel model (Bollen and
-Brand \protect\hyperlink{ref-Bollen2010}{2010}), also referred to as the
-`unobserved effects model' (Wooldridge
-\protect\hyperlink{ref-Wooldridge2012}{2012}; Croissant and Millo
-\protect\hyperlink{ref-R-plm_a}{2008}) (we will return to this model in
-the
+To begin, let us start by reviewing a general panel model
+\citep{Bollen2010}, also referred to as the `unobserved effects model'
+\citep{Wooldridge2012, R-plm_a} (we will return to this model in the
 \href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
-supplementary materials} of the article when we discuss loosening
-assumptions) \begin{align}
+supplementary materials} when we discuss loosening assumptions)
+\begin{align}
 y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \alpha_{i} + \varepsilon_{it} \label{eq:gpm}
@@
 shortly. Apart from that, we typically make the following assumptions
-(see, e.g., Wooldridge \protect\hyperlink{ref-Wooldridge2002}{2002};
-Schmidheiny \protect\hyperlink{ref-Schmidheiny2019}{2019}) about this
-model:
+about this model \citep[see, e.g.,][]{Wooldridge2002, Schmidheiny2019}:
 
@@
 
+Somewhat unintuitively, the modern approach to FE regression treats the
+unobserved effect \(\alpha_{i}\) as a \textit{random} variable, rather
+than a fixed parameter to be estimated for each of the cross sectional
+units \citep{Wooldridge2002}. And in fact, this view carries over to the
+way in which FE regression is applied in SEM. So, in order to help
+facilitate the discussion on the application of FE-SEM later on, it
+makes sense to spend some time discussing the basic idea of random
+effects regression (again, as we will see, the practical difference
+between the two models is very small).
+
 For the random effects (RE) model, we define a composite error term:
@@
 exogeneity assumption in the RE model implies \begin{align}
-\E[\varepsilon_{it} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = 0, \\
-\E[\alpha_{i} |\bm{X}_{i}, \bm{z}_{i}] = \E[\alpha_{i}] & = 0,
+\mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = 0, \\
+\mathop{\mathrm{\mathbb{E}}}[\alpha_{i} |\bm{X}_{i}, \bm{z}_{i}] = \mathop{\mathrm{\mathbb{E}}}[\alpha_{i}] & = 0,
 \end{align} where \(\bm{X}_{i} = \bm{x}_{i1}, ..., \bm{x}_{iT}\). For
@@
 uncorrelated, i.e.,
-\(\E[\bm{x}_{is}^{\intercal}\varepsilon_{it}] = \bm{0}\) and
-\(\E[\bm{z}_{i}^{\intercal}\varepsilon_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T\)
-(Wooldridge \protect\hyperlink{ref-Wooldridge2002}{2002}; Brüderl and
-Ludwig \protect\hyperlink{ref-Bruederl2015}{2015}). We assume the
-idiosyncratic errors are further independent of the individual effects,
-which implies \(\E[\alpha_{i}\varepsilon_{it}] = 0\).
+\(\mathop{\mathrm{\mathbb{E}}}[\bm{x}_{is}^{\intercal}\varepsilon_{it}] = \bm{0}\)
+and
+\(\mathop{\mathrm{\mathbb{E}}}[\bm{z}_{i}^{\intercal}\varepsilon_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T\)
+\citep{Wooldridge2002, Bruederl2015}. We assume the idiosyncratic errors
+are further independent of the individual effects, which implies
+\(\mathop{\mathrm{\mathbb{E}}}[\alpha_{i}\varepsilon_{it}] = 0\).
 
@@
 that the individual effects are uncorrelated with the independent
-variables. We can use an example to show why this is often
-controversial: If we are interested in the question of whether married
-men earn more than unmarried men, then the second part of the strict
-exogeneity assumption means that a man's marriage status would have to
-be uncorrelated with all the time-invariant characteristics that could
-potentially make that man an attractive marriage candidate in the first
-place; e.g., looks, personality, family's status, profession, etc.
-(Brüderl and Ludwig
-\protect\hyperlink{ref-Bruederl2015}{2015}).\footnote{Assuming, for the
-  sake of argument, that these characteristics are constant over time.}
+variables. We can use an intuitive concrete example to show why this is
+often controversial: If we are interested in the question of whether
+married men earn more than unmarried men, then the second part of the
+strict exogeneity assumption means that a man's marriage status would
+have to be uncorrelated with all the time-invariant characteristics that
+could potentially make that man an attractive marriage candidate in the
+first place; e.g., looks, personality, family's status, profession, etc.
+\citep{Bruederl2015}.\footnote{Assuming, for the sake of argument, that
+  these characteristics are constant over time.}
 
@@
 of the errors
-\(\bm{\Omega}_{i} = \E[\bm{\nu}_{i}\bm{\nu}_{i}^{\intercal}]\) can be
-constructed. However, the standard random effects model adds the
+\(\bm{\Omega}_{i} = \mathop{\mathrm{\mathbb{E}}}[\bm{\nu}_{i}\bm{\nu}_{i}^{\intercal}]\)
+can be constructed. However, the standard random effects model adds the
 additional assumptions \begin{alignat}{3}
-\E[\varepsilon_{it}^{2} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \E[\varepsilon_{it}^{2}] && = \sigma_{\varepsilon}^{2}, && \ t = 1, ..., T, \\
-\E[\varepsilon_{it}\varepsilon_{is}| \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \E[\varepsilon_{it}\varepsilon_{is}] && = 0, && \ \forall \ t \ne s 
+\mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}^{2} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}^{2}] && = \sigma_{\varepsilon}^{2}, && \ t = 1, ..., T, \\
+\mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}\varepsilon_{is}| \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}\varepsilon_{is}] && = 0, && \ \forall \ t \ne s 
 \end{alignat} i.e., the idiosyncratic errors are conditionally
 homoscedastic and serially uncorrelated and \begin{alignat}{2}
-\E[\alpha_{i}^{2}| \bm{X}_{i},\bm{z}_{i}] & = \E[\alpha_{i}^{2}] && = \sigma^{2}_{\alpha}
+\mathop{\mathrm{\mathbb{E}}}[\alpha_{i}^{2}| \bm{X}_{i},\bm{z}_{i}] & = \mathop{\mathrm{\mathbb{E}}}[\alpha_{i}^{2}] && = \sigma^{2}_{\alpha}
 \end{alignat} i.e., the individual effects are conditionally
 homoscedastic (they are necessarily serially correlated as long as
-\(\sigma_{\alpha}^{2} > 0\)) we arrive at the typical random effects
-structure of the \(NT \times NT\) matrix \(\bm{\Omega}\):
+\(\sigma_{\alpha}^{2} > 0\)). From that, we arrive at the typical random
+effects structure of the \(NT \times NT\) matrix \(\bm{\Omega}\):
 
@@
 \(\sigma^{2}_{\nu} = \sigma^{2}_{\alpha} + \sigma^{2}_{\varepsilon}\)
-(Wooldridge \protect\hyperlink{ref-Wooldridge2002}{2002}; Schmidheiny
-\protect\hyperlink{ref-Schmidheiny2019}{2019}). This means that in the
+\citep{Wooldridge2002, Schmidheiny2019}. This means that in the
 conditional covariance matrix of the errors, given the time-varying and
@@
 \(\hat{\bm{\Omega}}\) matrix. One solution is to use `partial-demeaning'
-to transform the data before performing simple OLS: \begin{align}
+to transform the data before performing simple POLS: \begin{align}
 (y_{it} - \theta\bar{y}_{i}) & = (\bm{x}_{it} - \theta\bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \theta \bar{\bm{z}}_{i})\bm{\gamma} + (\varepsilon_{it} - \theta \bar{\varepsilon}_{i}) \label{eq:partialdemean}
@@
 \(\bar{\varepsilon}_{i} = T^{-1}\sum_{t = 1}^{T}\varepsilon_{it}\)
-(Croissant and Millo \protect\hyperlink{ref-R-plm_a}{2008}).
+\citep{R-plm_a}.
 
 The RE model can also be estimated in the maximum likelihood framework,
-where in the associated literature panel models are generally referred
-to as either mixed models, hierarchical models or longitudinal models.
-The typical RE model discussed here is the equivalent to a mixed model
-with random intercepts and fixed slopes (Croissant and Millo
-\protect\hyperlink{ref-R-plm_a}{2008}). Under the assumption of
-normality, along with homoscedasticity and serially uncorrelated errors,
-the maximum likelihood estimator is the same as the OLS estimator. For
-more on the topic of mixed models, see for example Bates et al.
-(\protect\hyperlink{ref-R-lme4}{2015}), Bates
-(\protect\hyperlink{ref-Bates2010}{2010}).
+where in the associated literature on panel models are generally
+referred to as either mixed models, hierarchical models or longitudinal
+models. The typical RE model discussed here is the equivalent to a mixed
+model with random intercepts and fixed slopes \citep{R-plm_a}. Under the
+assumption of normality, along with homoscedasticity and serially
+uncorrelated errors, the maximum likelihood estimator is the same as the
+OLS estimator. For more on the topic of mixed models, see for example
+\citet{R-lme4}, \citet{Bates2010}.
 
@@
 \textit{not independent} of the model covariates, i.e.,
-\(\E[\alpha_{i}|\bm{X}_{i},\bm{z}_{i}] \ne \E[\alpha_{i}] \ne 0\). Under
-this assumption, grouping the individual effects in with the composite
-error will cause the coefficients of interest, here specifically
-\(\bm{\beta}\) to be inconsistent (Wooldridge
-\protect\hyperlink{ref-Wooldridge2002}{2002},
-\protect\hyperlink{ref-Wooldridge2012}{2012}). We write the model
-therefore again in terms of the general panel model in Equation
-\eqref{eq:gpm} with separate individual effects and idiosyncratic error
-terms. In order to drop assumptions involving the individual effects, a
-number of methods are available (e.g., differencing, least squares dummy
-variable regression), but the most common approach is to \textit{demean}
-the equation (Brüderl and Ludwig
-\protect\hyperlink{ref-Bruederl2015}{2015}). Demeaning is the same as
-the transformation applied in Equation \eqref{eq:partialdemean} in the
+\(\mathop{\mathrm{\mathbb{E}}}[\alpha_{i}|\bm{X}_{i},\bm{z}_{i}] \ne \mathop{\mathrm{\mathbb{E}}}[\alpha_{i}] \ne 0\).
+Under this assumption, grouping the individual effects in with the
+composite error will cause the coefficients of interest, here
+specifically \(\bm{\beta}\) to be inconsistent
+\citep{Wooldridge2002, Wooldridge2012}. We write the model therefore
+again in terms of the general panel model in Equation \eqref{eq:gpm}
+with separate individual effects and idiosyncratic error terms. In order
+to drop assumptions involving the individual effects, a number of
+methods are available (e.g., differencing, least squares dummy variable
+regression), but the most common approach is to \textit{demean} the
+equation \citep{Bruederl2015}. Demeaning is the same as the
+transformation applied in Equation \eqref{eq:partialdemean} in the
 special case where \(\theta = 1\). I.e., demeaning involves subtracting
@@
 on the idiosyncratic errors, i.e.,
-\(\E[\ddot{\varepsilon}_{it}|\bm{\ddot{x}}_{it}] = \E[\ddot{\varepsilon}_{it}] = 0\)
+\(\mathop{\mathrm{\mathbb{E}}}[\ddot{\varepsilon}_{it}|\bm{\ddot{x}}_{it}] = \mathop{\mathrm{\mathbb{E}}}[\ddot{\varepsilon}_{it}] = 0\)
 which also implies
-\(\E[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T\)
-(Brüderl and Ludwig \protect\hyperlink{ref-Bruederl2015}{2015};
-Wooldridge \protect\hyperlink{ref-Wooldridge2002}{2002}).
+\(\mathop{\mathrm{\mathbb{E}}}[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T\)
+\citep{Bruederl2015, Wooldridge2002}.
 
@@
 \bm{\beta}_{FE} & = (\ddot{\bm{X}}^{\intercal}\ddot{\bm{X}})^{-1}\ddot{\bm{X}}^{\intercal}\ddot{\bm{y}}
-\end{align} (Brüderl and Ludwig
-\protect\hyperlink{ref-Bruederl2015}{2015}). The downside to this
-approach is that no time-invariant predictors can be included in the
-model. However, there are alternative approaches in the random effects
-and mixed model frameworks that allow them to be included. These models
-are sometimes referred to as `within-between' or `hybrid' models, often
-based on the Chamberlain (\protect\hyperlink{ref-Chamberlain1980}{1980})
-and Mundlak (\protect\hyperlink{ref-Mundlak1978}{1978}) approaches, see
-for example Bell, Fairbrother, and Jones
-(\protect\hyperlink{ref-Bell2018}{2018}); Allison
-(\protect\hyperlink{ref-Allison2011}{2011}); Schunck
-(\protect\hyperlink{ref-Schunck2013}{2013}); Enders and Tofighi
-(\protect\hyperlink{ref-Enders2007}{2007}). In the
+\end{align} \citep{Bruederl2015}. The downside to this approach is that
+no time-invariant predictors can be included in the model. However,
+there are alternative approaches in the random effects and mixed model
+frameworks that allow them to be included. These models are sometimes
+referred to as `within-between' or `hybrid' models, often based on the
+Chamberlain \citeyearpar{Chamberlain1980} and Mundlak
+\citeyearpar{Mundlak1978} approaches, see for example \citet{Bell2018};
+\citet{Allison2011}; \citet{Schunck2013}; \citet{Enders2007}. In the
 \href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
@@
 hypothetical concepts, unobserved heterogeneity has historically been
-one of the main uses of latent variables in SEM (Skrondal and
-Rabe-Hesketh \protect\hyperlink{ref-Skrondal2004}{2004}).
+one of the main uses of latent variables in SEM \citep{Skrondal2004}.
 
@@
 covariance between any two of the new wide vectors
-\(\Cov(y_{ti},y_{si}) = \Var(\alpha_{i}), \ t \ne s\). Otherwise, when
-\(t = s\), the covariance
-\(\Cov(y_{ti},y_{ti}) = \Var(\alpha_{i}) + \Var(\varepsilon_{ti})\).
+\(\mathop{\mathrm{\mathrm{Cov}}}(y_{ti},y_{si}) = \mathop{\mathrm{\mathrm{Var}}}(\alpha_{i}), \ t \ne s\).
+Otherwise, when \(t = s\), the covariance
+\(\mathop{\mathrm{\mathrm{Cov}}}(y_{ti},y_{ti}) = \mathop{\mathrm{\mathrm{Var}}}(\alpha_{i}) + \mathop{\mathrm{\mathrm{Var}}}(\varepsilon_{ti})\).
 This is the structure we saw above in a typical element of
@@
 fixed to one to represent our assumption that the effect of the
-time-invariant unobserved heterogeneity is constant over time. It also
-means that the estimated variance of the latent variable is equal to the
+time-invariant unobserved heterogeneity is constant over time.\footnote{In
+  fact, the initial FE-SEM setup shown in the main article mimics the
+  POLS methods described above in that it assumes constant effects and
+  error variances over time. These assumptions can be loosened and
+  tested, as will be shown in the
+  \href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{supplementary
+  materials}. For now, for the sake of simplicity and comparability, we
+  retain the assumptions associated with the `pooled' models for the
+  most part.} It also means that the estimated variance of the latent
+variable is equal to the
 \textit{average covariance between the wide-format columns of the dependent variable over time}.
@@
 generating process, then the relationship between two units over time is
-just \(\alpha\), regardless of the time distance. Referring back to the
-random effects structure of \(\bm{\Omega}_{i}\) in Equality
-\eqref{eq:omeganui} for a generic unit \(i\), we see the covariance on
-all of the off-diagonals is \(\sigma^{2}_{\alpha}\). And, as we know,
-the average of something that does not change is that thing itself.
-I.e., if \(T(T-1)/2 = h\) is the number of elements on either the upper-
-or lower triangle of \(\bm{\Omega}_{i}\), then we have
+just \(\mathop{\mathrm{\mathrm{Var}}}(\alpha)\), regardless of the time
+distance. Referring back to the random effects structure of
+\(\bm{\Omega}_{i}\) in Equality \eqref{eq:omeganui} for a generic unit
+\(i\), we see the covariance on all of the off-diagonals is
+\(\sigma^{2}_{\alpha}\). And, as we know, the average of something that
+does not change is that thing itself. I.e., if \(T(T-1)/2 = h\) is the
+number of elements on either the upper- or lower triangle of
+\(\bm{\Omega}_{i}\), then we have
 \(h^{-1}\sum_{i=1}^{h}\sigma^{2}_{\alpha} = \frac{h \sigma^{2}_{\alpha}}{h} = \sigma^{2}_{\alpha}\).
 
-To show this, consider the following matrix equation of the variances
-and the nonredundant covariances in a three-wave intercept-only model
-that follows directly from Equation \eqref{eq:wide} (assuming
-\(\Cov(\varepsilon_{ti},\varepsilon_{si}) = 0, \ t \ne s\)), and which
-we can solve easily with least squares: \begin{align}
+To elaborate on this concept some more, consider the following matrix
+equation of the variances and the nonredundant covariances in a
+three-wave intercept-only model that follows directly from Equation
+\eqref{eq:wide} (assuming
+\(\mathop{\mathrm{\mathrm{Cov}}}(\varepsilon_{ti},\varepsilon_{si}) = 0, \ t \ne s\)),
+and which we can solve easily with least squares: \begin{align}
 \bm{A}\bm{x} & = \bm{b} \\
@@
 \begin{pmatrix}
-\Var(y_{1}) \\
-\Cov(y_{2},y_{1}) \\
-\Cov(y_{3},y_{1}) \\
-\Var(y_{2}) \\
-\Cov(y_{3},y_{2}) \\
-\Var(y_{3})
+\mathop{\mathrm{\mathrm{Var}}}(y_{1}) \\
+\mathop{\mathrm{\mathrm{Cov}}}(y_{2},y_{1}) \\
+\mathop{\mathrm{\mathrm{Cov}}}(y_{3},y_{1}) \\
+\mathop{\mathrm{\mathrm{Var}}}(y_{2}) \\
+\mathop{\mathrm{\mathrm{Cov}}}(y_{3},y_{2}) \\
+\mathop{\mathrm{\mathrm{Var}}}(y_{3})
 \end{pmatrix}
-\end{align} where \(\psi = \Var(\alpha)\),
-\(\phi_{t} = \Var(\varepsilon_{t})\). We can solve this equation to show
-\begin{align}
+\end{align} where \(\psi = \mathop{\mathrm{\mathrm{Var}}}(\alpha)\),
+\(\phi_{t} = \mathop{\mathrm{\mathrm{Var}}}(\varepsilon_{t})\). We can
+solve this equation to show \begin{align}
 \bm{A}\bm{x} & = \bm{b} \\
@@
 \begin{pmatrix}
-\frac{1}{3}\Cov(y_{2},y_{1}) + \frac{1}{3}\Cov(y_{3},y_{1}) + \frac{1}{3}\Cov(y_{3},y_{2}) \\
-\Var(y_{1}) - \hat{\psi} \\
-\Var(y_{2}) - \hat{\psi} \\
-\Var(y_{3}) - \hat{\psi}
+\frac{1}{3}\mathop{\mathrm{\mathrm{Cov}}}(y_{2},y_{1}) + \frac{1}{3}\mathop{\mathrm{\mathrm{Cov}}}(y_{3},y_{1}) + \frac{1}{3}\mathop{\mathrm{\mathrm{Cov}}}(y_{3},y_{2}) \\
+\mathop{\mathrm{\mathrm{Var}}}(y_{1}) - \hat{\psi} \\
+\mathop{\mathrm{\mathrm{Var}}}(y_{2}) - \hat{\psi} \\
+\mathop{\mathrm{\mathrm{Var}}}(y_{3}) - \hat{\psi}
 \end{pmatrix}.
-\end{align} This shows that if our assumption of constant covariances
-between the dependent variable across time holds, the estimated variance
-of \(\alpha\) is just what it should be: the average covariance between
-units of \(y\) over time. Once we add in observed covariates, the
-estimated covariance of \(\alpha\) then become the \textit{conditional}
-covariance of \(y\) over time, given those covariates.
+\end{align} So if, in fact the covariance between any two wide-format
+columns of \(y\) is
+\(\mathop{\mathrm{\mathrm{Cov}}}(y_{t},y_{s}) = \mathop{\mathrm{\mathrm{Var}}}(\alpha), \ \forall \ s \ne t\),
+then
+\(\hat{\psi} = \frac{1}{3}\mathop{\mathrm{\mathrm{Var}}}(\alpha) + \frac{1}{3}\mathop{\mathrm{\mathrm{Var}}}(\alpha) + \frac{1}{3}\mathop{\mathrm{\mathrm{Var}}}(\alpha) = \frac{3 \mathop{\mathrm{\mathrm{Var}}}(\alpha)}{3} = \mathop{\mathrm{\mathrm{Var}}}(\alpha)\).
+This shows that if our assumption about the underlying DGP is correct,
+i.e., \(y_{it} = \alpha_{i} + \varepsilon_{it}\) and
+\(\mathop{\mathrm{\mathrm{Cov}}}(\varepsilon_{t},\varepsilon_{s}) = 0, \ \forall \ t \ne s\),
+then the estimated variance of \(\alpha\) is just what it should be: the
+average covariance between units of \(y\) over time. Once we add in
+observed covariates, the estimated covariance of \(\alpha\) then become
+the \textit{conditional} covariance of \(y\) over time, given those
+covariates.
 
@@
 matrix notation. There are a number of different model notations (see,
-for example Bollen (\protect\hyperlink{ref-Bollen1989}{1989}) for an
-overview), but the one that will serve us best is one that was proposed
-by Graff (\protect\hyperlink{ref-Graff1979}{1979}): \begin{align}
+for example \citet{Bollen1989} for an overview), but the one that will
+serve us best is one that was proposed by \citet{Graff1979}:
+\begin{align}
 \bm{y}^{+} & = \bm{\Lambda_{y}}^{+} \bm{\eta}^{+}, \\
@@
 is \((p + q) \times (p + q + m + n)\) and \(\bm{B}\) is
-\((p + q + m + n) \times (p + q + m + n)\) (Bollen
-\protect\hyperlink{ref-Bollen1989}{1989}).
+\((p + q + m + n) \times (p + q + m + n)\) \citep{Bollen1989}.
 
@@
 between latent variables, observed ones are only used as indicators, see
-for example Bollen (\protect\hyperlink{ref-Bollen1989}{1989}), Kline
-(\protect\hyperlink{ref-Kline2016}{2016})). It also allows \(\bm{\xi}\),
-i.e., any latent exogenous variables, to influence \(\bm{y}\) directly.
-These two scenarios cover the traditional FE model with observed
-variables, and one in which latent variables are used to account for
-measurement error in the independent variables. It is also consistent
-with the notation used for these models in \texttt{lavaan}. In fact,
-\texttt{lavaan} switches automatically between matrix notations
+for example \citet{Bollen1989}, \citet{Kline2016}). It also allows
+\(\bm{\xi}\), i.e., any latent exogenous variables, to influence
+\(\bm{y}\) directly. These two scenarios cover the traditional FE model
+with observed variables, and one in which latent variables are used to
+account for measurement error in the independent variables. It is also
+consistent with the notation used for these models in \texttt{lavaan}.
+In fact, \texttt{lavaan} switches automatically between matrix notations
 depending on the specified model. That means the matrix representation
@@
 assumption of multivariate normality (otherwise just imposed on the
-dependent variables) also applies to the independent ones (Skrondal and
-Rabe-Hesketh \protect\hyperlink{ref-Skrondal2004}{2004}, 75). This can
-be problematic for noncontinuous independent variables like sex,
-nationality dummies, marriage status (married/unmarried), etc. See
-Skrondal and Rabe-Hesketh (\protect\hyperlink{ref-Skrondal2004}{2004})
-for more on this topic.
+dependent variables) also applies to the independent ones
+\citep[p.~75]{Skrondal2004}. This can be problematic for noncontinuous
+independent variables like sex, nationality dummies, marriage status
+(married/unmarried), etc. See \citet{Skrondal2004} for more on this
+topic.
 
@@
 
-What essentially differentiates an FE from a random effects (RE) model
-is our assumption concerning the relationship between the unobserved
-individual effects and the model covariates (Bollen and Brand
-\protect\hyperlink{ref-Bollen2010}{2010}). The FE model assumes that
-\(\E[\alpha x_{t}] \ne 0\). As such, if we fail to control for the
-correlation of the covariate and the time-invariant part of the error,
-then the coefficient of interest, here \(\beta\), will be biased. Our
-assumption regarding whether the individual effects are correlated with
-the model covariates occurs in
-\(\E[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] = \bm{\Psi}\), the
-covariance matrix of the errors \begin{align}
-\bm{y}^{+}\bm{y}^{+ \intercal} & = \E[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})^{\intercal}] \\
- & = \E[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\zeta}^{+ \intercal}(\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal})] \\
- & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \E[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal} \\
+What essentially differentiates an FE from an RE model is our assumption
+concerning the relationship between the unobserved individual effects
+and the model covariates \citep{Bollen2010}. The FE model assumes that
+\(\mathop{\mathrm{\mathbb{E}}}[\alpha x_{t}] \ne 0\). As such, if we
+fail to control for the correlation of the covariate and the
+time-invariant part of the error, then the coefficient of interest, here
+\(\beta\), will be biased. Our assumption regarding whether the
+individual effects are correlated with the model covariates occurs in
+\(\mathop{\mathrm{\mathbb{E}}}[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] = \bm{\Psi}\),
+the covariance matrix of the errors \begin{align}
+\bm{y}^{+}\bm{y}^{+ \intercal} & = \mathop{\mathrm{\mathbb{E}}}[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})^{\intercal}] \\
+ & = \mathop{\mathrm{\mathbb{E}}}[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\zeta}^{+ \intercal}(\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal})] \\
+ & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \mathop{\mathrm{\mathbb{E}}}[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal} \\
  & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \bm{\Psi} (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal}.
@@
 \begin{align}
-\bm{\Psi} & = \E
+\bm{\Psi} & = \mathop{\mathrm{\mathbb{E}}}
 \begin{blockarray}{cccccccc}
@@
 here and throughout mean-centered variables: \begin{align}
-\Cov(y_{t},x_{t}) & = \E[y_{t}x_{t}] \\
- & = \E[(\alpha + \beta x_{t} + \varepsilon_{t})x_{t}] \\
- & = \E[\alpha x_{t} + \beta x_{t}^{2} + \varepsilon_{t}x_{t}] \\
- & = \Cov(\alpha, x_{t}) + \beta \Var(x_{t}) \\
-\hat{\beta}_{FE-SEM} & = \frac{\Cov(y_{t},x_{t}) - \Cov(\alpha, x_{t})}{\Var(x_{t})}. 
+\mathop{\mathrm{\mathrm{Cov}}}(y_{t},x_{t}) & = \mathop{\mathrm{\mathbb{E}}}[y_{t}x_{t}] \\
+ & = \mathop{\mathrm{\mathbb{E}}}[(\alpha + \beta x_{t} + \varepsilon_{t})x_{t}] \\
+ & = \mathop{\mathrm{\mathbb{E}}}[\alpha x_{t} + \beta x_{t}^{2} + \varepsilon_{t}x_{t}] \\
+ & = \mathop{\mathrm{\mathrm{Cov}}}(\alpha, x_{t}) + \beta \mathop{\mathrm{\mathrm{Var}}}(x_{t}) \\
+\hat{\beta}_{FE-SEM} & = \frac{\mathop{\mathrm{\mathrm{Cov}}}(y_{t},x_{t}) - \mathop{\mathrm{\mathrm{Cov}}}(\alpha, x_{t})}{\mathop{\mathrm{\mathrm{Var}}}(x_{t})}. 
 \end{align} This should make intuitive sense. From the observed
@@
 dividing by the variance of the independent variable, as usual. For the
-RE model, we assume \(\E[\alpha x_{t}] = 0\) and the equation reduces to
-\(\hat{\beta}_{RE-SEM} = \Cov(y_{t},x_{t})/\Var(x_{t})\). The rest of
-the model-implied covariance matrix results from
+RE model, we assume \(\mathop{\mathrm{\mathbb{E}}}[\alpha x_{t}] = 0\)
+and the equation reduces to
+\(\hat{\beta}_{RE-SEM} = \mathop{\mathrm{\mathrm{Cov}}}(y_{t},x_{t})/\mathop{\mathrm{\mathrm{Var}}}(x_{t})\).
+The rest of the model-implied covariance matrix results from
 \(\bm{y}^{+}\bm{y}^{+ \intercal}\).
@@
 Specifying the most basic fixed effects model, like the one shown in
-Bollen and Brand (\protect\hyperlink{ref-Bollen2010}{2010}) (the same
-model as Equation \eqref{eq:fe} but with just one time-varying
-predictor) involves four components. First, we define the latent
-individual effects variable using the \texttt{=\textasciitilde{}}
-`measured by' or `manifested by' (Rosseel
-\protect\hyperlink{ref-R-lavaan}{2012}) operator at the same time
-constraining the factor loadings at each timepoint to one. I will call
-the latent variable \texttt{a} to stand for \(\alpha\). Constraining all
-of the factor loadings to one reflects our implicit assumption that the
-combined effect of the unit-specific unobserved factors is constant over
-time. This is the default behaviour of traditional POLS-based approaches
-to FE that use the stacked long-format data.
+\citet{Bollen2010} (the same model as Equation \eqref{eq:fe} but with
+just one time-varying predictor) involves four components. First, we
+define the latent individual effects variable using the
+\texttt{=\textasciitilde{}} `measured by' or `manifested by'
+\citep{R-lavaan} operator at the same time constraining the factor
+loadings at each timepoint to one. I will call the latent variable
+\texttt{a} to stand for \(\alpha\). Constraining all of the factor
+loadings to one reflects our implicit assumption that the combined
+effect of the unit-specific unobserved factors is constant over time.
+This is the default behaviour of traditional POLS-based approaches to FE
+that use the stacked long-format data.
 
@@
 about the relatedness of our covariate and the individual effects, i.e.,
-\(\E[x_{t}\alpha]\). For an FE model, we want to partial out any
-potential covariance between the independent variable and the individual
-effects. This accounts for any linear relationship between \(x_{t}\) and
-the unit-specific characteristics influencing the dependent variable.
-Further, allowing unrestricted covariances between the independent
-variable itself over time will not affect how the coefficient \(\beta\)
-is estimated, but will have an effect on the standard errors. To mimic
-the behaviour of a conventional FE model, we allow the independent
-variable to be correlated with the individual effects and itself over
-time. Covariances (including covariances between a variable and itself,
-i.e., variances) are specified using the
+\(\mathop{\mathrm{\mathbb{E}}}[x_{t}\alpha]\). For an FE model, we want
+to partial out any potential covariance between the independent variable
+and the individual effects. This accounts for any linear relationship
+between \(x_{t}\) and the unit-specific characteristics influencing the
+dependent variable. Further, allowing unrestricted covariances between
+the independent variable itself over time will not affect how the
+coefficient \(\beta\) is estimated, but will have an effect on the
+standard errors. To mimic the behaviour of a conventional FE model, we
+allow the independent variable to be correlated with the individual
+effects and itself over time. Covariances (including covariances between
+a variable and itself, i.e., variances) are specified using the
 \texttt{\textasciitilde{}\textasciitilde{}} operator:
@@
 \(\beta_{x,\alpha_{1}} = 0.50\). With the variances above set to one,
-the covariances will be roughly \(\Cov(x_{t},\alpha_{1}) = 0.85\) and
-\(\Cov(x_{t},\alpha_{2}) = 0.5\). The dependent variable is also
-influenced by the individual effects variables with
-\(\beta_{y_{t},\alpha_{1}} = 0.75\) and
+the covariances will be roughly
+\(\mathop{\mathrm{\mathrm{Cov}}}(x_{t},\alpha_{1}) = 0.85\) and
+\(\mathop{\mathrm{\mathrm{Cov}}}(x_{t},\alpha_{2}) = 0.5\). The
+dependent variable is also influenced by the individual effects
+variables with \(\beta_{y_{t},\alpha_{1}} = 0.75\) and
 \(\beta_{y_{t},\alpha_{2}} = 0.45\) These values were chosen
@@
   wrappers. In fact, both wrappers seem to be identical in terms of the
-  default settings, see Rosseel et al.
-  (\protect\hyperlink{ref-Rosseel2020}{2020}).}
+  default settings, see \citet{Rosseel2020}.}
 
@@
 Fixed effects regression in SEM has been outlined in well-known articles
-by (Allison \protect\hyperlink{ref-Allison2011}{2011}; Bollen and Brand
-\protect\hyperlink{ref-Bollen2010}{2010}; Teachman et al.
-\protect\hyperlink{ref-Teachman2001}{2001}). This article provides a
-focused look at the implementation of the basic model, as well as common
-extensions using the \texttt{lavaan} package in \texttt{R}.
+by \citep{Allison2011, Bollen2010, Teachman2001}. This article provides
+a focused look at the implementation of the basic model using the
+\texttt{lavaan} package in \texttt{R}. The
+\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
+supplementary materials} further discuss common extensions and some
+tools for evaluating and loosening model assumptions.
 
@@
 FE/RE model with ease. Further extensions, like measurement invariance
-testing (Schoot, Lugtig, and Hox
-\protect\hyperlink{ref-Schoot2012}{2012}; Millsap
-\protect\hyperlink{ref-Millsap2011}{2011}; Steenkamp and Baumgartner
-\protect\hyperlink{ref-Steenkamp1998}{1998}) as well as lagged dependent
-variables (Bollen and Brand \protect\hyperlink{ref-Bollen2010}{2010};
-Allison, Williams, and Moral-Benito
-\protect\hyperlink{ref-Allison2017}{2017}) for example, can also be
-implemented in a straightforward fashion.
+testing \citep{Schoot2012, Millsap2011, Steenkamp1998} as well as lagged
+dependent variables \citep{Bollen2010, Allison2017} for example, can
+also be implemented in a straightforward fashion.
 
@@
 currently popular extended models, such as Latent Curve Models in
-general (Curran and Bollen \protect\hyperlink{ref-Curran2001}{2001};
-Bollen and Curran \protect\hyperlink{ref-Bollen2004}{2004}), as well as
-special implementations like the Dynamic Panel Model (Allison, Williams,
-and Moral-Benito \protect\hyperlink{ref-Allison2017}{2017}), the
-Random-Intercept Cross-Lagged Panel Model (Hamaker, Kuiper, and Grasman
-\protect\hyperlink{ref-Hamaker2015}{2015}) and the Latent Curve Model
-with Structured Residuals (Curran et al.
-\protect\hyperlink{ref-Curran2014}{2014}). For this reason it is all the
-more important for researchers to have a good grasp on the method of
-applying FE-SEM. This article is meant to serve as a consolidated
-resource for researchers looking for concrete advise on specifying FE
-and more general panel models in SEM.
-
-\newpage
-
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-Allison2011}{}%
-Allison, Paul. 2011. \emph{Fixed Effects Regression Models}. Thousand
-Oaks: Sage Publications.
-
-\leavevmode\hypertarget{ref-Allison2017}{}%
-Allison, Paul, Richard Williams, and Enrique Moral-Benito. 2017.
-``Maximum Likelihood for Cross-Lagged Panel Models with Fixed Effects.''
-\emph{Socius} 3: 1--17. \url{https://doi.org/10.1177/2378023117710578}.
-
-\leavevmode\hypertarget{ref-Bates2010}{}%
-Bates, Douglas M. 2010. \emph{Lme4: Mixed-Effects Modeling with R}.
-Springer.
-
-\leavevmode\hypertarget{ref-R-lme4}{}%
-Bates, Douglas, Martin Mächler, Ben Bolker, and Steve Walker. 2015.
-``Fitting Linear Mixed-Effects Models Using lme4.'' \emph{Journal of
-Statistical Software} 67 (1): 1--48.
-\url{https://doi.org/10.18637/jss.v067.i01}.
-
-\leavevmode\hypertarget{ref-Bell2018}{}%
-Bell, Andrew, Malcolm Fairbrother, and Kelvyn Jones. 2018. ``Fixed and
-Random Effects Models: Making an Informed Choice.'' \emph{Quality and
-Quantity} 53: 1051--74.
-
-\leavevmode\hypertarget{ref-Bollen1989}{}%
-Bollen, Kenneth. 1989. \emph{Structural Equations with Latent
-Variables}. New York, Chichester: Wiley.
-
-\leavevmode\hypertarget{ref-Bollen2010}{}%
-Bollen, Kenneth, and Jennie Brand. 2010. ``A General Panel Model with
-Random and Fixed Effects: A Structural Equations Approach.''
-\emph{Social Forces} 89(1): 1--34.
-
-\leavevmode\hypertarget{ref-Bollen2004}{}%
-Bollen, Kenneth, and Patrick Curran. 2004. ``Autoregressive Latent
-Trajectory (ALT) Models: A Synthesis of Two Traditions.''
-\emph{Sociological Methods and Research} 32(3): 336--83.
-\url{https://doi.org/10.1177/0049124103260222}.
-
-\leavevmode\hypertarget{ref-Bruederl2015}{}%
-Brüderl, Josef, and Volker Ludwig. 2015. ``Fixed-Effects Panel
-Regression.'' In \emph{The Sage Handbook of Regression Analysis and
-Causal Inference}, edited by Henning Best and Christof Wolf, 327--57.
-London, Thousand Oaks: Sage Publications.
-
-\leavevmode\hypertarget{ref-Chamberlain1980}{}%
-Chamberlain, Gary. 1980. ``Analysis of Covariance with Qualitative
-Data.'' \emph{Review of Economic Studies} 47(1): 225--38.
-
-\leavevmode\hypertarget{ref-R-plm_a}{}%
-Croissant, Yves, and Giovanni Millo. 2008. ``Panel Data Econometrics in
-R: The plm Package.'' \emph{Journal of Statistical Software} 27 (2):
-1--43. \url{https://doi.org/10.18637/jss.v027.i02}.
-
-\leavevmode\hypertarget{ref-Curran2011}{}%
-Curran, Patrick, and Daniel Bauer. 2011. ``The Disaggregation of
-Within-Person and Between-Person Effects in Longitudinal Models of
-Change.'' \emph{Annual Review of Psychology} 62: 583--619.
-
-\leavevmode\hypertarget{ref-Curran2001}{}%
-Curran, Patrick, and Kenneth Bollen. 2001. ``The Best of Both Worlds:
-Combining Autoregressive and Latent Curve Models.'' In \emph{New Methods
-for the Analysis of Change}, edited by L. Collins and A. Sayer, 107--35.
-Washington, DC: American Psychological Press.
-\url{https://doi.org/\%20doi:10.1037/10409-004}.
-
-\leavevmode\hypertarget{ref-Curran2014}{}%
-Curran, Patrick, Andrea Howard, Sierra Bainter, Stephanie Lane, and
-James McGinley. 2014. ``The Separation of Between-Person and
-Within-Person Components of Individual Change over Time: A Latent Growth
-Curve Model with Structured Residuals.'' \emph{Journal of Consulting and
-Clinical Psychology} 82(5): 879--94.
-\url{https://doi.org/10.1037/a0035297}.
-
-\leavevmode\hypertarget{ref-Enders2007}{}%
-Enders, Craig, and Davood Tofighi. 2007. ``Centering Predictor Variables
-in Cross-Sectional Multilevel Models: A New Look at an Old Issue.''
-\emph{Psychological Methods} 12(2): 121--38.
-
-\leavevmode\hypertarget{ref-Graff1979}{}%
-Graff, J. 1979. ``Verallgemeinertes LISREL-Modell.'' Mannheim, Germany.
-
-\leavevmode\hypertarget{ref-Hamaker2015}{}%
-Hamaker, Ellen, Rebecca Kuiper, and Raoul Grasman. 2015. ``A Critique of
-the Cross-Lagged Panel Model.'' \emph{Psychological Methods} 20(1):
-102--16. \url{https://doi.org/10.1037/a0038889}.
-
-\leavevmode\hypertarget{ref-Kline2016}{}%
-Kline, Rex. 2016. \emph{Principles and Practice of Structural Equation
-Modeling. Fourth Edition}. New York: The Guilford Press.
-
-\leavevmode\hypertarget{ref-Millsap2011}{}%
-Millsap, Roger. 2011. \emph{Statistical Approaches to Measurement
-Invariance}. New York, London: Routledge.
-
-\leavevmode\hypertarget{ref-Mundlak1978}{}%
-Mundlak, Yair. 1978. ``On the Pooling of Time Series and Cross Section
-Data.'' \emph{Econometrica} 46(1): 69--85.
-
-\leavevmode\hypertarget{ref-Mplus}{}%
-Muthén, Linda, and Bengt Muthén. n.d. \emph{Mplus User's Guide. Eighth
-Edition}. Los Angeles, California: Muthén \& Muthén.
-
-\leavevmode\hypertarget{ref-R-base}{}%
-R Core Team. 2017. \emph{R: A Language and Environment for Statistical
-Computing}. Vienna, Austria: R Foundation for Statistical Computing.
-\url{https://www.R-project.org/}.
-
-\leavevmode\hypertarget{ref-R-lavaan}{}%
-Rosseel, Yves. 2012. ``lavaan: An R Package for Structural Equation
-Modeling.'' \emph{Journal of Statistical Software} 48 (2): 1--36.
-\url{http://www.jstatsoft.org/v48/i02/}.
-
-\leavevmode\hypertarget{ref-Rosseel2020}{}%
-Rosseel, Yves, Terrence D. Jorgensen, Daniel Oberski, Jarrett Byrnes,
-Leonard Vanbrabant, Victoria Savalei, Ed Merkle, et al. 2020.
-\emph{Lavaan: Latent Variable Analysis}.
-\url{https://CRAN.R-project.org/package=lavaan}.
-
-\leavevmode\hypertarget{ref-Schmidheiny2019}{}%
-Schmidheiny, Kurt. 2019. ``Panel Data: Fixed and Random Effects.''
-Unversität Basel.
-
-\leavevmode\hypertarget{ref-Schoot2012}{}%
-Schoot, Rens van de, Peter Lugtig, and Joop Hox. 2012. ``A Checklist for
-Testing Measurement Invariance.'' \emph{European Journal of
-Developmental Psychology} 9(4): 486--92.
-
-\leavevmode\hypertarget{ref-Schunck2013}{}%
-Schunck, Reinhard. 2013. ``Within and Between Estimates in
-Random-Effects Models: Advantages and Drawbacks of Random Effects and
-Hybrid Models.'' \emph{The Stata Journal} 13(1): 65--76.
-
-\leavevmode\hypertarget{ref-Skrondal2004}{}%
-Skrondal, Anders, and Sophia Rabe-Hesketh. 2004. \emph{Generalized
-Latent Variable Modeling}. Boca Raton, London, New York, Washington,
-D.C.: Chapman \& Hall/CRC.
-
-\leavevmode\hypertarget{ref-Steenkamp1998}{}%
-Steenkamp, Jan-Benedict, and Hans Baumgartner. 1998. ``Assessing
-Measurement Invariance in Cross-National Consumer Research.''
-\emph{Journal of Consumer Research} 25(1): 78--90.
-
-\leavevmode\hypertarget{ref-Teachman2001}{}%
-Teachman, Jay, Greg Duncan, Jean Yeung, and Dan Levy. 2001. ``Covariance
-Structure Models for Fixed and Random Effects.'' \emph{Sociological
-Methods and Research} 30(2): 271--88.
-
-\leavevmode\hypertarget{ref-Wooldridge2002}{}%
-Wooldridge, Jeffery. 2002. \emph{Econometric Analysis of Cross Sectional
-and Panel Data}. Cambridge, Massachusetts: The MIT Press.
-
-\leavevmode\hypertarget{ref-Wooldridge2012}{}%
----------. 2012. \emph{Introductory Econometrics: A Modern Approach,
-5\(^{th}\) Edition}. Mason, Ohio: Thomson South-Western.
+general \citep{Curran2001, Bollen2004}, as well as special
+implementations like the Dynamic Panel Model \citep{Allison2017}, the
+Random-Intercept Cross-Lagged Panel Model \citep{Hamaker2015} and the
+Latent Curve Model with Structured Residuals \citep{Curran2014}. For
+this reason it is all the more important for researchers to have a good
+grasp on the method of applying panel regression in SEM, and
+understanding the intuition of controlling for time-invariant
+confounders. This article is meant to serve as a consolidated resource
+for researchers looking for concrete advise on specifying FE and more
+general panel models in SEM.
+
+\bibliographystyle{tfcad}
+\bibliography{references2.bib}
 

### 2020-07-28 article_format/article_format_anon.Rmd
Changed the formating, made some minor edits

--- 2020-07-28 article_format/article_format.Rmd
@@
 author:
-  - name: Henrik Kenneth Andersen
+  - name: ""
     affil: a
-    email: henrik.andersen@soziologie.tu-chemnitz.de
+    email: ""
 affiliation:
@@
     address: |
-      Chemnitz University of Technology, Institute of Sociology, Chair for Empirical Social Research, Thüringer Weg 9, 09126 Chemnitz, Germany
 bibliography: references2.bib
@@
 output: rticles::tf_article
+keep_tex: true
 ---

### 2020-07-28 article_format/extensions_format.Rmd
Changed the formating, made some minor edits

--- 2020-07-22 extensions.Rmd
@@
 ---
-title: "Supplememtary examples for:"
-#author: "Henrik Kenneth Andersen"
-#date: "`r Sys.Date()`"
-output: 
-  pdf_document:
-    fig_caption: yes
-    number_sections: yes
-    includes: 
-      in_header: header.tex
-documentclass: article
-classoption: a4paper
-urlcolor: blue
-link-citations: yes
-bibliography: references.bib   
+title: |
+  | Supplementary materials for: 
+  | A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}
+#type: ARTICLE TEMPLATE
+author:
+  - name: Henrik Kenneth Andersen
+    affil: a
+    email: henrik.andersen@soziologie.tu-chemnitz.de
+affiliation:
+  - num: a
+    address: |
+      Chemnitz University of Technology, Institute of Sociology, Chair for Empirical Social Research, Thüringer Weg 9, 09126 Chemnitz, Germany
+bibliography: references2.bib
+# appendix: appendix.tex
+header-includes: |
+  \usepackage{hyperref}
+  \usepackage[utf8]{inputenc}
+  \usepackage{booktabs}
+  \usepackage{bm}
+  \usepackage{mathtools}
+  \usepackage{amssymb}
+  \usepackage{amsmath}
+  \usepackage{tikz}
+  \usetikzlibrary{arrows}
+  \usepackage[nofiglist]{endfloat}
+  \usepackage{blkarray}
+  \usepackage{setspace}
+  \usepackage{etoolbox}
+  \DeclareMathOperator{\E}{\mathbb{E}}
+  \DeclareMathOperator{\Var}{\mathrm{Var}}
+  \DeclareMathOperator{\Cov}{\mathrm{Cov}}
+  \DeclareMathOperator*{\argmax}{arg\,max}
+  \DeclareMathOperator*{\argmin}{arg\,min}
+  \mathtoolsset{showonlyrefs}
+  \BeforeBeginEnvironment{equation}{\begin{singlespace}\vspace*{-\baselineskip}}
+  \AfterEndEnvironment{equation}{\end{singlespace}\noindent\ignorespaces}
+  \BeforeBeginEnvironment{align}{\begin{singlespace}\vspace*{-\baselineskip}}
+  \AfterEndEnvironment{align}{\end{singlespace}\noindent\ignorespaces}
+  \def\tightlist{}
+  \pagenumbering{gobble}
+output: rticles::tf_article
 ---
 
-\pagenumbering{arabic} 
-\setcounter{page}{1}
-
 ```{r setup, include=FALSE}
+# Knitr options 
 knitr::opts_chunk$set(echo = TRUE)
 
-# ----- Packages
+# Packages (some are repeated below, add them here as well just to be sure)
+library(formatR)
+library(knitr)
+library(lavaan)
 
-library( formatR)
-library( knitr)
-library( lavaan)
-
-source("scripthooks.R")
+# Script hooks for chunk outputs
+source("../scripthooks.R")
 ```
 
-Andersen, Henrik K. (2020). A closer look at fixed-effects regression in structural equation modeling using lavaan.
+```{r echo=FALSE}
+df <- readRDS("../longData.Rda")
+dfw <- readRDS("../wideData.Rda")
 
-```{r echo=FALSE}
-df <- readRDS("longData.Rda")
-dfw <- readRDS("wideData.Rda")
+n <- readRDS("../n.Rda")
+waves <- readRDS("../waves.Rda")
 
-n <- readRDS("n.Rda")
-waves <- readRDS("waves.Rda")
-
-fe_sem <- readRDS("fe_sem.Rda")
-fe_sem.fit <- readRDS("fe_sem.fit.Rda")
+fe_sem <- readRDS("../fe_sem.Rda")
+fe_sem.fit <- readRDS("../fe_sem.fit.Rda")
 ```
 
-# Extensions {#exten}
+---
 
@@
 
-From this we see both models report the same estimated variance components, $\hat{\sigma}^{2}_{\alpha} = `r round(ercomp(fe2)[[1]][2], 3)`$ and $\hat{\sigma}^{2}_{\varepsilon} = `r round(ercomp(fe2)[[1]][1], 3)`$, telling us that about `r round(ercomp(fe2)[[1]][2]/(ercomp(fe2)[[1]][1] + ercomp(fe2)[[1]][2]), 3)`% of the residual variance is due to the differences between individuals (shown in the `share` column of the `ercomp()` output).
+From this we see both models report the same estimated variance components, $\hat{\sigma}^{2}_{\alpha} = `r round(ercomp(fe2)[[1]][2], 3)`$ and $\hat{\sigma}^{2}_{\varepsilon} = `r round(ercomp(fe2)[[1]][1], 3)`$, telling us that about `r round(ercomp(fe2)[[1]][2]/(ercomp(fe2)[[1]][1] + ercomp(fe2)[[1]][2]), 3)*100`% of the residual variance is due to the differences between individuals (shown in the `share` column of the `ercomp()` output).
 This is what is referred to as the intraclass correlation coefficient, or ICC [@Hox2010].
+
+# Extensions {#exten}
 
@@
 where $\theta$ is just the residual of a regression in which the true underlying variable is regressed on all other covariates.
-In this case, we need to regress $x_{t}$ on $\alpha_{1}$ and $\alpha_{2}$ for: $x_{t} = \gamma_{1}\alpha_{1} + \gamma_{2}\alpha_{2} + \theta_{t}$ [@Wooldridge2009, p.
+In this case, we need to regress $x_{t}$ on $\alpha_{1}$ and $\alpha_{2}$ for: $x_{t} = \tau + \gamma_{1}\alpha_{1} + \gamma_{2}\alpha_{2} + \theta_{t}$ where $\tau$ is the intercept, and $\gamma_{1}$, $\gamma_{2}$ are the regression coefficients and $\theta_{t}$ is the residual [@Wooldridge2009, p.
 318--320].
@@
 However, because we are working with simulated data, we have everything we need.
-Going back to the results above, we can get the residuals of $x_{t}$ by either running a regression and saving the residuals, or we could skip a step and get them directly using the 'residual maker' matrix [@Ruettenauer2020] which is $\bm{M} = \bm{I} - \bm{A}(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}$ and $\bm{A} = \begin{pmatrix}\bm{\alpha_{1}}, \bm{\alpha_{2}}\end{pmatrix}$ is the $n \times 2$ matrix of the covariates.
+Going back to the results above, we can get the residuals of $x_{t}$ by either running a regression and saving the residuals, or we could skip a step and get them directly using the 'residual maker' matrix [@Ruettenauer2020] which is $\bm{M} = \bm{I} - \bm{A}(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}$ and $\bm{A} = \begin{pmatrix}\bm{\iota}_{n} & \bm{\alpha_{1}} & \bm{\alpha_{2}}\end{pmatrix}$ is the $n \times 3$ matrix of the covariates (plus a constant).
 
@@
 
-\newpage
-
-# References {-}
-
-<div id="refs"></div>
+# References

### 2020-07-28 article_format/extensions_format.tex
Changed the formating, made some minor edits

% interactcadsample.tex
% v1.03 - April 2017

\documentclass[]{interact}

\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage{subfigure}% Support for small, `sub' figures and tables
%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required

\usepackage{natbib}% Citation support using natbib.sty
\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty

\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}

% see https://stackoverflow.com/a/47122900
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{booktabs}
\usepackage{bm}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{tikz}
\usetikzlibrary{arrows}
\usepackage[nofiglist]{endfloat}
\usepackage{blkarray}
\usepackage{setspace}
\usepackage{etoolbox}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\mathrm{Var}}
\DeclareMathOperator{\Cov}{\mathrm{Cov}}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\mathtoolsset{showonlyrefs}
\BeforeBeginEnvironment{equation}{\begin{singlespace}\vspace*{-\baselineskip}}
\AfterEndEnvironment{equation}{\end{singlespace}\noindent\ignorespaces}
\BeforeBeginEnvironment{align}{\begin{singlespace}\vspace*{-\baselineskip}}
\AfterEndEnvironment{align}{\end{singlespace}\noindent\ignorespaces}
\def\tightlist{}
\pagenumbering{gobble}

\begin{document}

\articletype{}

\title{Supplementary materials for:\\
A closer look at fixed effects regression in structural equation
modeling using \texttt{lavaan}}

\author{\name{Henrik Kenneth Andersen$^{a}$}
\affil{$^{a}$Chemnitz University of Technology, Institute of Sociology, Chair for
Empirical Social Research, Thüringer Weg 9, 09126 Chemnitz, Germany}
}

\thanks{CONTACT Henrik Kenneth Andersen. Email: \href{mailto:henrik.andersen@soziologie.tu-chemnitz.de}{\nolinkurl{henrik.andersen@soziologie.tu-chemnitz.de}}}

\maketitle

\singlespacing

\doublespacing

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

The following goes into some more detail on the comparison of
traditional FE and FE-SEM models and discusses several opportunities to
extend the basic FE model outlined in the main article, and provides
concrete guidance on the implementation in \texttt{lavaan}. First, I
verify that the FE-SEM model does, in fact, return essentially identical
results compared to the more traditional methods. Then, I go over a
number of possibilities to relax assumptions associated with the
traditional FE model. Then, I discuss the issue of measurement error and
show how we can use latent variables to deal with it and properly
estimate the coefficients of interest. Then, I show a type of hybrid
FE/RE model that allows us to control for time-invariant unobserved
heterogeneity while including time-invariant predictors in the model.

\hypertarget{a-comparison-with-non-sem-methods}{%
\section{A comparison with non-SEM
methods}\label{a-comparison-with-non-sem-methods}}

Just to be sure that the FE-SEM results do, in fact, line up with the
more traditional methods outlined in Section 2 of the main article, we
can use the long-format data (see
\href{https://github.com/henrik-andersen/FE-SEM}{the supplementary
materials at}) to run the typical FE model using the \texttt{plm}
package \citep{R-plm_a}. By default, the \texttt{plm} function assumes
the dataframe is structured so that the first two columns correspond to
the individual and time indices, see the
\href{https://cran.r-project.org/web/packages/plm/plm.pdf}{documentation}
or \citet{R-plm_a}.

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(plm)}

\CommentTok{# Run the FE model in plm }
\NormalTok{fe1 <-}\StringTok{ }\KeywordTok{plm}\NormalTok{(y }\OperatorTok{~}\StringTok{ }\NormalTok{x, }
           \DataTypeTok{effect =} \StringTok{"individual"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"within"}\NormalTok{, }
           \DataTypeTok{data =}\NormalTok{ df)}
\KeywordTok{summary}\NormalTok{(fe1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Oneway (individual) effect Within Model
## 
## Call:
## plm(formula = y ~ x, data = df, effect = "individual", model = "within")
## 
## Balanced Panel: n = 1000, T = 5, N = 5000
## 
## Residuals:
##      Min.   1st Qu.    Median   3rd Qu.      Max. 
## -3.720902 -0.601550 -0.021365  0.600833  3.238716 
## 
## Coefficients:
##   Estimate Std. Error t-value  Pr(>|t|)    
## x 0.293907   0.015635  18.798 < 2.2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Total Sum of Squares:    4452.6
## Residual Sum of Squares: 4091.1
## R-Squared:      0.081192
## Adj. R-Squared: -0.14857
## F-statistic: 353.377 on 1 and 3999 DF, p-value: < 2.22e-16
\end{verbatim}

\doublespacing

From this, we see that the results are, indeed, essentially identical,
with \(\hat{\beta}_{FE-SEM} =0.294 \ (0.016)\) and
\(\hat{\beta}_{FE} = 0.294 \ (0.016)\).

Other methods of estimating FE models work in the random or mixed
effects model framework. For example, we can include the cluster means
per individual of the time-varying independent variables, here \(x\), in
the equation to achieve within estimates
\citep{Mundlak1978, Chamberlain1980, Wooldridge2002}.

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Generate the cluster means for x per id }
\NormalTok{clusterMeanx <-}\StringTok{ }\KeywordTok{aggregate}\NormalTok{(df}\OperatorTok{$}\NormalTok{x, }\DataTypeTok{by =} \KeywordTok{list}\NormalTok{(df}\OperatorTok{$}\NormalTok{id), }\DataTypeTok{FUN =}\NormalTok{ mean)}
\CommentTok{# Rename the columns }
\KeywordTok{names}\NormalTok{(clusterMeanx) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"id"}\NormalTok{, }\StringTok{"xbar"}\NormalTok{)}

\CommentTok{# Add the cluster means back into df  }
\NormalTok{df <-}\StringTok{ }\KeywordTok{merge}\NormalTok{(df, clusterMeanx, }\DataTypeTok{by =} \StringTok{"id"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\doublespacing

Here using the \texttt{plm} function in the \texttt{random} setup:

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fe2 <-}\StringTok{ }\KeywordTok{plm}\NormalTok{(y }\OperatorTok{~}\StringTok{ }\NormalTok{x }\OperatorTok{+}\StringTok{ }\NormalTok{xbar, }
           \DataTypeTok{effect =} \StringTok{"individual"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"random"}\NormalTok{,}
           \DataTypeTok{data =}\NormalTok{ df)}
\KeywordTok{summary}\NormalTok{(fe2)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Oneway (individual) effect Random Effect Model 
##    (Swamy-Arora's transformation)
## 
## Call:
## plm(formula = y ~ x + xbar, data = df, effect = "individual", 
##     model = "random")
## 
## Balanced Panel: n = 1000, T = 5, N = 5000
## 
## Effects:
##                  var std.dev share
## idiosyncratic 1.0230  1.0115 0.862
## individual    0.1637  0.4046 0.138
## theta: 0.2546
## 
## Residuals:
##       Min.    1st Qu.     Median    3rd Qu.       Max. 
## -4.0697324 -0.6786344 -0.0094158  0.6678915  3.5329988 
## 
## Coefficients:
##              Estimate Std. Error z-value Pr(>|z|)    
## (Intercept) -0.022559   0.019192 -1.1754   0.2398    
## x            0.293907   0.015635 18.7983   <2e-16 ***
## xbar         0.757738   0.024015 31.5532   <2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Total Sum of Squares:    8878.9
## Residual Sum of Squares: 5112.1
## R-Squared:      0.42424
## Adj. R-Squared: 0.42401
## Chisq: 3682.01 on 2 DF, p-value: < 2.22e-16
\end{verbatim}

\doublespacing

And here using the \texttt{lmer} function of the \texttt{lme4} package
\citep{Bates2015} to estimate a mixed model:

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(lme4)}

\CommentTok{# Run the mixed model in lmer with the cluster means for x }
\NormalTok{mixed1 <-}\StringTok{ }\KeywordTok{lmer}\NormalTok{(y }\OperatorTok{~}\StringTok{ }\NormalTok{x }\OperatorTok{+}\StringTok{ }\NormalTok{xbar }\OperatorTok{+}\StringTok{ }\NormalTok{(}\DecValTok{1} \OperatorTok{|}\StringTok{ }\NormalTok{id), }\DataTypeTok{data =}\NormalTok{ df)}
\KeywordTok{summary}\NormalTok{(mixed1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Linear mixed model fit by REML ['lmerMod']
## Formula: y ~ x + xbar + (1 | id)
##    Data: df
## 
## REML criterion at convergence: 14906.7
## 
## Scaled residuals: 
##     Min      1Q  Median      3Q     Max 
## -3.9358 -0.6464 -0.0106  0.6349  3.2002 
## 
## Random effects:
##  Groups   Name        Variance Std.Dev.
##  id       (Intercept) 0.1637   0.4046  
##  Residual             1.0230   1.0115  
## Number of obs: 5000, groups:  id, 1000
## 
## Fixed effects:
##             Estimate Std. Error t value
## (Intercept) -0.02256    0.01919  -1.175
## x            0.29391    0.01563  18.798
## xbar         0.75774    0.02401  31.553
## 
## Correlation of Fixed Effects:
##      (Intr) x     
## x     0.000       
## xbar  0.010 -0.651
\end{verbatim}

\doublespacing

In both cases, the models return the same estimates as the FE and FE-SEM
models. Also, in both the \texttt{random} setup using the \texttt{plm}
function, and the mixed model using the \texttt{lmer} function, we get
estimates of the variance components, \(\hat{\sigma}^{2}_{\alpha}\) and
\(\hat{\sigma}^{2}_{\varepsilon}\):

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Print the variance components for the plm model  }
\KeywordTok{print}\NormalTok{(}\KeywordTok{ercomp}\NormalTok{(fe2), }
      \DataTypeTok{digits =} \DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                 var std.dev share
## idiosyncratic 1.023   1.011  0.86
## individual    0.164   0.405  0.14
## theta: 0.255
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Print the variance components for the lmer model }
\KeywordTok{print}\NormalTok{(}\KeywordTok{VarCorr}\NormalTok{(mixed1), }
      \DataTypeTok{comp =} \KeywordTok{c}\NormalTok{(}\StringTok{"Variance"}\NormalTok{, }\StringTok{"Std.Dev"}\NormalTok{), }
      \DataTypeTok{digits =} \DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  Groups   Name        Variance Std.Dev.
##  id       (Intercept) 0.164    0.405   
##  Residual             1.023    1.011
\end{verbatim}

\doublespacing

From this we see both models report the same estimated variance
components, \(\hat{\sigma}^{2}_{\alpha} = 0.164\) and
\(\hat{\sigma}^{2}_{\varepsilon} = 1.023\), telling us that about 13.8\%
of the residual variance is due to the differences between individuals
(shown in the \texttt{share} column of the \texttt{ercomp()} output).
This is what is referred to as the intraclass correlation coefficient,
or ICC \citep{Hox2010}.

\hypertarget{exten}{%
\section{Extensions}\label{exten}}

\hypertarget{relax}{%
\subsection{Relaxing assumptions meant to mimic traditional FE
models}\label{relax}}

There are a number of implicit assumptions attached to the typical FE
model that can be relaxed in SEM. Some of these assumptions have been
discussed already, and a fairly comprehenisve list of assumptions can be
found in \citet{Bollen2010}. Here, I will go over just a few,
concentrating on the implementation in \texttt{lavaan} and the
opportunity to empirically test whether the adjustments are justified or
not.

The assumptions we will discuss here pertain to the time-invariance of
the effects of both the latent individual effects and the observed
covariates, as well as a time-invariant error variance. We can also
empirically test the correlation between the individual effects and the
covariates to see whether a RE model is preferable to the FE model.

For example, we can rewrite the original FE equation as \begin{align}
y_{it} & = \beta_{t}x_{it} + \lambda_{t}\alpha_{i} + \varepsilon_{it}
\end{align} where \(\beta\) becomes \(\beta_{t}\) and the implicit
regression weight of one turns to \(\lambda_{t}\) to highlight the fact
that the effect of \(x\) as well as \(\alpha\) on \(y\) may vary over
time. We can furthermore easily relax the assumption of time-constant
error variance, i.e., \(\sigma^{2}_{\varepsilon_{t}}\). As noted in the
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/article.pdf}{main
article}, the assumption regarding
\(\mathop{\mathrm{\mathbb{E}}}[\alpha x_{t}]\) in \(\bm{\Psi}\)
determines whether we have an FE or RE model. We can set these to zero
and test whether the RE model would be preferable to the FE model. In
general, if the individual effects are truly uncorrelated with the model
covariates, it is advisable to switch to an RE model since because it
uses up less degrees of freedom, it will have smaller standard errors
\citep{Bollen2010}.

In the following \texttt{lavaan} code, we simply remove the factor
loadings of one for the latent individual effect variable which allows
them to be estimated freely at each timepoint. For the effect of the
covariate, we can either delete the constraints \texttt{b} in
\texttt{yt\ \textasciitilde{}\ b*xt} or give each regression a different
label, e.g., \texttt{b1}, \texttt{b2}, \texttt{b3}, etc. Similarly, to
allow the error variance to vary over time, we turn the constraints
\texttt{e} into simple labels, i.e., \texttt{e1}, \texttt{e2},
\texttt{e3}, etc., or again just delete them. In fact, regarding the
error variances, they will be estimated necessarily, and do not need to
be explicitly mentioned in the model syntax at all. Finally, to move
from an FE to an RE model, we could simply constrain the correlations
between the individual effects and the covariates to zero, i.e.,
\texttt{a\ \textasciitilde{}\textasciitilde{}\ 0*x1\ +\ 0*x2\ +\ 0*x3\ +\ 0*x4\ +\ 0*x5}.

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fe_sem_fullyrelaxed <-}\StringTok{ '}
\StringTok{# Define individual effects variable }
\StringTok{a =~ y1 + y2 + y3 + y4 + y5}
\StringTok{# Regressions, constrain coefficient to be equal over time}
\StringTok{y1 ~ b1*x1}
\StringTok{y2 ~ b2*x2 }
\StringTok{y3 ~ b3*x3}
\StringTok{y4 ~ b4*x4}
\StringTok{y5 ~ b5*x5}
\StringTok{# Allow unrestricted correlation between eta and covariates}
\StringTok{a ~~ x1 + x2 + x3 + x4 + x5}
\StringTok{# Alternatively: constrain all to 0 for RE model, or}
\StringTok{# just individual correlations}
\StringTok{# a ~~ 0*x1 + 0*x2 + 0*x3 + 0*x4 + 0*x5}
\StringTok{x1 ~~ x2 + x3 + x4 + x5}
\StringTok{x2 ~~ x3 + x4 + x5}
\StringTok{x3 ~~ x4 + x5}
\StringTok{x4 ~~ x5}
\StringTok{# Constrain residual variances to be equal over time}
\StringTok{y1 ~~ e1*y1}
\StringTok{y2 ~~ e2*y2}
\StringTok{y3 ~~ e3*y3}
\StringTok{y4 ~~ e4*y4}
\StringTok{y5 ~~ e5*y5}
\StringTok{'}
\NormalTok{fe_sem_fullyrelaxed.fit <-}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe_sem_fullyrelaxed, }
                                \DataTypeTok{data =}\NormalTok{ dfw, }
                                \DataTypeTok{estimator =} \StringTok{"ML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\doublespacing

\singlespacing

\doublespacing

As outlined in \citet{Bollen2010}, the researcher has the opportunity to
test each of the assumptions empirically and decide whether a more
parsimonious, i.e., restrictive model is justifiable. For each
assumption, a likelihood ratio test can be carried out to determine
whether the improvement to model fit resulting from the relaxation of
various assumptions is significant or whether the more parsimonious
model is preferable after all.

If we use the original model \texttt{fe\_sem.fit} (from the
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/article.pdf}{main
article}) as a starting point, the best strategy for testing these
assumptions is to work in a stepwise fashion, relaxing one assumption at
a time. We can begin by first constraining the correlation between
\(\alpha\) and \(x_{t}\) to zero (\texttt{re\_sem}) for an RE model. If
turning from an FE to an RE model does not significantly worsen model
fit, we can go forward with the rest of the steps with the RE model. If,
however, the fit does worsen significantly, it is likely better to stick
with the FE model; moving forward then with it to see if a less
restrictive FE model is preferable. We can perform a likelihood ratio
test in \texttt{R} using the \texttt{anova()} function:

\small
\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{anova}\NormalTok{( fe_sem.fit, re_sem.fit)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Chi-Squared Difference Test
## 
##            Df   AIC   BIC   Chisq Chisq diff Df diff Pr(>Chisq)    
## fe_sem.fit 32 30998 31111  30.137                                  
## re_sem.fit 37 31809 31897 850.928     820.79       5  < 2.2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

\doublespacing
\normalsize

The table that is generated shows a comparison of the nested models, in
decending order according to degrees of freedom. The RE model does not
estimate the correlations between the individual effects and the
covariates, so it is more parsimonious and thus listed at the bottom.
The \texttt{Chisq} column shows the \(\chi^{2}\) statistic for both
models and the \texttt{Chisq\ diff} column calculates the difference
between the two. Obviously, according to the DGP, the correlation
between the individual effects and \(x_{t}\) is not zero, so fixing
these to zero leads to a substantial amount of misfit. The last column
puts the \(\chi^{2}\) difference in relation to the difference in
degrees of freedom and gives a p-value for the probability that the
difference is solely due to chance. Here, the change in \(\chi^{2}\) is
highly significant, so the FE model should be retained.

After now having established once and for all that FE is our preferred
model, we can begin relaxing the rest of the assumptions. I show the
following merely as a demonstration of the procedure, we know already
from the DGP that the parsimonious model as specified in
\texttt{fe\_sem.fit} is appropriate. We can next allow the error
variances (\texttt{fe\_semb.fit}), the effect of \(x\) on \(y\)
(\texttt{fe\_semc.fit}) and finally the factor loadings of the
individual effects (\texttt{fe\_semd.fit}) all to vary over time.

\singlespacing

\doublespacing

\small
\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{anova}\NormalTok{( fe_sem.fit, fe_semb.fit, fe_semc.fit, fe_semd.fit)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Chi-Squared Difference Test
## 
##             Df   AIC   BIC  Chisq Chisq diff Df diff Pr(>Chisq)
## fe_semd.fit 20 31017 31189 25.140                              
## fe_semc.fit 24 31010 31162 25.764     0.6249       4     0.9603
## fe_semb.fit 28 31003 31135 26.686     0.9215       4     0.9215
## fe_sem.fit  32 30998 31111 30.137     3.4516       4     0.4853
\end{verbatim}

\doublespacing
\normalsize

Keep in mind that a less parsimonious model (fewer degrees of freedom)
can never fit worse than a more parsimonious one (more degrees of
freedom). I.e., chance variations due to sampling error mean that adding
constraints to a model will tend to always worsen fit, at least
minimally. The question here is whether the improvement to fit by
loosening constraints is meaningful or not. In the table above, we
should not expect any meaningful improvements moving from
\texttt{fe\_sem.fit} to \texttt{fe\_semd.fit}. Here, using simulated
data, we have the luxury of knowing that any significant differences in
\(\chi^{2}\) are due to chance. With real data, it is up to the
researcher to apply their best judgment and decide whether the results
are plausible or not.

\hypertarget{measerr}{%
\subsection{Measurement error}\label{measerr}}

What if the observed variables are not measured perfectly? Then what we
observe, call them \(\tilde{x}_{t}\) and \(\tilde{y}_{t}\) are
composites of the true score we are after, i.e., \(x_{t}\) and
\(y_{t}\), plus an additive measurement error portion: \begin{align}
\tilde{x}_{t} & = x_{t} + \upsilon_{t}, \\
\tilde{y}_{t} & = y_{t} + \nu_{t}.
\end{align} How does this affect our model? Well, first notice that
measurement error in the dependent variable is typically less of a
serious problem than measurement error in the independent variables. Let
us assume again mean-centered variables so that we can ignore the
intercept, and consider the following simple bivariate equation:
\begin{align}
y & = \beta x + \varepsilon
\end{align} if \(y\) is measured imperfectly and what we observe is
\(\tilde{y} = y + \nu\), then we can rewrite the equation as:
\begin{align}
(\tilde{y} - \nu) & = \beta x + \varepsilon \\
\tilde{y} & = \beta x + \varepsilon + \nu.
\end{align} The measurement error in \(y\) just gets added to the
regression error. As long as \(\nu\) is uncorrelated with \(x\), then
the regression coefficient will be unbiased
\citep{Pischke2007, Wooldridge2009}. However, this will increase the
error variance and thus make the estimates less precise.

We will look at the effect of measurement error in the dependent
variable using an example shortly. For now though, let us be safe in the
knowledge that the coefficient of interest is likely unbiased, and
concentrate on the more serious problem of error in the independent
variable.

The intuition behind the problem of measurement error in the independent
variable(s) can be explained as follows. Take
\(\tilde{x} = x + \upsilon\) and substitute this into the equation for
\(y\): \begin{align}
y & = \beta x + \varepsilon \\
 & = \beta(\tilde{x} - \upsilon) + \varepsilon \\
 & = \beta\tilde{x} + (\varepsilon - \beta\upsilon).
\end{align} Since \(\tilde{x}\) is obviously correlated with
\(\upsilon\) (unless the variance of \(\upsilon\) is so small so that
the correlation is essentially negligible), then the composite error in
this regression is also correlated with the independent variable and
thus the estimated coefficient of \(\beta\) will be biased.

\hypertarget{the-consequences-of-measurement-error}{%
\subsubsection{The consequences of measurement
error}\label{the-consequences-of-measurement-error}}

To demonstrate the effect of measurement error on the FE-SEM model, and
then provide a strategy for dealing with measurement error in SEM, the
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R}{simulated
dataset} generates multiple \emph{indicators} of the independent and
dependent variables that all measure the intended variable imprecisely.
Returning to our panel data, we have three indicators of each the
independent and dependent variable, per timepoint:\\
\begin{align}
\tilde{x}_{kt} & = x_{t} + \upsilon_{kt}, \\
\tilde{y}_{kt} & = y_{t} + \nu_{kt}
\end{align} where \(k = 1, 2, 3\) and \(t = 1, ..., T\). This is like
repeatedly presenting a respondent with a multi-item scale designed to
measure things like stress, depression, xenophobia, etc. over the course
of a panel study. To create the observed indicators, a random amount of
measurement error (ranging from
\(\{\sigma^{2}_{\upsilon_{k}}, \sigma^{2}_{\nu_{k}}\}\in \{1.0, 1.1, 1.2, 1.3, 1.4, 1.5\}\))
was added to the true variables, again see
\href{https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R}{the
simulation code}.

Let us first focus on the issue of imprecise measurements of the
independent variable of interest and run the same FE-SEM model above,
but this time we will use one of the measurement error sullied
indicators, here \(\tilde{x}_{1t}\), instead of the true independent
variable, \(x_{t}\). As for the naming conventions in the \texttt{R}
code, \texttt{x11} stands for the first indicator (\(k = 1\)) at the
first point in time (\(t = 1\)), whereas for example \texttt{x35} stands
for the third indicator (\(k = 3\)) at the fifth point in time
(\(t = 5\)).

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fe_sem2 <-}\StringTok{ '}
\StringTok{# Define individual effects variable }
\StringTok{a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5}
\StringTok{# Regressions, constrain coefficient to be equal over time}
\StringTok{# Now the imprecisely measured indicator tilde\{x\}_kt}
\StringTok{# instead of the true variable x_t}
\StringTok{y1 ~ b*x11 }
\StringTok{y2 ~ b*x12 }
\StringTok{y3 ~ b*x13}
\StringTok{y4 ~ b*x14}
\StringTok{y5 ~ b*x15}
\StringTok{# Allow unrestricted correlation between eta and covariates}
\StringTok{a ~~ x11 + x12 + x13 + x14 + x15}
\StringTok{x11 ~~ x12 + x13 + x14 + x15}
\StringTok{x12 ~~ x13 + x14 + x15}
\StringTok{x13 ~~ x14 + x15}
\StringTok{x14 ~~ x15}
\StringTok{# Constrain residual variances to be equal over time}
\StringTok{y1 ~~ e*y1}
\StringTok{y2 ~~ e*y2}
\StringTok{y3 ~~ e*y3}
\StringTok{y4 ~~ e*y4}
\StringTok{y5 ~~ e*y5}
\StringTok{'}
\NormalTok{fe_sem2.fit <-}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe_sem2, }
                    \DataTypeTok{data =}\NormalTok{ dfw, }
                    \DataTypeTok{estimator =} \StringTok{"ML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\doublespacing

Now, for the sake of brevity, let us look just at the estimated
coefficients for \(\beta\).

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{( fe_sem2.fit)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
...
##                    Estimate  Std.Err  z-value  P(>|z|)
##   y1 ~                                                
##     x11        (b)    0.091    0.009    9.611    0.000
##   y2 ~                                                
##     x12        (b)    0.091    0.009    9.611    0.000
##   y3 ~                                                
##     x13        (b)    0.091    0.009    9.611    0.000
##   y4 ~                                                
##     x14        (b)    0.091    0.009    9.611    0.000
##   y5 ~                                                
##     x15        (b)    0.091    0.009    9.611    0.000
## 
...
\end{verbatim}

\doublespacing

Obviously, the estimated coefficient \(\hat{\beta} =\) 0.091 is
substantially smaller than the true population coefficient of
\(\beta = 0.3\). And the discrepancy is not just due to sampling error.
In fact, we can derive the bias we are observing here.

For a simple bivariate regression model, it is straightforward to
quantify the bias due to measurement error. It will be \begin{align}
\mathop{\mathrm{\mathrm{Cov}}}(y, \tilde{x}) & = \mathop{\mathrm{\mathbb{E}}}[y \tilde{x}] \\
 & = \mathop{\mathrm{\mathbb{E}}}[(\beta\tilde{x} + \varepsilon)\tilde{x}] \\
 & = \mathop{\mathrm{\mathbb{E}}}[\beta\tilde{x}^{2} + \varepsilon \tilde{x}] \\
 & = \beta \mathop{\mathrm{\mathrm{Var}}}(\tilde{x}) \\
\hat{\beta} & = \frac{\mathop{\mathrm{\mathrm{Cov}}}(y, \tilde{x})}{\mathop{\mathrm{\mathrm{Var}}}(\tilde{x})} \\
 & = \frac{\mathop{\mathrm{\mathbb{E}}}[(\beta x + \varepsilon)(x + \upsilon)]}{\mathop{\mathrm{\mathbb{E}}}[(x + \upsilon)^{2}]} \\
 & = \frac{\mathop{\mathrm{\mathbb{E}}}[\beta x^{2} + \beta x \upsilon + \varepsilon x + \varepsilon \upsilon]}{\mathop{\mathrm{\mathbb{E}}}[x^{2} + 2 x \upsilon + \upsilon^{2}]} \\
 & = \beta \frac{\mathop{\mathrm{\mathrm{Var}}}(x)}{\mathop{\mathrm{\mathrm{Var}}}(x) + \mathop{\mathrm{\mathrm{Var}}}(\upsilon)}.
\end{align} which results if we assume that
\(\mathop{\mathrm{\mathbb{E}}}[x \upsilon] = 0\),
\(\mathop{\mathrm{\mathbb{E}}}[x \varepsilon] = 0\),
\(\mathop{\mathrm{\mathbb{E}}}[\tilde{x} \varepsilon] = 0\) and
\(\mathop{\mathrm{\mathbb{E}}}[\varepsilon \upsilon] = 0\)
\citep{Wooldridge2009}. However, the model we are interested is not a
bivariate model, so what was the point of showing the this? For one, it
points out that the bias will always move the estimated coefficient
closer to 0, since
\(\mathop{\mathrm{\mathrm{Var}}}(x) \le \mathop{\mathrm{\mathrm{Var}}}(x) + \mathop{\mathrm{\mathrm{Var}}}(\upsilon)\).
This means positive effects will be biased downwards and negative
effects biased upwards, always towards zero. This is why it is referred
to as \emph{attenuation bias}. Second, it will help to familiarize
ourselves with this equation to better understand the one for the
multivariate case.

Indeed, the magnitude of the bias in a multivariate model is somewhat
more complex to derive, but it will be \begin{align}
\hat{\beta} & = \beta \frac{\mathop{\mathrm{\mathrm{Var}}}(\theta)}{\mathop{\mathrm{\mathrm{Var}}}(\theta) + \mathop{\mathrm{\mathrm{Var}}}(\upsilon)}
\end{align} where \(\theta\) is just the residual of a regression in
which the true underlying variable is regressed on all other covariates.
In this case, we need to regress \(x_{t}\) on \(\alpha_{1}\) and
\(\alpha_{2}\) for:
\(x_{t} = \tau + \gamma_{1}\alpha_{1} + \gamma_{2}\alpha_{2} + \theta_{t}\)
where \(\tau\) is the intercept, and \(\gamma_{1}\), \(\gamma_{2}\) are
the regression coefficients and \(\theta_{t}\) is the residual
\citep[p.~318--320]{Wooldridge2009}.

Normally it is not possible to reconstruct the bias since in cases where
we have to rely on indicators, we would not know the true underlying
variable. Furthermore, in the case of a fixed-effects model, the
covariates are the unobserved time-invariant characteristics. However,
because we are working with simulated data, we have everything we need.
Going back to the results above, we can get the residuals of \(x_{t}\)
by either running a regression and saving the residuals, or we could
skip a step and get them directly using the `residual maker' matrix
\citep{Ruettenauer2020} which is
\(\bm{M} = \bm{I} - \bm{A}(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\)
and
\(\bm{A} = \begin{pmatrix}\bm{\iota}_{n} & \bm{\alpha_{1}} & \bm{\alpha_{2}}\end{pmatrix}\)
is the \(n \times 3\) matrix of the covariates (plus a constant).

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Make the n x n identity matrix}
\NormalTok{Id <-}\StringTok{ }\KeywordTok{diag}\NormalTok{( n)}

\CommentTok{# n x 2 matrix of covariates a1 and a2}
\NormalTok{A <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{( }\KeywordTok{c}\NormalTok{( }\KeywordTok{rep}\NormalTok{( }\DecValTok{1}\NormalTok{, n), dfw}\OperatorTok{$}\NormalTok{a1, dfw}\OperatorTok{$}\NormalTok{a2), }
             \DataTypeTok{nrow =}\NormalTok{ n, }\DataTypeTok{ncol =} \DecValTok{3}\NormalTok{)}

\CommentTok{# The residual maker matrix M = I - A(A'A)^-1 A'}
\NormalTok{M <-}\StringTok{ }\NormalTok{Id }\OperatorTok{-}\StringTok{ }\NormalTok{A }\OperatorTok{%*%}\StringTok{ }\KeywordTok{solve}\NormalTok{( }\KeywordTok{t}\NormalTok{( A) }\OperatorTok{%*%}\StringTok{ }\NormalTok{A) }\OperatorTok{%*%}\StringTok{ }\KeywordTok{t}\NormalTok{( A)}

\CommentTok{# Save the residuals, t for 'theta'}
\NormalTok{t <-}\StringTok{ }\NormalTok{M }\OperatorTok{%*%}\StringTok{ }\NormalTok{dfw}\OperatorTok{$}\NormalTok{x1}

\CommentTok{# Re-run the FE model from above with the 'true' }
\CommentTok{# independent variable for the correct estimate for beta }
\NormalTok{fe_sem.fit <-}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe_sem, }\DataTypeTok{data =}\NormalTok{ dfw, }\DataTypeTok{estimator =} \StringTok{"ML"}\NormalTok{)}

\CommentTok{# The equation for the biased beta }
\KeywordTok{lavInspect}\NormalTok{( fe_sem.fit, }\StringTok{"list"}\NormalTok{)[ }\DecValTok{6}\NormalTok{, }\DecValTok{14}\NormalTok{]}\OperatorTok{*}
\StringTok{  }\NormalTok{(( }\KeywordTok{var}\NormalTok{( t))}\OperatorTok{/}\NormalTok{( }\KeywordTok{var}\NormalTok{( t) }\OperatorTok{+}\StringTok{ }\KeywordTok{var}\NormalTok{( dfw}\OperatorTok{$}\NormalTok{x11 }\OperatorTok{-}\StringTok{ }\NormalTok{dfw}\OperatorTok{$}\NormalTok{x1)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##           [,1]
## [1,] 0.1027325
\end{verbatim}

\doublespacing

\singlespacing

\doublespacing

From this we can see that the biased estimate above of \(\hat{\beta} =\)
0.091 roughly comes from
\(\beta \frac{\mathop{\mathrm{\mathrm{Var}}}(\theta_{t})}{\mathop{\mathrm{\mathrm{Var}}}(\theta_{t}) + \mathop{\mathrm{\mathrm{Var}}}(\upsilon_{t})} = 0.294 \frac{1.054}{3.018}\)
= 0.103; `roughly' because the equation here is the population equation.
Due to sampling error, the estimates will tend vary slightly.

\hypertarget{using-latent-variables-to-deal-with-measurement-error}{%
\subsubsection{Using latent variables to deal with measurement
error}\label{using-latent-variables-to-deal-with-measurement-error}}

The way we deal with measurement error in SEM is surprisingly similar to
the logic of fixed-effects regression. Namely, if we have multiple
cross-sectional observations of the underlying construct of interest,
then we can define a latent variable that represents the common variance
across those multiple variables. Contrast this with the use of
longitudinal repeated measures to isolate the common variance across
time.

So, if we do in fact have multiple cross-sectional indicators for the
underlying variables of interest, then we can partition them into an
explained and unexplained portion:\\
\begin{align}
x_{kt} & = \lambda_{kt}^{x}\xi_{t} + \delta_{kt}, \\
y_{kt} & = \lambda_{kt}^{y}\eta_{t} + \varepsilon_{kt},
\end{align} where \(x_{kt}\) and \(y_{kt}\) are the \(k^{th}\)
indicators, \(\xi_{t}\) and \(\eta_{t}\) are latent factors representing
the common variance across the cross-sectional repeated measures, and
\(\delta_{kt}\) and \(\varepsilon_{kt}\) are the unexplained portions of
\(x_{t}\) and \(y_{t}\), respectively. The latent factors are linked to
the observed indicators through the factor loadings \(\lambda_{kt}\).

Thus, our FE regression equation changes from
\(y_{t} = \beta x_{t} + \alpha + \varepsilon_{t}\) to: \begin{align}
\eta_{t} & = \beta \xi_{t} + \alpha + \zeta_{t}
\end{align} where \(\zeta_{t}\) represents the disturbance, in other
words the residual of the latent dependent variable \(\eta_{t}\). First,
however, let us double-check that measurement error in the dependent
variable only increases the error variance (thus also increasing
standard errors and reducing \(R^{2}\)), but does not systematically
bias the coefficients of interest. The next model uses the indicators of
\(x\) and specifies latent variables (\(\xi_{t}\), \texttt{xi} in the
code) to represent the valid cross-sectional variance. The dependent
variable in the model is one of the impercisely measured indicators of
\(y\).

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fe_sem3 <-}\StringTok{ '}
\StringTok{# Define individual effects variable }
\StringTok{a =~ 1*y11 + 1*y12 + 1*y13 + 1*y14 + 1*y15}
\StringTok{# Measurement model for independent variables, xi }
\StringTok{xi1 =~ 1*x11 + x21 + x31 }
\StringTok{xi2 =~ 1*x12 + x22 + x32}
\StringTok{xi3 =~ 1*x13 + x23 + x33}
\StringTok{xi4 =~ 1*x14 + x24 + x34}
\StringTok{xi5 =~ 1*x15 + x25 + x35}
\StringTok{# Regressions, constrain coefficient to be equal over time}
\StringTok{y11 ~ b*xi1}
\StringTok{y12 ~ b*xi2 }
\StringTok{y13 ~ b*xi3}
\StringTok{y14 ~ b*xi4}
\StringTok{y15 ~ b*xi5}
\StringTok{# Allow unrestricted correlation between eta and covariates}
\StringTok{a ~~ xi1 + xi2 + xi3 + xi4 + xi5}
\StringTok{xi1 ~~ xi2 + xi3 + xi4 + xi5}
\StringTok{xi2 ~~ xi3 + xi4 + xi5}
\StringTok{xi3 ~~ xi4 + xi5}
\StringTok{xi4 ~~ xi5}
\StringTok{# Constrain residual variances to be equal over time}
\StringTok{y11 ~~ e*y11}
\StringTok{y12 ~~ e*y12}
\StringTok{y13 ~~ e*y13}
\StringTok{y14 ~~ e*y14}
\StringTok{y15 ~~ e*y15}
\StringTok{'}
\NormalTok{fe_sem3.fit <-}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe_sem3, }
                    \DataTypeTok{data =}\NormalTok{ dfw, }
                    \DataTypeTok{estimator =} \StringTok{"ML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\doublespacing

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{( fe_sem3.fit)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
...
## Regressions:
##                    Estimate  Std.Err  z-value  P(>|z|)
##   y11 ~                                               
##     xi1        (b)    0.299    0.029   10.302    0.000
##   y12 ~                                               
##     xi2        (b)    0.299    0.029   10.302    0.000
##   y13 ~                                               
##     xi3        (b)    0.299    0.029   10.302    0.000
##   y14 ~                                               
##     xi4        (b)    0.299    0.029   10.302    0.000
##   y15 ~                                               
##     xi5        (b)    0.299    0.029   10.302    0.000
...
\end{verbatim}

\doublespacing

The estimated coefficient here in model \texttt{fe\_sem3.fit} is
\(\hat{\beta}_{y_{1t},\xi_{t}} =\) 0.299 which is very close to the
estimated coefficient in the first, correctly specified model
\texttt{fe\_sem.fit}, where \(\hat{\beta}_{y_{t},x_{t}} =\) 0.294.
Notice, however, that the standard error of the estimate is
substantially larger, with 0.029 in \texttt{fe\_sem3.fit} vs.~0.016 in
\texttt{fe\_sem.fit} in which \(y\) was measured without error. The
explained variance (\(R^{2}\)) in the dependent variable was also much
higher in the first model:

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{lavInspect}\NormalTok{( fe_sem.fit, }\StringTok{"r2"}\NormalTok{)[ }\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##        y1        y2        y3        y4        y5 
## 0.5893174 0.5928913 0.5894869 0.5853815 0.5845328
\end{verbatim}

\doublespacing

compared to the current model:

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{lavInspect}\NormalTok{( fe_sem3.fit, }\StringTok{"r2"}\NormalTok{)[ }\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##       y11       y12       y13       y14       y15 
## 0.3901187 0.3914316 0.3828739 0.3850362 0.3703136
\end{verbatim}

\doublespacing

Finally, to see the benefits of removing measurement error from the
dependent variable in terms of standard errors and \(R^{2}\) statistics,
we can specify a model with latent variables representing the valid
cross-sectional variance in \(y\) (\texttt{n} for \(\eta\) in the code).

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fe_sem4 <-}\StringTok{ '}
\StringTok{# Measurement model for dependent variable, n for eta}
\StringTok{n1 =~ 1*y11 + y21 + y31}
\StringTok{n2 =~ 1*y12 + y22 + y32}
\StringTok{n3 =~ 1*y13 + y23 + y33}
\StringTok{n4 =~ 1*y14 + y24 + y34}
\StringTok{n5 =~ 1*y15 + y25 + y35}
\StringTok{# Define individual effects variable }
\StringTok{a =~ 1*n1 + 1*n2 + 1*n3 + 1*n4 + 1*n5}
\StringTok{# Measurement model for independent variables, xi }
\StringTok{xi1 =~ 1*x11 + x21 + x31 }
\StringTok{xi2 =~ 1*x12 + x22 + x32}
\StringTok{xi3 =~ 1*x13 + x23 + x33}
\StringTok{xi4 =~ 1*x14 + x24 + x34}
\StringTok{xi5 =~ 1*x15 + x25 + x35}
\StringTok{# Regressions, constrain coefficient to be equal over time}
\StringTok{n1 ~ b*xi1}
\StringTok{n2 ~ b*xi2 }
\StringTok{n3 ~ b*xi3}
\StringTok{n4 ~ b*xi4}
\StringTok{n5 ~ b*xi5}
\StringTok{# Allow unrestricted correlation between eta and covariates}
\StringTok{a ~~ xi1 + xi2 + xi3 + xi4 + xi5}
\StringTok{xi1 ~~ xi2 + xi3 + xi4 + xi5}
\StringTok{xi2 ~~ xi3 + xi4 + xi5}
\StringTok{xi3 ~~ xi4 + xi5}
\StringTok{xi4 ~~ xi5}
\StringTok{# Constrain residual variances to be equal over time}
\StringTok{n1 ~~ e*n1}
\StringTok{n2 ~~ e*n2}
\StringTok{n3 ~~ e*n3}
\StringTok{n4 ~~ e*n4}
\StringTok{n5 ~~ e*n5}
\StringTok{'}
\NormalTok{fe_sem4.fit <-}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe_sem4, }
                    \DataTypeTok{data =}\NormalTok{ dfw, }
                    \DataTypeTok{estimator =} \StringTok{"ML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\doublespacing

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{( fe_sem4.fit)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
...
## Regressions:
##                    Estimate  Std.Err  z-value  P(>|z|)
##   n1 ~                                                
##     xi1        (b)    0.264    0.023   11.515    0.000
##   n2 ~                                                
##     xi2        (b)    0.264    0.023   11.515    0.000
##   n3 ~                                                
##     xi3        (b)    0.264    0.023   11.515    0.000
##   n4 ~                                                
##     xi4        (b)    0.264    0.023   11.515    0.000
##   n5 ~                                                
##     xi5        (b)    0.264    0.023   11.515    0.000
...
\end{verbatim}

\doublespacing

Here, the effect \(\hat{\beta}_{\eta_{t},\xi_{t}}\) is somewhat further
off of the true effect of 0.3 than the preceding models. This will
depend on how the latent variables are estimated, which themselves will
depend on the underlying correlations between the indicators. Again, if
the main goal of the model is to avoid bias, it may be advisable to just
leave the manifest dependent variable as it is, and worry about
measurement error in the independent variables.

\hypertarget{time-invariant-predictors}{%
\subsection{Time-invariant predictors}\label{time-invariant-predictors}}

What if we do not just want to just control for the effects of all
time-invariant variables, but investigate some of them in detail? Many
time-invariant variables, like sex, birth cohort, nationality,
education, etc. can be interesting on their own. And typically, many of
these variables are readily available in a given dataset. The
traditional OLS-based FE model does not allow for this, as it wipes out
the effect of \emph{all} time-invariant variables, whether observed or
not.

In SEM, we can easily specify a type of \emph{hybrid} FE/RE model
\citep{Bollen2010} that allows us to control for time-invariant
unobserved heterogeneity while also investigating the effects of
specific observed time-invariant predictors.\footnote{These types of
  models have become well known outside of SEM as well, see for example
  \citet{Allison2011}; \citet{Schunck2013}; \citet{Bell2018}.}

In the next example, we continue with the most complex model we have
specified so far, \texttt{fe\_sem4.fit} in which measurement error in
both the independent and dependent variables is accounted for using
latent variables. Now, we would like as well to specifically investigate
the effect of \(\alpha_{2}\) on the dependent variable. The equation for
this model changes to:
\(\eta_{t} = \beta \xi_{t} + \alpha + \gamma \alpha_{2} + \zeta_{t}\).

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fe_sem5 <-}\StringTok{ '}
\StringTok{# Measurement model for dependent variable, n for eta}
\StringTok{n1 =~ 1*y11 + y21 + y31}
\StringTok{n2 =~ 1*y12 + y22 + y32}
\StringTok{n3 =~ 1*y13 + y23 + y33}
\StringTok{n4 =~ 1*y14 + y24 + y34}
\StringTok{n5 =~ 1*y15 + y25 + y35}
\StringTok{# Define individual effects variable }
\StringTok{a =~ 1*n1 + 1*n2 + 1*n3 + 1*n4 + 1*n5}
\StringTok{# Measurement model for independent variables, xi }
\StringTok{xi1 =~ 1*x11 + x21 + x31 }
\StringTok{xi2 =~ 1*x12 + x22 + x32}
\StringTok{xi3 =~ 1*x13 + x23 + x33}
\StringTok{xi4 =~ 1*x14 + x24 + x34}
\StringTok{xi5 =~ 1*x15 + x25 + x35}
\StringTok{# Regressions, constrain coefficient to be equal over time}
\StringTok{n1 ~ b*xi1 + g*a2}
\StringTok{n2 ~ b*xi2 + g*a2}
\StringTok{n3 ~ b*xi3 + g*a2}
\StringTok{n4 ~ b*xi4 + g*a2}
\StringTok{n5 ~ b*xi5 + g*a2}
\StringTok{# Allow unrestricted correlation between eta and covariates}
\StringTok{a ~~ xi1 + xi2 + xi3 + xi4 + xi5 + 0*a2}
\StringTok{a2 ~~ xi1 + xi2 + xi3 + xi4 + xi5}
\StringTok{xi1 ~~ xi2 + xi3 + xi4 + xi5}
\StringTok{xi2 ~~ xi3 + xi4 + xi5}
\StringTok{xi3 ~~ xi4 + xi5}
\StringTok{xi4 ~~ xi5}
\StringTok{# Constrain residual variances to be equal over time}
\StringTok{n1 ~~ e*n1}
\StringTok{n2 ~~ e*n2}
\StringTok{n3 ~~ e*n3}
\StringTok{n4 ~~ e*n4}
\StringTok{n5 ~~ e*n5}
\StringTok{'}
\NormalTok{fe_sem5.fit <-}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe_sem5, }
                    \DataTypeTok{data =}\NormalTok{ dfw, }
                    \DataTypeTok{estimator =} \StringTok{"ML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\doublespacing

Keep in mind, based on the DGP, the true parameters are \(\beta = 0.3\)
and \(\gamma = 0.45\).

\singlespacing

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{( fe_sem5.fit)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
...
## Regressions:
##                    Estimate  Std.Err  z-value  P(>|z|)
##   n1 ~                                                
##     xi1        (b)    0.265    0.023   11.515    0.000
##     a2         (g)    0.490    0.033   14.999    0.000
##   n2 ~                                                
##     xi2        (b)    0.265    0.023   11.515    0.000
##     a2         (g)    0.490    0.033   14.999    0.000
##   n3 ~                                                
##     xi3        (b)    0.265    0.023   11.515    0.000
##     a2         (g)    0.490    0.033   14.999    0.000
##   n4 ~                                                
##     xi4        (b)    0.265    0.023   11.515    0.000
##     a2         (g)    0.490    0.033   14.999    0.000
##   n5 ~                                                
##     xi5        (b)    0.265    0.023   11.515    0.000
##     a2         (g)    0.490    0.033   14.999    0.000
...
\end{verbatim}

\doublespacing

From this we can see that such a hybrid model is does a good job of
estimating the coefficients of interest, with \(\hat{\beta} =\) 0.265
(0.023) and \(\hat{\gamma} =\) 0.49 (0.033).

It is important, however, to realize that the unbiasedness of
\(\hat{\gamma}\) in this model is dependent on the assumption that
\(\mathop{\mathrm{\mathbb{E}}}[\zeta | \bm{\xi_{t}}, \alpha_{2}] = 0\).
In other words, the idiosyncratic error is mean independent of
\(\bm{\xi_{t}} = (\xi_{1}, \xi_{2}, ..., \xi_{T})\) as well as
\(\alpha_{2}\). The first part is easier to accept because we are
controlling for all potential time-invariant confounders that could
induce a relationship between the independent variable and the error.
The unbiasedness of \(\hat{\gamma}\), on the other hand rests on the
assumption that the time-invariant predictor is independent of the
error. If \(\alpha_{2}\) represented the respondent's intelligence and
\(\eta_{t}\), the dependent variable, represented the respondent's
income, for example, then \(\hat{\gamma}\) would be biased if both were
dependent on a third time-invariant variable, say level of schooling, if
it is not controlled for. For this reason, we need to treat the
regression on a time-invariant predictor like any other regular
multivariate regression model and look to include all plausible
potential confounders as controls in the model, or turn to other
methods, e.g., instrumental variables.

\bibliographystyle{tfcad}
\bibliography{references2.bib}

\end{document}

### 2020-07-28 article_format/extensions_format_anon.Rmd
Changed the formating, made some minor edits

--- 2020-07-28 article_format/extensions_format.Rmd
@@
 output: rticles::tf_article
+keep_tex: true
 ---
+
+\pagenumbering{arabic} 
+\setcounter{page}{1}
 

### 2020-07-28 article_format/fig1.tex
Changed the formating, made some minor edits

--- 2020-07-28 article_format/article_format.tex
@@
-% interactcadsample.tex
-% v1.03 - April 2017
+\documentclass[tikz]{standalone}
 
-\documentclass[]{interact}
-
-\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
-\usepackage{subfigure}% Support for small, `sub' figures and tables
-%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required
-
-\usepackage{natbib}% Citation support using natbib.sty
-\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
-\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty
-
-\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{proposition}[theorem]{Proposition}
-
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{example}[theorem]{Example}
-
-\theoremstyle{remark}
-\newtheorem{remark}{Remark}
-\newtheorem{notation}{Notation}
-
-% see https://stackoverflow.com/a/47122900
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-
-\usepackage{hyperref}
-\usepackage[utf8]{inputenc}
-\usepackage{booktabs}
-\usepackage{bm}
-\usepackage{mathtools}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{tikz}
 \usetikzlibrary{arrows}
-\usepackage[nofiglist]{endfloat}
-\usepackage{blkarray}
-\usepackage{setspace}
-\usepackage{etoolbox}
-\DeclareMathOperator{\E}{\mathbb{E}}
-\DeclareMathOperator{\Var}{\mathrm{Var}}
-\DeclareMathOperator{\Cov}{\mathrm{Cov}}
-\DeclareMathOperator*{\argmax}{arg\,max}
-\DeclareMathOperator*{\argmin}{arg\,min}
-\mathtoolsset{showonlyrefs}
-\BeforeBeginEnvironment{equation}{\begin{singlespace}\vspace*{-\baselineskip}}
-\AfterEndEnvironment{equation}{\end{singlespace}\noindent\ignorespaces}
-\BeforeBeginEnvironment{align}{\begin{singlespace}\vspace*{-\baselineskip}}
-\AfterEndEnvironment{align}{\end{singlespace}\noindent\ignorespaces}
-\def\tightlist{}
-\pagenumbering{gobble}
 
@@
 
-\articletype{}
-
-\title{A closer look at fixed effects regression in structural equation
-modeling using \texttt{lavaan}}
-
-\author{\name{Henrik Kenneth Andersen$^{a}$}
-\affil{$^{a}$Chemnitz University of Technology, Institute of Sociology, Chair for
-Empirical Social Research, Thüringer Weg 9, 09126 Chemnitz, Germany}
-}
-
-\thanks{CONTACT Henrik Kenneth Andersen. Email: \href{mailto:henrik.andersen@soziologie.tu-chemnitz.de}{\nolinkurl{henrik.andersen@soziologie.tu-chemnitz.de}}}
-
-\maketitle
-
-\begin{abstract}
-This article provides an in-depth look at fixed effects regression in
-the structural equation modeling (SEM) framework, specifically the
-application of fixed effects in the \texttt{lavaan} package for
-\texttt{R}. It is meant as a applied guide for researchers, covering the
-underlying model specification, syntax, and summary output. Online
-supplementary materials further discuss various common extensions to the
-basic fixed-effect model, demonstrating how to relax model assumptions,
-deal with measurement error in both the dependent and independent
-variables, and include time-invariant predictors in a type of hybrid
-fixed-/ random effects model.
-\end{abstract}
-
-\begin{keywords}
-Fixed effects, structural equation modeling, lavaan, R, panel analysis
-\end{keywords}
-
-\newpage
-
-\pagenumbering{arabic} 
-\setcounter{page}{1}
-
-\doublespacing
-
-\hypertarget{intro}{%
-\section{Introduction}\label{intro}}
-
-Several years ago, \citet{Curran2011} reflected positively on the
-growing use of panel studies in empirical social research. Some of the
-strengths of panel data are well-known, e.g., the ability to establish
-temporal precedence, increased statistical power and the reduction of
-potential alternative models. However, perhaps the greatest strength of
-panel data is that they allow for a more rigorous testing of substantive
-theories. Panel data, i.e., repeated measures of the same observed units
-(people, schools, firms, countries, etc.), allow researchers to
-decompose the error term into a part that stays constant within units
-and the part that changes over time. The part that does not change over
-time can be seen as the combined effect of all time-invariant influences
-(e.g., sex, date of birth, nationality) on the dependent variable. Fixed
-effects (FE) regression involves controlling for these time-invariant
-influences via a number of various methods. It thus accounts for a
-likely and common source of bias.
-
-Structural equation modeling (SEM) is a popular regression framework.
-One of its main strengths is its flexibility. Not only can complex
-causal structures with multiple dependent variables be tested
-simultaneously, but in longitudinal (and, more generally, hierarchical)
-studies both time-varying and invariant predictors can be included, and
-effects can easily be allowed to vary over time. Thus researchers can
-allow for and study effects that increase or fade over time, or that
-appear only in specific periods. Beyond that, with the use of latent
-variables, SEM provides a way to deal with measurement error and get
-closer to the true underlying constructs of interest.
-
-There are a number of articles describing basic concept of panel model
-regression, and FE regression in SEM
-\citep[e.g.,][]{Allison2011, Bollen2010, Teachman2001}. This article is
-intended as a \textit{practical guide} for researchers looking for
-in-depth help with specifying FE models in SEM. It focuses on the
-\texttt{lavaan} \citep{R-lavaan} package for \texttt{R} \citep{R-base}.
-While \texttt{Mplus} \citep{Mplus} is arguably the most robust SEM
-software currently available (in terms of features like alignment,
-latent variable interactions, for example), the \texttt{lavaan} package
-has many benefits. First, like \texttt{R} it is open source and
-completely free. For researchers dipping their toes into SEM, there is
-no financial barrier to try, and no risk if they decide it is not for
-them. Second, the implementation of \texttt{lavaan} in the larger
-\texttt{R} environment is an enormous advantage. Instead of poring over
-reams of plain text, copying out coefficients by hand, every part of the
-\texttt{lavaan} output is available as an object. This means that all
-aspects of the model, from fit indices, to coefficients and standard
-errors, to the model matrices, can be accessed and easily integrated
-into tables and plots. Furthermore, \texttt{R} can be used for a great
-deal of applications. It can be used to manage and manipulate as well as
-simulate data, perform symbolic algebra, run more traditional analyses
-(e.g., multiple regression, logistic regression, principal component
-analysis), etc. Once one is comfortable using \texttt{R}, there is no
-longer any need to switch between different software for data
-preparation and analysis.
-
-The following article outlines the basic idea of panel regression, the
-particularities of panel regression SEM, and shows its implementation in
-\texttt{lavaan}. The focus on FE panel regression stems from the fact
-that the assumptions needed to justify other common models, e.g., random
-effects, are often implausible. As we will see, the practical
-differences between the two are, in any event, very small. Using
-\href{https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R}{simulated
-data}, it demonstrates and annotates the code for the most basic FE
-model and provides an overview of the summary output. One of the main
-strengths of panel SEM compared to the more traditional methods of panel
-analysis is its flexibility. Therefore, a number of potential extensions
-to the basic model, including relaxing various assumptions, dealing with
-measurement error in both the independent and dependent variables, as
-well as the inclusion of time-invariant predictors in the form of a
-hybrid fixed-/ random effects model, are shown in detail in the form of
-\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
-supplementary materials}.
-
-\hypertarget{panel}{%
-\section{Panel models}\label{panel}}
-
-To begin, let us start by reviewing a general panel model
-\citep{Bollen2010}, also referred to as the `unobserved effects model'
-\citep{Wooldridge2012, R-plm_a} (we will return to this model in the
-\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
-supplementary materials} when we discuss loosening assumptions)
-\begin{align}
-y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \alpha_{i} + \varepsilon_{it} \label{eq:gpm}
-\end{align} where \(y_{it}\) is the dependent variable for unit
-\(i, \ i = 1, ..., N\) at time \(t, \ t = 1, ..., T\), \(\bm{x}_{it}\)
-is a \(1 \times K\) vector of time-varying covariates (which could
-include a constant) linked to the dependent variable by the
-\(K \times 1\) vector of coefficients \(\bm{\beta}\). \(\bm{z}_{i}\) is
-a \(1 \times M\) vector of time-invariant covariates linked to the
-dependent variable by the \(M \times 1\) vector of coefficients in
-\(\bm{\gamma}\), \(\alpha_{i}\) represents the combined effect of all
-unobserved time-constant variables affecting the dependent variable and
-\(\varepsilon_{it}\) is the idiosyncratic error.
-
-We can make stating some of the model assumptions easier by rewriting it
-in matrix notation \begin{align}
-\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\iota}_{T}\alpha_{i} + \bm{\varepsilon}_{i}
-\end{align} where \(\bm{y}_{i}\) and \(\bm{\varepsilon}_{i}\) are
-\(T \times 1\) vectors, \(\bm{X}_{i}\) and \(\bm{Z}_{i}\) are
-\(T \times K\) and \(T \times M\) matrices, respectively,
-\(\bm{\iota}_{T}\) is a \(T \times 1\) vector of ones and \(\alpha_{i}\)
-is a scalar. \(\bm{\beta}\) and \(\bm{\gamma}\) are unchanged from
-Equation \eqref{eq:gpm}.
-
-Consistency of the following models requires the assumption of strict
-exogeneity, although what constitutes strict exogeneity differs between
-the random and fixed effects setups. Each assumption will be discussed
-shortly. Apart from that, we typically make the following assumptions
-about this model \citep[see, e.g.,][]{Wooldridge2002, Schmidheiny2019}:
-
-\begin{itemize}
-\tightlist
-\item
-  Linearity: the model is linear in its parameters.
-\item
-  Independence: the observations are independent across individuals
-  (assured by random sampling in the cross-section), but not necessarily
-  across time.
-\item
-  The usual rank condition: we have more observations than independent
-  variables and there is no perfect collinearity between any of the
-  independent variables.
-\end{itemize}
-
-\hypertarget{re}{%
-\subsection{Random effects}\label{re}}
-
-Somewhat unintuitively, the modern approach to FE regression treats the
-unobserved effect \(\alpha_{i}\) as a \textit{random} variable, rather
-than a fixed parameter to be estimated for each of the cross sectional
-units \citep{Wooldridge2002}. And in fact, this view carries over to the
-way in which FE regression is applied in SEM. So, in order to help
-facilitate the discussion on the application of FE-SEM later on, it
-makes sense to spend some time discussing the basic idea of random
-effects regression (again, as we will see, the practical difference
-between the two models is very small).
-
-For the random effects (RE) model, we define a composite error term:
-\(\nu_{it} = \alpha_{i} + \varepsilon_{it}\) and rewrite the model in
-Equation \eqref{eq:gpm} as \begin{align}
-y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \nu_{it}, \ \text{or} \\
-\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\nu}_{i},
-\end{align} where
-\(\bm{\nu}_{i} = \alpha_{i} \bm{\iota}_{T} + \bm{\varepsilon}_{i}\) and
-\(\bm{\iota}_{T}\) is a \(T \times 1\) vector of ones. The strict
-exogeneity assumption in the RE model implies \begin{align}
-\mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = 0, \\
-\mathop{\mathrm{\mathbb{E}}}[\alpha_{i} |\bm{X}_{i}, \bm{z}_{i}] = \mathop{\mathrm{\mathbb{E}}}[\alpha_{i}] & = 0,
-\end{align} where \(\bm{X}_{i} = \bm{x}_{i1}, ..., \bm{x}_{iT}\). For
-both parts, the assumption that the unconditional expectations are 0 is
-unproblematic as long as a constant is included in the regression. The
-first part says the idiosyncratic errors at each timepoint are assumed
-to be independent of the explanatory variables at \textit{all}
-timepoints which is stronger than just assuming that they are
-\textit{contemporaneously} independent. This implies that they are also
-uncorrelated, i.e.,
-\(\mathop{\mathrm{\mathbb{E}}}[\bm{x}_{is}^{\intercal}\varepsilon_{it}] = \bm{0}\)
-and
-\(\mathop{\mathrm{\mathbb{E}}}[\bm{z}_{i}^{\intercal}\varepsilon_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T\)
-\citep{Wooldridge2002, Bruederl2015}. We assume the idiosyncratic errors
-are further independent of the individual effects, which implies
-\(\mathop{\mathrm{\mathbb{E}}}[\alpha_{i}\varepsilon_{it}] = 0\).
-
-The second part is the potentially controversial assumption: it states
-that the individual effects are uncorrelated with the independent
-variables. We can use an intuitive concrete example to show why this is
-often controversial: If we are interested in the question of whether
-married men earn more than unmarried men, then the second part of the
-strict exogeneity assumption means that a man's marriage status would
-have to be uncorrelated with all the time-invariant characteristics that
-could potentially make that man an attractive marriage candidate in the
-first place; e.g., looks, personality, family's status, profession, etc.
-\citep{Bruederl2015}.\footnote{Assuming, for the sake of argument, that
-  these characteristics are constant over time.}
-
-From what we have discussed so far, the \(T \times T\) covariance matrix
-of the errors
-\(\bm{\Omega}_{i} = \mathop{\mathrm{\mathbb{E}}}[\bm{\nu}_{i}\bm{\nu}_{i}^{\intercal}]\)
-can be constructed. However, the standard random effects model adds the
-additional assumptions \begin{alignat}{3}
-\mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}^{2} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}^{2}] && = \sigma_{\varepsilon}^{2}, && \ t = 1, ..., T, \\
-\mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}\varepsilon_{is}| \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}\varepsilon_{is}] && = 0, && \ \forall \ t \ne s 
-\end{alignat} i.e., the idiosyncratic errors are conditionally
-homoscedastic and serially uncorrelated and \begin{alignat}{2}
-\mathop{\mathrm{\mathbb{E}}}[\alpha_{i}^{2}| \bm{X}_{i},\bm{z}_{i}] & = \mathop{\mathrm{\mathbb{E}}}[\alpha_{i}^{2}] && = \sigma^{2}_{\alpha}
-\end{alignat} i.e., the individual effects are conditionally
-homoscedastic (they are necessarily serially correlated as long as
-\(\sigma_{\alpha}^{2} > 0\)). From that, we arrive at the typical random
-effects structure of the \(NT \times NT\) matrix \(\bm{\Omega}\):
-
-\begin{align}
-\bm{\Omega} & = 
-\begin{pmatrix}
-\bm{\Omega}_{1} & \hdots & 0 & \hdots & 0 \\
-\vdots & \ddots & \vdots & \ddots & \vdots \\
-0 & \hdots & \bm{\Omega}_{i} & \hdots & 0 \\
-\vdots & \ddots & \vdots & \ddots & \vdots \\
-0 & \hdots & 0 & \hdots & \bm{\Omega}_{N}
-\end{pmatrix}
-\end{align} with \(T \times T\) typical elements \begin{align}
-\bm{\Omega}_{i} & = 
-\begin{pmatrix}
-\sigma^{2}_{\nu} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\alpha} \\
-\sigma^{2}_{\alpha} & \sigma^{2}_{\nu} & \hdots & \sigma^{2}_{\alpha} \\
-\vdots & \vdots & \ddots & \vdots \\
-\sigma^{2}_{\alpha} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\nu}
-\end{pmatrix} \label{eq:omeganui}
-\end{align} where
-\(\sigma^{2}_{\nu} = \sigma^{2}_{\alpha} + \sigma^{2}_{\varepsilon}\)
-\citep{Wooldridge2002, Schmidheiny2019}. This means that in the
-conditional covariance matrix of the errors, given the time-varying and
--invariant covariates, units over time will be correlated due to the
-individual effects. We should keep the covariance structure of the
-errors in mind as it will help make sense of the use of latent variables
-to decompose the dependent variable into between- and within-variance
-components, discussed below in Section \ref{fe-sem}.
-
-Estimation of the RE model can be done using feasible generalized least
-squares (GLS) in which the two unknowns in \(\bm{\Omega}\),
-\(\sigma_{\alpha}^{2}\) and \(\sigma_{\nu}^{2}\), are first estimated
-using pooled ordinary least squares (pooled OLS or POLS),
-where\footnote{Normally a degrees-of-freedom correction is applied by
-  subtracting off the number of independent variables that is negligible
-  in large N samples. It is ignored here for the sake of simplicity.}
-\begin{align}
-\hat{\sigma}_{\nu}^{2} & = \frac{1}{NT}\sum_{i = 1}^{N}\sum_{t = 1}^{T}\hat{\nu}_{it}^{2}, \\
-\hat{\sigma}_{\varepsilon}^{2} & = \frac{1}{NT - N}\sum_{i = 1}^{N}\sum_{t = 1}^{T}(\hat{\nu}_{it} - \bar{\hat{\nu}}_{i})^{2}
-\end{align} and
-\(\hat{\nu}_{it} = y_{it} - \bm{x}_{it}\hat{\bm{\beta}}_{\tiny{POLS}} - \bm{z}_{i}\hat{\bm{\gamma}}_{\tiny{POLS}}\),
-\(\bar{\hat{\nu}}_{i} = T^{-1}\sum_{t = 1}^{T}\hat{\nu}_{it}\) and
-\(\hat{\sigma}^{2}_{\alpha} = \hat{\sigma}^{2}_{\nu} - \hat{\sigma}^{2}_{\varepsilon}\).
-Then, the coefficients are estimated using those estimates in the
-variance matrix \(\hat{\bm{\Omega}}\): \begin{align}
-\begin{pmatrix}
-\hat{\bm{\beta}}_{RE} \\
-\hat{\bm{\gamma}}_{RE}
-\end{pmatrix} & = 
-(\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{W})^{-1}\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{y}
-\end{align} where
-\(\bm{W} = \begin{pmatrix}\bm{X} & \bm{Z}\end{pmatrix}\) and \(\bm{X}\)
-is \(NT \times K\) and \(\bm{Z}\) is \(NT \times M\) and \(\bm{y}\) is
-\(NT \times 1\).
-
-In practice, however, computational problems can arise with large
-cross-sectional samples, where it can become difficult to invert the
-\(\hat{\bm{\Omega}}\) matrix. One solution is to use `partial-demeaning'
-to transform the data before performing simple POLS: \begin{align}
-(y_{it} - \theta\bar{y}_{i}) & = (\bm{x}_{it} - \theta\bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \theta \bar{\bm{z}}_{i})\bm{\gamma} + (\varepsilon_{it} - \theta \bar{\varepsilon}_{i}) \label{eq:partialdemean}
-\end{align} where
-\(\theta = 1 - [\sigma_{\alpha}^{2}/(\sigma_{\alpha}^{2} + T \sigma_{\varepsilon}^{2})]^{1/2}\),
-and \(\bar{y}_{i} = T^{-1}\sum_{t = 1}^{T}y_{it}\),
-\(\bar{\bm{x}}_{i} = T^{-1}\sum_{t = 1}^{T}\bm{x}_{it}\),
-\(\bar{\bm{z}}_{i} = T^{-1}\sum_{t = 1}^{T}\bm{z}_{i}\) and
-\(\bar{\varepsilon}_{i} = T^{-1}\sum_{t = 1}^{T}\varepsilon_{it}\)
-\citep{R-plm_a}.
-
-The RE model can also be estimated in the maximum likelihood framework,
-where in the associated literature on panel models are generally
-referred to as either mixed models, hierarchical models or longitudinal
-models. The typical RE model discussed here is the equivalent to a mixed
-model with random intercepts and fixed slopes \citep{R-plm_a}. Under the
-assumption of normality, along with homoscedasticity and serially
-uncorrelated errors, the maximum likelihood estimator is the same as the
-OLS estimator. For more on the topic of mixed models, see for example
-\citet{R-lme4}, \citet{Bates2010}.
-
-\hypertarget{fe}{%
-\subsection{Fixed effects}\label{fe}}
-
-For fixed effects, we assume that the individual effects are
-\textit{not independent} of the model covariates, i.e.,
-\(\mathop{\mathrm{\mathbb{E}}}[\alpha_{i}|\bm{X}_{i},\bm{z}_{i}] \ne \mathop{\mathrm{\mathbb{E}}}[\alpha_{i}] \ne 0\).
-Under this assumption, grouping the individual effects in with the
-composite error will cause the coefficients of interest, here
-specifically \(\bm{\beta}\) to be inconsistent
-\citep{Wooldridge2002, Wooldridge2012}. We write the model therefore
-again in terms of the general panel model in Equation \eqref{eq:gpm}
-with separate individual effects and idiosyncratic error terms. In order
-to drop assumptions involving the individual effects, a number of
-methods are available (e.g., differencing, least squares dummy variable
-regression), but the most common approach is to \textit{demean} the
-equation \citep{Bruederl2015}. Demeaning is the same as the
-transformation applied in Equation \eqref{eq:partialdemean} in the
-special case where \(\theta = 1\). I.e., demeaning involves subtracting
-the per-unit, over-time average from each of the model terms, i.e.,
-\begin{align}
-(y_{it} - \bar{y}_{i}) & = (\bm{x}_{it} - \bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \bar{\bm{z}}_{i})\bm{\gamma} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
-\ddot{y}_{it} & = \bm{\ddot{x}}_{it}\bm{\beta} + \ddot{\varepsilon}_{it}, \ \text{or} \\
-\bm{\ddot{y}}_{i} & = \bm{\ddot{X}}_{i}\bm{\beta} + \bm{\ddot{\varepsilon}}_{i} \label{eq:fe}
-\end{align} where the over-time averages are calculated the same as
-above, and the variables with the dots above them represent the demeaned
-versions. Because the average of something that does not change is that
-thing itself, the individual effects, along with any time-invariant
-predictors, get wiped out by the demeaning. This means that no
-assumptions about the relatedness of the model covariates and the
-unit-specific portion of the error are needed. Consistency of the
-estimates is related solely to the strict exogeneity assumption imposed
-on the idiosyncratic errors, i.e.,
-\(\mathop{\mathrm{\mathbb{E}}}[\ddot{\varepsilon}_{it}|\bm{\ddot{x}}_{it}] = \mathop{\mathrm{\mathbb{E}}}[\ddot{\varepsilon}_{it}] = 0\)
-which also implies
-\(\mathop{\mathrm{\mathbb{E}}}[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T\)
-\citep{Bruederl2015, Wooldridge2002}.
-
-Having demeaned the data, the typical FE estimator is POLS on the
-transformed data \begin{align}
-\bm{\beta}_{FE} & = (\ddot{\bm{X}}^{\intercal}\ddot{\bm{X}})^{-1}\ddot{\bm{X}}^{\intercal}\ddot{\bm{y}}
-\end{align} \citep{Bruederl2015}. The downside to this approach is that
-no time-invariant predictors can be included in the model. However,
-there are alternative approaches in the random effects and mixed model
-frameworks that allow them to be included. These models are sometimes
-referred to as `within-between' or `hybrid' models, often based on the
-Chamberlain \citeyearpar{Chamberlain1980} and Mundlak
-\citeyearpar{Mundlak1978} approaches, see for example \citet{Bell2018};
-\citet{Allison2011}; \citet{Schunck2013}; \citet{Enders2007}. In the
-\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
-appendix}, it will be discussed how to also get around this restriction
-using SEM.
-
-\hypertarget{fe-sem}{%
-\section{Fixed effects in structural equation modeling}\label{fe-sem}}
-
-Moving from the conventional methods outlined above to SEM, we must
-state the FE model in a different way. We turn to latent variables to
-account for time-invariant unobserved heterogeneity. In fact, besides
-accounting for measurement error and the representation of abstract
-hypothetical concepts, unobserved heterogeneity has historically been
-one of the main uses of latent variables in SEM \citep{Skrondal2004}.
-
-\hypertarget{modeling-time-invariant-unobserved-heterogeneity-as-a-latent-variable}{%
-\subsection{Modeling time-invariant unobserved heterogeneity as a latent
-variable}\label{modeling-time-invariant-unobserved-heterogeneity-as-a-latent-variable}}
-
-We first need to convert the data from stacked, long-format vectors of
-length \(NT\) into \(T\) individual vectors of length \(N\). To see why
-this is necessary, consider what effect this has on the vector of
-responses \(y_{it}\). Let us, for a minute ignore any covariates and
-focus just on the dependent variable (a so-called `intercept-only' or
-`null' model) so that we have
-\(y_{it} = \alpha_{i} + \varepsilon_{it}\). When we convert the data to
-wide-format, we get \(T\) individual equations, \begin{align}
-\bm{y}_{t} & = \bm{\alpha} + \bm{\varepsilon}_{t} \\
-\begin{bmatrix}
-y_{1t} \\
-y_{2t} \\
-\vdots \\
-y_{Nt}\end{bmatrix} & = 
-\begin{bmatrix}
-\alpha_{1} \\
-\alpha_{2} \\
-\vdots \\
-\alpha_{N}
-\end{bmatrix} + 
-\begin{bmatrix}
-\varepsilon_{1t} \\
-\varepsilon_{2t} \\
-\vdots \\
-\varepsilon_{Nt}
-\end{bmatrix} \label{eq:wide}
-\end{align} for each \(t = 1, 2, ..., T\). Because the idiosyncratic
-errors are assumed to be uncorrelated across units and across time, the
-covariance between any two of the new wide vectors
-\(\mathop{\mathrm{\mathrm{Cov}}}(y_{ti},y_{si}) = \mathop{\mathrm{\mathrm{Var}}}(\alpha_{i}), \ t \ne s\).
-Otherwise, when \(t = s\), the covariance
-\(\mathop{\mathrm{\mathrm{Cov}}}(y_{ti},y_{ti}) = \mathop{\mathrm{\mathrm{Var}}}(\alpha_{i}) + \mathop{\mathrm{\mathrm{Var}}}(\varepsilon_{ti})\).
-This is the structure we saw above in a typical element of
-\(\bm{\Omega}\).
-
-And in fact this is exactly how a latent variable is used to account for
-time-invariant unobserved heterogeneity. The dependent variable at each
-timepoint is regressed onto the latent variable, see Figure
-\ref{fig:fesem}. Here, the regression weights or `factor loadings' are
-fixed to one to represent our assumption that the effect of the
-time-invariant unobserved heterogeneity is constant over time.\footnote{In
-  fact, the initial FE-SEM setup shown in the main article mimics the
-  POLS methods described above in that it assumes constant effects and
-  error variances over time. These assumptions can be loosened and
-  tested, as will be shown in the
-  \href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{supplementary
-  materials}. For now, for the sake of simplicity and comparability, we
-  retain the assumptions associated with the `pooled' models for the
-  most part.} It also means that the estimated variance of the latent
-variable is equal to the
-\textit{average covariance between the wide-format columns of the dependent variable over time}.
-If \(y_{it} = \alpha_{i} + \varepsilon_{it}\) is the true data
-generating process, then the relationship between two units over time is
-just \(\mathop{\mathrm{\mathrm{Var}}}(\alpha)\), regardless of the time
-distance. Referring back to the random effects structure of
-\(\bm{\Omega}_{i}\) in Equality \eqref{eq:omeganui} for a generic unit
-\(i\), we see the covariance on all of the off-diagonals is
-\(\sigma^{2}_{\alpha}\). And, as we know, the average of something that
-does not change is that thing itself. I.e., if \(T(T-1)/2 = h\) is the
-number of elements on either the upper- or lower triangle of
-\(\bm{\Omega}_{i}\), then we have
-\(h^{-1}\sum_{i=1}^{h}\sigma^{2}_{\alpha} = \frac{h \sigma^{2}_{\alpha}}{h} = \sigma^{2}_{\alpha}\).
-
-To elaborate on this concept some more, consider the following matrix
-equation of the variances and the nonredundant covariances in a
-three-wave intercept-only model that follows directly from Equation
-\eqref{eq:wide} (assuming
-\(\mathop{\mathrm{\mathrm{Cov}}}(\varepsilon_{ti},\varepsilon_{si}) = 0, \ t \ne s\)),
-and which we can solve easily with least squares: \begin{align}
-\bm{A}\bm{x} & = \bm{b} \\
-\begin{pmatrix}
-1 & 1 & 0 & 0 \\
-1 & 0 & 0 & 0 \\
-1 & 0 & 0 & 0 \\
-1 & 0 & 1 & 0 \\
-1 & 0 & 0 & 0 \\
-1 & 0 & 0 & 1
-\end{pmatrix}
-\begin{pmatrix}
-\psi \\
-\phi_{1} \\
-\phi_{2} \\
-\phi_{3}
-\end{pmatrix} & = 
-\begin{pmatrix}
-\mathop{\mathrm{\mathrm{Var}}}(y_{1}) \\
-\mathop{\mathrm{\mathrm{Cov}}}(y_{2},y_{1}) \\
-\mathop{\mathrm{\mathrm{Cov}}}(y_{3},y_{1}) \\
-\mathop{\mathrm{\mathrm{Var}}}(y_{2}) \\
-\mathop{\mathrm{\mathrm{Cov}}}(y_{3},y_{2}) \\
-\mathop{\mathrm{\mathrm{Var}}}(y_{3})
-\end{pmatrix}
-\end{align} where \(\psi = \mathop{\mathrm{\mathrm{Var}}}(\alpha)\),
-\(\phi_{t} = \mathop{\mathrm{\mathrm{Var}}}(\varepsilon_{t})\). We can
-solve this equation to show \begin{align}
-\bm{A}\bm{x} & = \bm{b} \\
-(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{A}\bm{x} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} \\
-\hat{\bm{x}} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} \\
-\begin{pmatrix}
-\hat{\psi} \\
-\hat{\phi}_{1} \\
-\hat{\phi}_{2} \\
-\hat{\phi}_{3}
-\end{pmatrix} & = 
-\begin{pmatrix}
-\frac{1}{3}\mathop{\mathrm{\mathrm{Cov}}}(y_{2},y_{1}) + \frac{1}{3}\mathop{\mathrm{\mathrm{Cov}}}(y_{3},y_{1}) + \frac{1}{3}\mathop{\mathrm{\mathrm{Cov}}}(y_{3},y_{2}) \\
-\mathop{\mathrm{\mathrm{Var}}}(y_{1}) - \hat{\psi} \\
-\mathop{\mathrm{\mathrm{Var}}}(y_{2}) - \hat{\psi} \\
-\mathop{\mathrm{\mathrm{Var}}}(y_{3}) - \hat{\psi}
-\end{pmatrix}.
-\end{align} So if, in fact the covariance between any two wide-format
-columns of \(y\) is
-\(\mathop{\mathrm{\mathrm{Cov}}}(y_{t},y_{s}) = \mathop{\mathrm{\mathrm{Var}}}(\alpha), \ \forall \ s \ne t\),
-then
-\(\hat{\psi} = \frac{1}{3}\mathop{\mathrm{\mathrm{Var}}}(\alpha) + \frac{1}{3}\mathop{\mathrm{\mathrm{Var}}}(\alpha) + \frac{1}{3}\mathop{\mathrm{\mathrm{Var}}}(\alpha) = \frac{3 \mathop{\mathrm{\mathrm{Var}}}(\alpha)}{3} = \mathop{\mathrm{\mathrm{Var}}}(\alpha)\).
-This shows that if our assumption about the underlying DGP is correct,
-i.e., \(y_{it} = \alpha_{i} + \varepsilon_{it}\) and
-\(\mathop{\mathrm{\mathrm{Cov}}}(\varepsilon_{t},\varepsilon_{s}) = 0, \ \forall \ t \ne s\),
-then the estimated variance of \(\alpha\) is just what it should be: the
-average covariance between units of \(y\) over time. Once we add in
-observed covariates, the estimated covariance of \(\alpha\) then become
-the \textit{conditional} covariance of \(y\) over time, given those
-covariates.
-
-\hypertarget{model-notation}{%
-\subsection{Model notation}\label{model-notation}}
-
-Having explained how a latent variable is used to estimate the
-individual effects, we need to state the FE model using SEM-compatible
-matrix notation. There are a number of different model notations (see,
-for example \citet{Bollen1989} for an overview), but the one that will
-serve us best is one that was proposed by \citet{Graff1979}:
-\begin{align}
-\bm{y}^{+} & = \bm{\Lambda_{y}}^{+} \bm{\eta}^{+}, \\
-\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+}, 
-\end{align} where
-\(\bm{\eta}^{+} = (\bm{y}, \bm{x}, \bm{\eta}, \bm{\xi})^{\intercal}\),
-\(\bm{\zeta}^{+} = (\bm{\varepsilon}, \bm{\delta}, \bm{\zeta}, \bm{\xi})^{\intercal}\),
-\(\bm{y}^{+} = (\bm{y}, \bm{x})^{\intercal}\). \(\bm{y}\) is a vector of
-observed dependent variables and \(\bm{x}\) is a vector of observed
-independent variables. \(\bm{\eta}\) is a vector of the latent dependent
-variables and \(\bm{\xi}\) is a vector of latent independent variables.
-\(\bm{\varepsilon}\) and \(\bm{\delta}\) are vectors of the errors of
-the observed dependent and independent variables, respectively, and
-\(\bm{\zeta}\) is a vector of the errors, or disturbances, of the latent
-variables. Notice the \(^{+}\) symbol is just meant to differentiate the
-vectors with them from those without them. That means, \(\bm{\eta}^{+}\)
-is a vector that holds the observed and latent variables, both dependent
-(in SEM they are referred to as `endogenous') and independent (i.e.,
-`exogeneous')\footnote{For our purposes, the terms endogenous and
-  dependent, on the one hand, and exogenous and independent, on the
-  other, can be used interchangeably.}, \(\bm{\zeta}^{+}\) holds the
-errors for the observed variables and the disturbances of the latent
-variables. \(\bm{y}^{+}\) holds just the observed variables, both
-dependent and independent, and \(\bm{\Lambda_{y}}^{+}\) is a matrix of
-ones and zeros that selects the observed variables from
-\(\bm{\eta}^{+}\). Lastly, \(\bm{B}\) is a matrix that holds the
-regression coefficients.
-
-If we say that \(p\) and \(q\) stand for the number of observed
-dependent and independent variables, respectively, and \(m\) and \(n\)
-stand for the number of latent dependent and independent variables,
-respectively, then \(\bm{\eta}^{+}\) and \(\bm{\zeta}^{+}\) are
-\(p + q + m + n\), \(\bm{y}^{+}\) is \(p + q\), \(\bm{\Lambda_{y}}^{+}\)
-is \((p + q) \times (p + q + m + n)\) and \(\bm{B}\) is
-\((p + q + m + n) \times (p + q + m + n)\) \citep{Bollen1989}.
-
-This notation may be confusing at first, but it has advantages. First,
-it allows us the flexibility we need for the models. For example, it
-allows observed \(\bm{x}\) to directly influence observed \(\bm{y}\)
-(more common notation assumes that substantive effects occur only
-between latent variables, observed ones are only used as indicators, see
-for example \citet{Bollen1989}, \citet{Kline2016}). It also allows
-\(\bm{\xi}\), i.e., any latent exogenous variables, to influence
-\(\bm{y}\) directly. These two scenarios cover the traditional FE model
-with observed variables, and one in which latent variables are used to
-account for measurement error in the independent variables. It is also
-consistent with the notation used for these models in \texttt{lavaan}.
-In fact, \texttt{lavaan} switches automatically between matrix notations
-depending on the specified model. That means the matrix representation
-of the model one sees if they type in
-\texttt{lavInspect(model,\ what\ =\ "est")} after specifying their model
-in \texttt{lavaan} will match up with the notation used here.\footnote{In
-  \texttt{Mplus}, the model matrices can be requested by including
-  \texttt{OUTPUT:\ TECH1} in the input file.} It does have a potential
-disadvantage however. Besides being less intuitive than the typical
-\(\bm{\eta} = \bm{B}\bm{\eta} + \bm{\Gamma}\bm{\xi} + \bm{\zeta}\)
-notation, it means that by including the observed covariates in the
-stacked long vector \(\bm{y}^{+}\), they are treated as another response
-variable with variances and covariances to be estimated by the model
-(instead of just using the sample statistics). This means that the
-assumption of multivariate normality (otherwise just imposed on the
-dependent variables) also applies to the independent ones
-\citep[p.~75]{Skrondal2004}. This can be problematic for noncontinuous
-independent variables like sex, nationality dummies, marriage status
-(married/unmarried), etc. See \citet{Skrondal2004} for more on this
-topic.
-
-\begin{figure}
-\begin{center}
-\resizebox{0.75\textwidth}{!}{%
 \begin{tikzpicture}
@@
 \end{tikzpicture}
-}
-\caption{Typical three-wave FE-SEM model with contemporary effects\label{fig:fesem}}
-\end{center}
-\end{figure}
-
-Let us, however, make things more concrete and take a look at a simple,
-three-wave version of the typical FE-SEM using this notation (shown
-graphically in Figure \ref{fig:fesem}). For that, we have the following
-matrix notation (with labels on the outside of the matrices):
-\begin{align}
-\begin{split}
-\bm{y}^{+} & = \bm{\Lambda_{y}}^{+}\bm{\eta}^{+} \\
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3}
-\end{pmatrix} & = 
-\begin{blockarray}{cccccccc}
- & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
- \begin{block}{c(ccccccc)}
- y_{1} & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
- y_{2} & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
- y_{3} & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\ 
- x_{1} & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
- x_{2} & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
- x_{3} & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
- \end{block}
-\end{blockarray}
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix},
-\end{split} \label{eq:y}
-\end{align} \begin{align}
-\begin{split}
-\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+} \\
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix} & = 
-\begin{blockarray}{cccccccc}
- & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
- \begin{block}{c(ccccccc)}
- y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
- y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
- y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
- x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- \end{block}
-\end{blockarray}
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix} + 
-\begin{pmatrix}
-\varepsilon_{1} \\
-\varepsilon_{2} \\
-\varepsilon_{3} \\
-x_{1} = \delta_{1} \\
-x_{2} = \delta_{2} \\
-x_{3} = \delta_{3} \\
-\alpha = \xi \\
-\end{pmatrix}. 
-\end{split} \label{eq:eta}
-\end{align} Notice in \(\bm{\zeta}^{+}\), for the independent variables
-we could either write, for example \(x_{t}\) or \(\delta_{t}\). As
-mentioned above, this is due to the model notation treating the
-independent variables like dependent variables with
-variances/covariances to be estimated. For the sake of simplicity, we
-will ignore this subtlety and refer to the observed variable from now
-on, keeping in mind that if the multivariate normality assumption holds,
-the estimated statistics will likely be sufficiently close to the sample
-ones for it to not make much of a difference.
-
-Admittedly, Equations \eqref{eq:y} and \eqref{eq:eta} may not look like
-much yet. We can remedy this by first putting the equation for
-\(\bm{\eta}^{+}\) in reduced form, i.e., by getting rid of the dependent
-variable on the r.h.s.: \begin{align}
-\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+} \\
-\bm{\eta}^{+} - \bm{B}\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
-(\bm{I} - \bm{B})\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
-\bm{\eta}^{+} & = (\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+},
-\end{align} where \(\bm{I}\) is the identity matrix. By substituting
-this back into the equation for the observed variables we get
-\(\bm{y}^{+} = \bm{\Lambda_{y}}^{+}[(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+}]\),
-which works out to: \begin{align}
-\begin{pmatrix} 
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\end{pmatrix} & = 
-\begin{pmatrix}
-\alpha + \beta x_{1} + \varepsilon_{1} \\
-\alpha + \beta x_{2} + \varepsilon_{2} \\
-\alpha + \beta x_{3} + \varepsilon_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3}
-\end{pmatrix},
-\end{align} which is of course exactly what we should expect given
-Equation \eqref{eq:fe}\footnote{We can use the \texttt{sympy} package in
-  \texttt{python} to verify and show the steps for this and other
-  examples, see the
-  \href{https://github.com/henrik-andersen/FE-SEM/blob/master/sympy-doublecheck-matrixnotation.py}{supplementary
-  materials} for the code.}.
-
-\hypertarget{assumptions}{%
-\subsection{Assumptions}\label{assumptions}}
-
-What essentially differentiates an FE from an RE model is our assumption
-concerning the relationship between the unobserved individual effects
-and the model covariates \citep{Bollen2010}. The FE model assumes that
-\(\mathop{\mathrm{\mathbb{E}}}[\alpha x_{t}] \ne 0\). As such, if we
-fail to control for the correlation of the covariate and the
-time-invariant part of the error, then the coefficient of interest, here
-\(\beta\), will be biased. Our assumption regarding whether the
-individual effects are correlated with the model covariates occurs in
-\(\mathop{\mathrm{\mathbb{E}}}[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] = \bm{\Psi}\),
-the covariance matrix of the errors \begin{align}
-\bm{y}^{+}\bm{y}^{+ \intercal} & = \mathop{\mathrm{\mathbb{E}}}[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})^{\intercal}] \\
- & = \mathop{\mathrm{\mathbb{E}}}[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\zeta}^{+ \intercal}(\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal})] \\
- & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \mathop{\mathrm{\mathbb{E}}}[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal} \\
- & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \bm{\Psi} (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal}.
-\end{align} In the case of an FE model, \(\bm{\Psi}\) will reflect our
-belief that the individual effects are correlated with the model
-covariates, here again for demonstration the three-wave model:
-\begin{align}
-\bm{\Psi} & = \mathop{\mathrm{\mathbb{E}}}
-\begin{blockarray}{cccccccc}
- & \varepsilon_{1} & \varepsilon_{2} & \varepsilon_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
- \begin{block}{c(ccccccc)}
- \varepsilon_{1} & \varepsilon_{1}^{2} &                     &                     &              &              &              & \\
- \varepsilon_{2} & 0                   & \varepsilon_{2}^{2} &                     &              &              &              & \\
- \varepsilon_{3} & 0                   & 0                   & \varepsilon_{3}^{2} &              &              &              & \\
- x_{1}           & 0                   & 0                   & 0                   & x_{1}^{2}    &              &              & \\
- x_{2}           & 0                   & 0                   & 0                   & x_{2}x_{1}   & x_{2}^{2}    &              & \\
- x_{3}           & 0                   & 0                   & 0                   & x_{3}x_{1}   & x_{3}x_{2}   & x_{3}^{2}    & \\
- \alpha          & 0                   & 0                   & 0                   & \alpha x_{1} & \alpha x_{2} & \alpha x_{3} & \alpha^{2} \\
- \end{block}
-\end{blockarray}.
-\end{align} Knowing this, we can work out the equation for the
-coefficient of interest, \(\beta\). For the sake of simplicity, assume
-here and throughout mean-centered variables: \begin{align}
-\mathop{\mathrm{\mathrm{Cov}}}(y_{t},x_{t}) & = \mathop{\mathrm{\mathbb{E}}}[y_{t}x_{t}] \\
- & = \mathop{\mathrm{\mathbb{E}}}[(\alpha + \beta x_{t} + \varepsilon_{t})x_{t}] \\
- & = \mathop{\mathrm{\mathbb{E}}}[\alpha x_{t} + \beta x_{t}^{2} + \varepsilon_{t}x_{t}] \\
- & = \mathop{\mathrm{\mathrm{Cov}}}(\alpha, x_{t}) + \beta \mathop{\mathrm{\mathrm{Var}}}(x_{t}) \\
-\hat{\beta}_{FE-SEM} & = \frac{\mathop{\mathrm{\mathrm{Cov}}}(y_{t},x_{t}) - \mathop{\mathrm{\mathrm{Cov}}}(\alpha, x_{t})}{\mathop{\mathrm{\mathrm{Var}}}(x_{t})}. 
-\end{align} This should make intuitive sense. From the observed
-covariance between the dependent and the independent variable, we are
-partialling out the part that is due to the covariance between the
-independent variable and the individual effects per unit, and then
-dividing by the variance of the independent variable, as usual. For the
-RE model, we assume \(\mathop{\mathrm{\mathbb{E}}}[\alpha x_{t}] = 0\)
-and the equation reduces to
-\(\hat{\beta}_{RE-SEM} = \mathop{\mathrm{\mathrm{Cov}}}(y_{t},x_{t})/\mathop{\mathrm{\mathrm{Var}}}(x_{t})\).
-The rest of the model-implied covariance matrix results from
-\(\bm{y}^{+}\bm{y}^{+ \intercal}\).
-
-\hypertarget{fesem}{%
-\section{Fixed effects in lavaan}\label{fesem}}
-
-The package \texttt{lavaan} needs to be installed once with
-\texttt{install.packages("lavaan")}. To be able to use it, we need to
-load it for every new \texttt{R} session:
-
-\singlespacing
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{( lavaan)}
-\end{Highlighting}
-\end{Shaded}
-
-\doublespacing
-
-For users unfamiliar with \texttt{R}, SEM analyses can be carried out
-with almost no knowledge of the language. Typically, someone unfamiliar
-with \texttt{R} would prepare their data using some other statistical
-software, and then save the intended dataset as a \texttt{.csv},
-\texttt{.xlsx}, \texttt{.dta}, \texttt{.sav}, etc. file. The user must
-then import the data, preferably as a dataframe, and the rest occurs
-using the \texttt{lavaan} syntax.\footnote{There are many online
-  tutorials for importing data in various formats, see, for example some
-  from
-  \href{https://www.datacamp.com/community/tutorials/r-data-import-tutorial}{datacamp}
-  or
-  \href{https://www.statmethods.net/input/importingdata.html}{Quick-R},
-  or any of the many posts on
-  \href{https://stackoverflow.com/search?q=r+import+data}{stackoverflow}.}
-
-Specifying the most basic fixed effects model, like the one shown in
-\citet{Bollen2010} (the same model as Equation \eqref{eq:fe} but with
-just one time-varying predictor) involves four components. First, we
-define the latent individual effects variable using the
-\texttt{=\textasciitilde{}} `measured by' or `manifested by'
-\citep{R-lavaan} operator at the same time constraining the factor
-loadings at each timepoint to one. I will call the latent variable
-\texttt{a} to stand for \(\alpha\). Constraining all of the factor
-loadings to one reflects our implicit assumption that the combined
-effect of the unit-specific unobserved factors is constant over time.
-This is the default behaviour of traditional POLS-based approaches to FE
-that use the stacked long-format data.
-
-\singlespacing
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{a =}\ErrorTok{~}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y1 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y2 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y3 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y4 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y5}
-\end{Highlighting}
-\end{Shaded}
-
-\doublespacing
-
-Second, we regress the dependent variable on the independent variable
-using the \texttt{\textasciitilde{}} regression operator. With stacked,
-long-format data, only one regression coefficient is estimated over all
-observed timepoints. To have our FE-SEM model mimic this behaviour, we
-need to constrain the the estimated coefficient to equal over time. We
-do so by adding the same label to the regression coefficient at every
-time point. We will use the label \texttt{b} (this label was chosen
-arbitrarily, we could have used any letter or string of characters) and
-have it act as an equality constraint for the regression coefficient of
-interest \(\beta\):
-
-\singlespacing
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{y1 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x1}
-\NormalTok{y2 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x2 }
-\NormalTok{y3 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x3}
-\NormalTok{y4 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x4}
-\NormalTok{y5 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x5}
-\end{Highlighting}
-\end{Shaded}
-
-\doublespacing
-
-The key to a FE model, as opposed to an RE model are our assumptions
-about the relatedness of our covariate and the individual effects, i.e.,
-\(\mathop{\mathrm{\mathbb{E}}}[x_{t}\alpha]\). For an FE model, we want
-to partial out any potential covariance between the independent variable
-and the individual effects. This accounts for any linear relationship
-between \(x_{t}\) and the unit-specific characteristics influencing the
-dependent variable. Further, allowing unrestricted covariances between
-the independent variable itself over time will not affect how the
-coefficient \(\beta\) is estimated, but will have an effect on the
-standard errors. To mimic the behaviour of a conventional FE model, we
-allow the independent variable to be correlated with the individual
-effects and itself over time. Covariances (including covariances between
-a variable and itself, i.e., variances) are specified using the
-\texttt{\textasciitilde{}\textasciitilde{}} operator:
-
-\singlespacing
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{a }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x1 }\OperatorTok{+}\StringTok{ }\NormalTok{x2 }\OperatorTok{+}\StringTok{ }\NormalTok{x3 }\OperatorTok{+}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
-\NormalTok{x1 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x2 }\OperatorTok{+}\StringTok{ }\NormalTok{x3 }\OperatorTok{+}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
-\NormalTok{x2 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x3 }\OperatorTok{+}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
-\NormalTok{x3 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
-\NormalTok{x4 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x5}
-\end{Highlighting}
-\end{Shaded}
-
-\doublespacing
-
-The last component of our code involves the variances of the residuals.
-This component is optional, but we can constrain the residual variances
-to be equal over time to again mimic the behaviour of a conventional FE
-model using POLS on stacked data. Here, again, we use labels to make
-equality constraints. Because \(y_{t}\) is endogenous, the
-\texttt{\textasciitilde{}\textasciitilde{}} operator specifies the
-variances of \emph{residuals}, i.e., \(\varepsilon_{t}\).
-
-\singlespacing
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{y1 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y1}
-\NormalTok{y2 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y2}
-\NormalTok{y3 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y3}
-\NormalTok{y4 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y4}
-\NormalTok{y5 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y5}
-\end{Highlighting}
-\end{Shaded}
-
-\doublespacing
-
-\hypertarget{ex1}{%
-\section{A simulated example}\label{ex1}}
-
-\singlespacing
-
-\doublespacing
-
-To demonstrate the application of FE models in SEM, a dataset can be
-simulated that embodies the FE assumptions. Again, the code for data
-simulation can be found in the
-\href{https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R}{online
-supplementary materials}.
-
-To show that the latent individual effects variables represent the
-\emph{combined} effect of all time-invariant characteristics, the
-dependent variable will be influenced by two separate unit-specific
-variables, which we can call \(\alpha_{1}\) and \(\alpha_{2}\). We will
-construct the simulated data such that the independent variable is
-correlated with both of the time-invariant variables. This means that
-approaches that fail to account for this confounding influence, such as
-POLS or RE, will be biased.
-
-The wide-format equations for the data generating process can be
-described as: \begin{align}
-\bm{x_{t}} & = \bm{\alpha_{1}}\beta_{x_{t},\alpha_{1}} + \bm{\alpha_{2}}\beta_{x_{t},\alpha_{2}} + \bm{\delta_{t}}, \\
-\bm{y_{t}} & = \bm{x_{t}}\beta_{y_{t},x_{t}} + \bm{\alpha_{1}}\beta_{y_{t},\alpha_{1}} + \bm{\alpha_{2}} \beta_{y_{t},\alpha_{2}} + \bm{\varepsilon_{t}} 
-\end{align} where, for the sake of simplicity, \(\bm{\alpha_{1}}\),
-\(\bm{\alpha_{2}}\), \(\bm{\delta_{t}}\) and \(\bm{\varepsilon_{t}}\)
-are \(\sim N(0,1)\).
-
-For the following example, a sample size of 1,000, observed over five
-waves, was chosen. The unique variance of \(\bm{x}\), as well as both
-the individual-effect variables is also \(\sim N(0,1)\). The coefficient
-of interest, \(\beta_{y,x}\) is set to be equal to \(0.3\). A
-correlation between \(\bm{x}\) and the individual effects is induced
-through \(\beta_{x,\alpha_{1}} = 0.85\) and
-\(\beta_{x,\alpha_{1}} = 0.50\). With the variances above set to one,
-the covariances will be roughly
-\(\mathop{\mathrm{\mathrm{Cov}}}(x_{t},\alpha_{1}) = 0.85\) and
-\(\mathop{\mathrm{\mathrm{Cov}}}(x_{t},\alpha_{2}) = 0.5\). The
-dependent variable is also influenced by the individual effects
-variables with \(\beta_{y_{t},\alpha_{1}} = 0.75\) and
-\(\beta_{y_{t},\alpha_{2}} = 0.45\) These values were chosen
-arbitrarily.
-
-Now, we run the FE-SEM in \texttt{lavaan}.
-
-\singlespacing
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{fe_sem <-}\StringTok{ '}
-\StringTok{# Define individual effects variable }
-\StringTok{a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5}
-\StringTok{# Regressions, constrain coefficient to be equal over time}
-\StringTok{y1 ~ b*x1}
-\StringTok{y2 ~ b*x2 }
-\StringTok{y3 ~ b*x3}
-\StringTok{y4 ~ b*x4}
-\StringTok{y5 ~ b*x5}
-\StringTok{# Allow unrestricted correlation between eta and covariates}
-\StringTok{a ~~ x1 + x2 + x3 + x4 + x5}
-\StringTok{x1 ~~ x2 + x3 + x4 + x5}
-\StringTok{x2 ~~ x3 + x4 + x5}
-\StringTok{x3 ~~ x4 + x5}
-\StringTok{x4 ~~ x5}
-\StringTok{# Constrain residual variances to be equal over time}
-\StringTok{y1 ~~ e*y1}
-\StringTok{y2 ~~ e*y2}
-\StringTok{y3 ~~ e*y3}
-\StringTok{y4 ~~ e*y4}
-\StringTok{y5 ~~ e*y5}
-\StringTok{'}
-\NormalTok{fe_sem.fit <-}\StringTok{ }\KeywordTok{sem}\NormalTok{(}\DataTypeTok{model =}\NormalTok{ fe_sem, }
-                  \DataTypeTok{data =}\NormalTok{ dfw, }
-                  \DataTypeTok{estimator =} \StringTok{"ML"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\doublespacing
-
-\singlespacing
-
-\doublespacing
-
-We can get a summary of the model with \texttt{summary()}. The first
-portion of the summary output gives an overview of some basic
-information and fit statistics. The maximum likelihood estimator is the
-default, so it did not have to be explicitly selected in the fitting
-function call. Other estimators are available, including generalized and
-unweighted least squares (\texttt{GLS} and \texttt{ULS}, respectively),
-robust standard errors maximum likelihood (\texttt{MLM}) and several
-others (see \href{https://lavaan.ugent.be/tutorial/est.html}{the lavaan
-online tutorial for more}).
-
-This part of the summary output also tells us that the analysis is based
-on 1,000 observations (missings would be shown here as well if there
-were any), and that the \(\chi^{2}\) statistic is 30.138 based on 32
-degrees of freedom (55 observed covariances minus 1 error variance, 1
-coefficient, 1 latent variable variance, 5 exogenous variable variances
-and 15 covariances for \(55 - 23 = 32\) df). The p-value on the
-\(\chi^{2}\) statistic is not significant with \(p =\) 0.561 which tells
-us the differences between the model-implied and observed covariance
-matrices are likely due to chance, and that the model fits the data well
-(given how the data was generated, it would be surprising if this were
-not the case). Other fit measures including typical comparative fit
-indices can be requested by either adding \texttt{fit.measures\ =\ TRUE}
-as a secondary argument to the \texttt{summary()} call, or by asking for
-a complete list of all available fit statistics using
-\texttt{lavInspect(model,\ "fit")} where \texttt{model} stands for the
-name of the fitted model, in this case \texttt{fe\_sem.fit}.
-
-\singlespacing
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{summary}\NormalTok{(fe_sem.fit)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## lavaan 0.6-6 ended normally after 37 iterations
-## 
-##   Estimator                                         ML
-##   Optimization method                           NLMINB
-##   Number of free parameters                         31
-##   Number of equality constraints                     8
-##                                                       
-##   Number of observations                          1000
-##                                                       
-## Model Test User Model:
-##                                                       
-##   Test statistic                                30.138
-##   Degrees of freedom                                32
-##   P-value (Chi-square)                           0.561
-## 
-## Parameter Estimates:
-## 
-##   Standard errors                             Standard
-##   Information                                 Expected
-##   Information saturated (h1) model          Structured
-...
-\end{verbatim}
-
-\doublespacing
-
-Next the summary output shows the measurement models for the latent
-variables, if any. In this case the latent variable \texttt{a} for
-\(\alpha\) is measured by each of the five observed dependent variables
-with factor loadings fixed to 1.0.
-
-\singlespacing
-
-\begin{verbatim}
-...
-## Latent Variables:
-##                    Estimate  Std.Err  z-value  P(>|z|)
-##   a =~                                                
-##     y1                1.000                           
-##     y2                1.000                           
-##     y3                1.000                           
-##     y4                1.000                           
-##     y5                1.000                           
-...
-\end{verbatim}
-
-\doublespacing
-
-The regressions are shown next. Here, because we have constrained the
-regression coefficients to be equal over time (the equality constraint
-label \texttt{(b)} is listed to the left of the estimates), the estimate
-of \(\beta =\) 0.294 (0.016) is repeated five times. The corresponding
-z- and p-values show that the coefficient is, unsurprisingly,
-significant.
-
-\singlespacing
-
-\begin{verbatim}
-...
-## Regressions:
-##                    Estimate  Std.Err  z-value  P(>|z|)
-##   y1 ~                                                
-##     x1         (b)    0.294    0.016   18.809    0.000
-##   y2 ~                                                
-##     x2         (b)    0.294    0.016   18.809    0.000
-##   y3 ~                                                
-##     x3         (b)    0.294    0.016   18.809    0.000
-##   y4 ~                                                
-##     x4         (b)    0.294    0.016   18.809    0.000
-##   y5 ~                                                
-##     x5         (b)    0.294    0.016   18.809    0.000
-...
-\end{verbatim}
-
-\doublespacing
-
-Next, the covariance estimates are listed. First, the covariances
-between the latent individual effects variable and the independent
-variable over time are shown, and then the covariances between the
-independent variable with itself over time.
-
-One should always take care to double-check that there are no unintended
-covariances listed here. Like \texttt{Mplus}, the \texttt{lavaan}
-package estimates some covariances per default, without the user
-explicitly having to add them to the model syntax. For example,
-covariances between latent variables are estimated per default. If one
-does not wish for them to covary, it must be explicitly stated, e.g.,
-with \texttt{f1\ \textasciitilde{}\textasciitilde{}\ 0*f2}, assuming the
-latent variables are called \texttt{f1} and \texttt{f2}, or by
-overriding the default behaviour for the entire model by adding
-\texttt{orthogonal\ =\ TRUE} (which sets the correlation between all
-latent variables to zero) to the fitting call.\footnote{This is at least
-  the current behaviour of both the \texttt{cfa} and \texttt{sem}
-  wrappers. In fact, both wrappers seem to be identical in terms of the
-  default settings, see \citet{Rosseel2020}.}
-
-\singlespacing
-
-\begin{verbatim}
-...
-## Covariances:
-##                    Estimate  Std.Err  z-value  P(>|z|)
-##   a ~~                                                
-##     x1                0.844    0.055   15.355    0.000
-##     x2                0.867    0.056   15.441    0.000
-##     x3                0.845    0.055   15.400    0.000
-##     x4                0.822    0.053   15.455    0.000
-##     x5                0.820    0.053   15.572    0.000
-##   x1 ~~                                               
-##     x2                0.908    0.070   12.900    0.000
-##     x3                0.935    0.069   13.466    0.000
-##     x4                0.921    0.067   13.661    0.000
-##     x5                0.914    0.067   13.716    0.000
-##   x2 ~~                                               
-##     x3                0.889    0.070   12.675    0.000
-##     x4                0.922    0.069   13.423    0.000
-##     x5                0.889    0.068   13.165    0.000
-##   x3 ~~                                               
-##     x4                0.865    0.067   12.976    0.000
-##     x5                0.901    0.066   13.554    0.000
-##   x4 ~~                                               
-##     x5                0.850    0.064   13.285    0.000
-...
-\end{verbatim}
-
-\doublespacing
-
-Finally, the variance estimates are listed. Here, we see that in order
-to mimic the behaviour of a traditional FE model, the error variances
-over time were specified to be equal using the equality constraint
-\texttt{(e)}. Notice the \texttt{.} beside \texttt{y1}, \texttt{y2},
-etc.: this indicates that the listed variance refers to an endogenous
-variable, and that it is thus an error variance. In this case, these
-refer to the variances of \(\varepsilon_{t}\). After that, the variances
-of the exogenous variables, both observed and unobserved are listed.
-
-\singlespacing
-
-\begin{verbatim}
-...
-## Variances:
-##                    Estimate  Std.Err  z-value  P(>|z|)
-##    .y1         (e)    1.022    0.023   44.721    0.000
-##    .y2         (e)    1.022    0.023   44.721    0.000
-##    .y3         (e)    1.022    0.023   44.721    0.000
-##    .y4         (e)    1.022    0.023   44.721    0.000
-##    .y5         (e)    1.022    0.023   44.721    0.000
-##     x1                1.986    0.089   22.361    0.000
-##     x2                2.079    0.093   22.361    0.000
-##     x3                1.987    0.089   22.361    0.000
-##     x4                1.860    0.083   22.361    0.000
-##     x5                1.814    0.081   22.361    0.000
-##     a                 0.799    0.052   15.310    0.000
-\end{verbatim}
-
-\doublespacing
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-Fixed effects regression in SEM has been outlined in well-known articles
-by \citep{Allison2011, Bollen2010, Teachman2001}. This article provides
-a focused look at the implementation of the basic model using the
-\texttt{lavaan} package in \texttt{R}. The
-\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
-supplementary materials} further discuss common extensions and some
-tools for evaluating and loosening model assumptions.
-
-The benefits of FE-SEM as opposed to traditional OLS-based FE-models are
-largely the same ones that apply to the SEM framework in general: for
-one, SEM allows for a great deal of flexibility. For example, it is easy
-to loosen model constraints as necessary. Measurement error in both the
-dependent and independent variables can be dealt with using latent
-variables to achieve unbiased and more efficient results. Researchers
-interested in time-invariant predictors can integrate them into a hybrid
-FE/RE model with ease. Further extensions, like measurement invariance
-testing \citep{Schoot2012, Millsap2011, Steenkamp1998} as well as lagged
-dependent variables \citep{Bollen2010, Allison2017} for example, can
-also be implemented in a straightforward fashion.
-
-The most basic FE-SEM is furthermore the basis for a variety of
-currently popular extended models, such as Latent Curve Models in
-general \citep{Curran2001, Bollen2004}, as well as special
-implementations like the Dynamic Panel Model \citep{Allison2017}, the
-Random-Intercept Cross-Lagged Panel Model \citep{Hamaker2015} and the
-Latent Curve Model with Structured Residuals \citep{Curran2014}. For
-this reason it is all the more important for researchers to have a good
-grasp on the method of applying panel regression in SEM, and
-understanding the intuition of controlling for time-invariant
-confounders. This article is meant to serve as a consolidated resource
-for researchers looking for concrete advise on specifying FE and more
-general panel models in SEM.
-
-\bibliographystyle{tfcad}
-\bibliography{references2.bib}
 

### 2020-07-28 article_format/title.Rmd
Changed the formating, made some minor edits

--- 2020-07-28 article_format/article_format.Rmd
@@
 ---
-
-```{r setup, include=FALSE}
-# Knitr options 
-knitr::opts_chunk$set(echo = TRUE)
-
-# Packages (some are repeated below, add them here as well just to be sure)
-library(formatR)
-library(knitr)
-library(lavaan)
-
-# Script hooks for chunk outputs
-source("../scripthooks.R")
-```
-
-\newpage
-
-\pagenumbering{arabic} 
-\setcounter{page}{1}
-
-\doublespacing
-
-# Introduction {#intro}
-
-Several years ago, @Curran2011 reflected positively on the growing use of panel studies in empirical social research.
-Some of the strengths of panel data are well-known, e.g., the ability to establish temporal precedence, increased statistical power and the reduction of potential alternative models.
-However, perhaps the greatest strength of panel data is that they allow for a more rigorous testing of substantive theories.
-Panel data, i.e., repeated measures of the same observed units (people, schools, firms, countries, etc.), allow researchers to decompose the error term into a part that stays constant within units and the part that changes over time.
-The part that does not change over time can be seen as the combined effect of all time-invariant influences (e.g., sex, date of birth, nationality) on the dependent variable.
-Fixed effects (FE) regression involves controlling for these time-invariant influences via a number of various methods.
-It thus accounts for a likely and common source of bias.
-
-Structural equation modeling (SEM) is a popular regression framework.
-One of its main strengths is its flexibility.
-Not only can complex causal structures with multiple dependent variables be tested simultaneously, but in longitudinal (and, more generally, hierarchical) studies both time-varying and invariant predictors can be included, and effects can easily be allowed to vary over time.
-Thus researchers can allow for and study effects that increase or fade over time, or that appear only in specific periods.
-Beyond that, with the use of latent variables, SEM provides a way to deal with measurement error and get closer to the true underlying constructs of interest.
-
-There are a number of articles describing basic concept of panel model regression, and FE regression in SEM [e.g., @Allison2011; @Bollen2010; @Teachman2001].
-This article is intended as a \textit{practical guide} for researchers looking for in-depth help with specifying FE models in SEM.
-It focuses on the `lavaan` [@R-lavaan] package for `R` [@R-base].
-While `Mplus` [@Mplus] is arguably the most robust SEM software currently available (in terms of features like alignment, latent variable interactions, for example), the `lavaan` package has many benefits.
-First, like `R` it is open source and completely free.
-For researchers dipping their toes into SEM, there is no financial barrier to try, and no risk if they decide it is not for them.
-Second, the implementation of `lavaan` in the larger `R` environment is an enormous advantage.
-Instead of poring over reams of plain text, copying out coefficients by hand, every part of the `lavaan` output is available as an object.
-This means that all aspects of the model, from fit indices, to coefficients and standard errors, to the model matrices, can be accessed and easily integrated into tables and plots.
-Furthermore, `R` can be used for a great deal of applications.
-It can be used to manage and manipulate as well as simulate data, perform symbolic algebra, run more traditional analyses (e.g., multiple regression, logistic regression, principal component analysis), etc.
-Once one is comfortable using `R`, there is no longer any need to switch between different software for data preparation and analysis.
-
-The following article outlines the basic idea of panel regression, the particularities of panel regression SEM, and shows its implementation in `lavaan`.
-The focus on FE panel regression stems from the fact that the assumptions needed to justify other common models, e.g., random effects, are often implausible.
-As we will see, the practical differences between the two are, in any event, very small.
-Using [simulated data](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R), it demonstrates and annotates the code for the most basic FE model and provides an overview of the summary output.
-One of the main strengths of panel SEM compared to the more traditional methods of panel analysis is its flexibility.
-Therefore, a number of potential extensions to the basic model, including relaxing various assumptions, dealing with measurement error in both the independent and dependent variables, as well as the inclusion of time-invariant predictors in the form of a hybrid fixed-/ random effects model, are shown in detail in the form of [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf).
-
-# Panel models {#panel}
-
-To begin, let us start by reviewing a general panel model [@Bollen2010], also referred to as the 'unobserved effects model' [@Wooldridge2012; @R-plm_a] (we will return to this model in the [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf) when we discuss loosening assumptions)
-\begin{align}
-y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \alpha_{i} + \varepsilon_{it} \label{eq:gpm}
-\end{align}
-where $y_{it}$ is the dependent variable for unit $i, \ i = 1, ..., N$ at time $t, \ t = 1, ..., T$, $\bm{x}_{it}$ is a $1 \times K$ vector of time-varying covariates (which could include a constant) linked to the dependent variable by the $K \times 1$ vector of coefficients $\bm{\beta}$.
-$\bm{z}_{i}$ is a $1 \times M$ vector of time-invariant covariates linked to the dependent variable by the $M \times 1$ vector of coefficients in $\bm{\gamma}$, $\alpha_{i}$ represents the combined effect of all unobserved time-constant variables affecting the dependent variable and $\varepsilon_{it}$ is the idiosyncratic error.
-
-We can make stating some of the model assumptions easier by rewriting it in matrix notation
-\begin{align}
-\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\iota}_{T}\alpha_{i} + \bm{\varepsilon}_{i}
-\end{align}
-where $\bm{y}_{i}$ and $\bm{\varepsilon}_{i}$ are $T \times 1$ vectors, $\bm{X}_{i}$ and $\bm{Z}_{i}$ are $T \times K$ and $T \times M$ matrices, respectively, $\bm{\iota}_{T}$ is a $T \times 1$ vector of ones and $\alpha_{i}$ is a scalar.
-$\bm{\beta}$ and $\bm{\gamma}$ are unchanged from Equation \eqref{eq:gpm}.
-
-Consistency of the following models requires the assumption of strict exogeneity, although what constitutes strict exogeneity differs between the random and fixed effects setups.
-Each assumption will be discussed shortly.
-Apart from that, we typically make the following assumptions about this model [see, e.g., @Wooldridge2002; @Schmidheiny2019]:
-
-- Linearity: the model is linear in its parameters. 
-- Independence: the observations are independent across individuals (assured by random sampling in the cross-section), but not necessarily across time. 
-- The usual rank condition: we have more observations than independent variables and there is no perfect collinearity between any of the independent variables.
-
-## Random effects {#re}
-
-Somewhat unintuitively, the modern approach to FE regression treats the unobserved effect $\alpha_{i}$ as a \textit{random} variable, rather than a fixed parameter to be estimated for each of the cross sectional units [@Wooldridge2002].
-And in fact, this view carries over to the way in which FE regression is applied in SEM.
-So, in order to help facilitate the discussion on the application of FE-SEM later on, it makes sense to spend some time discussing the basic idea of random effects regression (again, as we will see, the practical difference between the two models is very small).
-
-For the random effects (RE) model, we define a composite error term: $\nu_{it} = \alpha_{i} + \varepsilon_{it}$ and rewrite the model in Equation \eqref{eq:gpm} as 
-\begin{align}
-y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \nu_{it}, \ \text{or} \\
-\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\nu}_{i},
-\end{align}
-where $\bm{\nu}_{i} = \alpha_{i} \bm{\iota}_{T} + \bm{\varepsilon}_{i}$ and $\bm{\iota}_{T}$ is a $T \times 1$ vector of ones. The strict exogeneity assumption in the RE model implies
-\begin{align}
-\E[\varepsilon_{it} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = 0, \\
-\E[\alpha_{i} |\bm{X}_{i}, \bm{z}_{i}] = \E[\alpha_{i}] & = 0,
-\end{align}
-where $\bm{X}_{i} = \bm{x}_{i1}, ..., \bm{x}_{iT}$.
-For both parts, the assumption that the unconditional expectations are 0 is unproblematic as long as a constant is included in the regression.
-The first part says the idiosyncratic errors at each timepoint are assumed to be independent of the explanatory variables at \textit{all} timepoints which is stronger than just assuming that they are \textit{contemporaneously} independent.
-This implies that they are also uncorrelated, i.e., $\E[\bm{x}_{is}^{\intercal}\varepsilon_{it}] = \bm{0}$ and $\E[\bm{z}_{i}^{\intercal}\varepsilon_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T$ [@Wooldridge2002; @Bruederl2015].
-We assume the idiosyncratic errors are further independent of the individual effects, which implies $\E[\alpha_{i}\varepsilon_{it}] = 0$.
-
-The second part is the potentially controversial assumption: it states that the individual effects are uncorrelated with the independent variables.
-We can use an intuitive concrete example to show why this is often controversial: If we are interested in the question of whether married men earn more than unmarried men, then the second part of the strict exogeneity assumption means that a man's marriage status would have to be uncorrelated with all the time-invariant characteristics that could potentially make that man an attractive marriage candidate in the first place; e.g., looks, personality, family's status, profession, etc.
-[@Bruederl2015].^[Assuming, for the sake of argument, that these characteristics are constant over time.]
-
-From what we have discussed so far, the $T \times T$ covariance matrix of the errors $\bm{\Omega}_{i} = \E[\bm{\nu}_{i}\bm{\nu}_{i}^{\intercal}]$ can be constructed.
-However, the standard random effects model adds the additional assumptions
-\begin{alignat}{3}
-\E[\varepsilon_{it}^{2} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \E[\varepsilon_{it}^{2}] && = \sigma_{\varepsilon}^{2}, && \ t = 1, ..., T, \\
-\E[\varepsilon_{it}\varepsilon_{is}| \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \E[\varepsilon_{it}\varepsilon_{is}] && = 0, && \ \forall \ t \ne s 
-\end{alignat}
-i.e., the idiosyncratic errors are conditionally homoscedastic and serially uncorrelated and 
-\begin{alignat}{2}
-\E[\alpha_{i}^{2}| \bm{X}_{i},\bm{z}_{i}] & = \E[\alpha_{i}^{2}] && = \sigma^{2}_{\alpha}
-\end{alignat}
-i.e., the individual effects are conditionally homoscedastic (they are necessarily serially correlated as long as $\sigma_{\alpha}^{2} > 0$).
-From that, we arrive at the typical random effects structure of the $NT \times NT$ matrix $\bm{\Omega}$:
-
-\begin{align}
-\bm{\Omega} & = 
-\begin{pmatrix}
-\bm{\Omega}_{1} & \hdots & 0 & \hdots & 0 \\
-\vdots & \ddots & \vdots & \ddots & \vdots \\
-0 & \hdots & \bm{\Omega}_{i} & \hdots & 0 \\
-\vdots & \ddots & \vdots & \ddots & \vdots \\
-0 & \hdots & 0 & \hdots & \bm{\Omega}_{N}
-\end{pmatrix}
-\end{align}
-with $T \times T$ typical elements 
-\begin{align}
-\bm{\Omega}_{i} & = 
-\begin{pmatrix}
-\sigma^{2}_{\nu} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\alpha} \\
-\sigma^{2}_{\alpha} & \sigma^{2}_{\nu} & \hdots & \sigma^{2}_{\alpha} \\
-\vdots & \vdots & \ddots & \vdots \\
-\sigma^{2}_{\alpha} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\nu}
-\end{pmatrix} \label{eq:omeganui}
-\end{align}
-where $\sigma^{2}_{\nu} = \sigma^{2}_{\alpha} + \sigma^{2}_{\varepsilon}$ [@Wooldridge2002; @Schmidheiny2019].
-This means that in the conditional covariance matrix of the errors, given the time-varying and -invariant covariates, units over time will be correlated due to the individual effects.
-We should keep the covariance structure of the errors in mind as it will help make sense of the use of latent variables to decompose the dependent variable into between- and within-variance components, discussed below in Section \ref{fe-sem}.
-
-Estimation of the RE model can be done using feasible generalized least squares (GLS) in which the two unknowns in $\bm{\Omega}$, $\sigma_{\alpha}^{2}$ and $\sigma_{\nu}^{2}$, are first estimated using pooled ordinary least squares (pooled OLS or POLS), where^[Normally a degrees-of-freedom correction is applied by subtracting off the number of independent variables that is negligible in large N samples.
-It is ignored here for the sake of simplicity.]
-\begin{align}
-\hat{\sigma}_{\nu}^{2} & = \frac{1}{NT}\sum_{i = 1}^{N}\sum_{t = 1}^{T}\hat{\nu}_{it}^{2}, \\
-\hat{\sigma}_{\varepsilon}^{2} & = \frac{1}{NT - N}\sum_{i = 1}^{N}\sum_{t = 1}^{T}(\hat{\nu}_{it} - \bar{\hat{\nu}}_{i})^{2}
-\end{align}
-and $\hat{\nu}_{it} = y_{it} - \bm{x}_{it}\hat{\bm{\beta}}_{\tiny{POLS}} - \bm{z}_{i}\hat{\bm{\gamma}}_{\tiny{POLS}}$, $\bar{\hat{\nu}}_{i} = T^{-1}\sum_{t = 1}^{T}\hat{\nu}_{it}$ and $\hat{\sigma}^{2}_{\alpha} = \hat{\sigma}^{2}_{\nu} - \hat{\sigma}^{2}_{\varepsilon}$.
-Then, the coefficients are estimated using those estimates in the variance matrix $\hat{\bm{\Omega}}$:
-\begin{align}
-\begin{pmatrix}
-\hat{\bm{\beta}}_{RE} \\
-\hat{\bm{\gamma}}_{RE}
-\end{pmatrix} & = 
-(\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{W})^{-1}\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{y}
-\end{align}
-where $\bm{W} = \begin{pmatrix}\bm{X} & \bm{Z}\end{pmatrix}$ and $\bm{X}$ is $NT \times K$ and $\bm{Z}$ is $NT \times M$ and $\bm{y}$ is $NT \times 1$.
-
-In practice, however, computational problems can arise with large cross-sectional samples, where it can become difficult to invert the $\hat{\bm{\Omega}}$ matrix.
-One solution is to use 'partial-demeaning' to transform the data before performing simple POLS:
-\begin{align}
-(y_{it} - \theta\bar{y}_{i}) & = (\bm{x}_{it} - \theta\bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \theta \bar{\bm{z}}_{i})\bm{\gamma} + (\varepsilon_{it} - \theta \bar{\varepsilon}_{i}) \label{eq:partialdemean}
-\end{align} 
-where $\theta = 1 - [\sigma_{\alpha}^{2}/(\sigma_{\alpha}^{2} + T \sigma_{\varepsilon}^{2})]^{1/2}$, and $\bar{y}_{i} = T^{-1}\sum_{t = 1}^{T}y_{it}$, $\bar{\bm{x}}_{i} = T^{-1}\sum_{t = 1}^{T}\bm{x}_{it}$, $\bar{\bm{z}}_{i} = T^{-1}\sum_{t = 1}^{T}\bm{z}_{i}$ and $\bar{\varepsilon}_{i} = T^{-1}\sum_{t = 1}^{T}\varepsilon_{it}$ [@R-plm_a].
-
-The RE model can also be estimated in the maximum likelihood framework, where in the associated literature on panel models are generally referred to as either mixed models, hierarchical models or longitudinal models.
-The typical RE model discussed here is the equivalent to a mixed model with random intercepts and fixed slopes [@R-plm_a].
-Under the assumption of normality, along with homoscedasticity and serially uncorrelated errors, the maximum likelihood estimator is the same as the OLS estimator.
-For more on the topic of mixed models, see for example @R-lme4, @Bates2010.
-
-## Fixed effects {#fe}
-
-For fixed effects, we assume that the individual effects are \textit{not independent} of the model covariates, i.e., $\E[\alpha_{i}|\bm{X}_{i},\bm{z}_{i}] \ne \E[\alpha_{i}] \ne 0$.
-Under this assumption, grouping the individual effects in with the composite error will cause the coefficients of interest, here specifically $\bm{\beta}$ to be inconsistent [@Wooldridge2002; @Wooldridge2012].
-We write the model therefore again in terms of the general panel model in Equation \eqref{eq:gpm} with separate individual effects and idiosyncratic error terms.
-In order to drop assumptions involving the individual effects, a number of methods are available (e.g., differencing, least squares dummy variable regression), but the most common approach is to \textit{demean} the equation [@Bruederl2015].
-Demeaning is the same as the transformation applied in Equation \eqref{eq:partialdemean} in the special case where $\theta = 1$.
-I.e., demeaning involves subtracting the per-unit, over-time average from each of the model terms, i.e.,
-\begin{align}
-(y_{it} - \bar{y}_{i}) & = (\bm{x}_{it} - \bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \bar{\bm{z}}_{i})\bm{\gamma} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
-\ddot{y}_{it} & = \bm{\ddot{x}}_{it}\bm{\beta} + \ddot{\varepsilon}_{it}, \ \text{or} \\
-\bm{\ddot{y}}_{i} & = \bm{\ddot{X}}_{i}\bm{\beta} + \bm{\ddot{\varepsilon}}_{i} \label{eq:fe}
-\end{align}
-where the over-time averages are calculated the same as above, and the variables with the dots above them represent the demeaned versions. 
-Because the average of something that does not change is that thing itself, the individual effects, along with any time-invariant predictors, get wiped out by the demeaning.
-This means that no assumptions about the relatedness of the model covariates and the unit-specific portion of the error are needed.
-Consistency of the estimates is related solely to the strict exogeneity assumption imposed on the idiosyncratic errors, i.e., $\E[\ddot{\varepsilon}_{it}|\bm{\ddot{x}}_{it}] = \E[\ddot{\varepsilon}_{it}] = 0$ which also implies $\E[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T$ [@Bruederl2015; @Wooldridge2002].
-
-Having demeaned the data, the typical FE estimator is POLS on the transformed data
-\begin{align}
-\bm{\beta}_{FE} & = (\ddot{\bm{X}}^{\intercal}\ddot{\bm{X}})^{-1}\ddot{\bm{X}}^{\intercal}\ddot{\bm{y}}
-\end{align}
-[@Bruederl2015].
-The downside to this approach is that no time-invariant predictors can be included in the model.
-However, there are alternative approaches in the random effects and mixed model frameworks that allow them to be included.
-These models are sometimes referred to as 'within-between' or 'hybrid' models, often based on the Chamberlain -@Chamberlain1980 and Mundlak -@Mundlak1978 approaches, see for example @Bell2018; @Allison2011; @Schunck2013; @Enders2007.
-In the [online appendix](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf), it will be discussed how to also get around this restriction using SEM.
-
-# Fixed effects in structural equation modeling {#fe-sem}
-
-Moving from the conventional methods outlined above to SEM, we must state the FE model in a different way.
-We turn to latent variables to account for time-invariant unobserved heterogeneity.
-In fact, besides accounting for measurement error and the representation of abstract hypothetical concepts, unobserved heterogeneity has historically been one of the main uses of latent variables in SEM [@Skrondal2004].
-
-## Modeling time-invariant unobserved heterogeneity as a latent variable
-
-We first need to convert the data from stacked, long-format vectors of length $NT$ into $T$ individual vectors of length $N$.
-To see why this is necessary, consider what effect this has on the vector of responses $y_{it}$.
-Let us, for a minute ignore any covariates and focus just on the dependent variable (a so-called 'intercept-only' or 'null' model) so that we have $y_{it} = \alpha_{i} + \varepsilon_{it}$.
-When we convert the data to wide-format, we get $T$ individual equations,
-\begin{align}
-\bm{y}_{t} & = \bm{\alpha} + \bm{\varepsilon}_{t} \\
-\begin{bmatrix}
-y_{1t} \\
-y_{2t} \\
-\vdots \\
-y_{Nt}\end{bmatrix} & = 
-\begin{bmatrix}
-\alpha_{1} \\
-\alpha_{2} \\
-\vdots \\
-\alpha_{N}
-\end{bmatrix} + 
-\begin{bmatrix}
-\varepsilon_{1t} \\
-\varepsilon_{2t} \\
-\vdots \\
-\varepsilon_{Nt}
-\end{bmatrix} \label{eq:wide}
-\end{align}
-for each $t = 1, 2, ..., T$.
-Because the idiosyncratic errors are assumed to be uncorrelated across units and across time, the covariance between any two of the new wide vectors $\Cov(y_{ti},y_{si}) = \Var(\alpha_{i}), \ t \ne s$.
-Otherwise, when $t = s$, the covariance $\Cov(y_{ti},y_{ti}) = \Var(\alpha_{i}) + \Var(\varepsilon_{ti})$.
-This is the structure we saw above in a typical element of $\bm{\Omega}$.
-
-And in fact this is exactly how a latent variable is used to account for time-invariant unobserved heterogeneity.
-The dependent variable at each timepoint is regressed onto the latent variable, see Figure \ref{fig:fesem}.
-Here, the regression weights or 'factor loadings' are fixed to one to represent our assumption that the effect of the time-invariant unobserved heterogeneity is constant over time.^[In fact, the initial FE-SEM setup shown in the main article mimics the POLS methods described above in that it assumes constant effects and error variances over time.
-These assumptions can be loosened and tested, as will be shown in the [supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf).
-For now, for the sake of simplicity and comparability, we retain the assumptions associated with the 'pooled' models for the most part.] It also means that the estimated variance of the latent variable is equal to the \textit{average covariance between the wide-format columns of the dependent variable over time}.
-If $y_{it} = \alpha_{i} + \varepsilon_{it}$ is the true data generating process, then the relationship between two units over time is just $\Var(\alpha)$, regardless of the time distance.
-Referring back to the random effects structure of $\bm{\Omega}_{i}$ in Equality \eqref{eq:omeganui} for a generic unit $i$, we see the covariance on all of the off-diagonals is $\sigma^{2}_{\alpha}$.
-And, as we know, the average of something that does not change is that thing itself.
-I.e., if $T(T-1)/2 = h$ is the number of elements on either the upper- or lower triangle of $\bm{\Omega}_{i}$, then we have $h^{-1}\sum_{i=1}^{h}\sigma^{2}_{\alpha} = \frac{h \sigma^{2}_{\alpha}}{h} = \sigma^{2}_{\alpha}$.
-
-To elaborate on this concept some more, consider the following matrix equation of the variances and the nonredundant covariances in a three-wave intercept-only model that follows directly from Equation \eqref{eq:wide} (assuming $\Cov(\varepsilon_{ti},\varepsilon_{si}) = 0, \ t \ne s$), and which we can solve easily with least squares:
-\begin{align}
-\bm{A}\bm{x} & = \bm{b} \\
-\begin{pmatrix}
-1 & 1 & 0 & 0 \\
-1 & 0 & 0 & 0 \\
-1 & 0 & 0 & 0 \\
-1 & 0 & 1 & 0 \\
-1 & 0 & 0 & 0 \\
-1 & 0 & 0 & 1
-\end{pmatrix}
-\begin{pmatrix}
-\psi \\
-\phi_{1} \\
-\phi_{2} \\
-\phi_{3}
-\end{pmatrix} & = 
-\begin{pmatrix}
-\Var(y_{1}) \\
-\Cov(y_{2},y_{1}) \\
-\Cov(y_{3},y_{1}) \\
-\Var(y_{2}) \\
-\Cov(y_{3},y_{2}) \\
-\Var(y_{3})
-\end{pmatrix}
-\end{align}
-where $\psi = \Var(\alpha)$, $\phi_{t} = \Var(\varepsilon_{t})$. We can solve this equation to show
-\begin{align}
-\bm{A}\bm{x} & = \bm{b} \\
-(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{A}\bm{x} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} \\
-\hat{\bm{x}} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} \\
-\begin{pmatrix}
-\hat{\psi} \\
-\hat{\phi}_{1} \\
-\hat{\phi}_{2} \\
-\hat{\phi}_{3}
-\end{pmatrix} & = 
-\begin{pmatrix}
-\frac{1}{3}\Cov(y_{2},y_{1}) + \frac{1}{3}\Cov(y_{3},y_{1}) + \frac{1}{3}\Cov(y_{3},y_{2}) \\
-\Var(y_{1}) - \hat{\psi} \\
-\Var(y_{2}) - \hat{\psi} \\
-\Var(y_{3}) - \hat{\psi}
-\end{pmatrix}.
-\end{align}
-So if, in fact the covariance between any two wide-format columns of $y$ is $\Cov(y_{t},y_{s}) = \Var(\alpha), \ \forall \ s \ne t$, then $\hat{\psi} = \frac{1}{3}\Var(\alpha) + \frac{1}{3}\Var(\alpha) + \frac{1}{3}\Var(\alpha) = \frac{3 \Var(\alpha)}{3} = \Var(\alpha)$.
-This shows that if our assumption about the underlying DGP is correct, i.e., $y_{it} = \alpha_{i} + \varepsilon_{it}$ and $\Cov(\varepsilon_{t},\varepsilon_{s}) = 0, \ \forall \ t \ne s$, then the estimated variance of $\alpha$ is just what it should be: the average covariance between units of $y$ over time.
-Once we add in observed covariates, the estimated covariance of $\alpha$ then become the \textit{conditional} covariance of $y$ over time, given those covariates.
-
-## Model notation
-
-Having explained how a latent variable is used to estimate the individual effects, we need to state the FE model using SEM-compatible matrix notation.
-There are a number of different model notations (see, for example @Bollen1989 for an overview), but the one that will serve us best is one that was proposed by @Graff1979:
-\begin{align}
-\bm{y}^{+} & = \bm{\Lambda_{y}}^{+} \bm{\eta}^{+}, \\
-\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+}, 
-\end{align}
-where $\bm{\eta}^{+} = (\bm{y}, \bm{x}, \bm{\eta}, \bm{\xi})^{\intercal}$, $\bm{\zeta}^{+} = (\bm{\varepsilon}, \bm{\delta}, \bm{\zeta}, \bm{\xi})^{\intercal}$, $\bm{y}^{+} = (\bm{y}, \bm{x})^{\intercal}$.
-$\bm{y}$ is a vector of observed dependent variables and $\bm{x}$ is a vector of observed independent variables.
-$\bm{\eta}$ is a vector of the latent dependent variables and $\bm{\xi}$ is a vector of latent independent variables.
-$\bm{\varepsilon}$ and $\bm{\delta}$ are vectors of the errors of the observed dependent and independent variables, respectively, and $\bm{\zeta}$ is a vector of the errors, or disturbances, of the latent variables.
-Notice the $^{+}$ symbol is just meant to differentiate the vectors with them from those without them.
-That means, $\bm{\eta}^{+}$ is a vector that holds the observed and latent variables, both dependent (in SEM they are referred to as 'endogenous') and independent (i.e., 'exogeneous')^[For our purposes, the terms endogenous and dependent, on the one hand, and exogenous and independent, on the other, can be used interchangeably.], $\bm{\zeta}^{+}$ holds the errors for the observed variables and the disturbances of the latent variables.
-$\bm{y}^{+}$ holds just the observed variables, both dependent and independent, and $\bm{\Lambda_{y}}^{+}$ is a matrix of ones and zeros that selects the observed variables from $\bm{\eta}^{+}$.
-Lastly, $\bm{B}$ is a matrix that holds the regression coefficients.
-
-If we say that $p$ and $q$ stand for the number of observed dependent and independent variables, respectively, and $m$ and $n$ stand for the number of latent dependent and independent variables, respectively, then $\bm{\eta}^{+}$ and $\bm{\zeta}^{+}$ are $p + q + m + n$, $\bm{y}^{+}$ is $p + q$, $\bm{\Lambda_{y}}^{+}$ is $(p + q) \times (p + q + m + n)$ and $\bm{B}$ is $(p + q + m + n) \times (p + q + m + n)$ [@Bollen1989].
-
-This notation may be confusing at first, but it has advantages.
-First, it allows us the flexibility we need for the models.
-For example, it allows observed $\bm{x}$ to directly influence observed $\bm{y}$ (more common notation assumes that substantive effects occur only between latent variables, observed ones are only used as indicators, see for example @Bollen1989, @Kline2016).
-It also allows $\bm{\xi}$, i.e., any latent exogenous variables, to influence $\bm{y}$ directly.
-These two scenarios cover the traditional FE model with observed variables, and one in which latent variables are used to account for measurement error in the independent variables.
-It is also consistent with the notation used for these models in `lavaan`.
-In fact, `lavaan` switches automatically between matrix notations depending on the specified model.
-That means the matrix representation of the model one sees if they type in `lavInspect(model, what = "est")` after specifying their model in `lavaan` will match up with the notation used here.^[In `Mplus`, the model matrices can be requested by including `OUTPUT: TECH1` in the input file.] It does have a potential disadvantage however.
-Besides being less intuitive than the typical $\bm{\eta} = \bm{B}\bm{\eta} + \bm{\Gamma}\bm{\xi} + \bm{\zeta}$ notation, it means that by including the observed covariates in the stacked long vector $\bm{y}^{+}$, they are treated as another response variable with variances and covariances to be estimated by the model (instead of just using the sample statistics).
-This means that the assumption of multivariate normality (otherwise just imposed on the dependent variables) also applies to the independent ones [@Skrondal2004, p.
-75].
-This can be problematic for noncontinuous independent variables like sex, nationality dummies, marriage status (married/unmarried), etc.
-See @Skrondal2004 for more on this topic.
-
-\begin{figure}
-\begin{center}
-\resizebox{0.75\textwidth}{!}{%
-\begin{tikzpicture}
-% Node styles ---
-\tikzstyle{man} = [rectangle, thick, minimum size = 1cm, draw = black!80, fill = white!100, font = \sffamily] % Manifest variables
-\tikzstyle{lat} = [circle, thick, minimum size = 1cm, draw = black!80, fill = white!100, font = \sffamily] % Latent variables
-\tikzstyle{err} = [circle, draw = black!80, fill = white!100, font = \sffamily] % Errors 
-% Edge styles
-\tikzstyle{con} = [-latex, font = \sffamily] % Effects
-\tikzstyle{cons} = [-latex, font = \sffamily\small] % Effects with smaller font 
-\tikzstyle{cor} = [latex-latex, font = \sffamily] % Correlations
-% Begin Figure ---
-% Nodes 
-\node at (+0.0,+0.0) [man] (x1) {$x_{1}$};
-\node at (+2.0,+0.0) [man] (x2) {$x_{2}$};
-\node at (+6.0,+0.0) [man] (x3) {$x_{3}$};
-\node at (+4.0,+0.0) [lat] (eta) {$\alpha$};
-\node at (+0.0,-2.0) [man] (y1) {$y_{1}$};
-\node at (+2.0,-2.0) [man] (y2) {$y_{2}$};
-\node at (+6.0,-2.0) [man] (y3) {$y_{3}$};
-\node at (+0.0,-3.5) [err] (e1) {$\varepsilon_{1}$};
-\node at (+2.0,-3.5) [err] (e2) {$\varepsilon_{2}$};
-\node at (+6.0,-3.5) [err] (e3) {$\varepsilon_{3}$};
-% Paths
-\path (x1) edge [con, draw = black!100, left] node {$\beta$} (y1);
-\path (x2) edge [con, draw = black!100, left, near start] node {$\beta$} (y2);
-\path (x3) edge [con, draw = black!100, right] node {$\beta$} (y3);
-\path (e1) edge [con, draw = black!100, left] node {1} (y1);
-\path (e2) edge [con, draw = black!100, left] node {1} (y2);
-\path (e3) edge [con, draw = black!100, right] node {1} (y3);
-\path (eta) edge [con, draw = black!100, above, near start] node {1} (y1);
-\path (eta) edge [con, draw = black!100, below] node {1} (y2);
-\path (eta) edge [con, draw = black!100, above] node {1} (y3);
-% correlations
-\path (x1) edge [cor, bend left] node {} (x2);
-\path (x1) edge [cor, bend left] node {} (eta);
-\path (x1) edge [cor, bend left] node {} (x3);
-\path (x2) edge [cor, bend left] node {} (eta);
-\path (x2) edge [cor, bend left] node {} (x3);
-\path (x3) edge [cor, bend right] node {} (eta);
-\end{tikzpicture}
-}
-\caption{Typical three-wave FE-SEM model with contemporary effects\label{fig:fesem}}
-\end{center}
-\end{figure}
-
-Let us, however, make things more concrete and take a look at a simple, three-wave version of the typical FE-SEM using this notation (shown graphically in Figure \ref{fig:fesem}).
-For that, we have the following matrix notation (with labels on the outside of the matrices):
-\begin{align}
-\begin{split}
-\bm{y}^{+} & = \bm{\Lambda_{y}}^{+}\bm{\eta}^{+} \\
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3}
-\end{pmatrix} & = 
-\begin{blockarray}{cccccccc}
- & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
- \begin{block}{c(ccccccc)}
- y_{1} & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
- y_{2} & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
- y_{3} & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\ 
- x_{1} & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
- x_{2} & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
- x_{3} & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
- \end{block}
-\end{blockarray}
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix},
-\end{split} \label{eq:y}
-\end{align}
-\begin{align}
-\begin{split}
-\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+} \\
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix} & = 
-\begin{blockarray}{cccccccc}
- & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
- \begin{block}{c(ccccccc)}
- y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
- y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
- y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
- x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- \end{block}
-\end{blockarray}
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix} + 
-\begin{pmatrix}
-\varepsilon_{1} \\
-\varepsilon_{2} \\
-\varepsilon_{3} \\
-x_{1} = \delta_{1} \\
-x_{2} = \delta_{2} \\
-x_{3} = \delta_{3} \\
-\alpha = \xi \\
-\end{pmatrix}. 
-\end{split} \label{eq:eta}
-\end{align} 
-Notice in $\bm{\zeta}^{+}$, for the independent variables we could either write, for example $x_{t}$ or $\delta_{t}$.
-As mentioned above, this is due to the model notation treating the independent variables like dependent variables with variances/covariances to be estimated.
-For the sake of simplicity, we will ignore this subtlety and refer to the observed variable from now on, keeping in mind that if the multivariate normality assumption holds, the estimated statistics will likely be sufficiently close to the sample ones for it to not make much of a difference.
-
-Admittedly, Equations \eqref{eq:y} and \eqref{eq:eta} may not look like much yet.
-We can remedy this by first putting the equation for $\bm{\eta}^{+}$ in reduced form, i.e., by getting rid of the dependent variable on the r.h.s.:
-\begin{align}
-\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+} \\
-\bm{\eta}^{+} - \bm{B}\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
-(\bm{I} - \bm{B})\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
-\bm{\eta}^{+} & = (\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+},
-\end{align}
-where $\bm{I}$ is the identity matrix.
-By substituting this back into the equation for the observed variables we get $\bm{y}^{+} = \bm{\Lambda_{y}}^{+}[(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+}]$, which works out to:
-\begin{align}
-\begin{pmatrix} 
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\end{pmatrix} & = 
-\begin{pmatrix}
-\alpha + \beta x_{1} + \varepsilon_{1} \\
-\alpha + \beta x_{2} + \varepsilon_{2} \\
-\alpha + \beta x_{3} + \varepsilon_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3}
-\end{pmatrix},
-\end{align}
-which is of course exactly what we should expect given Equation \eqref{eq:fe}^[We can use the `sympy` package in `python` to verify and show the steps for this and other examples, see the [supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/sympy-doublecheck-matrixnotation.py) for the code.].
-
-## Assumptions
-
-What essentially differentiates an FE from an RE model is our assumption concerning the relationship between the unobserved individual effects and the model covariates [@Bollen2010].
-The FE model assumes that $\E[\alpha x_{t}] \ne 0$.
-As such, if we fail to control for the correlation of the covariate and the time-invariant part of the error, then the coefficient of interest, here $\beta$, will be biased.
-Our assumption regarding whether the individual effects are correlated with the model covariates occurs in $\E[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] = \bm{\Psi}$, the covariance matrix of the errors
-\begin{align}
-\bm{y}^{+}\bm{y}^{+ \intercal} & = \E[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})^{\intercal}] \\
- & = \E[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\zeta}^{+ \intercal}(\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal})] \\
- & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \E[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal} \\
- & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \bm{\Psi} (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal}.
-\end{align}
-In the case of an FE model, $\bm{\Psi}$ will reflect our belief that the individual effects are correlated with the model covariates, here again for demonstration the three-wave model:
-\begin{align}
-\bm{\Psi} & = \E
-\begin{blockarray}{cccccccc}
- & \varepsilon_{1} & \varepsilon_{2} & \varepsilon_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
- \begin{block}{c(ccccccc)}
- \varepsilon_{1} & \varepsilon_{1}^{2} &                     &                     &              &              &              & \\
- \varepsilon_{2} & 0                   & \varepsilon_{2}^{2} &                     &              &              &              & \\
- \varepsilon_{3} & 0                   & 0                   & \varepsilon_{3}^{2} &              &              &              & \\
- x_{1}           & 0                   & 0                   & 0                   & x_{1}^{2}    &              &              & \\
- x_{2}           & 0                   & 0                   & 0                   & x_{2}x_{1}   & x_{2}^{2}    &              & \\
- x_{3}           & 0                   & 0                   & 0                   & x_{3}x_{1}   & x_{3}x_{2}   & x_{3}^{2}    & \\
- \alpha          & 0                   & 0                   & 0                   & \alpha x_{1} & \alpha x_{2} & \alpha x_{3} & \alpha^{2} \\
- \end{block}
-\end{blockarray}.
-\end{align}
-Knowing this, we can work out the equation for the coefficient of interest, $\beta$. For the sake of simplicity, assume here and throughout mean-centered variables: 
-\begin{align}
-\Cov(y_{t},x_{t}) & = \E[y_{t}x_{t}] \\
- & = \E[(\alpha + \beta x_{t} + \varepsilon_{t})x_{t}] \\
- & = \E[\alpha x_{t} + \beta x_{t}^{2} + \varepsilon_{t}x_{t}] \\
- & = \Cov(\alpha, x_{t}) + \beta \Var(x_{t}) \\
-\hat{\beta}_{FE-SEM} & = \frac{\Cov(y_{t},x_{t}) - \Cov(\alpha, x_{t})}{\Var(x_{t})}. 
-\end{align}
-This should make intuitive sense.
-From the observed covariance between the dependent and the independent variable, we are partialling out the part that is due to the covariance between the independent variable and the individual effects per unit, and then dividing by the variance of the independent variable, as usual.
-For the RE model, we assume $\E[\alpha x_{t}] = 0$ and the equation reduces to $\hat{\beta}_{RE-SEM} = \Cov(y_{t},x_{t})/\Var(x_{t})$.
-The rest of the model-implied covariance matrix results from $\bm{y}^{+}\bm{y}^{+ \intercal}$.
-
-# Fixed effects in lavaan {#fesem}
-
-The package `lavaan` needs to be installed once with `install.packages("lavaan")`. To be able to use it, we need to load it for every new `R` session:
-
-```{r, message=FALSE, warning=FALSE, error=FALSE}
-library( lavaan)
-```
-
-For users unfamiliar with `R`, SEM analyses can be carried out with almost no knowledge of the language.
-Typically, someone unfamiliar with `R` would prepare their data using some other statistical software, and then save the intended dataset as a `.csv`, `.xlsx`, `.dta`, `.sav`, etc.
-file.
-The user must then import the data, preferably as a dataframe, and the rest occurs using the `lavaan` syntax.^[There are many online tutorials for importing data in various formats, see, for example some from  [datacamp](https://www.datacamp.com/community/tutorials/r-data-import-tutorial) or [Quick-R](https://www.statmethods.net/input/importingdata.html), or any of the many posts on [stackoverflow](https://stackoverflow.com/search?q=r+import+data).]
-
-Specifying the most basic fixed effects model, like the one shown in @Bollen2010 (the same model as Equation \eqref{eq:fe} but with just one time-varying predictor) involves four components.
-First, we define the latent individual effects variable using the `=~` 'measured by' or 'manifested by' [@R-lavaan] operator at the same time constraining the factor loadings at each timepoint to one.
-I will call the latent variable `a` to stand for $\alpha$.
-Constraining all of the factor loadings to one reflects our implicit assumption that the combined effect of the unit-specific unobserved factors is constant over time.
-This is the default behaviour of traditional POLS-based approaches to FE that use the stacked long-format data.
-
-```{r, eval=FALSE}
-a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
-```
-
-Second, we regress the dependent variable on the independent variable using the `~` regression operator.
-With stacked, long-format data, only one regression coefficient is estimated over all observed timepoints.
-To have our FE-SEM model mimic this behaviour, we need to constrain the the estimated coefficient to equal over time.
-We do so by adding the same label to the regression coefficient at every time point.
-We will use the label `b` (this label was chosen arbitrarily, we could have used any letter or string of characters) and have it act as an equality constraint for the regression coefficient of interest $\beta$:
-
-```{r eval=FALSE}
-y1 ~ b*x1
-y2 ~ b*x2 
-y3 ~ b*x3
-y4 ~ b*x4
-y5 ~ b*x5
-```
-
-The key to a FE model, as opposed to an RE model are our assumptions about the relatedness of our covariate and the individual effects, i.e., $\E[x_{t}\alpha]$.
-For an FE model, we want to partial out any potential covariance between the independent variable and the individual effects.
-This accounts for any linear relationship between $x_{t}$ and the unit-specific characteristics influencing the dependent variable.
-Further, allowing unrestricted covariances between the independent variable itself over time will not affect how the coefficient $\beta$ is estimated, but will have an effect on the standard errors.
-To mimic the behaviour of a conventional FE model, we allow the independent variable to be correlated with the individual effects and itself over time.
-Covariances (including covariances between a variable and itself, i.e., variances) are specified using the `~~` operator:
-
-```{r eval=FALSE}
-a ~~ x1 + x2 + x3 + x4 + x5
-x1 ~~ x2 + x3 + x4 + x5
-x2 ~~ x3 + x4 + x5
-x3 ~~ x4 + x5
-x4 ~~ x5
-```
-
-The last component of our code involves the variances of the residuals.
-This component is optional, but we can constrain the residual variances to be equal over time to again mimic the behaviour of a conventional FE model using POLS on stacked data.
-Here, again, we use labels to make equality constraints.
-Because $y_{t}$ is endogenous, the `~~` operator specifies the variances of *residuals*, i.e., $\varepsilon_{t}$.
-
-```{r eval=FALSE}
-y1 ~~ e*y1
-y2 ~~ e*y2
-y3 ~~ e*y3
-y4 ~~ e*y4
-y5 ~~ e*y5
-```
-
-# A simulated example {#ex1}
-
-```{r echo=FALSE}
-df <- readRDS("../longData.Rda")
-dfw <- readRDS("../wideData.Rda")
-```
-
-To demonstrate the application of FE models in SEM, a dataset can be simulated that embodies the FE assumptions.
-Again, the code for data simulation can be found in the [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R).
-
-To show that the latent individual effects variables represent the *combined* effect of all time-invariant characteristics, the dependent variable will be influenced by two separate unit-specific variables, which we can call $\alpha_{1}$ and $\alpha_{2}$.
-We will construct the simulated data such that the independent variable is correlated with both of the time-invariant variables.
-This means that approaches that fail to account for this confounding influence, such as POLS or RE, will be biased.
-
-The wide-format equations for the data generating process can be described as:
-\begin{align}
-\bm{x_{t}} & = \bm{\alpha_{1}}\beta_{x_{t},\alpha_{1}} + \bm{\alpha_{2}}\beta_{x_{t},\alpha_{2}} + \bm{\delta_{t}}, \\
-\bm{y_{t}} & = \bm{x_{t}}\beta_{y_{t},x_{t}} + \bm{\alpha_{1}}\beta_{y_{t},\alpha_{1}} + \bm{\alpha_{2}} \beta_{y_{t},\alpha_{2}} + \bm{\varepsilon_{t}} 
-\end{align}
-where, for the sake of simplicity, $\bm{\alpha_{1}}$, $\bm{\alpha_{2}}$, $\bm{\delta_{t}}$ and $\bm{\varepsilon_{t}}$ are $\sim N(0,1)$.
-
-For the following example, a sample size of 1,000, observed over five waves, was chosen.
-The unique variance of $\bm{x}$, as well as both the individual-effect variables is also $\sim N(0,1)$.
-The coefficient of interest, $\beta_{y,x}$ is set to be equal to $0.3$.
-A correlation between $\bm{x}$ and the individual effects is induced through $\beta_{x,\alpha_{1}} = 0.85$ and $\beta_{x,\alpha_{1}} = 0.50$.
-With the variances above set to one, the covariances will be roughly $\Cov(x_{t},\alpha_{1}) = 0.85$ and $\Cov(x_{t},\alpha_{2}) = 0.5$.
-The dependent variable is also influenced by the individual effects variables with $\beta_{y_{t},\alpha_{1}} = 0.75$ and $\beta_{y_{t},\alpha_{2}} = 0.45$ These values were chosen arbitrarily.
-
-Now, we run the FE-SEM in `lavaan`.
-
-```{r}
-fe_sem <- '
-# Define individual effects variable 
-a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
-# Regressions, constrain coefficient to be equal over time
-y1 ~ b*x1
-y2 ~ b*x2 
-y3 ~ b*x3
-y4 ~ b*x4
-y5 ~ b*x5
-# Allow unrestricted correlation between eta and covariates
-a ~~ x1 + x2 + x3 + x4 + x5
-x1 ~~ x2 + x3 + x4 + x5
-x2 ~~ x3 + x4 + x5
-x3 ~~ x4 + x5
-x4 ~~ x5
-# Constrain residual variances to be equal over time
-y1 ~~ e*y1
-y2 ~~ e*y2
-y3 ~~ e*y3
-y4 ~~ e*y4
-y5 ~~ e*y5
-'
-fe_sem.fit <- sem(model = fe_sem, 
-                  data = dfw, 
-                  estimator = "ML")
-```
-
-```{r, echo=FALSE, message=FALSE, error=FALSE, warning=FALSE}
-saveRDS(fe_sem, "../fe_sem.Rda")
-saveRDS(fe_sem.fit, "../fe_sem.fit.Rda")
-```
-
-We can get a summary of the model with `summary()`.
-The first portion of the summary output gives an overview of some basic information and fit statistics.
-The maximum likelihood estimator is the default, so it did not have to be explicitly selected in the fitting function call.
-Other estimators are available, including generalized and unweighted least squares (`GLS` and `ULS`, respectively), robust standard errors maximum likelihood (`MLM`) and several others (see [the lavaan online tutorial for more](https://lavaan.ugent.be/tutorial/est.html)).
-
-This part of the summary output also tells us that the analysis is based on 1,000 observations (missings would be shown here as well if there were any), and that the $\chi^{2}$ statistic is `r round( lavInspect( fe_sem.fit, "fit")[ "chisq"], 3)` based on `r lavInspect( fe_sem.fit, "fit")[ "df"]` degrees of freedom (55 observed covariances minus 1 error variance, 1 coefficient, 1 latent variable variance, 5 exogenous variable variances and 15 covariances for $55 - 23 = 32$ df).
-The p-value on the $\chi^{2}$ statistic is not significant with $p =$ `r round( lavInspect( fe_sem.fit, "fit")[ "pvalue"], 3)` which tells us the differences between the model-implied and observed covariance matrices are likely due to chance, and that the model fits the data well (given how the data was generated, it would be surprising if this were not the case).
-Other fit measures including typical comparative fit indices can be requested by either adding `fit.measures = TRUE` as a secondary argument to the `summary()` call, or by asking for a complete list of all available fit statistics using `lavInspect(model, "fit")` where `model` stands for the name of the fitted model, in this case `fe_sem.fit`.
-
-```{r restate model, output.lines=1:20}
-summary(fe_sem.fit)
-```
-
-Next the summary output shows the measurement models for the latent variables, if any.
-In this case the latent variable `a` for $\alpha$ is measured by each of the five observed dependent variables with factor loadings fixed to 1.0.
-
-```{r measurement, echo=FALSE, output.lines=22:29}
-summary(fe_sem.fit)
-```
-
-The regressions are shown next.
-Here, because we have constrained the regression coefficients to be equal over time (the equality constraint label `(b)` is listed to the left of the estimates), the estimate of $\beta =$ `r round( lavInspect( fe_sem.fit, "list")[6, 14], 3)` (`r round( lavInspect( fe_sem.fit, "list")[6, 15], 3)`) is repeated five times.
-The corresponding z- and p-values show that the coefficient is, unsurprisingly, significant.
-
-```{r regression, echo=FALSE, output.lines=31:42}
-summary(fe_sem.fit)
-```
-
-Next, the covariance estimates are listed.
-First, the covariances between the latent individual effects variable and the independent variable over time are shown, and then the covariances between the independent variable with itself over time.
-
-One should always take care to double-check that there are no unintended covariances listed here.
-Like `Mplus`, the `lavaan` package estimates some covariances per default, without the user explicitly having to add them to the model syntax.
-For example, covariances between latent variables are estimated per default.
-If one does not wish for them to covary, it must be explicitly stated, e.g., with `f1 ~~ 0*f2`, assuming the latent variables are called `f1` and `f2`, or by overriding the default behaviour for the entire model by adding `orthogonal = TRUE` (which sets the correlation between all latent variables to zero) to the fitting call.^[This is at least the current behaviour of both the `cfa` and `sem` wrappers.
-In fact, both wrappers seem to be identical in terms of the default settings, see @Rosseel2020.]
-
-```{r covariance, echo=FALSE, output.lines=44:65}
-summary(fe_sem.fit)
-```
-
-Finally, the variance estimates are listed.
-Here, we see that in order to mimic the behaviour of a traditional FE model, the error variances over time were specified to be equal using the equality constraint `(e)`.
-Notice the `.` beside `y1`, `y2`, etc.: this indicates that the listed variance refers to an endogenous variable, and that it is thus an error variance.
-In this case, these refer to the variances of $\varepsilon_{t}$.
-After that, the variances of the exogenous variables, both observed and unobserved are listed.
-
-```{r variance, echo=FALSE, output.lines=67:79}
-summary(fe_sem.fit)
-```
-
-# Conclusion
-
-Fixed effects regression in SEM has been outlined in well-known articles by [@Allison2011; @Bollen2010; @Teachman2001].
-This article provides a focused look at the implementation of the basic model using the `lavaan` package in `R`.
-The [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf) further discuss common extensions and some tools for evaluating and loosening model assumptions.
-
-The benefits of FE-SEM as opposed to traditional OLS-based FE-models are largely the same ones that apply to the SEM framework in general: for one, SEM allows for a great deal of flexibility.
-For example, it is easy to loosen model constraints as necessary.
-Measurement error in both the dependent and independent variables can be dealt with using latent variables to achieve unbiased and more efficient results.
-Researchers interested in time-invariant predictors can integrate them into a hybrid FE/RE model with ease.
-Further extensions, like measurement invariance testing [@Schoot2012; @Millsap2011; @Steenkamp1998] as well as lagged dependent variables [@Bollen2010; @Allison2017] for example, can also be implemented in a straightforward fashion.
-
-The most basic FE-SEM is furthermore the basis for a variety of currently popular extended models, such as Latent Curve Models in general [@Curran2001; @Bollen2004], as well as special implementations like the Dynamic Panel Model [@Allison2017], the Random-Intercept Cross-Lagged Panel Model [@Hamaker2015] and the Latent Curve Model with Structured Residuals [@Curran2014].
-For this reason it is all the more important for researchers to have a good grasp on the method of applying panel regression in SEM, and understanding the intuition of controlling for time-invariant confounders.
-This article is meant to serve as a consolidated resource for researchers looking for concrete advise on specifying FE and more general panel models in SEM.
-
-# References

### 2020-07-28 article_format/title.tex
Changed the formating, made some minor edits

--- 2020-07-28 article_format/article_format.tex
@@
 % see https://stackoverflow.com/a/47122900
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 
@@
 
-\newpage
-
-\pagenumbering{arabic} 
-\setcounter{page}{1}
-
-\doublespacing
-
-\hypertarget{intro}{%
-\section{Introduction}\label{intro}}
-
-Several years ago, \citet{Curran2011} reflected positively on the
-growing use of panel studies in empirical social research. Some of the
-strengths of panel data are well-known, e.g., the ability to establish
-temporal precedence, increased statistical power and the reduction of
-potential alternative models. However, perhaps the greatest strength of
-panel data is that they allow for a more rigorous testing of substantive
-theories. Panel data, i.e., repeated measures of the same observed units
-(people, schools, firms, countries, etc.), allow researchers to
-decompose the error term into a part that stays constant within units
-and the part that changes over time. The part that does not change over
-time can be seen as the combined effect of all time-invariant influences
-(e.g., sex, date of birth, nationality) on the dependent variable. Fixed
-effects (FE) regression involves controlling for these time-invariant
-influences via a number of various methods. It thus accounts for a
-likely and common source of bias.
-
-Structural equation modeling (SEM) is a popular regression framework.
-One of its main strengths is its flexibility. Not only can complex
-causal structures with multiple dependent variables be tested
-simultaneously, but in longitudinal (and, more generally, hierarchical)
-studies both time-varying and invariant predictors can be included, and
-effects can easily be allowed to vary over time. Thus researchers can
-allow for and study effects that increase or fade over time, or that
-appear only in specific periods. Beyond that, with the use of latent
-variables, SEM provides a way to deal with measurement error and get
-closer to the true underlying constructs of interest.
-
-There are a number of articles describing basic concept of panel model
-regression, and FE regression in SEM
-\citep[e.g.,][]{Allison2011, Bollen2010, Teachman2001}. This article is
-intended as a \textit{practical guide} for researchers looking for
-in-depth help with specifying FE models in SEM. It focuses on the
-\texttt{lavaan} \citep{R-lavaan} package for \texttt{R} \citep{R-base}.
-While \texttt{Mplus} \citep{Mplus} is arguably the most robust SEM
-software currently available (in terms of features like alignment,
-latent variable interactions, for example), the \texttt{lavaan} package
-has many benefits. First, like \texttt{R} it is open source and
-completely free. For researchers dipping their toes into SEM, there is
-no financial barrier to try, and no risk if they decide it is not for
-them. Second, the implementation of \texttt{lavaan} in the larger
-\texttt{R} environment is an enormous advantage. Instead of poring over
-reams of plain text, copying out coefficients by hand, every part of the
-\texttt{lavaan} output is available as an object. This means that all
-aspects of the model, from fit indices, to coefficients and standard
-errors, to the model matrices, can be accessed and easily integrated
-into tables and plots. Furthermore, \texttt{R} can be used for a great
-deal of applications. It can be used to manage and manipulate as well as
-simulate data, perform symbolic algebra, run more traditional analyses
-(e.g., multiple regression, logistic regression, principal component
-analysis), etc. Once one is comfortable using \texttt{R}, there is no
-longer any need to switch between different software for data
-preparation and analysis.
-
-The following article outlines the basic idea of panel regression, the
-particularities of panel regression SEM, and shows its implementation in
-\texttt{lavaan}. The focus on FE panel regression stems from the fact
-that the assumptions needed to justify other common models, e.g., random
-effects, are often implausible. As we will see, the practical
-differences between the two are, in any event, very small. Using
-\href{https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R}{simulated
-data}, it demonstrates and annotates the code for the most basic FE
-model and provides an overview of the summary output. One of the main
-strengths of panel SEM compared to the more traditional methods of panel
-analysis is its flexibility. Therefore, a number of potential extensions
-to the basic model, including relaxing various assumptions, dealing with
-measurement error in both the independent and dependent variables, as
-well as the inclusion of time-invariant predictors in the form of a
-hybrid fixed-/ random effects model, are shown in detail in the form of
-\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
-supplementary materials}.
-
-\hypertarget{panel}{%
-\section{Panel models}\label{panel}}
-
-To begin, let us start by reviewing a general panel model
-\citep{Bollen2010}, also referred to as the `unobserved effects model'
-\citep{Wooldridge2012, R-plm_a} (we will return to this model in the
-\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
-supplementary materials} when we discuss loosening assumptions)
-\begin{align}
-y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \alpha_{i} + \varepsilon_{it} \label{eq:gpm}
-\end{align} where \(y_{it}\) is the dependent variable for unit
-\(i, \ i = 1, ..., N\) at time \(t, \ t = 1, ..., T\), \(\bm{x}_{it}\)
-is a \(1 \times K\) vector of time-varying covariates (which could
-include a constant) linked to the dependent variable by the
-\(K \times 1\) vector of coefficients \(\bm{\beta}\). \(\bm{z}_{i}\) is
-a \(1 \times M\) vector of time-invariant covariates linked to the
-dependent variable by the \(M \times 1\) vector of coefficients in
-\(\bm{\gamma}\), \(\alpha_{i}\) represents the combined effect of all
-unobserved time-constant variables affecting the dependent variable and
-\(\varepsilon_{it}\) is the idiosyncratic error.
-
-We can make stating some of the model assumptions easier by rewriting it
-in matrix notation \begin{align}
-\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\iota}_{T}\alpha_{i} + \bm{\varepsilon}_{i}
-\end{align} where \(\bm{y}_{i}\) and \(\bm{\varepsilon}_{i}\) are
-\(T \times 1\) vectors, \(\bm{X}_{i}\) and \(\bm{Z}_{i}\) are
-\(T \times K\) and \(T \times M\) matrices, respectively,
-\(\bm{\iota}_{T}\) is a \(T \times 1\) vector of ones and \(\alpha_{i}\)
-is a scalar. \(\bm{\beta}\) and \(\bm{\gamma}\) are unchanged from
-Equation \eqref{eq:gpm}.
-
-Consistency of the following models requires the assumption of strict
-exogeneity, although what constitutes strict exogeneity differs between
-the random and fixed effects setups. Each assumption will be discussed
-shortly. Apart from that, we typically make the following assumptions
-about this model \citep[see, e.g.,][]{Wooldridge2002, Schmidheiny2019}:
-
-\begin{itemize}
-\tightlist
-\item
-  Linearity: the model is linear in its parameters.
-\item
-  Independence: the observations are independent across individuals
-  (assured by random sampling in the cross-section), but not necessarily
-  across time.
-\item
-  The usual rank condition: we have more observations than independent
-  variables and there is no perfect collinearity between any of the
-  independent variables.
-\end{itemize}
-
-\hypertarget{re}{%
-\subsection{Random effects}\label{re}}
-
-Somewhat unintuitively, the modern approach to FE regression treats the
-unobserved effect \(\alpha_{i}\) as a \textit{random} variable, rather
-than a fixed parameter to be estimated for each of the cross sectional
-units \citep{Wooldridge2002}. And in fact, this view carries over to the
-way in which FE regression is applied in SEM. So, in order to help
-facilitate the discussion on the application of FE-SEM later on, it
-makes sense to spend some time discussing the basic idea of random
-effects regression (again, as we will see, the practical difference
-between the two models is very small).
-
-For the random effects (RE) model, we define a composite error term:
-\(\nu_{it} = \alpha_{i} + \varepsilon_{it}\) and rewrite the model in
-Equation \eqref{eq:gpm} as \begin{align}
-y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \nu_{it}, \ \text{or} \\
-\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\nu}_{i},
-\end{align} where
-\(\bm{\nu}_{i} = \alpha_{i} \bm{\iota}_{T} + \bm{\varepsilon}_{i}\) and
-\(\bm{\iota}_{T}\) is a \(T \times 1\) vector of ones. The strict
-exogeneity assumption in the RE model implies \begin{align}
-\mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = 0, \\
-\mathop{\mathrm{\mathbb{E}}}[\alpha_{i} |\bm{X}_{i}, \bm{z}_{i}] = \mathop{\mathrm{\mathbb{E}}}[\alpha_{i}] & = 0,
-\end{align} where \(\bm{X}_{i} = \bm{x}_{i1}, ..., \bm{x}_{iT}\). For
-both parts, the assumption that the unconditional expectations are 0 is
-unproblematic as long as a constant is included in the regression. The
-first part says the idiosyncratic errors at each timepoint are assumed
-to be independent of the explanatory variables at \textit{all}
-timepoints which is stronger than just assuming that they are
-\textit{contemporaneously} independent. This implies that they are also
-uncorrelated, i.e.,
-\(\mathop{\mathrm{\mathbb{E}}}[\bm{x}_{is}^{\intercal}\varepsilon_{it}] = \bm{0}\)
-and
-\(\mathop{\mathrm{\mathbb{E}}}[\bm{z}_{i}^{\intercal}\varepsilon_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T\)
-\citep{Wooldridge2002, Bruederl2015}. We assume the idiosyncratic errors
-are further independent of the individual effects, which implies
-\(\mathop{\mathrm{\mathbb{E}}}[\alpha_{i}\varepsilon_{it}] = 0\).
-
-The second part is the potentially controversial assumption: it states
-that the individual effects are uncorrelated with the independent
-variables. We can use an intuitive concrete example to show why this is
-often controversial: If we are interested in the question of whether
-married men earn more than unmarried men, then the second part of the
-strict exogeneity assumption means that a man's marriage status would
-have to be uncorrelated with all the time-invariant characteristics that
-could potentially make that man an attractive marriage candidate in the
-first place; e.g., looks, personality, family's status, profession, etc.
-\citep{Bruederl2015}.\footnote{Assuming, for the sake of argument, that
-  these characteristics are constant over time.}
-
-From what we have discussed so far, the \(T \times T\) covariance matrix
-of the errors
-\(\bm{\Omega}_{i} = \mathop{\mathrm{\mathbb{E}}}[\bm{\nu}_{i}\bm{\nu}_{i}^{\intercal}]\)
-can be constructed. However, the standard random effects model adds the
-additional assumptions \begin{alignat}{3}
-\mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}^{2} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}^{2}] && = \sigma_{\varepsilon}^{2}, && \ t = 1, ..., T, \\
-\mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}\varepsilon_{is}| \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \mathop{\mathrm{\mathbb{E}}}[\varepsilon_{it}\varepsilon_{is}] && = 0, && \ \forall \ t \ne s 
-\end{alignat} i.e., the idiosyncratic errors are conditionally
-homoscedastic and serially uncorrelated and \begin{alignat}{2}
-\mathop{\mathrm{\mathbb{E}}}[\alpha_{i}^{2}| \bm{X}_{i},\bm{z}_{i}] & = \mathop{\mathrm{\mathbb{E}}}[\alpha_{i}^{2}] && = \sigma^{2}_{\alpha}
-\end{alignat} i.e., the individual effects are conditionally
-homoscedastic (they are necessarily serially correlated as long as
-\(\sigma_{\alpha}^{2} > 0\)). From that, we arrive at the typical random
-effects structure of the \(NT \times NT\) matrix \(\bm{\Omega}\):
-
-\begin{align}
-\bm{\Omega} & = 
-\begin{pmatrix}
-\bm{\Omega}_{1} & \hdots & 0 & \hdots & 0 \\
-\vdots & \ddots & \vdots & \ddots & \vdots \\
-0 & \hdots & \bm{\Omega}_{i} & \hdots & 0 \\
-\vdots & \ddots & \vdots & \ddots & \vdots \\
-0 & \hdots & 0 & \hdots & \bm{\Omega}_{N}
-\end{pmatrix}
-\end{align} with \(T \times T\) typical elements \begin{align}
-\bm{\Omega}_{i} & = 
-\begin{pmatrix}
-\sigma^{2}_{\nu} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\alpha} \\
-\sigma^{2}_{\alpha} & \sigma^{2}_{\nu} & \hdots & \sigma^{2}_{\alpha} \\
-\vdots & \vdots & \ddots & \vdots \\
-\sigma^{2}_{\alpha} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\nu}
-\end{pmatrix} \label{eq:omeganui}
-\end{align} where
-\(\sigma^{2}_{\nu} = \sigma^{2}_{\alpha} + \sigma^{2}_{\varepsilon}\)
-\citep{Wooldridge2002, Schmidheiny2019}. This means that in the
-conditional covariance matrix of the errors, given the time-varying and
--invariant covariates, units over time will be correlated due to the
-individual effects. We should keep the covariance structure of the
-errors in mind as it will help make sense of the use of latent variables
-to decompose the dependent variable into between- and within-variance
-components, discussed below in Section \ref{fe-sem}.
-
-Estimation of the RE model can be done using feasible generalized least
-squares (GLS) in which the two unknowns in \(\bm{\Omega}\),
-\(\sigma_{\alpha}^{2}\) and \(\sigma_{\nu}^{2}\), are first estimated
-using pooled ordinary least squares (pooled OLS or POLS),
-where\footnote{Normally a degrees-of-freedom correction is applied by
-  subtracting off the number of independent variables that is negligible
-  in large N samples. It is ignored here for the sake of simplicity.}
-\begin{align}
-\hat{\sigma}_{\nu}^{2} & = \frac{1}{NT}\sum_{i = 1}^{N}\sum_{t = 1}^{T}\hat{\nu}_{it}^{2}, \\
-\hat{\sigma}_{\varepsilon}^{2} & = \frac{1}{NT - N}\sum_{i = 1}^{N}\sum_{t = 1}^{T}(\hat{\nu}_{it} - \bar{\hat{\nu}}_{i})^{2}
-\end{align} and
-\(\hat{\nu}_{it} = y_{it} - \bm{x}_{it}\hat{\bm{\beta}}_{\tiny{POLS}} - \bm{z}_{i}\hat{\bm{\gamma}}_{\tiny{POLS}}\),
-\(\bar{\hat{\nu}}_{i} = T^{-1}\sum_{t = 1}^{T}\hat{\nu}_{it}\) and
-\(\hat{\sigma}^{2}_{\alpha} = \hat{\sigma}^{2}_{\nu} - \hat{\sigma}^{2}_{\varepsilon}\).
-Then, the coefficients are estimated using those estimates in the
-variance matrix \(\hat{\bm{\Omega}}\): \begin{align}
-\begin{pmatrix}
-\hat{\bm{\beta}}_{RE} \\
-\hat{\bm{\gamma}}_{RE}
-\end{pmatrix} & = 
-(\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{W})^{-1}\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{y}
-\end{align} where
-\(\bm{W} = \begin{pmatrix}\bm{X} & \bm{Z}\end{pmatrix}\) and \(\bm{X}\)
-is \(NT \times K\) and \(\bm{Z}\) is \(NT \times M\) and \(\bm{y}\) is
-\(NT \times 1\).
-
-In practice, however, computational problems can arise with large
-cross-sectional samples, where it can become difficult to invert the
-\(\hat{\bm{\Omega}}\) matrix. One solution is to use `partial-demeaning'
-to transform the data before performing simple POLS: \begin{align}
-(y_{it} - \theta\bar{y}_{i}) & = (\bm{x}_{it} - \theta\bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \theta \bar{\bm{z}}_{i})\bm{\gamma} + (\varepsilon_{it} - \theta \bar{\varepsilon}_{i}) \label{eq:partialdemean}
-\end{align} where
-\(\theta = 1 - [\sigma_{\alpha}^{2}/(\sigma_{\alpha}^{2} + T \sigma_{\varepsilon}^{2})]^{1/2}\),
-and \(\bar{y}_{i} = T^{-1}\sum_{t = 1}^{T}y_{it}\),
-\(\bar{\bm{x}}_{i} = T^{-1}\sum_{t = 1}^{T}\bm{x}_{it}\),
-\(\bar{\bm{z}}_{i} = T^{-1}\sum_{t = 1}^{T}\bm{z}_{i}\) and
-\(\bar{\varepsilon}_{i} = T^{-1}\sum_{t = 1}^{T}\varepsilon_{it}\)
-\citep{R-plm_a}.
-
-The RE model can also be estimated in the maximum likelihood framework,
-where in the associated literature on panel models are generally
-referred to as either mixed models, hierarchical models or longitudinal
-models. The typical RE model discussed here is the equivalent to a mixed
-model with random intercepts and fixed slopes \citep{R-plm_a}. Under the
-assumption of normality, along with homoscedasticity and serially
-uncorrelated errors, the maximum likelihood estimator is the same as the
-OLS estimator. For more on the topic of mixed models, see for example
-\citet{R-lme4}, \citet{Bates2010}.
-
-\hypertarget{fe}{%
-\subsection{Fixed effects}\label{fe}}
-
-For fixed effects, we assume that the individual effects are
-\textit{not independent} of the model covariates, i.e.,
-\(\mathop{\mathrm{\mathbb{E}}}[\alpha_{i}|\bm{X}_{i},\bm{z}_{i}] \ne \mathop{\mathrm{\mathbb{E}}}[\alpha_{i}] \ne 0\).
-Under this assumption, grouping the individual effects in with the
-composite error will cause the coefficients of interest, here
-specifically \(\bm{\beta}\) to be inconsistent
-\citep{Wooldridge2002, Wooldridge2012}. We write the model therefore
-again in terms of the general panel model in Equation \eqref{eq:gpm}
-with separate individual effects and idiosyncratic error terms. In order
-to drop assumptions involving the individual effects, a number of
-methods are available (e.g., differencing, least squares dummy variable
-regression), but the most common approach is to \textit{demean} the
-equation \citep{Bruederl2015}. Demeaning is the same as the
-transformation applied in Equation \eqref{eq:partialdemean} in the
-special case where \(\theta = 1\). I.e., demeaning involves subtracting
-the per-unit, over-time average from each of the model terms, i.e.,
-\begin{align}
-(y_{it} - \bar{y}_{i}) & = (\bm{x}_{it} - \bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \bar{\bm{z}}_{i})\bm{\gamma} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
-\ddot{y}_{it} & = \bm{\ddot{x}}_{it}\bm{\beta} + \ddot{\varepsilon}_{it}, \ \text{or} \\
-\bm{\ddot{y}}_{i} & = \bm{\ddot{X}}_{i}\bm{\beta} + \bm{\ddot{\varepsilon}}_{i} \label{eq:fe}
-\end{align} where the over-time averages are calculated the same as
-above, and the variables with the dots above them represent the demeaned
-versions. Because the average of something that does not change is that
-thing itself, the individual effects, along with any time-invariant
-predictors, get wiped out by the demeaning. This means that no
-assumptions about the relatedness of the model covariates and the
-unit-specific portion of the error are needed. Consistency of the
-estimates is related solely to the strict exogeneity assumption imposed
-on the idiosyncratic errors, i.e.,
-\(\mathop{\mathrm{\mathbb{E}}}[\ddot{\varepsilon}_{it}|\bm{\ddot{x}}_{it}] = \mathop{\mathrm{\mathbb{E}}}[\ddot{\varepsilon}_{it}] = 0\)
-which also implies
-\(\mathop{\mathrm{\mathbb{E}}}[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T\)
-\citep{Bruederl2015, Wooldridge2002}.
-
-Having demeaned the data, the typical FE estimator is POLS on the
-transformed data \begin{align}
-\bm{\beta}_{FE} & = (\ddot{\bm{X}}^{\intercal}\ddot{\bm{X}})^{-1}\ddot{\bm{X}}^{\intercal}\ddot{\bm{y}}
-\end{align} \citep{Bruederl2015}. The downside to this approach is that
-no time-invariant predictors can be included in the model. However,
-there are alternative approaches in the random effects and mixed model
-frameworks that allow them to be included. These models are sometimes
-referred to as `within-between' or `hybrid' models, often based on the
-Chamberlain \citeyearpar{Chamberlain1980} and Mundlak
-\citeyearpar{Mundlak1978} approaches, see for example \citet{Bell2018};
-\citet{Allison2011}; \citet{Schunck2013}; \citet{Enders2007}. In the
-\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
-appendix}, it will be discussed how to also get around this restriction
-using SEM.
-
-\hypertarget{fe-sem}{%
-\section{Fixed effects in structural equation modeling}\label{fe-sem}}
-
-Moving from the conventional methods outlined above to SEM, we must
-state the FE model in a different way. We turn to latent variables to
-account for time-invariant unobserved heterogeneity. In fact, besides
-accounting for measurement error and the representation of abstract
-hypothetical concepts, unobserved heterogeneity has historically been
-one of the main uses of latent variables in SEM \citep{Skrondal2004}.
-
-\hypertarget{modeling-time-invariant-unobserved-heterogeneity-as-a-latent-variable}{%
-\subsection{Modeling time-invariant unobserved heterogeneity as a latent
-variable}\label{modeling-time-invariant-unobserved-heterogeneity-as-a-latent-variable}}
-
-We first need to convert the data from stacked, long-format vectors of
-length \(NT\) into \(T\) individual vectors of length \(N\). To see why
-this is necessary, consider what effect this has on the vector of
-responses \(y_{it}\). Let us, for a minute ignore any covariates and
-focus just on the dependent variable (a so-called `intercept-only' or
-`null' model) so that we have
-\(y_{it} = \alpha_{i} + \varepsilon_{it}\). When we convert the data to
-wide-format, we get \(T\) individual equations, \begin{align}
-\bm{y}_{t} & = \bm{\alpha} + \bm{\varepsilon}_{t} \\
-\begin{bmatrix}
-y_{1t} \\
-y_{2t} \\
-\vdots \\
-y_{Nt}\end{bmatrix} & = 
-\begin{bmatrix}
-\alpha_{1} \\
-\alpha_{2} \\
-\vdots \\
-\alpha_{N}
-\end{bmatrix} + 
-\begin{bmatrix}
-\varepsilon_{1t} \\
-\varepsilon_{2t} \\
-\vdots \\
-\varepsilon_{Nt}
-\end{bmatrix} \label{eq:wide}
-\end{align} for each \(t = 1, 2, ..., T\). Because the idiosyncratic
-errors are assumed to be uncorrelated across units and across time, the
-covariance between any two of the new wide vectors
-\(\mathop{\mathrm{\mathrm{Cov}}}(y_{ti},y_{si}) = \mathop{\mathrm{\mathrm{Var}}}(\alpha_{i}), \ t \ne s\).
-Otherwise, when \(t = s\), the covariance
-\(\mathop{\mathrm{\mathrm{Cov}}}(y_{ti},y_{ti}) = \mathop{\mathrm{\mathrm{Var}}}(\alpha_{i}) + \mathop{\mathrm{\mathrm{Var}}}(\varepsilon_{ti})\).
-This is the structure we saw above in a typical element of
-\(\bm{\Omega}\).
-
-And in fact this is exactly how a latent variable is used to account for
-time-invariant unobserved heterogeneity. The dependent variable at each
-timepoint is regressed onto the latent variable, see Figure
-\ref{fig:fesem}. Here, the regression weights or `factor loadings' are
-fixed to one to represent our assumption that the effect of the
-time-invariant unobserved heterogeneity is constant over time.\footnote{In
-  fact, the initial FE-SEM setup shown in the main article mimics the
-  POLS methods described above in that it assumes constant effects and
-  error variances over time. These assumptions can be loosened and
-  tested, as will be shown in the
-  \href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{supplementary
-  materials}. For now, for the sake of simplicity and comparability, we
-  retain the assumptions associated with the `pooled' models for the
-  most part.} It also means that the estimated variance of the latent
-variable is equal to the
-\textit{average covariance between the wide-format columns of the dependent variable over time}.
-If \(y_{it} = \alpha_{i} + \varepsilon_{it}\) is the true data
-generating process, then the relationship between two units over time is
-just \(\mathop{\mathrm{\mathrm{Var}}}(\alpha)\), regardless of the time
-distance. Referring back to the random effects structure of
-\(\bm{\Omega}_{i}\) in Equality \eqref{eq:omeganui} for a generic unit
-\(i\), we see the covariance on all of the off-diagonals is
-\(\sigma^{2}_{\alpha}\). And, as we know, the average of something that
-does not change is that thing itself. I.e., if \(T(T-1)/2 = h\) is the
-number of elements on either the upper- or lower triangle of
-\(\bm{\Omega}_{i}\), then we have
-\(h^{-1}\sum_{i=1}^{h}\sigma^{2}_{\alpha} = \frac{h \sigma^{2}_{\alpha}}{h} = \sigma^{2}_{\alpha}\).
-
-To elaborate on this concept some more, consider the following matrix
-equation of the variances and the nonredundant covariances in a
-three-wave intercept-only model that follows directly from Equation
-\eqref{eq:wide} (assuming
-\(\mathop{\mathrm{\mathrm{Cov}}}(\varepsilon_{ti},\varepsilon_{si}) = 0, \ t \ne s\)),
-and which we can solve easily with least squares: \begin{align}
-\bm{A}\bm{x} & = \bm{b} \\
-\begin{pmatrix}
-1 & 1 & 0 & 0 \\
-1 & 0 & 0 & 0 \\
-1 & 0 & 0 & 0 \\
-1 & 0 & 1 & 0 \\
-1 & 0 & 0 & 0 \\
-1 & 0 & 0 & 1
-\end{pmatrix}
-\begin{pmatrix}
-\psi \\
-\phi_{1} \\
-\phi_{2} \\
-\phi_{3}
-\end{pmatrix} & = 
-\begin{pmatrix}
-\mathop{\mathrm{\mathrm{Var}}}(y_{1}) \\
-\mathop{\mathrm{\mathrm{Cov}}}(y_{2},y_{1}) \\
-\mathop{\mathrm{\mathrm{Cov}}}(y_{3},y_{1}) \\
-\mathop{\mathrm{\mathrm{Var}}}(y_{2}) \\
-\mathop{\mathrm{\mathrm{Cov}}}(y_{3},y_{2}) \\
-\mathop{\mathrm{\mathrm{Var}}}(y_{3})
-\end{pmatrix}
-\end{align} where \(\psi = \mathop{\mathrm{\mathrm{Var}}}(\alpha)\),
-\(\phi_{t} = \mathop{\mathrm{\mathrm{Var}}}(\varepsilon_{t})\). We can
-solve this equation to show \begin{align}
-\bm{A}\bm{x} & = \bm{b} \\
-(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{A}\bm{x} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} \\
-\hat{\bm{x}} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} \\
-\begin{pmatrix}
-\hat{\psi} \\
-\hat{\phi}_{1} \\
-\hat{\phi}_{2} \\
-\hat{\phi}_{3}
-\end{pmatrix} & = 
-\begin{pmatrix}
-\frac{1}{3}\mathop{\mathrm{\mathrm{Cov}}}(y_{2},y_{1}) + \frac{1}{3}\mathop{\mathrm{\mathrm{Cov}}}(y_{3},y_{1}) + \frac{1}{3}\mathop{\mathrm{\mathrm{Cov}}}(y_{3},y_{2}) \\
-\mathop{\mathrm{\mathrm{Var}}}(y_{1}) - \hat{\psi} \\
-\mathop{\mathrm{\mathrm{Var}}}(y_{2}) - \hat{\psi} \\
-\mathop{\mathrm{\mathrm{Var}}}(y_{3}) - \hat{\psi}
-\end{pmatrix}.
-\end{align} So if, in fact the covariance between any two wide-format
-columns of \(y\) is
-\(\mathop{\mathrm{\mathrm{Cov}}}(y_{t},y_{s}) = \mathop{\mathrm{\mathrm{Var}}}(\alpha), \ \forall \ s \ne t\),
-then
-\(\hat{\psi} = \frac{1}{3}\mathop{\mathrm{\mathrm{Var}}}(\alpha) + \frac{1}{3}\mathop{\mathrm{\mathrm{Var}}}(\alpha) + \frac{1}{3}\mathop{\mathrm{\mathrm{Var}}}(\alpha) = \frac{3 \mathop{\mathrm{\mathrm{Var}}}(\alpha)}{3} = \mathop{\mathrm{\mathrm{Var}}}(\alpha)\).
-This shows that if our assumption about the underlying DGP is correct,
-i.e., \(y_{it} = \alpha_{i} + \varepsilon_{it}\) and
-\(\mathop{\mathrm{\mathrm{Cov}}}(\varepsilon_{t},\varepsilon_{s}) = 0, \ \forall \ t \ne s\),
-then the estimated variance of \(\alpha\) is just what it should be: the
-average covariance between units of \(y\) over time. Once we add in
-observed covariates, the estimated covariance of \(\alpha\) then become
-the \textit{conditional} covariance of \(y\) over time, given those
-covariates.
-
-\hypertarget{model-notation}{%
-\subsection{Model notation}\label{model-notation}}
-
-Having explained how a latent variable is used to estimate the
-individual effects, we need to state the FE model using SEM-compatible
-matrix notation. There are a number of different model notations (see,
-for example \citet{Bollen1989} for an overview), but the one that will
-serve us best is one that was proposed by \citet{Graff1979}:
-\begin{align}
-\bm{y}^{+} & = \bm{\Lambda_{y}}^{+} \bm{\eta}^{+}, \\
-\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+}, 
-\end{align} where
-\(\bm{\eta}^{+} = (\bm{y}, \bm{x}, \bm{\eta}, \bm{\xi})^{\intercal}\),
-\(\bm{\zeta}^{+} = (\bm{\varepsilon}, \bm{\delta}, \bm{\zeta}, \bm{\xi})^{\intercal}\),
-\(\bm{y}^{+} = (\bm{y}, \bm{x})^{\intercal}\). \(\bm{y}\) is a vector of
-observed dependent variables and \(\bm{x}\) is a vector of observed
-independent variables. \(\bm{\eta}\) is a vector of the latent dependent
-variables and \(\bm{\xi}\) is a vector of latent independent variables.
-\(\bm{\varepsilon}\) and \(\bm{\delta}\) are vectors of the errors of
-the observed dependent and independent variables, respectively, and
-\(\bm{\zeta}\) is a vector of the errors, or disturbances, of the latent
-variables. Notice the \(^{+}\) symbol is just meant to differentiate the
-vectors with them from those without them. That means, \(\bm{\eta}^{+}\)
-is a vector that holds the observed and latent variables, both dependent
-(in SEM they are referred to as `endogenous') and independent (i.e.,
-`exogeneous')\footnote{For our purposes, the terms endogenous and
-  dependent, on the one hand, and exogenous and independent, on the
-  other, can be used interchangeably.}, \(\bm{\zeta}^{+}\) holds the
-errors for the observed variables and the disturbances of the latent
-variables. \(\bm{y}^{+}\) holds just the observed variables, both
-dependent and independent, and \(\bm{\Lambda_{y}}^{+}\) is a matrix of
-ones and zeros that selects the observed variables from
-\(\bm{\eta}^{+}\). Lastly, \(\bm{B}\) is a matrix that holds the
-regression coefficients.
-
-If we say that \(p\) and \(q\) stand for the number of observed
-dependent and independent variables, respectively, and \(m\) and \(n\)
-stand for the number of latent dependent and independent variables,
-respectively, then \(\bm{\eta}^{+}\) and \(\bm{\zeta}^{+}\) are
-\(p + q + m + n\), \(\bm{y}^{+}\) is \(p + q\), \(\bm{\Lambda_{y}}^{+}\)
-is \((p + q) \times (p + q + m + n)\) and \(\bm{B}\) is
-\((p + q + m + n) \times (p + q + m + n)\) \citep{Bollen1989}.
-
-This notation may be confusing at first, but it has advantages. First,
-it allows us the flexibility we need for the models. For example, it
-allows observed \(\bm{x}\) to directly influence observed \(\bm{y}\)
-(more common notation assumes that substantive effects occur only
-between latent variables, observed ones are only used as indicators, see
-for example \citet{Bollen1989}, \citet{Kline2016}). It also allows
-\(\bm{\xi}\), i.e., any latent exogenous variables, to influence
-\(\bm{y}\) directly. These two scenarios cover the traditional FE model
-with observed variables, and one in which latent variables are used to
-account for measurement error in the independent variables. It is also
-consistent with the notation used for these models in \texttt{lavaan}.
-In fact, \texttt{lavaan} switches automatically between matrix notations
-depending on the specified model. That means the matrix representation
-of the model one sees if they type in
-\texttt{lavInspect(model,\ what\ =\ "est")} after specifying their model
-in \texttt{lavaan} will match up with the notation used here.\footnote{In
-  \texttt{Mplus}, the model matrices can be requested by including
-  \texttt{OUTPUT:\ TECH1} in the input file.} It does have a potential
-disadvantage however. Besides being less intuitive than the typical
-\(\bm{\eta} = \bm{B}\bm{\eta} + \bm{\Gamma}\bm{\xi} + \bm{\zeta}\)
-notation, it means that by including the observed covariates in the
-stacked long vector \(\bm{y}^{+}\), they are treated as another response
-variable with variances and covariances to be estimated by the model
-(instead of just using the sample statistics). This means that the
-assumption of multivariate normality (otherwise just imposed on the
-dependent variables) also applies to the independent ones
-\citep[p.~75]{Skrondal2004}. This can be problematic for noncontinuous
-independent variables like sex, nationality dummies, marriage status
-(married/unmarried), etc. See \citet{Skrondal2004} for more on this
-topic.
-
-\begin{figure}
-\begin{center}
-\resizebox{0.75\textwidth}{!}{%
-\begin{tikzpicture}
-% Node styles ---
-\tikzstyle{man} = [rectangle, thick, minimum size = 1cm, draw = black!80, fill = white!100, font = \sffamily] % Manifest variables
-\tikzstyle{lat} = [circle, thick, minimum size = 1cm, draw = black!80, fill = white!100, font = \sffamily] % Latent variables
-\tikzstyle{err} = [circle, draw = black!80, fill = white!100, font = \sffamily] % Errors 
-% Edge styles
-\tikzstyle{con} = [-latex, font = \sffamily] % Effects
-\tikzstyle{cons} = [-latex, font = \sffamily\small] % Effects with smaller font 
-\tikzstyle{cor} = [latex-latex, font = \sffamily] % Correlations
-% Begin Figure ---
-% Nodes 
-\node at (+0.0,+0.0) [man] (x1) {$x_{1}$};
-\node at (+2.0,+0.0) [man] (x2) {$x_{2}$};
-\node at (+6.0,+0.0) [man] (x3) {$x_{3}$};
-\node at (+4.0,+0.0) [lat] (eta) {$\alpha$};
-\node at (+0.0,-2.0) [man] (y1) {$y_{1}$};
-\node at (+2.0,-2.0) [man] (y2) {$y_{2}$};
-\node at (+6.0,-2.0) [man] (y3) {$y_{3}$};
-\node at (+0.0,-3.5) [err] (e1) {$\varepsilon_{1}$};
-\node at (+2.0,-3.5) [err] (e2) {$\varepsilon_{2}$};
-\node at (+6.0,-3.5) [err] (e3) {$\varepsilon_{3}$};
-% Paths
-\path (x1) edge [con, draw = black!100, left] node {$\beta$} (y1);
-\path (x2) edge [con, draw = black!100, left, near start] node {$\beta$} (y2);
-\path (x3) edge [con, draw = black!100, right] node {$\beta$} (y3);
-\path (e1) edge [con, draw = black!100, left] node {1} (y1);
-\path (e2) edge [con, draw = black!100, left] node {1} (y2);
-\path (e3) edge [con, draw = black!100, right] node {1} (y3);
-\path (eta) edge [con, draw = black!100, above, near start] node {1} (y1);
-\path (eta) edge [con, draw = black!100, below] node {1} (y2);
-\path (eta) edge [con, draw = black!100, above] node {1} (y3);
-% correlations
-\path (x1) edge [cor, bend left] node {} (x2);
-\path (x1) edge [cor, bend left] node {} (eta);
-\path (x1) edge [cor, bend left] node {} (x3);
-\path (x2) edge [cor, bend left] node {} (eta);
-\path (x2) edge [cor, bend left] node {} (x3);
-\path (x3) edge [cor, bend right] node {} (eta);
-\end{tikzpicture}
-}
-\caption{Typical three-wave FE-SEM model with contemporary effects\label{fig:fesem}}
-\end{center}
-\end{figure}
-
-Let us, however, make things more concrete and take a look at a simple,
-three-wave version of the typical FE-SEM using this notation (shown
-graphically in Figure \ref{fig:fesem}). For that, we have the following
-matrix notation (with labels on the outside of the matrices):
-\begin{align}
-\begin{split}
-\bm{y}^{+} & = \bm{\Lambda_{y}}^{+}\bm{\eta}^{+} \\
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3}
-\end{pmatrix} & = 
-\begin{blockarray}{cccccccc}
- & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
- \begin{block}{c(ccccccc)}
- y_{1} & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
- y_{2} & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
- y_{3} & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\ 
- x_{1} & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
- x_{2} & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
- x_{3} & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
- \end{block}
-\end{blockarray}
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix},
-\end{split} \label{eq:y}
-\end{align} \begin{align}
-\begin{split}
-\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+} \\
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix} & = 
-\begin{blockarray}{cccccccc}
- & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
- \begin{block}{c(ccccccc)}
- y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
- y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
- y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
- x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- \end{block}
-\end{blockarray}
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix} + 
-\begin{pmatrix}
-\varepsilon_{1} \\
-\varepsilon_{2} \\
-\varepsilon_{3} \\
-x_{1} = \delta_{1} \\
-x_{2} = \delta_{2} \\
-x_{3} = \delta_{3} \\
-\alpha = \xi \\
-\end{pmatrix}. 
-\end{split} \label{eq:eta}
-\end{align} Notice in \(\bm{\zeta}^{+}\), for the independent variables
-we could either write, for example \(x_{t}\) or \(\delta_{t}\). As
-mentioned above, this is due to the model notation treating the
-independent variables like dependent variables with
-variances/covariances to be estimated. For the sake of simplicity, we
-will ignore this subtlety and refer to the observed variable from now
-on, keeping in mind that if the multivariate normality assumption holds,
-the estimated statistics will likely be sufficiently close to the sample
-ones for it to not make much of a difference.
-
-Admittedly, Equations \eqref{eq:y} and \eqref{eq:eta} may not look like
-much yet. We can remedy this by first putting the equation for
-\(\bm{\eta}^{+}\) in reduced form, i.e., by getting rid of the dependent
-variable on the r.h.s.: \begin{align}
-\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+} \\
-\bm{\eta}^{+} - \bm{B}\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
-(\bm{I} - \bm{B})\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
-\bm{\eta}^{+} & = (\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+},
-\end{align} where \(\bm{I}\) is the identity matrix. By substituting
-this back into the equation for the observed variables we get
-\(\bm{y}^{+} = \bm{\Lambda_{y}}^{+}[(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+}]\),
-which works out to: \begin{align}
-\begin{pmatrix} 
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\end{pmatrix} & = 
-\begin{pmatrix}
-\alpha + \beta x_{1} + \varepsilon_{1} \\
-\alpha + \beta x_{2} + \varepsilon_{2} \\
-\alpha + \beta x_{3} + \varepsilon_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3}
-\end{pmatrix},
-\end{align} which is of course exactly what we should expect given
-Equation \eqref{eq:fe}\footnote{We can use the \texttt{sympy} package in
-  \texttt{python} to verify and show the steps for this and other
-  examples, see the
-  \href{https://github.com/henrik-andersen/FE-SEM/blob/master/sympy-doublecheck-matrixnotation.py}{supplementary
-  materials} for the code.}.
-
-\hypertarget{assumptions}{%
-\subsection{Assumptions}\label{assumptions}}
-
-What essentially differentiates an FE from an RE model is our assumption
-concerning the relationship between the unobserved individual effects
-and the model covariates \citep{Bollen2010}. The FE model assumes that
-\(\mathop{\mathrm{\mathbb{E}}}[\alpha x_{t}] \ne 0\). As such, if we
-fail to control for the correlation of the covariate and the
-time-invariant part of the error, then the coefficient of interest, here
-\(\beta\), will be biased. Our assumption regarding whether the
-individual effects are correlated with the model covariates occurs in
-\(\mathop{\mathrm{\mathbb{E}}}[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] = \bm{\Psi}\),
-the covariance matrix of the errors \begin{align}
-\bm{y}^{+}\bm{y}^{+ \intercal} & = \mathop{\mathrm{\mathbb{E}}}[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})^{\intercal}] \\
- & = \mathop{\mathrm{\mathbb{E}}}[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\zeta}^{+ \intercal}(\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal})] \\
- & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \mathop{\mathrm{\mathbb{E}}}[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal} \\
- & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \bm{\Psi} (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal}.
-\end{align} In the case of an FE model, \(\bm{\Psi}\) will reflect our
-belief that the individual effects are correlated with the model
-covariates, here again for demonstration the three-wave model:
-\begin{align}
-\bm{\Psi} & = \mathop{\mathrm{\mathbb{E}}}
-\begin{blockarray}{cccccccc}
- & \varepsilon_{1} & \varepsilon_{2} & \varepsilon_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
- \begin{block}{c(ccccccc)}
- \varepsilon_{1} & \varepsilon_{1}^{2} &                     &                     &              &              &              & \\
- \varepsilon_{2} & 0                   & \varepsilon_{2}^{2} &                     &              &              &              & \\
- \varepsilon_{3} & 0                   & 0                   & \varepsilon_{3}^{2} &              &              &              & \\
- x_{1}           & 0                   & 0                   & 0                   & x_{1}^{2}    &              &              & \\
- x_{2}           & 0                   & 0                   & 0                   & x_{2}x_{1}   & x_{2}^{2}    &              & \\
- x_{3}           & 0                   & 0                   & 0                   & x_{3}x_{1}   & x_{3}x_{2}   & x_{3}^{2}    & \\
- \alpha          & 0                   & 0                   & 0                   & \alpha x_{1} & \alpha x_{2} & \alpha x_{3} & \alpha^{2} \\
- \end{block}
-\end{blockarray}.
-\end{align} Knowing this, we can work out the equation for the
-coefficient of interest, \(\beta\). For the sake of simplicity, assume
-here and throughout mean-centered variables: \begin{align}
-\mathop{\mathrm{\mathrm{Cov}}}(y_{t},x_{t}) & = \mathop{\mathrm{\mathbb{E}}}[y_{t}x_{t}] \\
- & = \mathop{\mathrm{\mathbb{E}}}[(\alpha + \beta x_{t} + \varepsilon_{t})x_{t}] \\
- & = \mathop{\mathrm{\mathbb{E}}}[\alpha x_{t} + \beta x_{t}^{2} + \varepsilon_{t}x_{t}] \\
- & = \mathop{\mathrm{\mathrm{Cov}}}(\alpha, x_{t}) + \beta \mathop{\mathrm{\mathrm{Var}}}(x_{t}) \\
-\hat{\beta}_{FE-SEM} & = \frac{\mathop{\mathrm{\mathrm{Cov}}}(y_{t},x_{t}) - \mathop{\mathrm{\mathrm{Cov}}}(\alpha, x_{t})}{\mathop{\mathrm{\mathrm{Var}}}(x_{t})}. 
-\end{align} This should make intuitive sense. From the observed
-covariance between the dependent and the independent variable, we are
-partialling out the part that is due to the covariance between the
-independent variable and the individual effects per unit, and then
-dividing by the variance of the independent variable, as usual. For the
-RE model, we assume \(\mathop{\mathrm{\mathbb{E}}}[\alpha x_{t}] = 0\)
-and the equation reduces to
-\(\hat{\beta}_{RE-SEM} = \mathop{\mathrm{\mathrm{Cov}}}(y_{t},x_{t})/\mathop{\mathrm{\mathrm{Var}}}(x_{t})\).
-The rest of the model-implied covariance matrix results from
-\(\bm{y}^{+}\bm{y}^{+ \intercal}\).
-
-\hypertarget{fesem}{%
-\section{Fixed effects in lavaan}\label{fesem}}
-
-The package \texttt{lavaan} needs to be installed once with
-\texttt{install.packages("lavaan")}. To be able to use it, we need to
-load it for every new \texttt{R} session:
-
-\singlespacing
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{( lavaan)}
-\end{Highlighting}
-\end{Shaded}
-
-\doublespacing
-
-For users unfamiliar with \texttt{R}, SEM analyses can be carried out
-with almost no knowledge of the language. Typically, someone unfamiliar
-with \texttt{R} would prepare their data using some other statistical
-software, and then save the intended dataset as a \texttt{.csv},
-\texttt{.xlsx}, \texttt{.dta}, \texttt{.sav}, etc. file. The user must
-then import the data, preferably as a dataframe, and the rest occurs
-using the \texttt{lavaan} syntax.\footnote{There are many online
-  tutorials for importing data in various formats, see, for example some
-  from
-  \href{https://www.datacamp.com/community/tutorials/r-data-import-tutorial}{datacamp}
-  or
-  \href{https://www.statmethods.net/input/importingdata.html}{Quick-R},
-  or any of the many posts on
-  \href{https://stackoverflow.com/search?q=r+import+data}{stackoverflow}.}
-
-Specifying the most basic fixed effects model, like the one shown in
-\citet{Bollen2010} (the same model as Equation \eqref{eq:fe} but with
-just one time-varying predictor) involves four components. First, we
-define the latent individual effects variable using the
-\texttt{=\textasciitilde{}} `measured by' or `manifested by'
-\citep{R-lavaan} operator at the same time constraining the factor
-loadings at each timepoint to one. I will call the latent variable
-\texttt{a} to stand for \(\alpha\). Constraining all of the factor
-loadings to one reflects our implicit assumption that the combined
-effect of the unit-specific unobserved factors is constant over time.
-This is the default behaviour of traditional POLS-based approaches to FE
-that use the stacked long-format data.
-
-\singlespacing
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{a =}\ErrorTok{~}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y1 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y2 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y3 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y4 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y5}
-\end{Highlighting}
-\end{Shaded}
-
-\doublespacing
-
-Second, we regress the dependent variable on the independent variable
-using the \texttt{\textasciitilde{}} regression operator. With stacked,
-long-format data, only one regression coefficient is estimated over all
-observed timepoints. To have our FE-SEM model mimic this behaviour, we
-need to constrain the the estimated coefficient to equal over time. We
-do so by adding the same label to the regression coefficient at every
-time point. We will use the label \texttt{b} (this label was chosen
-arbitrarily, we could have used any letter or string of characters) and
-have it act as an equality constraint for the regression coefficient of
-interest \(\beta\):
-
-\singlespacing
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{y1 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x1}
-\NormalTok{y2 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x2 }
-\NormalTok{y3 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x3}
-\NormalTok{y4 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x4}
-\NormalTok{y5 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x5}
-\end{Highlighting}
-\end{Shaded}
-
-\doublespacing
-
-The key to a FE model, as opposed to an RE model are our assumptions
-about the relatedness of our covariate and the individual effects, i.e.,
-\(\mathop{\mathrm{\mathbb{E}}}[x_{t}\alpha]\). For an FE model, we want
-to partial out any potential covariance between the independent variable
-and the individual effects. This accounts for any linear relationship
-between \(x_{t}\) and the unit-specific characteristics influencing the
-dependent variable. Further, allowing unrestricted covariances between
-the independent variable itself over time will not affect how the
-coefficient \(\beta\) is estimated, but will have an effect on the
-standard errors. To mimic the behaviour of a conventional FE model, we
-allow the independent variable to be correlated with the individual
-effects and itself over time. Covariances (including covariances between
-a variable and itself, i.e., variances) are specified using the
-\texttt{\textasciitilde{}\textasciitilde{}} operator:
-
-\singlespacing
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{a }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x1 }\OperatorTok{+}\StringTok{ }\NormalTok{x2 }\OperatorTok{+}\StringTok{ }\NormalTok{x3 }\OperatorTok{+}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
-\NormalTok{x1 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x2 }\OperatorTok{+}\StringTok{ }\NormalTok{x3 }\OperatorTok{+}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
-\NormalTok{x2 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x3 }\OperatorTok{+}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
-\NormalTok{x3 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
-\NormalTok{x4 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x5}
-\end{Highlighting}
-\end{Shaded}
-
-\doublespacing
-
-The last component of our code involves the variances of the residuals.
-This component is optional, but we can constrain the residual variances
-to be equal over time to again mimic the behaviour of a conventional FE
-model using POLS on stacked data. Here, again, we use labels to make
-equality constraints. Because \(y_{t}\) is endogenous, the
-\texttt{\textasciitilde{}\textasciitilde{}} operator specifies the
-variances of \emph{residuals}, i.e., \(\varepsilon_{t}\).
-
-\singlespacing
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{y1 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y1}
-\NormalTok{y2 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y2}
-\NormalTok{y3 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y3}
-\NormalTok{y4 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y4}
-\NormalTok{y5 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y5}
-\end{Highlighting}
-\end{Shaded}
-
-\doublespacing
-
-\hypertarget{ex1}{%
-\section{A simulated example}\label{ex1}}
-
-\singlespacing
-
-\doublespacing
-
-To demonstrate the application of FE models in SEM, a dataset can be
-simulated that embodies the FE assumptions. Again, the code for data
-simulation can be found in the
-\href{https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R}{online
-supplementary materials}.
-
-To show that the latent individual effects variables represent the
-\emph{combined} effect of all time-invariant characteristics, the
-dependent variable will be influenced by two separate unit-specific
-variables, which we can call \(\alpha_{1}\) and \(\alpha_{2}\). We will
-construct the simulated data such that the independent variable is
-correlated with both of the time-invariant variables. This means that
-approaches that fail to account for this confounding influence, such as
-POLS or RE, will be biased.
-
-The wide-format equations for the data generating process can be
-described as: \begin{align}
-\bm{x_{t}} & = \bm{\alpha_{1}}\beta_{x_{t},\alpha_{1}} + \bm{\alpha_{2}}\beta_{x_{t},\alpha_{2}} + \bm{\delta_{t}}, \\
-\bm{y_{t}} & = \bm{x_{t}}\beta_{y_{t},x_{t}} + \bm{\alpha_{1}}\beta_{y_{t},\alpha_{1}} + \bm{\alpha_{2}} \beta_{y_{t},\alpha_{2}} + \bm{\varepsilon_{t}} 
-\end{align} where, for the sake of simplicity, \(\bm{\alpha_{1}}\),
-\(\bm{\alpha_{2}}\), \(\bm{\delta_{t}}\) and \(\bm{\varepsilon_{t}}\)
-are \(\sim N(0,1)\).
-
-For the following example, a sample size of 1,000, observed over five
-waves, was chosen. The unique variance of \(\bm{x}\), as well as both
-the individual-effect variables is also \(\sim N(0,1)\). The coefficient
-of interest, \(\beta_{y,x}\) is set to be equal to \(0.3\). A
-correlation between \(\bm{x}\) and the individual effects is induced
-through \(\beta_{x,\alpha_{1}} = 0.85\) and
-\(\beta_{x,\alpha_{1}} = 0.50\). With the variances above set to one,
-the covariances will be roughly
-\(\mathop{\mathrm{\mathrm{Cov}}}(x_{t},\alpha_{1}) = 0.85\) and
-\(\mathop{\mathrm{\mathrm{Cov}}}(x_{t},\alpha_{2}) = 0.5\). The
-dependent variable is also influenced by the individual effects
-variables with \(\beta_{y_{t},\alpha_{1}} = 0.75\) and
-\(\beta_{y_{t},\alpha_{2}} = 0.45\) These values were chosen
-arbitrarily.
-
-Now, we run the FE-SEM in \texttt{lavaan}.
-
-\singlespacing
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{fe_sem <-}\StringTok{ '}
-\StringTok{# Define individual effects variable }
-\StringTok{a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5}
-\StringTok{# Regressions, constrain coefficient to be equal over time}
-\StringTok{y1 ~ b*x1}
-\StringTok{y2 ~ b*x2 }
-\StringTok{y3 ~ b*x3}
-\StringTok{y4 ~ b*x4}
-\StringTok{y5 ~ b*x5}
-\StringTok{# Allow unrestricted correlation between eta and covariates}
-\StringTok{a ~~ x1 + x2 + x3 + x4 + x5}
-\StringTok{x1 ~~ x2 + x3 + x4 + x5}
-\StringTok{x2 ~~ x3 + x4 + x5}
-\StringTok{x3 ~~ x4 + x5}
-\StringTok{x4 ~~ x5}
-\StringTok{# Constrain residual variances to be equal over time}
-\StringTok{y1 ~~ e*y1}
-\StringTok{y2 ~~ e*y2}
-\StringTok{y3 ~~ e*y3}
-\StringTok{y4 ~~ e*y4}
-\StringTok{y5 ~~ e*y5}
-\StringTok{'}
-\NormalTok{fe_sem.fit <-}\StringTok{ }\KeywordTok{sem}\NormalTok{(}\DataTypeTok{model =}\NormalTok{ fe_sem, }
-                  \DataTypeTok{data =}\NormalTok{ dfw, }
-                  \DataTypeTok{estimator =} \StringTok{"ML"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\doublespacing
-
-\singlespacing
-
-\doublespacing
-
-We can get a summary of the model with \texttt{summary()}. The first
-portion of the summary output gives an overview of some basic
-information and fit statistics. The maximum likelihood estimator is the
-default, so it did not have to be explicitly selected in the fitting
-function call. Other estimators are available, including generalized and
-unweighted least squares (\texttt{GLS} and \texttt{ULS}, respectively),
-robust standard errors maximum likelihood (\texttt{MLM}) and several
-others (see \href{https://lavaan.ugent.be/tutorial/est.html}{the lavaan
-online tutorial for more}).
-
-This part of the summary output also tells us that the analysis is based
-on 1,000 observations (missings would be shown here as well if there
-were any), and that the \(\chi^{2}\) statistic is 30.138 based on 32
-degrees of freedom (55 observed covariances minus 1 error variance, 1
-coefficient, 1 latent variable variance, 5 exogenous variable variances
-and 15 covariances for \(55 - 23 = 32\) df). The p-value on the
-\(\chi^{2}\) statistic is not significant with \(p =\) 0.561 which tells
-us the differences between the model-implied and observed covariance
-matrices are likely due to chance, and that the model fits the data well
-(given how the data was generated, it would be surprising if this were
-not the case). Other fit measures including typical comparative fit
-indices can be requested by either adding \texttt{fit.measures\ =\ TRUE}
-as a secondary argument to the \texttt{summary()} call, or by asking for
-a complete list of all available fit statistics using
-\texttt{lavInspect(model,\ "fit")} where \texttt{model} stands for the
-name of the fitted model, in this case \texttt{fe\_sem.fit}.
-
-\singlespacing
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{summary}\NormalTok{(fe_sem.fit)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## lavaan 0.6-6 ended normally after 37 iterations
-## 
-##   Estimator                                         ML
-##   Optimization method                           NLMINB
-##   Number of free parameters                         31
-##   Number of equality constraints                     8
-##                                                       
-##   Number of observations                          1000
-##                                                       
-## Model Test User Model:
-##                                                       
-##   Test statistic                                30.138
-##   Degrees of freedom                                32
-##   P-value (Chi-square)                           0.561
-## 
-## Parameter Estimates:
-## 
-##   Standard errors                             Standard
-##   Information                                 Expected
-##   Information saturated (h1) model          Structured
-...
-\end{verbatim}
-
-\doublespacing
-
-Next the summary output shows the measurement models for the latent
-variables, if any. In this case the latent variable \texttt{a} for
-\(\alpha\) is measured by each of the five observed dependent variables
-with factor loadings fixed to 1.0.
-
-\singlespacing
-
-\begin{verbatim}
-...
-## Latent Variables:
-##                    Estimate  Std.Err  z-value  P(>|z|)
-##   a =~                                                
-##     y1                1.000                           
-##     y2                1.000                           
-##     y3                1.000                           
-##     y4                1.000                           
-##     y5                1.000                           
-...
-\end{verbatim}
-
-\doublespacing
-
-The regressions are shown next. Here, because we have constrained the
-regression coefficients to be equal over time (the equality constraint
-label \texttt{(b)} is listed to the left of the estimates), the estimate
-of \(\beta =\) 0.294 (0.016) is repeated five times. The corresponding
-z- and p-values show that the coefficient is, unsurprisingly,
-significant.
-
-\singlespacing
-
-\begin{verbatim}
-...
-## Regressions:
-##                    Estimate  Std.Err  z-value  P(>|z|)
-##   y1 ~                                                
-##     x1         (b)    0.294    0.016   18.809    0.000
-##   y2 ~                                                
-##     x2         (b)    0.294    0.016   18.809    0.000
-##   y3 ~                                                
-##     x3         (b)    0.294    0.016   18.809    0.000
-##   y4 ~                                                
-##     x4         (b)    0.294    0.016   18.809    0.000
-##   y5 ~                                                
-##     x5         (b)    0.294    0.016   18.809    0.000
-...
-\end{verbatim}
-
-\doublespacing
-
-Next, the covariance estimates are listed. First, the covariances
-between the latent individual effects variable and the independent
-variable over time are shown, and then the covariances between the
-independent variable with itself over time.
-
-One should always take care to double-check that there are no unintended
-covariances listed here. Like \texttt{Mplus}, the \texttt{lavaan}
-package estimates some covariances per default, without the user
-explicitly having to add them to the model syntax. For example,
-covariances between latent variables are estimated per default. If one
-does not wish for them to covary, it must be explicitly stated, e.g.,
-with \texttt{f1\ \textasciitilde{}\textasciitilde{}\ 0*f2}, assuming the
-latent variables are called \texttt{f1} and \texttt{f2}, or by
-overriding the default behaviour for the entire model by adding
-\texttt{orthogonal\ =\ TRUE} (which sets the correlation between all
-latent variables to zero) to the fitting call.\footnote{This is at least
-  the current behaviour of both the \texttt{cfa} and \texttt{sem}
-  wrappers. In fact, both wrappers seem to be identical in terms of the
-  default settings, see \citet{Rosseel2020}.}
-
-\singlespacing
-
-\begin{verbatim}
-...
-## Covariances:
-##                    Estimate  Std.Err  z-value  P(>|z|)
-##   a ~~                                                
-##     x1                0.844    0.055   15.355    0.000
-##     x2                0.867    0.056   15.441    0.000
-##     x3                0.845    0.055   15.400    0.000
-##     x4                0.822    0.053   15.455    0.000
-##     x5                0.820    0.053   15.572    0.000
-##   x1 ~~                                               
-##     x2                0.908    0.070   12.900    0.000
-##     x3                0.935    0.069   13.466    0.000
-##     x4                0.921    0.067   13.661    0.000
-##     x5                0.914    0.067   13.716    0.000
-##   x2 ~~                                               
-##     x3                0.889    0.070   12.675    0.000
-##     x4                0.922    0.069   13.423    0.000
-##     x5                0.889    0.068   13.165    0.000
-##   x3 ~~                                               
-##     x4                0.865    0.067   12.976    0.000
-##     x5                0.901    0.066   13.554    0.000
-##   x4 ~~                                               
-##     x5                0.850    0.064   13.285    0.000
-...
-\end{verbatim}
-
-\doublespacing
-
-Finally, the variance estimates are listed. Here, we see that in order
-to mimic the behaviour of a traditional FE model, the error variances
-over time were specified to be equal using the equality constraint
-\texttt{(e)}. Notice the \texttt{.} beside \texttt{y1}, \texttt{y2},
-etc.: this indicates that the listed variance refers to an endogenous
-variable, and that it is thus an error variance. In this case, these
-refer to the variances of \(\varepsilon_{t}\). After that, the variances
-of the exogenous variables, both observed and unobserved are listed.
-
-\singlespacing
-
-\begin{verbatim}
-...
-## Variances:
-##                    Estimate  Std.Err  z-value  P(>|z|)
-##    .y1         (e)    1.022    0.023   44.721    0.000
-##    .y2         (e)    1.022    0.023   44.721    0.000
-##    .y3         (e)    1.022    0.023   44.721    0.000
-##    .y4         (e)    1.022    0.023   44.721    0.000
-##    .y5         (e)    1.022    0.023   44.721    0.000
-##     x1                1.986    0.089   22.361    0.000
-##     x2                2.079    0.093   22.361    0.000
-##     x3                1.987    0.089   22.361    0.000
-##     x4                1.860    0.083   22.361    0.000
-##     x5                1.814    0.081   22.361    0.000
-##     a                 0.799    0.052   15.310    0.000
-\end{verbatim}
-
-\doublespacing
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-Fixed effects regression in SEM has been outlined in well-known articles
-by \citep{Allison2011, Bollen2010, Teachman2001}. This article provides
-a focused look at the implementation of the basic model using the
-\texttt{lavaan} package in \texttt{R}. The
-\href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{online
-supplementary materials} further discuss common extensions and some
-tools for evaluating and loosening model assumptions.
-
-The benefits of FE-SEM as opposed to traditional OLS-based FE-models are
-largely the same ones that apply to the SEM framework in general: for
-one, SEM allows for a great deal of flexibility. For example, it is easy
-to loosen model constraints as necessary. Measurement error in both the
-dependent and independent variables can be dealt with using latent
-variables to achieve unbiased and more efficient results. Researchers
-interested in time-invariant predictors can integrate them into a hybrid
-FE/RE model with ease. Further extensions, like measurement invariance
-testing \citep{Schoot2012, Millsap2011, Steenkamp1998} as well as lagged
-dependent variables \citep{Bollen2010, Allison2017} for example, can
-also be implemented in a straightforward fashion.
-
-The most basic FE-SEM is furthermore the basis for a variety of
-currently popular extended models, such as Latent Curve Models in
-general \citep{Curran2001, Bollen2004}, as well as special
-implementations like the Dynamic Panel Model \citep{Allison2017}, the
-Random-Intercept Cross-Lagged Panel Model \citep{Hamaker2015} and the
-Latent Curve Model with Structured Residuals \citep{Curran2014}. For
-this reason it is all the more important for researchers to have a good
-grasp on the method of applying panel regression in SEM, and
-understanding the intuition of controlling for time-invariant
-confounders. This article is meant to serve as a consolidated resource
-for researchers looking for concrete advise on specifying FE and more
-general panel models in SEM.
-
 \bibliographystyle{tfcad}

### 2021-01-14 article_format/article_format.tex
Started major changes for resubmission

@@
 
+% Pandoc citation processing
+
 \usepackage{hyperref}
@@
 \author{\name{Henrik Kenneth Andersen$^{a}$}
-\affil{$^{a}$Chemnitz University of Technology, Institute of Sociology, Chair for
-Empirical Social Research, Thüringer Weg 9, 09126 Chemnitz, Germany}
+\affil{$^{a}$Chemnitz University of Technology, Institute of Sociology,
+Chair for Empirical Social Research, Thüringer Weg 9, 09126 Chemnitz,
+Germany}
 }
 
-\thanks{CONTACT Henrik Kenneth Andersen. Email: \href{mailto:henrik.andersen@soziologie.tu-chemnitz.de}{\nolinkurl{henrik.andersen@soziologie.tu-chemnitz.de}}}
+\thanks{CONTACT Henrik Kenneth
+Andersen. Email: \href{mailto:henrik.andersen@soziologie.tu-chemnitz.de}{\nolinkurl{henrik.andersen@soziologie.tu-chemnitz.de}}}
 
@@
 fixed to one to represent our assumption that the effect of the
-time-invariant unobserved heterogeneity is constant over time.\footnote{In
-  fact, the initial FE-SEM setup shown in the main article mimics the
-  POLS methods described above in that it assumes constant effects and
-  error variances over time. These assumptions can be loosened and
-  tested, as will be shown in the
+time-invariant unobserved heterogeneity is constant over
+time.\footnote{In fact, the initial FE-SEM setup shown in the main
+  article mimics the POLS methods described above in that it assumes
+  constant effects and error variances over time. These assumptions can
+  be loosened and tested, as will be shown in the
   \href{https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf}{supplementary
@@
 \begin{Highlighting}[]
-\NormalTok{a =}\ErrorTok{~}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y1 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y2 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y3 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y4 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y5}
+\NormalTok{a =}\ErrorTok{\textasciitilde{}}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y1 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y2 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y3 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y4 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y5}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{y1 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x1}
-\NormalTok{y2 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x2 }
-\NormalTok{y3 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x3}
-\NormalTok{y4 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x4}
-\NormalTok{y5 }\OperatorTok{~}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x5}
+\NormalTok{y1 }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x1}
+\NormalTok{y2 }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x2 }
+\NormalTok{y3 }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x3}
+\NormalTok{y4 }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x4}
+\NormalTok{y5 }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x5}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{a }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x1 }\OperatorTok{+}\StringTok{ }\NormalTok{x2 }\OperatorTok{+}\StringTok{ }\NormalTok{x3 }\OperatorTok{+}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
-\NormalTok{x1 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x2 }\OperatorTok{+}\StringTok{ }\NormalTok{x3 }\OperatorTok{+}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
-\NormalTok{x2 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x3 }\OperatorTok{+}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
-\NormalTok{x3 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
-\NormalTok{x4 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{x5}
+\NormalTok{a }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{x1 }\OperatorTok{+}\StringTok{ }\NormalTok{x2 }\OperatorTok{+}\StringTok{ }\NormalTok{x3 }\OperatorTok{+}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
+\NormalTok{x1 }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{x2 }\OperatorTok{+}\StringTok{ }\NormalTok{x3 }\OperatorTok{+}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
+\NormalTok{x2 }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{x3 }\OperatorTok{+}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
+\NormalTok{x3 }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
+\NormalTok{x4 }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{x5}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{y1 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y1}
-\NormalTok{y2 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y2}
-\NormalTok{y3 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y3}
-\NormalTok{y4 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y4}
-\NormalTok{y5 }\OperatorTok{~}\ErrorTok{~}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y5}
+\NormalTok{y1 }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y1}
+\NormalTok{y2 }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y2}
+\NormalTok{y3 }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y3}
+\NormalTok{y4 }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y4}
+\NormalTok{y5 }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y5}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{fe_sem <-}\StringTok{ '}
-\StringTok{# Define individual effects variable }
-\StringTok{a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5}
-\StringTok{# Regressions, constrain coefficient to be equal over time}
-\StringTok{y1 ~ b*x1}
-\StringTok{y2 ~ b*x2 }
-\StringTok{y3 ~ b*x3}
-\StringTok{y4 ~ b*x4}
-\StringTok{y5 ~ b*x5}
-\StringTok{# Allow unrestricted correlation between eta and covariates}
-\StringTok{a ~~ x1 + x2 + x3 + x4 + x5}
-\StringTok{x1 ~~ x2 + x3 + x4 + x5}
-\StringTok{x2 ~~ x3 + x4 + x5}
-\StringTok{x3 ~~ x4 + x5}
-\StringTok{x4 ~~ x5}
-\StringTok{# Constrain residual variances to be equal over time}
-\StringTok{y1 ~~ e*y1}
-\StringTok{y2 ~~ e*y2}
-\StringTok{y3 ~~ e*y3}
-\StringTok{y4 ~~ e*y4}
-\StringTok{y5 ~~ e*y5}
-\StringTok{'}
-\NormalTok{fe_sem.fit <-}\StringTok{ }\KeywordTok{sem}\NormalTok{(}\DataTypeTok{model =}\NormalTok{ fe_sem, }
+\NormalTok{fe\_sem \textless{}{-}}\StringTok{ \textquotesingle{}}
+\StringTok{\# Define individual effects variable }
+\StringTok{a =\textasciitilde{} 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5}
+\StringTok{\# Regressions, constrain coefficient to be equal over time}
+\StringTok{y1 \textasciitilde{} b*x1}
+\StringTok{y2 \textasciitilde{} b*x2 }
+\StringTok{y3 \textasciitilde{} b*x3}
+\StringTok{y4 \textasciitilde{} b*x4}
+\StringTok{y5 \textasciitilde{} b*x5}
+\StringTok{\# Allow unrestricted correlation between eta and covariates}
+\StringTok{a \textasciitilde{}\textasciitilde{} x1 + x2 + x3 + x4 + x5}
+\StringTok{x1 \textasciitilde{}\textasciitilde{} x2 + x3 + x4 + x5}
+\StringTok{x2 \textasciitilde{}\textasciitilde{} x3 + x4 + x5}
+\StringTok{x3 \textasciitilde{}\textasciitilde{} x4 + x5}
+\StringTok{x4 \textasciitilde{}\textasciitilde{} x5}
+\StringTok{\# Constrain residual variances to be equal over time}
+\StringTok{y1 \textasciitilde{}\textasciitilde{} e*y1}
+\StringTok{y2 \textasciitilde{}\textasciitilde{} e*y2}
+\StringTok{y3 \textasciitilde{}\textasciitilde{} e*y3}
+\StringTok{y4 \textasciitilde{}\textasciitilde{} e*y4}
+\StringTok{y5 \textasciitilde{}\textasciitilde{} e*y5}
+\StringTok{\textquotesingle{}}
+\NormalTok{fe\_sem.fit \textless{}{-}}\StringTok{ }\KeywordTok{sem}\NormalTok{(}\DataTypeTok{model =}\NormalTok{ fe\_sem, }
                   \DataTypeTok{data =}\NormalTok{ dfw, }
@@
 \begin{Highlighting}[]
-\KeywordTok{summary}\NormalTok{(fe_sem.fit)}
+\KeywordTok{summary}\NormalTok{(fe\_sem.fit)}
 \end{Highlighting}
@@
 \begin{verbatim}
-## lavaan 0.6-6 ended normally after 37 iterations
+## lavaan 0.6-7 ended normally after 37 iterations
 ## 

### 2021-01-14 article_format/article_format_revision.Rmd
Started major changes for resubmission

--- 2020-07-28 article_format/article_format.Rmd
@@
 title: |
-  A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}
-#type: ARTICLE TEMPLATE
+  A closer look at random and fixed effects panel regression in structural equation modeling using \texttt{lavaan}
+type: TEACHER'S CORNER
 author:
@@
 abstract: |
-  This article provides an in-depth look at fixed effects regression in the structural equation modeling (SEM) framework, specifically the application of fixed effects in the `lavaan` package for `R`.
+  This article provides an in-depth look at random and fixed effects panel regression in the structural equation modeling (SEM) framework, specifically the application of fixed effects in the `lavaan` package for `R`.
 It is meant as a applied guide for researchers, covering the underlying model specification, syntax, and summary output.
-Online supplementary materials further discuss various common extensions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of hybrid fixed-/ random effects model.
 keywords: |
-  Fixed effects, structural equation modeling, lavaan, R, panel analysis
+  Fixed effects, structural equation modeling, lavaan, R, panel regression, longitudinal data 
 header-includes: |
@@
 The part that does not change over time can be seen as the combined effect of all time-invariant influences (e.g., sex, date of birth, nationality) on the dependent variable.
-Fixed effects (FE) regression involves controlling for these time-invariant influences via a number of various methods.
-It thus accounts for a likely and common source of bias.
+Random effects (RE) and fixed effects (FE) regression involve accounting for these often unobserved time-invariant influences via a number of methods.
+In the case of RE models, it is assumed that the stable characteristics are unrelated to the other model covariates.
+In that case, the stable characteristics, or individual effects as they are often referred to as, will not affect point estimates but are a source of error variance that, if ignored, lead to larger standard errors and increase the likelihood of type II errors (failure to reject a false null hypothesis).
+FE models are useful when the individual effects are expected to be related to one or more of the model covariates.
+In that case, these unobserved stable characteristics will act as confounders and lead to biased point estimates.
+In controlling for these individual effects, FE regression thus accounts for a likely and common source of bias.
 
@@
 Thus researchers can allow for and study effects that increase or fade over time, or that appear only in specific periods.
-Beyond that, with the use of latent variables, SEM provides a way to deal with measurement error and get closer to the true underlying constructs of interest.
+Beyond that, with the use of latent variables, SEM provides a way to deal with measurement error and get closer to the theoretical constructs of interest.
 
-There are a number of articles describing basic concept of panel model regression, and FE regression in SEM [e.g., @Allison2011; @Bollen2010; @Teachman2001].
-This article is intended as a \textit{practical guide} for researchers looking for in-depth help with specifying FE models in SEM.
-It focuses on the `lavaan` [@R-lavaan] package for `R` [@R-base].
+There are a number of articles describing basic concept of panel model regression, including RE and FE regression in SEM [e.g., @Allison2011; @Bollen2010; @Teachman2001].
+This article is intended as a \textit{practical guide} for researchers looking for help specifying panel regression models in SEM.
+It assumes some basic knowledge of SEM (i.e., minimizing the difference between observed and model-implied covariance matrices and mean vectors, maximum likelihood estimation, etc.);^[For an introduction to SEM, @Bollen1989 is a classic for good reason, and @Ferron2007 lay out exactly how SEM works in one easy to follow article.] perhaps the reader is an experienced SEM-user looking for a quick guide to estimating the static panel models discussed here; or, perhaps the reader is new to SEM, potentially because an adviser or superior has suggested it to them.
+For the latter case, this article reviews the assumptions associated with the more tradititonal least squares-based approach to RE and FE models, so that the implementation in SEM is more easily to follow.
+The article focuses on the `lavaan` [@R-lavaan] package for `R` [@R-base].
 While `Mplus` [@Mplus] is arguably the most robust SEM software currently available (in terms of features like alignment, latent variable interactions, for example), the `lavaan` package has many benefits.
@@
 The following article outlines the basic idea of panel regression, the particularities of panel regression SEM, and shows its implementation in `lavaan`.
-The focus on FE panel regression stems from the fact that the assumptions needed to justify other common models, e.g., random effects, are often implausible.
-As we will see, the practical differences between the two are, in any event, very small.
 Using [simulated data](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R), it demonstrates and annotates the code for the most basic FE model and provides an overview of the summary output.
@@
 
-# Panel models {#panel}
+# Random and Fixed Effects Panel models {#panel}
 
-To begin, let us start by reviewing a general panel model [@Bollen2010], also referred to as the 'unobserved effects model' [@Wooldridge2012; @R-plm_a] (we will return to this model in the [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf) when we discuss loosening assumptions)
+It is typically the case that the values for a given unit on a variable at one point in time will tend to tell us something about that unit's values at another point in time.
+There are two main explanations for this 'empirical regularity' [@Heckman1981; @Hsiao2014, p.
+261; @Bruederl2015; @Bianconcini2018].
+First, it could be that an experience at one point in time has a causal^[Here the word causal is used to mean 'non-spurious'.] effect on future experiences.
+In other words, experiencing an event could change the probability of the same or a similar event taking place in the future.
+For example, if employment increases wages, then the incentive to continue working should increase over time, thereby making it more likely that someone who was employed at one point in time will continue to be employed in the future [@Heckman1981].
+When past experiences causally impact future events, the empirical regularity is referred to as 'state dependence'.
+So-called 'dynamic' panel models that include the lagged dependent variable in the equation for the current dependent variable, like autoregressive and autoregressive cross-lagged models, are examples of panel models that account for state dependence.
+The second explanation is that correlations over time are due to stable unobservables, like sex, place of birth, motivation, ability or personality characteristics.^[From here on out the units of interest are assumed to be individuals, but other units of observation like schools, companies, countries, etc., are possible.] In other words, stable unit-specific characteristics might predispose individuals to experiencing events with a certain likelihood over time.
+For example, stable characteristics like an individual's sex and motivation could be part of the reason why some individuals tend to be continuously employed, while others experience spells of unemployment, or are habitually unemployed.
+This second source of empirical regularity is referred to as individual heterogeneity or, more generally, unobserved heterogeneity [@Wooldridge2012].
+So-called 'static' panel models (that do not include the lagged dependent variable in the equation for the current one) focus on this second source of empirical regularity.
+
+The random and fixed effects models that will be discussed here are examples of models that attempt to account for unobserved heterogeneity as a source of empirical regularity.
+In the case of both the random and fixed effects models, it is assumed that there exist stable individual characteristics that affect the dependent variable of interest over time.
+Failure to account for this source of stable variance will reduce model fit.
+If any of the stable characteristics are related to the model covariates (a plausible scenario in many cases where the independent variable(s) of interest are not fixed), then they must also be considered confounders, and failure to account for them will further lead to biased point estimates.^[Whether or not the stable individual characteristics are related to the model covariates is the modern differentiator between random effects, where the individual effects are assumed independent of the covariates, and fixed effects, where the individual effects are assumed to be related to one or more of them, see @Wooldridge2012, p.
+251.
+This can be a cause for confusion for those who are more familiar with the mixed or multilevel model framework.
+There, 'fixed' effects are those coefficients that apply to the entire sample, 'random' effects are the variance components for the higher level grouping variables, see for example @Hox2010; @Bates2015; @R-lme4.]
+
+In this section, the basic random and fixed effects models, as well as their implementation in SEM, will be discussed.
+Later on in the article, it will be discussed how to extend the model, by relaxing some assumptions and introducing other time-invariant predictors in a type of 'hybrid' random and fixed effects model.
+The outlook will describe further extensions, such as the inclusion of lagged dependent variable for a dynamic panel model with individual effects.
+Let us begin with a simple linear 'unobserved effects model' [@Wooldridge2012; @R-plm_a]
+\begin{align}
+y_{it} & = \beta_{0} + \beta_{1} x_{it} + \alpha_{i} + \varepsilon_{it}, \ i = 1, \ldots, N, \ t = 1, \ldots, T \label{eq:panel-model}
+\end{align}
+where $y_{it}$ is the dependent variable for unit $i$ at time $t$ and $x_{it}$ is a single covariate with associated scalar coefficient $\beta_{1}$.
+$\alpha_{i}$ represents the combined effect of all unobserved time-constant variables affecting the dependent variable and $\varepsilon_{it}$ is the idiosyncratic error.
+This model could of course be extended by turning $x_{it}$ into a vector of covariates $\bm{x}_{it} = (x_{1_{it}}, x_{2_{it}}, \ldots, x_{k_{it}})$, or by introducing time-invariant covariates, $\bm{z}_{i} = (z_{1_{i}}, z_{2_{i}}, \ldots, z_{m_{i}})$, but we want to first focus on assumptions and mechanics before moving on to discussing possibilities for extending the model.
+
+Equation \eqref{eq:panel-model} is often written in matrix form for unit $i$ as 
+\begin{align}
+\bm{y}_{i} & = \bm{\iota}_{T}\beta_{0} + \bm{x}_{i}\beta_{1} + \bm{\iota}_{T} \alpha_{i} + \bm{\varepsilon}_{i}, \ i = 1, \ldots, N \label{eq:matrix-panel-model}
+\end{align}
+where $\bm{y}_{i} = (y_{i1}, y_{i2}, \ldots, y_{iT})^{\intercal}$ and $\bm{x}_{i} = (x_{i1}, x_{i2}, \ldots, x_{iT})^{\intercal}$ are $T \times 1$ vectors of each of the $t$ observations for unit $i$, $\bm{\iota}_{T}$ is a $T \times 1$ vector of ones and $\bm{\varepsilon}_{i}  = (\varepsilon_{i1}, \varepsilon_{i2}, \ldots, \varepsilon_{iT})^{\intercal}$ is likewise a $T \times 1$ vector of errors.
+It may be helpful to keep this Equation \eqref{eq:matrix-panel-model} in mind later when we restate the model for estimation using SEM.
+There, we will take the long-format data and transform it into wide-format.
+
+## Assumptions in least squares-based approaches
+
+One of the main strengths of SEM is the flexibility it affords.
+Nearly every part of the model is specified 'by hand', so to speak (although modern SEM software has come a long way from, say, `EQS` [@Bentler2006], where the errors and disturbances had to be mentioned in the syntax and now comes with a variety of default settings for ease of use), which means model assumptions are conveyed directly, by the user specifying paths and correlations and setting other relations to zero.
+Common least squares-based statistical packages [like the `plm` package in `R`, @R-plm_a] require less input from the user (usually it is enough to input the model variables and specify some top-level options).
+The assumptions in the least squares-based models are widely known [@Wooldridge2012; @Bruederl2015; @Greene2012; @Angrist2009] but do not come into play prominently while specifying the model.
+So here we will discuss the 'typical' assumptions associated with least squares-based random and fixed effects models.
+Later, when we examine the implementation in SEM, we will then be able to relate back to this discussion.^[What constitutes a 'typical' least squares-based approach to these panel regression models is likely debatable.
+Here, this refers broadly to the assumptions discussed in the literature and default settings one encounters in statistical packages intended for this purpose.]
+
+We want to relate the SEM-based approach to random and fixed effects to the least squares-based approach which is likely more widely known and more straightforward to implement, i.e., statistical packages for static panel regression models exist (e.g., the `plm` package in `R`, see @R-plm_a), and typically only require the user to enter the model variables and specify some top-level options (random effects vs.
+fixed effects, individual vs.
+individual and time effects, etc.).
+The model assumptions for random and fixed effects models are widely known [@Wooldridge2012; @Bruederl2015; @Greene2012; @Angrist2009] but are not prominent in th
+
+The least squares-based approach to such a panel model typically makes the following base assumptions:^[What constitutes a 'typical' least squares-based approach to these panel regression models is likely debatable.
+Here, this refers broadly to the default assumptions and settings one encounters in statistical packages intended for this purpose, like the `plm` package in `R`, see @R-plm_a.] first, the mean function is linear, i.e., $\E(y_{it} | x_{it}, \alpha_{i}) = \beta_{0} + \beta_{1} x_{it} + \alpha_{i}$, and second, that the cross-sectional observations are independent, which is assured by random sampling [@Wooldridge2012].
+Further, and just as with any other regression model, the idiosyncratic error is assumed to be mean independent of the model covariates and the individual effects, i.e., $\E(\varepsilon_{it} | x_{it}, \alpha_{i}) = \E(\varepsilon_{it}) = 0$.^[If there is an overall intercept, $\beta_{0}$, then it is safe to assume the unconditional expectation of the idiosyncratic error is zero, i.e., $\E(\varepsilon_{it}) = 0$.] The mean independence assumption also implies the assumption of a zero correlation between the idiosyncratic error and the model covariates, i.e., $\E(x_{it}\varepsilon_{it}) = 0$.
+
+The (modern) difference between random and fixed effects boils down to the assumption of the relatedness of the model covariates and the individual effects.
+The random effects model assumes the model covariates are unrelated to the individual effects, i.e., in this case $\E(\alpha_{i} | x_{it}) = \E(\alpha_{i})$ and thus $\E(\alpha_{i}x_{it}) = 0$.
+As such, we could put $\alpha_{i}$ into the error term to form the composite error, $\nu_{it} = \alpha_{i} + \varepsilon_{it}$, and the point estimates of the coefficients of interest, here $\beta_{1}$, would be consistent (assuming the model is otherwise correctly specified).
+The problem with putting the individual effects into the error term is that it would induce dependency between the compostie errors within the same individual, which goes against the independence assumption [@Wooldridge2012; @Schmidheiny2019].
+We will discuss that in more detail below.
+In the fixed effects model, one or more of the model covariates are assumed to be related to the individual effects, here $\E(\alpha_{i} | x_{it}) \ne \E(\alpha_{i})$ and thus $\E(\alpha_{i}x_{it}) \ne 0$.
+Thus, ignoring the individual heterogeneity and putting $\alpha_{i}$ into the error term would mean that point estimates of the coefficients of interest would be inconsistent.
+
+The traditional least squares-based approaches to random and fixed effects often assume as well that the errors are homoskedastictic over time, i.e., $\E(\varepsilon_{it}^{2}) = \sigma_{\varepsilon_{t}}^{2} = \sigma_{\varepsilon}^{2}$.
+The individual effects are assumed to have a constant partial effect on the dependent variable over time [@Wooldridge2012, p.
+248].
+Also, typically least squares-based approaches stack each of the $N$ repeated measures per individual $i$ and calculate one single coefficient per variable for the entire sample.
+I.e., it is generally not intended that the estimated coefficient should vary across $t$ or $i$.
+
+# Random and fixed effects in SEM
+
+For panel regression in SEM, we typically convert any stacked, long-format vectors of observations of length $NT$ into $T$ separate vectors of length $N$.
+In doing so, we get $T$ individual equations for each time point
+\begin{align}
+\bm{y}_{t} & = \bm{\iota}_{N}\beta_{0_{t}} + \bm{x}_{t}\beta_{1_{t}} + \bm{\alpha} + \bm{\varepsilon}_{t}, \ t = 1, \ldots, T \\
+\begin{bmatrix}
+y_{1_{t}} \\
+y_{2_{t}} \\
+\vdots \\
+y_{N_{t}}
+\end{bmatrix} & = 
+\begin{bmatrix}
+1 \\
+1 \\ 
+\vdots \\
+1
+\end{bmatrix} \beta_{0_{t}} + 
+\begin{bmatrix}
+x_{1_{t}} \\
+x_{2_{t}} \\
+\vdots \\
+x_{N_{t}}
+\end{bmatrix} \beta_{1_{t}} + 
+\begin{bmatrix}
+\alpha_{1} \\
+\alpha_{2} \\
+\vdots \\
+\alpha_{N}
+\end{bmatrix} + 
+\begin{bmatrix}
+\varepsilon_{1_{t}} \\
+\varepsilon_{2_{t}} \\
+\vdots \\
+\varepsilon_{N_{t}}
+\end{bmatrix}
+\end{align}
+
+Let us discuss the random and fixed effects models in SEM by examining a path diagram.
+Figure \ref{fig:panel-sem} shows a four-wave panel model with individual effects.
+As usual, rectangles indicate observed variables and circles indicate latent variables.
+One-sided errors indicate effects and two-headed errors indicate correlations or covariances.
+
+We assume all of the observed variables are grand mean centered, so that we can drop the overall intercept from the model.
+
+\begin{figure}
+\begin{center}
+\resizebox{0.75\textwidth}{!}{%
+\begin{tikzpicture}
+% Node styles ---
+\tikzstyle{man} = [rectangle, thick, minimum size = 0.5cm, draw = black!100, fill = white!100, font = \sffamily] 
+\tikzstyle{lat} = [circle, thick, minimum size = 0.5cm, draw = black!100, fill = white!100, font = \sffamily] 
+\tikzstyle{err} = [rectangle, minimum size = 0.5cm, draw = white!100, fill = white!100, font = \sffamily] 
+\tikzstyle{pha} = [rectangle, minimum size = 0.001cm, draw = white, fill = white]
+\tikzstyle{con} = [-latex, draw = black!100, font = \sffamily] 
+\tikzstyle{seqex} = [latex-latex, draw = black!100, font = \sffamily, dashed]
+\tikzstyle{cons} = [-latex, draw = black!100, font = \sffamily\small] 
+\tikzstyle{cor} = [latex-latex, font = \sffamily] 
+% Begin Figure ---
+% Nodes 
+\node at (+0.0,+0.0) [man] (x1) {$x_{1}$};
+\node at (+2.0,+0.0) [man] (x2) {$x_{2}$};
+\node at (+6.0,+0.0) [man] (x3) {$x_{3}$};
+\node at (+8.0,+0.0) [man] (x4) {$x_{4}$};
+\node at (+4.0,+0.0) [lat] (alp) {$\alpha$};
+\node at (+0.0,+2.1) [man] (y1) {$y_{1}$};
+\node at (+2.0,+2.1) [man] (y2) {$y_{2}$};
+\node at (+6.0,+2.1) [man] (y3) {$y_{3}$};
+\node at (+8.0,+2.1) [man] (y4) {$y_{4}$};
+\node at (+0.0,+3.0) [err] (e1) {\footnotesize $\varepsilon_{1}$};
+\node at (+2.0,+3.0) [err] (e2) {\footnotesize $\varepsilon_{2}$};
+\node at (+6.0,+3.0) [err] (e3) {\footnotesize $\varepsilon_{3}$};
+\node at (+8.0,+3.0) [err] (e4) {\footnotesize $\varepsilon_{4}$};
+% Paths
+\path (alp) edge [con] node [near start, fill=white] {1} (y1);
+\path (alp) edge [con] node [midway, fill=white] {1} (y2);
+\path (alp) edge [con] node [midway, fill=white] {1} (y3);
+\path (alp) edge [con] node [near start, fill=white] {1} (y4);
+\path (x1) edge [con] node [near start, fill=white] {$\beta$} (y1);
+\path (x2) edge [con] node [near start, fill=white] {$\beta$} (y2);
+\path (x3) edge [con] node [near start, fill=white] {$\beta$} (y3);
+\path (x4) edge [con] node [near start, fill=white] {$\beta$} (y4);
+\path (e1) edge [con] node {} (y1);
+\path (e2) edge [con] node {} (y2);
+\path (e3) edge [con] node {} (y3);
+\path (e4) edge [con] node {} (y4);
+% correlations
+\path (x1) edge [cor, bend right] node {} (x2);
+\path (x1) edge [seqex, bend right] node {} (alp);
+\path (x1) edge [cor, bend right] node {} (x3);
+\path (x1) edge [cor, bend right] node {} (x4);
+\path (x2) edge [seqex, bend right] node {} (alp);
+\path (x2) edge [cor, bend right] node {} (x3);
+\path (x2) edge [cor, bend right] node {} (x4);
+\path (x3) edge [seqex, bend left] node {} (alp);
+\path (x4) edge [seqex, bend left] node {} (alp); 
+\end{tikzpicture}
+}
+\caption{Four-wave static panel model\label{fig:panel-sem}}
+\end{center}
+\end{figure}
+
+To begin, let us review a general panel model [@Bollen2010], also referred to as the 'unobserved effects model' [@Wooldridge2012; @R-plm_a] (we will return to this model in the [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf) when we discuss loosening assumptions)
 \begin{align}
@@
 ## Random effects {#re}
-
-Somewhat unintuitively, the modern approach to FE regression treats the unobserved effect $\alpha_{i}$ as a \textit{random} variable, rather than a fixed parameter to be estimated for each of the cross sectional units [@Wooldridge2002].
-And in fact, this view carries over to the way in which FE regression is applied in SEM.
-So, in order to help facilitate the discussion on the application of FE-SEM later on, it makes sense to spend some time discussing the basic idea of random effects regression (again, as we will see, the practical difference between the two models is very small).
 
@@
 and $\hat{\nu}_{it} = y_{it} - \bm{x}_{it}\hat{\bm{\beta}}_{\tiny{POLS}} - \bm{z}_{i}\hat{\bm{\gamma}}_{\tiny{POLS}}$, $\bar{\hat{\nu}}_{i} = T^{-1}\sum_{t = 1}^{T}\hat{\nu}_{it}$ and $\hat{\sigma}^{2}_{\alpha} = \hat{\sigma}^{2}_{\nu} - \hat{\sigma}^{2}_{\varepsilon}$.
+Here, $\bm{\hat{\beta}}_{\tiny{POLS}}$ and $\hat{\bm{\gamma}}_{\tiny{POLS}}$ are the POLS estimates for the coefficients of the time-varying and -invariant covariates, respectively.
 Then, the coefficients are estimated using those estimates in the variance matrix $\hat{\bm{\Omega}}$:
@@
 And, as we know, the average of something that does not change is that thing itself.
-I.e., if $T(T-1)/2 = h$ is the number of elements on either the upper- or lower triangle of $\bm{\Omega}_{i}$, then we have $h^{-1}\sum_{i=1}^{h}\sigma^{2}_{\alpha} = \frac{h \sigma^{2}_{\alpha}}{h} = \sigma^{2}_{\alpha}$.
 
@@
 So if, in fact the covariance between any two wide-format columns of $y$ is $\Cov(y_{t},y_{s}) = \Var(\alpha), \ \forall \ s \ne t$, then $\hat{\psi} = \frac{1}{3}\Var(\alpha) + \frac{1}{3}\Var(\alpha) + \frac{1}{3}\Var(\alpha) = \frac{3 \Var(\alpha)}{3} = \Var(\alpha)$.
-This shows that if our assumption about the underlying DGP is correct, i.e., $y_{it} = \alpha_{i} + \varepsilon_{it}$ and $\Cov(\varepsilon_{t},\varepsilon_{s}) = 0, \ \forall \ t \ne s$, then the estimated variance of $\alpha$ is just what it should be: the average covariance between units of $y$ over time.
+This shows that if our assumption about the underlying data generating process (DGP) is correct, i.e., $y_{it} = \alpha_{i} + \varepsilon_{it}$ and $\Cov(\varepsilon_{t},\varepsilon_{s}) = 0, \ \forall \ t \ne s$, then the estimated variance of $\alpha$ is just what it should be: the average covariance between units of $y$ over time.
 Once we add in observed covariates, the estimated covariance of $\alpha$ then become the \textit{conditional} covariance of $y$ over time, given those covariates.
 
-## Model notation
-
-Having explained how a latent variable is used to estimate the individual effects, we need to state the FE model using SEM-compatible matrix notation.
-There are a number of different model notations (see, for example @Bollen1989 for an overview), but the one that will serve us best is one that was proposed by @Graff1979:
-\begin{align}
-\bm{y}^{+} & = \bm{\Lambda_{y}}^{+} \bm{\eta}^{+}, \\
-\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+}, 
-\end{align}
-where $\bm{\eta}^{+} = (\bm{y}, \bm{x}, \bm{\eta}, \bm{\xi})^{\intercal}$, $\bm{\zeta}^{+} = (\bm{\varepsilon}, \bm{\delta}, \bm{\zeta}, \bm{\xi})^{\intercal}$, $\bm{y}^{+} = (\bm{y}, \bm{x})^{\intercal}$.
-$\bm{y}$ is a vector of observed dependent variables and $\bm{x}$ is a vector of observed independent variables.
-$\bm{\eta}$ is a vector of the latent dependent variables and $\bm{\xi}$ is a vector of latent independent variables.
-$\bm{\varepsilon}$ and $\bm{\delta}$ are vectors of the errors of the observed dependent and independent variables, respectively, and $\bm{\zeta}$ is a vector of the errors, or disturbances, of the latent variables.
-Notice the $^{+}$ symbol is just meant to differentiate the vectors with them from those without them.
-That means, $\bm{\eta}^{+}$ is a vector that holds the observed and latent variables, both dependent (in SEM they are referred to as 'endogenous') and independent (i.e., 'exogeneous')^[For our purposes, the terms endogenous and dependent, on the one hand, and exogenous and independent, on the other, can be used interchangeably.], $\bm{\zeta}^{+}$ holds the errors for the observed variables and the disturbances of the latent variables.
-$\bm{y}^{+}$ holds just the observed variables, both dependent and independent, and $\bm{\Lambda_{y}}^{+}$ is a matrix of ones and zeros that selects the observed variables from $\bm{\eta}^{+}$.
-Lastly, $\bm{B}$ is a matrix that holds the regression coefficients.
-
-If we say that $p$ and $q$ stand for the number of observed dependent and independent variables, respectively, and $m$ and $n$ stand for the number of latent dependent and independent variables, respectively, then $\bm{\eta}^{+}$ and $\bm{\zeta}^{+}$ are $p + q + m + n$, $\bm{y}^{+}$ is $p + q$, $\bm{\Lambda_{y}}^{+}$ is $(p + q) \times (p + q + m + n)$ and $\bm{B}$ is $(p + q + m + n) \times (p + q + m + n)$ [@Bollen1989].
-
-This notation may be confusing at first, but it has advantages.
-First, it allows us the flexibility we need for the models.
-For example, it allows observed $\bm{x}$ to directly influence observed $\bm{y}$ (more common notation assumes that substantive effects occur only between latent variables, observed ones are only used as indicators, see for example @Bollen1989, @Kline2016).
-It also allows $\bm{\xi}$, i.e., any latent exogenous variables, to influence $\bm{y}$ directly.
-These two scenarios cover the traditional FE model with observed variables, and one in which latent variables are used to account for measurement error in the independent variables.
-It is also consistent with the notation used for these models in `lavaan`.
-In fact, `lavaan` switches automatically between matrix notations depending on the specified model.
-That means the matrix representation of the model one sees if they type in `lavInspect(model, what = "est")` after specifying their model in `lavaan` will match up with the notation used here.^[In `Mplus`, the model matrices can be requested by including `OUTPUT: TECH1` in the input file.] It does have a potential disadvantage however.
-Besides being less intuitive than the typical $\bm{\eta} = \bm{B}\bm{\eta} + \bm{\Gamma}\bm{\xi} + \bm{\zeta}$ notation, it means that by including the observed covariates in the stacked long vector $\bm{y}^{+}$, they are treated as another response variable with variances and covariances to be estimated by the model (instead of just using the sample statistics).
-This means that the assumption of multivariate normality (otherwise just imposed on the dependent variables) also applies to the independent ones [@Skrondal2004, p.
-75].
-This can be problematic for noncontinuous independent variables like sex, nationality dummies, marriage status (married/unmarried), etc.
-See @Skrondal2004 for more on this topic.
+\begin{figure}
+\begin{center}
+\resizebox{0.75\textwidth}{!}{%
+\begin{tikzpicture}
+% Node styles ---
+\tikzstyle{man} = [rectangle, thick, minimum size = 1cm, draw = black!80, fill = white!100, font = \sffamily] % Manifest variables
+\tikzstyle{lat} = [circle, thick, minimum size = 1cm, draw = black!80, fill = white!100, font = \sffamily] % Latent variables
+\tikzstyle{err} = [circle, draw = black!80, fill = white!100, font = \sffamily] % Errors 
+% Edge styles
+\tikzstyle{con} = [-latex, font = \sffamily] % Effects
+\tikzstyle{cons} = [-latex, font = \sffamily\small] % Effects with smaller font 
+\tikzstyle{cor} = [latex-latex, font = \sffamily] % Correlations
+% Begin Figure ---
+% Nodes 
+\node at (+0.0,+0.0) [man] (x1) {$x_{1}$};
+\node at (+2.0,+0.0) [man] (x2) {$x_{2}$};
+\node at (+6.0,+0.0) [man] (x3) {$x_{3}$};
+\node at (+4.0,+0.0) [lat] (eta) {$\alpha$};
+\node at (+0.0,-2.0) [man] (y1) {$y_{1}$};
+\node at (+2.0,-2.0) [man] (y2) {$y_{2}$};
+\node at (+6.0,-2.0) [man] (y3) {$y_{3}$};
+\node at (+0.0,-3.5) [err] (e1) {$\varepsilon_{1}$};
+\node at (+2.0,-3.5) [err] (e2) {$\varepsilon_{2}$};
+\node at (+6.0,-3.5) [err] (e3) {$\varepsilon_{3}$};
+% Paths
+\path (x1) edge [con, draw = black!100, left] node {$\beta$} (y1);
+\path (x2) edge [con, draw = black!100, left, near start] node {$\beta$} (y2);
+\path (x3) edge [con, draw = black!100, right] node {$\beta$} (y3);
+\path (e1) edge [con, draw = black!100, left] node {1} (y1);
+\path (e2) edge [con, draw = black!100, left] node {1} (y2);
+\path (e3) edge [con, draw = black!100, right] node {1} (y3);
+\path (eta) edge [con, draw = black!100, above, near start] node {1} (y1);
+\path (eta) edge [con, draw = black!100, below] node {1} (y2);
+\path (eta) edge [con, draw = black!100, above] node {1} (y3);
+% correlations
+\path (x1) edge [cor, bend left] node {} (x2);
+%\path (x1) edge [cor, bend left] node {} (eta);
+\path (x1) edge [cor, bend left] node {} (x3);
+%\path (x2) edge [cor, bend left] node {} (eta);
+\path (x2) edge [cor, bend left] node {} (x3);
+%\path (x3) edge [cor, bend right] node {} (eta);
+\end{tikzpicture}
+}
+\caption{Typical three-wave RE-SEM model with contemporary effects\label{fig:resem}}
+\end{center}
+\end{figure}
 
@@
 
-Let us, however, make things more concrete and take a look at a simple, three-wave version of the typical FE-SEM using this notation (shown graphically in Figure \ref{fig:fesem}).
-For that, we have the following matrix notation (with labels on the outside of the matrices):
-\begin{align}
-\begin{split}
-\bm{y}^{+} & = \bm{\Lambda_{y}}^{+}\bm{\eta}^{+} \\
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3}
-\end{pmatrix} & = 
-\begin{blockarray}{cccccccc}
- & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
- \begin{block}{c(ccccccc)}
- y_{1} & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
- y_{2} & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
- y_{3} & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\ 
- x_{1} & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
- x_{2} & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
- x_{3} & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
- \end{block}
-\end{blockarray}
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix},
-\end{split} \label{eq:y}
-\end{align}
-\begin{align}
-\begin{split}
-\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+} \\
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix} & = 
-\begin{blockarray}{cccccccc}
- & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
- \begin{block}{c(ccccccc)}
- y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
- y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
- y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
- x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- \end{block}
-\end{blockarray}
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix} + 
-\begin{pmatrix}
-\varepsilon_{1} \\
-\varepsilon_{2} \\
-\varepsilon_{3} \\
-x_{1} = \delta_{1} \\
-x_{2} = \delta_{2} \\
-x_{3} = \delta_{3} \\
-\alpha = \xi \\
-\end{pmatrix}. 
-\end{split} \label{eq:eta}
-\end{align} 
-Notice in $\bm{\zeta}^{+}$, for the independent variables we could either write, for example $x_{t}$ or $\delta_{t}$.
-As mentioned above, this is due to the model notation treating the independent variables like dependent variables with variances/covariances to be estimated.
-For the sake of simplicity, we will ignore this subtlety and refer to the observed variable from now on, keeping in mind that if the multivariate normality assumption holds, the estimated statistics will likely be sufficiently close to the sample ones for it to not make much of a difference.
+# Random and fixed effects in lavaan {#fesem}
 
-Admittedly, Equations \eqref{eq:y} and \eqref{eq:eta} may not look like much yet.
-We can remedy this by first putting the equation for $\bm{\eta}^{+}$ in reduced form, i.e., by getting rid of the dependent variable on the r.h.s.:
-\begin{align}
-\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+} \\
-\bm{\eta}^{+} - \bm{B}\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
-(\bm{I} - \bm{B})\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
-\bm{\eta}^{+} & = (\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+},
-\end{align}
-where $\bm{I}$ is the identity matrix.
-By substituting this back into the equation for the observed variables we get $\bm{y}^{+} = \bm{\Lambda_{y}}^{+}[(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+}]$, which works out to:
-\begin{align}
-\begin{pmatrix} 
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\end{pmatrix} & = 
-\begin{pmatrix}
-\alpha + \beta x_{1} + \varepsilon_{1} \\
-\alpha + \beta x_{2} + \varepsilon_{2} \\
-\alpha + \beta x_{3} + \varepsilon_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3}
-\end{pmatrix},
-\end{align}
-which is of course exactly what we should expect given Equation \eqref{eq:fe}^[We can use the `sympy` package in `python` to verify and show the steps for this and other examples, see the [supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/sympy-doublecheck-matrixnotation.py) for the code.].
+The basic panel models discussed in this tutorial consist of three main parts.
+First, we want to create a latent variable representing the individual effects.
+Second, we regress the time-varying dependent variable on the time-varying (and time-invariant, see Section \ref{time-invariant}) covariates.
+Third, we specify the correlations depending on our assumptions.
+If we believe the individual effects are unrelated to the time-varying covariates (we must assume they are unrelated to the time-invariant ones, otherwise the model is not identified), then we apply the RE assumptions and constrain the covariances between the individual effects and the time-varying covariates to zero.
+If we believe the individual effects are indeed related to the model covariates (or more realistic assumption in many circumstances), then we must specify these correlations between the individual effects and the time-varying covariates.
+Finally, in an optional step, we can constrain the residual variances to be equal over time, if we want or need to, potentially in order to save degrees of freedom.
+The following section will explain these steps in `lavaan` in detail.
 
-## Assumptions
+## The `lavaan` package in `R`
 
-What essentially differentiates an FE from an RE model is our assumption concerning the relationship between the unobserved individual effects and the model covariates [@Bollen2010].
-The FE model assumes that $\E[\alpha x_{t}] \ne 0$.
-As such, if we fail to control for the correlation of the covariate and the time-invariant part of the error, then the coefficient of interest, here $\beta$, will be biased.
-Our assumption regarding whether the individual effects are correlated with the model covariates occurs in $\E[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] = \bm{\Psi}$, the covariance matrix of the errors
-\begin{align}
-\bm{y}^{+}\bm{y}^{+ \intercal} & = \E[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})^{\intercal}] \\
- & = \E[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\zeta}^{+ \intercal}(\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal})] \\
- & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \E[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal} \\
- & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \bm{\Psi} (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal}.
-\end{align}
-In the case of an FE model, $\bm{\Psi}$ will reflect our belief that the individual effects are correlated with the model covariates, here again for demonstration the three-wave model:
-\begin{align}
-\bm{\Psi} & = \E
-\begin{blockarray}{cccccccc}
- & \varepsilon_{1} & \varepsilon_{2} & \varepsilon_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
- \begin{block}{c(ccccccc)}
- \varepsilon_{1} & \varepsilon_{1}^{2} &                     &                     &              &              &              & \\
- \varepsilon_{2} & 0                   & \varepsilon_{2}^{2} &                     &              &              &              & \\
- \varepsilon_{3} & 0                   & 0                   & \varepsilon_{3}^{2} &              &              &              & \\
- x_{1}           & 0                   & 0                   & 0                   & x_{1}^{2}    &              &              & \\
- x_{2}           & 0                   & 0                   & 0                   & x_{2}x_{1}   & x_{2}^{2}    &              & \\
- x_{3}           & 0                   & 0                   & 0                   & x_{3}x_{1}   & x_{3}x_{2}   & x_{3}^{2}    & \\
- \alpha          & 0                   & 0                   & 0                   & \alpha x_{1} & \alpha x_{2} & \alpha x_{3} & \alpha^{2} \\
- \end{block}
-\end{blockarray}.
-\end{align}
-Knowing this, we can work out the equation for the coefficient of interest, $\beta$. For the sake of simplicity, assume here and throughout mean-centered variables: 
-\begin{align}
-\Cov(y_{t},x_{t}) & = \E[y_{t}x_{t}] \\
- & = \E[(\alpha + \beta x_{t} + \varepsilon_{t})x_{t}] \\
- & = \E[\alpha x_{t} + \beta x_{t}^{2} + \varepsilon_{t}x_{t}] \\
- & = \Cov(\alpha, x_{t}) + \beta \Var(x_{t}) \\
-\hat{\beta}_{FE-SEM} & = \frac{\Cov(y_{t},x_{t}) - \Cov(\alpha, x_{t})}{\Var(x_{t})}. 
-\end{align}
-This should make intuitive sense.
-From the observed covariance between the dependent and the independent variable, we are partialling out the part that is due to the covariance between the independent variable and the individual effects per unit, and then dividing by the variance of the independent variable, as usual.
-For the RE model, we assume $\E[\alpha x_{t}] = 0$ and the equation reduces to $\hat{\beta}_{RE-SEM} = \Cov(y_{t},x_{t})/\Var(x_{t})$.
-The rest of the model-implied covariance matrix results from $\bm{y}^{+}\bm{y}^{+ \intercal}$.
-
-# Fixed effects in lavaan {#fesem}
-
-The package `lavaan` needs to be installed once with `install.packages("lavaan")`. To be able to use it, we need to load it for every new `R` session:
+The package `lavaan` needs to be installed once with `install.packages("lavaan")`.
+This can be entered directly into the console or at the top of the script.
+To be able to use the package, we need to load it for every new `R` session:
 
@@
 
-Specifying the most basic fixed effects model, like the one shown in @Bollen2010 (the same model as Equation \eqref{eq:fe} but with just one time-varying predictor) involves four components.
-First, we define the latent individual effects variable using the `=~` 'measured by' or 'manifested by' [@R-lavaan] operator at the same time constraining the factor loadings at each timepoint to one.
+To use `lavaan`, we create an `R` object using the assignment operator `<-`, see the model syntax example below.
+Here, the object has been called `fe_sem` because the option for the individual effects to covary with the time-varying covariates has been chosen (more on that in a moment).
+The object can be named anything that complies with object naming conventions in `R` (e.g., the object name must start with a letter or dot, separate using underscores or dots, etc.).
+The model syntax is enclosed in quotes, either single `''` or double `""`.
+This means that the model syntax is essentially a string that the `lavaan` package interprets in a second step.
+Once the model has been specified, we use the `sem()` function to 'fit' the model.
+Notice a second object is made out of the fitted `lavaan` object.
+Here the fitted `lavaan` object has been named `fe_sem.fit`.
+That is, we specify the SEM by writing the model syntax as a string and saving it as an object.
+Then, in a second step, we run the `sem()` function on that object.
+The `sem()` function requires at least two arguments: `model`, i.e., the model object (here: `fe_sem`), and `data`, i.e., the dataframe or covariance matrix (along with the mean vector, if desired).
+That is, at a bare minimum, we must tell `lavaan` how the model is specified and where the data is.
+There are a number of other optional arguments that can be included.
+If they are not, the defaults of the `sem()` wrapper are used.^[See [further details on the `sem()` wrapper defaults](https://rdrr.io/cran/lavaan/man/sem.html), or enter `lavOptions()` into the console to get a full list of defaults.
+An explanation of the optional arguments can be found by entering `?lavOptions` in the console.
+There are other 'wrappers' with slightly different default options, like `cfa()` for example, see [the `lavaan` tutorial website](https://lavaan.ugent.be/tutorial/cfa.html).] Many default options are uninteresting for the current tutorial, but it is important to note that the `sem()` wrapper allows all latent exogenous variables to covary,
+
+For this example, even though it is the default estimator, `estimator = "ML"` has been included as an optional argument to emphasize the fact.
+The [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R) provide some guidance on using robust estimators and full information maximum likelihood (FIML) to deal with non-normal data and missing values; both of which can be accessed with optional arguments in the `sem()` function call.
+
+## Model syntax
+
+Again, specifying the most basic random or fixed effects model, like the one shown in @Bollen2010 (the same model as Equation \eqref{eq:fe} but with just one time-varying predictor) involves three to four components.
+First, we define the latent individual effects variable using the `=~` 'measured by' or 'manifested by' [@R-lavaan] operator at the same time constraining the factor loadings at each timepoint to one with `1*` (see line 3 in the model code below).
 I will call the latent variable `a` to stand for $\alpha$.
 Constraining all of the factor loadings to one reflects our implicit assumption that the combined effect of the unit-specific unobserved factors is constant over time.
-This is the default behaviour of traditional POLS-based approaches to FE that use the stacked long-format data.
+This is the default behaviour of traditional POLS-based approaches to RE and FE that use the stacked long-format data.
 
-```{r, eval=FALSE}
-a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
+Second, we regress the dependent variable on the independent variable using the `~` regression operator (see lines 5--9).
+With stacked, long-format data, only one regression coefficient is estimated over all observed timepoints.
+To have our model mimic this behaviour, we need to constrain the the estimated coefficient to equal over time.
+We do so by adding the same label to the regression coefficient at every time point.
+We will use the label `b` (this label was chosen arbitrarily, we could have used any letter or string of characters) and have it act as an equality constraint for the regression coefficient of interest $\beta$.
+
+The key to a FE model, as opposed to an RE model are our assumptions about the relatedness of our time-varying covariates and the individual effects, i.e., $\E[x_{t}\alpha]$.
+For an FE model, we want to partial out any potential covariance between the independent variable and the individual effects.
+This accounts for any linear relationship between $x_{t}$ and the unit-specific characteristics influencing the dependent variable.
+Further, allowing unrestricted covariances between the independent variable itself over time will not affect how the coefficient $\beta$ is estimated, but will have an effect on the standard errors (see lines 12--17).
+To mimic the behaviour of a conventional FE model, we allow the independent variable to be correlated with the individual effects and itself over time.
+Covariances (including covariances between a variable and itself, i.e., variances) are specified using the `~~` operator.
+The alternative RE model is achieved by replacing line 11 with line 12, which is currently commented out.
+In line 12, the covariances between the individual effects and the time-varying covariate are constrained to zero; the RE assumption.
+This is done in the same way as fixing the factor loadings to one in line 3.
+Here, we fix the covariances to zero with `0*`.
+
+The last component of our code involves the variances of the residuals (lines 19--23).
+This component is optional, but we can constrain the residual variances to be equal over time to again mimic the behaviour of a conventional RE/FE model using POLS on stacked data.
+Here, again, we use labels to make equality constraints.
+Because $y_{t}$ is endogenous, the `~~` operator specifies the variances of *residuals*, i.e., $\varepsilon_{t}$.
+
+```{r echo=FALSE}
+df <- readRDS("../longData.Rda")
+dfw <- readRDS("../wideData.Rda")
 ```
 
-Second, we regress the dependent variable on the independent variable using the `~` regression operator.
-With stacked, long-format data, only one regression coefficient is estimated over all observed timepoints.
-To have our FE-SEM model mimic this behaviour, we need to constrain the the estimated coefficient to equal over time.
-We do so by adding the same label to the regression coefficient at every time point.
-We will use the label `b` (this label was chosen arbitrarily, we could have used any letter or string of characters) and have it act as an equality constraint for the regression coefficient of interest $\beta$:
-
-```{r eval=FALSE}
+```{r, attr.source=".numberLines"}
+fe_sem <- '
+# Define individual effects variable 
+a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
+# Regressions, constrain coefficient to be equal over time
 y1 ~ b*x1
@@
 y5 ~ b*x5
-```
-
-The key to a FE model, as opposed to an RE model are our assumptions about the relatedness of our covariate and the individual effects, i.e., $\E[x_{t}\alpha]$.
-For an FE model, we want to partial out any potential covariance between the independent variable and the individual effects.
-This accounts for any linear relationship between $x_{t}$ and the unit-specific characteristics influencing the dependent variable.
-Further, allowing unrestricted covariances between the independent variable itself over time will not affect how the coefficient $\beta$ is estimated, but will have an effect on the standard errors.
-To mimic the behaviour of a conventional FE model, we allow the independent variable to be correlated with the individual effects and itself over time.
-Covariances (including covariances between a variable and itself, i.e., variances) are specified using the `~~` operator:
-
-```{r eval=FALSE}
-a ~~ x1 + x2 + x3 + x4 + x5
+# Correlations, individual effects are related to time-varying 
+# covariates depending on RE or FE assumptions 
+a ~~ x1 + x2 + x3 + x4 + x5             # FE Model 
+# a ~~ 0*x1 + 0*x2 + 0*x3 + 0*x4 + 0*x5 # RE Model 
 x1 ~~ x2 + x3 + x4 + x5
@@
 x4 ~~ x5
-```
-
-The last component of our code involves the variances of the residuals.
-This component is optional, but we can constrain the residual variances to be equal over time to again mimic the behaviour of a conventional FE model using POLS on stacked data.
-Here, again, we use labels to make equality constraints.
-Because $y_{t}$ is endogenous, the `~~` operator specifies the variances of *residuals*, i.e., $\varepsilon_{t}$.
-
-```{r eval=FALSE}
+# Constrain residual variances to be equal over time
 y1 ~~ e*y1
@@
 y5 ~~ e*y5
+'
+fe_sem.fit <- sem(model = fe_sem, 
+                  data = dfw, 
+                  estimator = "ML")
 ```
@@
 # A simulated example {#ex1}
-
-```{r echo=FALSE}
-df <- readRDS("../longData.Rda")
-dfw <- readRDS("../wideData.Rda")
-```
 
@@
 The dependent variable is also influenced by the individual effects variables with $\beta_{y_{t},\alpha_{1}} = 0.75$ and $\beta_{y_{t},\alpha_{2}} = 0.45$ These values were chosen arbitrarily.
-
-Now, we run the FE-SEM in `lavaan`.
-
-```{r}
-fe_sem <- '
-# Define individual effects variable 
-a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
-# Regressions, constrain coefficient to be equal over time
-y1 ~ b*x1
-y2 ~ b*x2 
-y3 ~ b*x3
-y4 ~ b*x4
-y5 ~ b*x5
-# Allow unrestricted correlation between eta and covariates
-a ~~ x1 + x2 + x3 + x4 + x5
-x1 ~~ x2 + x3 + x4 + x5
-x2 ~~ x3 + x4 + x5
-x3 ~~ x4 + x5
-x4 ~~ x5
-# Constrain residual variances to be equal over time
-y1 ~~ e*y1
-y2 ~~ e*y2
-y3 ~~ e*y3
-y4 ~~ e*y4
-y5 ~~ e*y5
-'
-fe_sem.fit <- sem(model = fe_sem, 
-                  data = dfw, 
-                  estimator = "ML")
-```
 
@@
 
+# Extensions {#extensions}
+
+## Relaxing assumptions meant to mimic traditional FE models {#relax}
+
+There are a number of implicit assumptions attached to the typical FE model that can be relaxed in SEM.
+Some of these assumptions have been discussed already, and a fairly comprehenisve list of assumptions can be found in @Bollen2010.
+Here, I will go over just a few, concentrating on the implementation in `lavaan` and the opportunity to empirically test whether the adjustments are justified or not.
+
+The assumptions we will discuss here pertain to the time-invariance of the effects of both the latent individual effects and the observed covariates, as well as a time-invariant error variance.
+We can also empirically test the correlation between the individual effects and the covariates to see whether a RE model is preferable to the FE model.
+
+For example, we can rewrite the original FE equation as
+\begin{align}
+y_{it} & = \beta_{t}x_{it} + \lambda_{t}\alpha_{i} + \varepsilon_{it}
+\end{align}
+where $\beta$ becomes $\beta_{t}$ and the implicit regression weight of one turns to $\lambda_{t}$ to highlight the fact that the effect of $x$ as well as $\alpha$ on $y$ may vary over time.
+We can furthermore easily relax the assumption of time-constant error variance, i.e., $\sigma^{2}_{\varepsilon_{t}}$.
+As noted in the [main article](https://github.com/henrik-andersen/FE-SEM/blob/master/article.pdf), the assumption regarding $\E[\alpha x_{t}]$ in $\bm{\Psi}$ determines whether we have an FE or RE model.
+We can set these to zero and test whether the RE model would be preferable to the FE model.
+In general, if the individual effects are truly uncorrelated with the model covariates, it is advisable to switch to an RE model since because it uses up less degrees of freedom, it will have smaller standard errors [@Bollen2010].
+
+In the following `lavaan` code, we simply remove the factor loadings of one for the latent individual effect variable which allows them to be estimated freely at each timepoint.
+For the effect of the covariate, we can either delete the constraints `b` in `yt ~ b*xt` or give each regression a different label, e.g., `b1`, `b2`, `b3`, etc.
+Similarly, to allow the error variance to vary over time, we turn the constraints `e` into simple labels, i.e., `e1`, `e2`, `e3`, etc., or again just delete them.
+In fact, regarding the error variances, they will be estimated necessarily, and do not need to be explicitly mentioned in the model syntax at all.
+Finally, to move from an FE to an RE model, we could simply constrain the correlations between the individual effects and the covariates to zero, i.e., `a ~~ 0*x1 + 0*x2 + 0*x3 + 0*x4 + 0*x5`.
+
+```{r eval=FALSE, attr.source=".numberLines"}
+fe_sem_fullyrelaxed <- '
+# Define individual effects variable 
+a =~ y1 + y2 + y3 + y4 + y5
+# Regressions, constrain coefficient to be equal over time
+y1 ~ b1*x1
+y2 ~ b2*x2 
+y3 ~ b3*x3
+y4 ~ b4*x4
+y5 ~ b5*x5
+# Allow unrestricted correlation between eta and covariates
+a ~~ x1 + x2 + x3 + x4 + x5
+# Alternatively: constrain all to 0 for RE model, or
+# just individual correlations
+# a ~~ 0*x1 + 0*x2 + 0*x3 + 0*x4 + 0*x5
+x1 ~~ x2 + x3 + x4 + x5
+x2 ~~ x3 + x4 + x5
+x3 ~~ x4 + x5
+x4 ~~ x5
+# Constrain residual variances to be equal over time
+y1 ~~ e1*y1
+y2 ~~ e2*y2
+y3 ~~ e3*y3
+y4 ~~ e4*y4
+y5 ~~ e5*y5
+'
+fe_sem_fullyrelaxed.fit <- sem( model = fe_sem_fullyrelaxed, 
+                                data = dfw, 
+                                estimator = "ML")
+```
+
+```{r echo=FALSE, warning=FALSE, message=FALSE, error=FALSE, attr.source=".numberLines"}
+re_sem <- '
+# Define individual effects variable 
+a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
+# Regressions, constrain coefficient to be equal over time
+y1 ~ b*x1
+y2 ~ b*x2 
+y3 ~ b*x3
+y4 ~ b*x4
+y5 ~ b*x5
+# Allow unrestricted correlation between eta and covariates
+a ~~ 0*x1 + 0*x2 + 0*x3 + 0*x4 + 0*x5
+x1 ~~ x2 + x3 + x4 + x5
+x2 ~~ x3 + x4 + x5
+x3 ~~ x4 + x5
+x4 ~~ x5
+# Constrain residual variances to be equal over time
+y1 ~~ e*y1
+y2 ~~ e*y2
+y3 ~~ e*y3
+y4 ~~ e*y4
+y5 ~~ e*y5
+'
+re_sem.fit <- sem( model = re_sem, 
+                   data = dfw, 
+                   estimator = "ML")
+```
+
+As outlined in @Bollen2010, the researcher has the opportunity to test each of the assumptions empirically and decide whether a more parsimonious, i.e., restrictive model is justifiable.
+For each assumption, a likelihood ratio test can be carried out to determine whether the improvement to model fit resulting from the relaxation of various assumptions is significant or whether the more parsimonious model is preferable after all.
+
+If we use the original model `fe_sem.fit` (from the [main article](https://github.com/henrik-andersen/FE-SEM/blob/master/article.pdf)) as a starting point, the best strategy for testing these assumptions is to work in a stepwise fashion, relaxing one assumption at a time.
+We can begin by first constraining the correlation between $\alpha$ and $x_{t}$ to zero (`re_sem`) for an RE model.
+If turning from an FE to an RE model does not significantly worsen model fit, we can go forward with the rest of the steps with the RE model.
+If, however, the fit does worsen significantly, it is likely better to stick with the FE model; moving forward then with it to see if a less restrictive FE model is preferable.
+We can perform a likelihood ratio test in `R` using the `anova()` function:
+
+\small
+```{r}
+anova( fe_sem.fit, re_sem.fit)
+```
+\normalsize
+
+The table that is generated shows a comparison of the nested models, in decending order according to degrees of freedom.
+The RE model does not estimate the correlations between the individual effects and the covariates, so it is more parsimonious and thus listed at the bottom.
+The `Chisq` column shows the $\chi^{2}$ statistic for both models and the `Chisq diff` column calculates the difference between the two.
+Obviously, according to the DGP, the correlation between the individual effects and $x_{t}$ is not zero, so fixing these to zero leads to a substantial amount of misfit.
+The last column puts the $\chi^{2}$ difference in relation to the difference in degrees of freedom and gives a p-value for the probability that the difference is solely due to chance.
+Here, the change in $\chi^{2}$ is highly significant, so the FE model should be retained.
+
+After now having established, based on the likelihood ratio test, that FE is our preferred model, we can begin relaxing the rest of the assumptions.
+I show the following merely as a demonstration of the procedure, we know already from the DGP that the parsimonious model as specified in `fe_sem.fit` is appropriate.
+We can next allow the error variances (`fe_semb.fit`), the effect of $x$ on $y$ (`fe_semc.fit`) and finally the factor loadings of the individual effects (`fe_semd.fit`) all to vary over time.
+
+```{r echo=FALSE}
+fe_semb <- '
+# Define individual effects variable 
+a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
+# Regressions, constrain coefficient to be equal over time
+y1 ~ b*x1
+y2 ~ b*x2 
+y3 ~ b*x3
+y4 ~ b*x4
+y5 ~ b*x5
+# Allow unrestricted correlation between eta and covariates
+a ~~ x1 + x2 + x3 + x4 + x5
+x1 ~~ x2 + x3 + x4 + x5
+x2 ~~ x3 + x4 + x5
+x3 ~~ x4 + x5
+x4 ~~ x5
+# Constrain residual variances to be equal over time
+y1 ~~ e1*y1
+y2 ~~ e2*y2
+y3 ~~ e3*y3
+y4 ~~ e4*y4
+y5 ~~ e5*y5
+'
+fe_semb.fit <- sem( model = fe_semb, 
+                   data = dfw, 
+                   estimator = "ML")
+
+fe_semc <- '
+# Define individual effects variable 
+a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
+# Regressions, constrain coefficient to be equal over time
+y1 ~ b1*x1
+y2 ~ b2*x2 
+y3 ~ b3*x3
+y4 ~ b4*x4
+y5 ~ b5*x5
+# Allow unrestricted correlation between eta and covariates
+a ~~ x1 + x2 + x3 + x4 + x5
+x1 ~~ x2 + x3 + x4 + x5
+x2 ~~ x3 + x4 + x5
+x3 ~~ x4 + x5
+x4 ~~ x5
+# Constrain residual variances to be equal over time
+y1 ~~ e1*y1
+y2 ~~ e2*y2
+y3 ~~ e3*y3
+y4 ~~ e4*y4
+y5 ~~ e5*y5
+'
+fe_semc.fit <- sem( model = fe_semc, 
+                   data = dfw, 
+                   estimator = "ML")
+
+fe_semd <- '
+# Define individual effects variable 
+a =~ y1 + y2 + y3 + y4 + y5
+# Regressions, constrain coefficient to be equal over time
+y1 ~ b1*x1
+y2 ~ b2*x2 
+y3 ~ b3*x3
+y4 ~ b4*x4
+y5 ~ b5*x5
+# Allow unrestricted correlation between eta and covariates
+a ~~ x1 + x2 + x3 + x4 + x5
+x1 ~~ x2 + x3 + x4 + x5
+x2 ~~ x3 + x4 + x5
+x3 ~~ x4 + x5
+x4 ~~ x5
+# Constrain residual variances to be equal over time
+y1 ~~ e1*y1
+y2 ~~ e2*y2
+y3 ~~ e3*y3
+y4 ~~ e4*y4
+y5 ~~ e5*y5
+'
+fe_semd.fit <- sem( model = fe_semd, 
+                   data = dfw, 
+                   estimator = "ML")
+```
+
+\small
+```{r}
+anova( fe_sem.fit, fe_semb.fit, fe_semc.fit, fe_semd.fit)
+```
+\normalsize
+
+Keep in mind that a less parsimonious model (fewer degrees of freedom) can never fit worse than a more parsimonious one (more degrees of freedom).
+I.e., chance variations due to sampling error mean that adding constraints to a model will tend to always worsen fit, at least minimally.
+The question here is whether the improvement to fit by loosening constraints is meaningful or not.
+In the table above, we should not expect any meaningful improvements moving from `fe_sem.fit` to `fe_semd.fit`.
+Here, using simulated data, we have the luxury of knowing that any significant differences in $\chi^{2}$ are due to chance.
+With real data, it is up to the researcher to apply their best judgment and decide whether the results are plausible or not.
+
+## Time-invariant predictors {#time-invariant}
+
+What if we do not just want to just control for the effects of all time-invariant variables, but investigate some of them in detail?
+Many time-invariant variables, like sex, birth cohort, nationality, education, etc.
+can be interesting on their own.
+And typically, many of these variables are readily available in a given dataset.
+The traditional OLS-based FE model does not allow for this, as it wipes out the effect of *all* time-invariant variables, whether observed or not.
+
+In SEM, we can easily specify a type of *hybrid* FE/RE model [@Bollen2010] that allows us to control for time-invariant unobserved heterogeneity while also investigating the effects of specific observed time-invariant predictors.^[These types of models have become well known outside of SEM as well, see for example @Allison2011; @Schunck2013; @Bell2018.]
+
+In the next example, we continue with the most complex model we have specified so far, `fe_sem4.fit` in which measurement error in both the independent and dependent variables is accounted for using latent variables.
+Now, we would like as well to specifically investigate the effect of $\alpha_{2}$ on the dependent variable.
+The equation for this model changes to: $\eta_{t} = \beta \xi_{t} + \alpha + \gamma \alpha_{2} + \zeta_{t}$.
+
+```{r eval=TRUE, attr.source=".numberLines"}
+fe_sem5 <- '
+# Measurement model for dependent variable, n for eta
+n1 =~ 1*y11 + y21 + y31
+n2 =~ 1*y12 + y22 + y32
+n3 =~ 1*y13 + y23 + y33
+n4 =~ 1*y14 + y24 + y34
+n5 =~ 1*y15 + y25 + y35
+# Define individual effects variable 
+a =~ 1*n1 + 1*n2 + 1*n3 + 1*n4 + 1*n5
+# Measurement model for independent variables, xi 
+xi1 =~ 1*x11 + x21 + x31 
+xi2 =~ 1*x12 + x22 + x32
+xi3 =~ 1*x13 + x23 + x33
+xi4 =~ 1*x14 + x24 + x34
+xi5 =~ 1*x15 + x25 + x35
+# Regressions, constrain coefficient to be equal over time
+n1 ~ b*xi1 + g*a2
+n2 ~ b*xi2 + g*a2
+n3 ~ b*xi3 + g*a2
+n4 ~ b*xi4 + g*a2
+n5 ~ b*xi5 + g*a2
+# Allow unrestricted correlation between eta and covariates
+a ~~ xi1 + xi2 + xi3 + xi4 + xi5 + 0*a2
+a2 ~~ xi1 + xi2 + xi3 + xi4 + xi5
+xi1 ~~ xi2 + xi3 + xi4 + xi5
+xi2 ~~ xi3 + xi4 + xi5
+xi3 ~~ xi4 + xi5
+xi4 ~~ xi5
+# Constrain residual variances to be equal over time
+n1 ~~ e*n1
+n2 ~~ e*n2
+n3 ~~ e*n3
+n4 ~~ e*n4
+n5 ~~ e*n5
+'
+fe_sem5.fit <- sem( model = fe_sem5, 
+                    data = dfw, 
+                    estimator = "ML")
+```
+
+Keep in mind, based on the DGP, the true parameters are $\beta = 0.3$ and $\gamma = 0.45$.
+
+```{r, output.lines=71:87, eval=TRUE}
+summary(fe_sem5.fit)
+```
+
+From this we can see that such a hybrid model is does a good job of estimating the coefficients of interest, with $\hat{\beta} =$ `r round( lavInspect( fe_sem5.fit, "list")[ 36, 14], 3)` (`r round( lavInspect( fe_sem5.fit, "list")[ 36, 15], 3)`) and $\hat{\gamma} =$ `r round( lavInspect( fe_sem5.fit, "list")[ 37, 14], 3)` (`r round( lavInspect( fe_sem5.fit, "list")[ 37, 15], 3)`).
+
+It is important, however, to realize that the unbiasedness of $\hat{\gamma}$ in this model is dependent on the assumption that $\E[\zeta | \bm{\xi_{t}}, \alpha_{2}] = 0$.
+In other words, the idiosyncratic error is mean independent of $\bm{\xi_{t}} = (\xi_{1}, \xi_{2}, ..., \xi_{T})$ as well as $\alpha_{2}$.
+The first part is easier to accept because we are controlling for all potential time-invariant confounders that could induce a relationship between the independent variable and the error.
+The unbiasedness of $\hat{\gamma}$, on the other hand rests on the assumption that the time-invariant predictor is independent of the error.
+If $\alpha_{2}$ represented the respondent's intelligence and $\eta_{t}$, the dependent variable, represented the respondent's income, for example, then $\hat{\gamma}$ would be biased if both were dependent on a third time-invariant variable, say level of schooling, if it is not controlled for.
+For this reason, we need to treat the regression on a time-invariant predictor like any other regular multivariate regression model and look to include all plausible potential confounders as controls in the model, or turn to other methods, e.g., instrumental variables.
+
 # Conclusion

### 2021-01-14 article_format/article_format_revision.tex
Started major changes for resubmission

--- 2021-01-14 article_format/article_format.tex
@@
 
-\articletype{}
-
-\title{A closer look at fixed effects regression in structural equation
-modeling using \texttt{lavaan}}
+\articletype{TEACHER'S CORNER}
+
+\title{A closer look at random and fixed effects panel regression in
+structural equation modeling using \texttt{lavaan}}
 
@@
 \begin{abstract}
-This article provides an in-depth look at fixed effects regression in
-the structural equation modeling (SEM) framework, specifically the
-application of fixed effects in the \texttt{lavaan} package for
-\texttt{R}. It is meant as a applied guide for researchers, covering the
-underlying model specification, syntax, and summary output. Online
-supplementary materials further discuss various common extensions to the
-basic fixed-effect model, demonstrating how to relax model assumptions,
-deal with measurement error in both the dependent and independent
-variables, and include time-invariant predictors in a type of hybrid
-fixed-/ random effects model.
+This article provides an in-depth look at random and fixed effects panel
+regression in the structural equation modeling (SEM) framework,
+specifically the application of fixed effects in the \texttt{lavaan}
+package for \texttt{R}. It is meant as a applied guide for researchers,
+covering the underlying model specification, syntax, and summary output.
 \end{abstract}
@@
 \begin{keywords}
-Fixed effects, structural equation modeling, lavaan, R, panel analysis
+Fixed effects, structural equation modeling, lavaan, R, panel
+regression, longitudinal data
 \end{keywords}
@@
 time can be seen as the combined effect of all time-invariant influences
-(e.g., sex, date of birth, nationality) on the dependent variable. Fixed
-effects (FE) regression involves controlling for these time-invariant
-influences via a number of various methods. It thus accounts for a
-likely and common source of bias.
+(e.g., sex, date of birth, nationality) on the dependent variable.
+Random effects (RE) and fixed effects (FE) regression involve accounting
+for these often unobserved time-invariant influences via a number of
+methods. In the case of RE models, it is assumed that the stable
+characteristics are unrelated to the other model covariates. In that
+case, the stable characteristics, or individual effects as they are
+often referred to as, will not affect point estimates but are a source
+of error variance that, if ignored, lead to larger standard errors and
+increase the likelihood of type II errors (failure to reject a false
+null hypothesis). FE models are useful when the individual effects are
+expected to be related to one or more of the model covariates. In that
+case, these unobserved stable characteristics will act as confounders
+and lead to biased point estimates. In controlling for these individual
+effects, FE regression thus accounts for a likely and common source of
+bias.
 
@@
 variables, SEM provides a way to deal with measurement error and get
-closer to the true underlying constructs of interest.
+closer to the theoretical constructs of interest.
 
 There are a number of articles describing basic concept of panel model
-regression, and FE regression in SEM
+regression, including RE and FE regression in SEM
 \citep[e.g.,][]{Allison2011, Bollen2010, Teachman2001}. This article is
-intended as a \textit{practical guide} for researchers looking for
-in-depth help with specifying FE models in SEM. It focuses on the
+intended as a \textit{practical guide} for researchers looking for help
+specifying panel regression models in SEM. It assumes some basic
+knowledge of SEM (i.e., minimizing the difference between observed and
+model-implied covariance matrices and mean vectors, maximum likelihood
+estimation, etc.);\footnote{For an introduction to SEM,
+  \citet{Bollen1989} is a classic for good reason, and
+  \citet{Ferron2007} lay out exactly how SEM works in one easy to follow
+  article.} perhaps the reader is an experienced SEM-user looking for a
+quick guide to estimating the static panel models discussed here; or,
+perhaps the reader is new to SEM, potentially because an adviser or
+superior has suggested it to them. For the latter case, this article
+reviews the assumptions associated with the more tradititonal least
+squares-based approach to RE and FE models, so that the implementation
+in SEM is more easily to follow. The article focuses on the
 \texttt{lavaan} \citep{R-lavaan} package for \texttt{R} \citep{R-base}.
@@
 particularities of panel regression SEM, and shows its implementation in
-\texttt{lavaan}. The focus on FE panel regression stems from the fact
-that the assumptions needed to justify other common models, e.g., random
-effects, are often implausible. As we will see, the practical
-differences between the two are, in any event, very small. Using
+\texttt{lavaan}. Using
 \href{https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R}{simulated
@@
 \hypertarget{panel}{%
-\section{Panel models}\label{panel}}
-
-To begin, let us start by reviewing a general panel model
-\citep{Bollen2010}, also referred to as the `unobserved effects model'
+\section{Random and Fixed Effects Panel models}\label{panel}}
+
+It is typically the case that the values for a given unit on a variable
+at one point in time will tend to tell us something about that unit's
+values at another point in time. There are two main explanations for
+this `empirical regularity'
+\citetext{\citealp{Heckman1981}; \citealp[p.~261]{Hsiao2014}; \citealp{Bruederl2015}; \citealp{Bianconcini2018}}.
+First, it could be that an experience at one point in time has a
+causal\footnote{Here the word causal is used to mean `non-spurious'.}
+effect on future experiences. In other words, experiencing an event
+could change the probability of the same or a similar event taking place
+in the future. For example, if employment increases wages, then the
+incentive to continue working should increase over time, thereby making
+it more likely that someone who was employed at one point in time will
+continue to be employed in the future \citep{Heckman1981}. When past
+experiences causally impact future events, the empirical regularity is
+referred to as `state dependence'. So-called `dynamic' panel models that
+include the lagged dependent variable in the equation for the current
+dependent variable, like autoregressive and autoregressive cross-lagged
+models, are examples of panel models that account for state dependence.
+The second explanation is that correlations over time are due to stable
+unobservables, like sex, place of birth, motivation, ability or
+personality characteristics.\footnote{From here on out the units of
+  interest are assumed to be individuals, but other units of observation
+  like schools, companies, countries, etc., are possible.} In other
+words, stable unit-specific characteristics might predispose individuals
+to experiencing events with a certain likelihood over time. For example,
+stable characteristics like an individual's sex and motivation could be
+part of the reason why some individuals tend to be continuously
+employed, while others experience spells of unemployment, or are
+habitually unemployed. This second source of empirical regularity is
+referred to as individual heterogeneity or, more generally, unobserved
+heterogeneity \citep{Wooldridge2012}. So-called `static' panel models
+(that do not include the lagged dependent variable in the equation for
+the current one) focus on this second source of empirical regularity.
+
+The random and fixed effects models that will be discussed here are
+examples of models that attempt to account for unobserved heterogeneity
+as a source of empirical regularity. In the case of both the random and
+fixed effects models, it is assumed that there exist stable individual
+characteristics that affect the dependent variable of interest over
+time. Failure to account for this source of stable variance will reduce
+model fit. If any of the stable characteristics are related to the model
+covariates (a plausible scenario in many cases where the independent
+variable(s) of interest are not fixed), then they must also be
+considered confounders, and failure to account for them will further
+lead to biased point estimates.\footnote{Whether or not the stable
+  individual characteristics are related to the model covariates is the
+  modern differentiator between random effects, where the individual
+  effects are assumed independent of the covariates, and fixed effects,
+  where the individual effects are assumed to be related to one or more
+  of them, see \citet{Wooldridge2012}, p.~251. This can be a cause for
+  confusion for those who are more familiar with the mixed or multilevel
+  model framework. There, `fixed' effects are those coefficients that
+  apply to the entire sample, `random' effects are the variance
+  components for the higher level grouping variables, see for example
+  \citet{Hox2010}; \citet{Bates2015}; \citet{R-lme4}.}
+
+In this section, the basic random and fixed effects models, as well as
+their implementation in SEM, will be discussed. Later on in the article,
+it will be discussed how to extend the model, by relaxing some
+assumptions and introducing other time-invariant predictors in a type of
+`hybrid' random and fixed effects model. The outlook will describe
+further extensions, such as the inclusion of lagged dependent variable
+for a dynamic panel model with individual effects.\\
+Let us begin with a simple linear `unobserved effects model'
+\citep{Wooldridge2012, R-plm_a} \begin{align}
+y_{it} & = \beta_{0} + \beta_{1} x_{it} + \alpha_{i} + \varepsilon_{it}, \ i = 1, \ldots, N, \ t = 1, \ldots, T \label{eq:panel-model}
+\end{align} where \(y_{it}\) is the dependent variable for unit \(i\) at
+time \(t\) and \(x_{it}\) is a single covariate with associated scalar
+coefficient \(\beta_{1}\). \(\alpha_{i}\) represents the combined effect
+of all unobserved time-constant variables affecting the dependent
+variable and \(\varepsilon_{it}\) is the idiosyncratic error. This model
+could of course be extended by turning \(x_{it}\) into a vector of
+covariates
+\(\bm{x}_{it} = (x_{1_{it}}, x_{2_{it}}, \ldots, x_{k_{it}})\), or by
+introducing time-invariant covariates,
+\(\bm{z}_{i} = (z_{1_{i}}, z_{2_{i}}, \ldots, z_{m_{i}})\), but we want
+to first focus on assumptions and mechanics before moving on to
+discussing possibilities for extending the model.
+
+Equation \eqref{eq:panel-model} is often written in matrix form for unit
+\(i\) as \begin{align}
+\bm{y}_{i} & = \bm{\iota}_{T}\beta_{0} + \bm{x}_{i}\beta_{1} + \bm{\iota}_{T} \alpha_{i} + \bm{\varepsilon}_{i}, \ i = 1, \ldots, N \label{eq:matrix-panel-model}
+\end{align} where
+\(\bm{y}_{i} = (y_{i1}, y_{i2}, \ldots, y_{iT})^{\intercal}\) and
+\(\bm{x}_{i} = (x_{i1}, x_{i2}, \ldots, x_{iT})^{\intercal}\) are
+\(T \times 1\) vectors of each of the \(t\) observations for unit \(i\),
+\(\bm{\iota}_{T}\) is a \(T \times 1\) vector of ones and
+\(\bm{\varepsilon}_{i} = (\varepsilon_{i1}, \varepsilon_{i2}, \ldots, \varepsilon_{iT})^{\intercal}\)
+is likewise a \(T \times 1\) vector of errors. It may be helpful to keep
+this Equation \eqref{eq:matrix-panel-model} in mind later when we
+restate the model for estimation using SEM. There, we will take the
+long-format data and transform it into wide-format.
+
+\hypertarget{assumptions-in-least-squares-based-approaches}{%
+\subsection{Assumptions in least squares-based
+approaches}\label{assumptions-in-least-squares-based-approaches}}
+
+One of the main strengths of SEM is the flexibility it affords. Nearly
+every part of the model is specified `by hand', so to speak (although
+modern SEM software has come a long way from, say, \texttt{EQS}
+\citep{Bentler2006}, where the errors and disturbances had to be
+mentioned in the syntax and now comes with a variety of default settings
+for ease of use), which means model assumptions are conveyed directly,
+by the user specifying paths and correlations and setting other
+relations to zero. Common least squares-based statistical packages
+\citep[like the \texttt{plm} package in \texttt{R},][]{R-plm_a} require
+less input from the user (usually it is enough to input the model
+variables and specify some top-level options). The assumptions in the
+least squares-based models are widely known
+\citep{Wooldridge2012, Bruederl2015, Greene2012, Angrist2009} but do not
+come into play prominently while specifying the model. So here we will
+discuss the `typical' assumptions associated with least squares-based
+random and fixed effects models. Later, when we examine the
+implementation in SEM, we will then be able to relate back to this
+discussion.\footnote{What constitutes a `typical' least squares-based
+  approach to these panel regression models is likely debatable. Here,
+  this refers broadly to the assumptions discussed in the literature and
+  default settings one encounters in statistical packages intended for
+  this purpose.}
+
+We want to relate the SEM-based approach to random and fixed effects to
+the least squares-based approach which is likely more widely known and
+more straightforward to implement, i.e., statistical packages for static
+panel regression models exist (e.g., the \texttt{plm} package in
+\texttt{R}, see \citet{R-plm_a}), and typically only require the user to
+enter the model variables and specify some top-level options (random
+effects vs.~fixed effects, individual vs.~individual and time effects,
+etc.). The model assumptions for random and fixed effects models are
+widely known
+\citep{Wooldridge2012, Bruederl2015, Greene2012, Angrist2009} but are
+not prominent in th
+
+The least squares-based approach to such a panel model typically makes
+the following base assumptions:\footnote{What constitutes a `typical'
+  least squares-based approach to these panel regression models is
+  likely debatable. Here, this refers broadly to the default assumptions
+  and settings one encounters in statistical packages intended for this
+  purpose, like the \texttt{plm} package in \texttt{R}, see
+  \citet{R-plm_a}.} first, the mean function is linear, i.e.,
+\(\mathop{\mathrm{\mathbb{E}}}(y_{it} | x_{it}, \alpha_{i}) = \beta_{0} + \beta_{1} x_{it} + \alpha_{i}\),
+and second, that the cross-sectional observations are independent, which
+is assured by random sampling \citep{Wooldridge2012}. Further, and just
+as with any other regression model, the idiosyncratic error is assumed
+to be mean independent of the model covariates and the individual
+effects, i.e.,
+\(\mathop{\mathrm{\mathbb{E}}}(\varepsilon_{it} | x_{it}, \alpha_{i}) = \mathop{\mathrm{\mathbb{E}}}(\varepsilon_{it}) = 0\).\footnote{If
+  there is an overall intercept, \(\beta_{0}\), then it is safe to
+  assume the unconditional expectation of the idiosyncratic error is
+  zero, i.e., \(\mathop{\mathrm{\mathbb{E}}}(\varepsilon_{it}) = 0\).}
+The mean independence assumption also implies the assumption of a zero
+correlation between the idiosyncratic error and the model covariates,
+i.e., \(\mathop{\mathrm{\mathbb{E}}}(x_{it}\varepsilon_{it}) = 0\).
+
+The (modern) difference between random and fixed effects boils down to
+the assumption of the relatedness of the model covariates and the
+individual effects. The random effects model assumes the model
+covariates are unrelated to the individual effects, i.e., in this case
+\(\mathop{\mathrm{\mathbb{E}}}(\alpha_{i} | x_{it}) = \mathop{\mathrm{\mathbb{E}}}(\alpha_{i})\)
+and thus \(\mathop{\mathrm{\mathbb{E}}}(\alpha_{i}x_{it}) = 0\). As
+such, we could put \(\alpha_{i}\) into the error term to form the
+composite error, \(\nu_{it} = \alpha_{i} + \varepsilon_{it}\), and the
+point estimates of the coefficients of interest, here \(\beta_{1}\),
+would be consistent (assuming the model is otherwise correctly
+specified). The problem with putting the individual effects into the
+error term is that it would induce dependency between the compostie
+errors within the same individual, which goes against the independence
+assumption \citep{Wooldridge2012, Schmidheiny2019}. We will discuss that
+in more detail below. In the fixed effects model, one or more of the
+model covariates are assumed to be related to the individual effects,
+here
+\(\mathop{\mathrm{\mathbb{E}}}(\alpha_{i} | x_{it}) \ne \mathop{\mathrm{\mathbb{E}}}(\alpha_{i})\)
+and thus \(\mathop{\mathrm{\mathbb{E}}}(\alpha_{i}x_{it}) \ne 0\). Thus,
+ignoring the individual heterogeneity and putting \(\alpha_{i}\) into
+the error term would mean that point estimates of the coefficients of
+interest would be inconsistent.
+
+The traditional least squares-based approaches to random and fixed
+effects often assume as well that the errors are homoskedastictic over
+time, i.e.,
+\(\mathop{\mathrm{\mathbb{E}}}(\varepsilon_{it}^{2}) = \sigma_{\varepsilon_{t}}^{2} = \sigma_{\varepsilon}^{2}\).
+The individual effects are assumed to have a constant partial effect on
+the dependent variable over time \citep[p.~248]{Wooldridge2012}. Also,
+typically least squares-based approaches stack each of the \(N\)
+repeated measures per individual \(i\) and calculate one single
+coefficient per variable for the entire sample. I.e., it is generally
+not intended that the estimated coefficient should vary across \(t\) or
+\(i\).
+
+\hypertarget{random-and-fixed-effects-in-sem}{%
+\section{Random and fixed effects in
+SEM}\label{random-and-fixed-effects-in-sem}}
+
+For panel regression in SEM, we typically convert any stacked,
+long-format vectors of observations of length \(NT\) into \(T\) separate
+vectors of length \(N\). In doing so, we get \(T\) individual equations
+for each time point \begin{align}
+\bm{y}_{t} & = \bm{\iota}_{N}\beta_{0_{t}} + \bm{x}_{t}\beta_{1_{t}} + \bm{\alpha} + \bm{\varepsilon}_{t}, \ t = 1, \ldots, T \\
+\begin{bmatrix}
+y_{1_{t}} \\
+y_{2_{t}} \\
+\vdots \\
+y_{N_{t}}
+\end{bmatrix} & = 
+\begin{bmatrix}
+1 \\
+1 \\ 
+\vdots \\
+1
+\end{bmatrix} \beta_{0_{t}} + 
+\begin{bmatrix}
+x_{1_{t}} \\
+x_{2_{t}} \\
+\vdots \\
+x_{N_{t}}
+\end{bmatrix} \beta_{1_{t}} + 
+\begin{bmatrix}
+\alpha_{1} \\
+\alpha_{2} \\
+\vdots \\
+\alpha_{N}
+\end{bmatrix} + 
+\begin{bmatrix}
+\varepsilon_{1_{t}} \\
+\varepsilon_{2_{t}} \\
+\vdots \\
+\varepsilon_{N_{t}}
+\end{bmatrix}
+\end{align}
+
+Let us discuss the random and fixed effects models in SEM by examining a
+path diagram. Figure \ref{fig:panel-sem} shows a four-wave panel model
+with individual effects. As usual, rectangles indicate observed
+variables and circles indicate latent variables. One-sided errors
+indicate effects and two-headed errors indicate correlations or
+covariances.
+
+We assume all of the observed variables are grand mean centered, so that
+we can drop the overall intercept from the model.
+
+\begin{figure}
+\begin{center}
+\resizebox{0.75\textwidth}{!}{%
+\begin{tikzpicture}
+% Node styles ---
+\tikzstyle{man} = [rectangle, thick, minimum size = 0.5cm, draw = black!100, fill = white!100, font = \sffamily] 
+\tikzstyle{lat} = [circle, thick, minimum size = 0.5cm, draw = black!100, fill = white!100, font = \sffamily] 
+\tikzstyle{err} = [rectangle, minimum size = 0.5cm, draw = white!100, fill = white!100, font = \sffamily] 
+\tikzstyle{pha} = [rectangle, minimum size = 0.001cm, draw = white, fill = white]
+\tikzstyle{con} = [-latex, draw = black!100, font = \sffamily] 
+\tikzstyle{seqex} = [latex-latex, draw = black!100, font = \sffamily, dashed]
+\tikzstyle{cons} = [-latex, draw = black!100, font = \sffamily\small] 
+\tikzstyle{cor} = [latex-latex, font = \sffamily] 
+% Begin Figure ---
+% Nodes 
+\node at (+0.0,+0.0) [man] (x1) {$x_{1}$};
+\node at (+2.0,+0.0) [man] (x2) {$x_{2}$};
+\node at (+6.0,+0.0) [man] (x3) {$x_{3}$};
+\node at (+8.0,+0.0) [man] (x4) {$x_{4}$};
+\node at (+4.0,+0.0) [lat] (alp) {$\alpha$};
+\node at (+0.0,+2.1) [man] (y1) {$y_{1}$};
+\node at (+2.0,+2.1) [man] (y2) {$y_{2}$};
+\node at (+6.0,+2.1) [man] (y3) {$y_{3}$};
+\node at (+8.0,+2.1) [man] (y4) {$y_{4}$};
+\node at (+0.0,+3.0) [err] (e1) {\footnotesize $\varepsilon_{1}$};
+\node at (+2.0,+3.0) [err] (e2) {\footnotesize $\varepsilon_{2}$};
+\node at (+6.0,+3.0) [err] (e3) {\footnotesize $\varepsilon_{3}$};
+\node at (+8.0,+3.0) [err] (e4) {\footnotesize $\varepsilon_{4}$};
+% Paths
+\path (alp) edge [con] node [near start, fill=white] {1} (y1);
+\path (alp) edge [con] node [midway, fill=white] {1} (y2);
+\path (alp) edge [con] node [midway, fill=white] {1} (y3);
+\path (alp) edge [con] node [near start, fill=white] {1} (y4);
+\path (x1) edge [con] node [near start, fill=white] {$\beta$} (y1);
+\path (x2) edge [con] node [near start, fill=white] {$\beta$} (y2);
+\path (x3) edge [con] node [near start, fill=white] {$\beta$} (y3);
+\path (x4) edge [con] node [near start, fill=white] {$\beta$} (y4);
+\path (e1) edge [con] node {} (y1);
+\path (e2) edge [con] node {} (y2);
+\path (e3) edge [con] node {} (y3);
+\path (e4) edge [con] node {} (y4);
+% correlations
+\path (x1) edge [cor, bend right] node {} (x2);
+\path (x1) edge [seqex, bend right] node {} (alp);
+\path (x1) edge [cor, bend right] node {} (x3);
+\path (x1) edge [cor, bend right] node {} (x4);
+\path (x2) edge [seqex, bend right] node {} (alp);
+\path (x2) edge [cor, bend right] node {} (x3);
+\path (x2) edge [cor, bend right] node {} (x4);
+\path (x3) edge [seqex, bend left] node {} (alp);
+\path (x4) edge [seqex, bend left] node {} (alp); 
+\end{tikzpicture}
+}
+\caption{Four-wave static panel model\label{fig:panel-sem}}
+\end{center}
+\end{figure}
+
+To begin, let us review a general panel model \citep{Bollen2010}, also
+referred to as the `unobserved effects model'
 \citep{Wooldridge2012, R-plm_a} (we will return to this model in the
@@
 \subsection{Random effects}\label{re}}
-
-Somewhat unintuitively, the modern approach to FE regression treats the
-unobserved effect \(\alpha_{i}\) as a \textit{random} variable, rather
-than a fixed parameter to be estimated for each of the cross sectional
-units \citep{Wooldridge2002}. And in fact, this view carries over to the
-way in which FE regression is applied in SEM. So, in order to help
-facilitate the discussion on the application of FE-SEM later on, it
-makes sense to spend some time discussing the basic idea of random
-effects regression (again, as we will see, the practical difference
-between the two models is very small).
 
@@
 \(\hat{\sigma}^{2}_{\alpha} = \hat{\sigma}^{2}_{\nu} - \hat{\sigma}^{2}_{\varepsilon}\).
-Then, the coefficients are estimated using those estimates in the
-variance matrix \(\hat{\bm{\Omega}}\): \begin{align}
+Here, \(\bm{\hat{\beta}}_{\tiny{POLS}}\) and
+\(\hat{\bm{\gamma}}_{\tiny{POLS}}\) are the POLS estimates for the
+coefficients of the time-varying and -invariant covariates,
+respectively. Then, the coefficients are estimated using those estimates
+in the variance matrix \(\hat{\bm{\Omega}}\): \begin{align}
 \begin{pmatrix}
@@
 \(\sigma^{2}_{\alpha}\). And, as we know, the average of something that
-does not change is that thing itself. I.e., if \(T(T-1)/2 = h\) is the
-number of elements on either the upper- or lower triangle of
-\(\bm{\Omega}_{i}\), then we have
-\(h^{-1}\sum_{i=1}^{h}\sigma^{2}_{\alpha} = \frac{h \sigma^{2}_{\alpha}}{h} = \sigma^{2}_{\alpha}\).
+does not change is that thing itself.
 
@@
 \(\hat{\psi} = \frac{1}{3}\mathop{\mathrm{\mathrm{Var}}}(\alpha) + \frac{1}{3}\mathop{\mathrm{\mathrm{Var}}}(\alpha) + \frac{1}{3}\mathop{\mathrm{\mathrm{Var}}}(\alpha) = \frac{3 \mathop{\mathrm{\mathrm{Var}}}(\alpha)}{3} = \mathop{\mathrm{\mathrm{Var}}}(\alpha)\).
-This shows that if our assumption about the underlying DGP is correct,
-i.e., \(y_{it} = \alpha_{i} + \varepsilon_{it}\) and
+This shows that if our assumption about the underlying data generating
+process (DGP) is correct, i.e.,
+\(y_{it} = \alpha_{i} + \varepsilon_{it}\) and
 \(\mathop{\mathrm{\mathrm{Cov}}}(\varepsilon_{t},\varepsilon_{s}) = 0, \ \forall \ t \ne s\),
@@
 
-\hypertarget{model-notation}{%
-\subsection{Model notation}\label{model-notation}}
-
-Having explained how a latent variable is used to estimate the
-individual effects, we need to state the FE model using SEM-compatible
-matrix notation. There are a number of different model notations (see,
-for example \citet{Bollen1989} for an overview), but the one that will
-serve us best is one that was proposed by \citet{Graff1979}:
-\begin{align}
-\bm{y}^{+} & = \bm{\Lambda_{y}}^{+} \bm{\eta}^{+}, \\
-\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+}, 
-\end{align} where
-\(\bm{\eta}^{+} = (\bm{y}, \bm{x}, \bm{\eta}, \bm{\xi})^{\intercal}\),
-\(\bm{\zeta}^{+} = (\bm{\varepsilon}, \bm{\delta}, \bm{\zeta}, \bm{\xi})^{\intercal}\),
-\(\bm{y}^{+} = (\bm{y}, \bm{x})^{\intercal}\). \(\bm{y}\) is a vector of
-observed dependent variables and \(\bm{x}\) is a vector of observed
-independent variables. \(\bm{\eta}\) is a vector of the latent dependent
-variables and \(\bm{\xi}\) is a vector of latent independent variables.
-\(\bm{\varepsilon}\) and \(\bm{\delta}\) are vectors of the errors of
-the observed dependent and independent variables, respectively, and
-\(\bm{\zeta}\) is a vector of the errors, or disturbances, of the latent
-variables. Notice the \(^{+}\) symbol is just meant to differentiate the
-vectors with them from those without them. That means, \(\bm{\eta}^{+}\)
-is a vector that holds the observed and latent variables, both dependent
-(in SEM they are referred to as `endogenous') and independent (i.e.,
-`exogeneous')\footnote{For our purposes, the terms endogenous and
-  dependent, on the one hand, and exogenous and independent, on the
-  other, can be used interchangeably.}, \(\bm{\zeta}^{+}\) holds the
-errors for the observed variables and the disturbances of the latent
-variables. \(\bm{y}^{+}\) holds just the observed variables, both
-dependent and independent, and \(\bm{\Lambda_{y}}^{+}\) is a matrix of
-ones and zeros that selects the observed variables from
-\(\bm{\eta}^{+}\). Lastly, \(\bm{B}\) is a matrix that holds the
-regression coefficients.
-
-If we say that \(p\) and \(q\) stand for the number of observed
-dependent and independent variables, respectively, and \(m\) and \(n\)
-stand for the number of latent dependent and independent variables,
-respectively, then \(\bm{\eta}^{+}\) and \(\bm{\zeta}^{+}\) are
-\(p + q + m + n\), \(\bm{y}^{+}\) is \(p + q\), \(\bm{\Lambda_{y}}^{+}\)
-is \((p + q) \times (p + q + m + n)\) and \(\bm{B}\) is
-\((p + q + m + n) \times (p + q + m + n)\) \citep{Bollen1989}.
-
-This notation may be confusing at first, but it has advantages. First,
-it allows us the flexibility we need for the models. For example, it
-allows observed \(\bm{x}\) to directly influence observed \(\bm{y}\)
-(more common notation assumes that substantive effects occur only
-between latent variables, observed ones are only used as indicators, see
-for example \citet{Bollen1989}, \citet{Kline2016}). It also allows
-\(\bm{\xi}\), i.e., any latent exogenous variables, to influence
-\(\bm{y}\) directly. These two scenarios cover the traditional FE model
-with observed variables, and one in which latent variables are used to
-account for measurement error in the independent variables. It is also
-consistent with the notation used for these models in \texttt{lavaan}.
-In fact, \texttt{lavaan} switches automatically between matrix notations
-depending on the specified model. That means the matrix representation
-of the model one sees if they type in
-\texttt{lavInspect(model,\ what\ =\ "est")} after specifying their model
-in \texttt{lavaan} will match up with the notation used here.\footnote{In
-  \texttt{Mplus}, the model matrices can be requested by including
-  \texttt{OUTPUT:\ TECH1} in the input file.} It does have a potential
-disadvantage however. Besides being less intuitive than the typical
-\(\bm{\eta} = \bm{B}\bm{\eta} + \bm{\Gamma}\bm{\xi} + \bm{\zeta}\)
-notation, it means that by including the observed covariates in the
-stacked long vector \(\bm{y}^{+}\), they are treated as another response
-variable with variances and covariances to be estimated by the model
-(instead of just using the sample statistics). This means that the
-assumption of multivariate normality (otherwise just imposed on the
-dependent variables) also applies to the independent ones
-\citep[p.~75]{Skrondal2004}. This can be problematic for noncontinuous
-independent variables like sex, nationality dummies, marriage status
-(married/unmarried), etc. See \citet{Skrondal2004} for more on this
-topic.
+\begin{figure}
+\begin{center}
+\resizebox{0.75\textwidth}{!}{%
+\begin{tikzpicture}
+% Node styles ---
+\tikzstyle{man} = [rectangle, thick, minimum size = 1cm, draw = black!80, fill = white!100, font = \sffamily] % Manifest variables
+\tikzstyle{lat} = [circle, thick, minimum size = 1cm, draw = black!80, fill = white!100, font = \sffamily] % Latent variables
+\tikzstyle{err} = [circle, draw = black!80, fill = white!100, font = \sffamily] % Errors 
+% Edge styles
+\tikzstyle{con} = [-latex, font = \sffamily] % Effects
+\tikzstyle{cons} = [-latex, font = \sffamily\small] % Effects with smaller font 
+\tikzstyle{cor} = [latex-latex, font = \sffamily] % Correlations
+% Begin Figure ---
+% Nodes 
+\node at (+0.0,+0.0) [man] (x1) {$x_{1}$};
+\node at (+2.0,+0.0) [man] (x2) {$x_{2}$};
+\node at (+6.0,+0.0) [man] (x3) {$x_{3}$};
+\node at (+4.0,+0.0) [lat] (eta) {$\alpha$};
+\node at (+0.0,-2.0) [man] (y1) {$y_{1}$};
+\node at (+2.0,-2.0) [man] (y2) {$y_{2}$};
+\node at (+6.0,-2.0) [man] (y3) {$y_{3}$};
+\node at (+0.0,-3.5) [err] (e1) {$\varepsilon_{1}$};
+\node at (+2.0,-3.5) [err] (e2) {$\varepsilon_{2}$};
+\node at (+6.0,-3.5) [err] (e3) {$\varepsilon_{3}$};
+% Paths
+\path (x1) edge [con, draw = black!100, left] node {$\beta$} (y1);
+\path (x2) edge [con, draw = black!100, left, near start] node {$\beta$} (y2);
+\path (x3) edge [con, draw = black!100, right] node {$\beta$} (y3);
+\path (e1) edge [con, draw = black!100, left] node {1} (y1);
+\path (e2) edge [con, draw = black!100, left] node {1} (y2);
+\path (e3) edge [con, draw = black!100, right] node {1} (y3);
+\path (eta) edge [con, draw = black!100, above, near start] node {1} (y1);
+\path (eta) edge [con, draw = black!100, below] node {1} (y2);
+\path (eta) edge [con, draw = black!100, above] node {1} (y3);
+% correlations
+\path (x1) edge [cor, bend left] node {} (x2);
+%\path (x1) edge [cor, bend left] node {} (eta);
+\path (x1) edge [cor, bend left] node {} (x3);
+%\path (x2) edge [cor, bend left] node {} (eta);
+\path (x2) edge [cor, bend left] node {} (x3);
+%\path (x3) edge [cor, bend right] node {} (eta);
+\end{tikzpicture}
+}
+\caption{Typical three-wave RE-SEM model with contemporary effects\label{fig:resem}}
+\end{center}
+\end{figure}
 
@@
 
-Let us, however, make things more concrete and take a look at a simple,
-three-wave version of the typical FE-SEM using this notation (shown
-graphically in Figure \ref{fig:fesem}). For that, we have the following
-matrix notation (with labels on the outside of the matrices):
-\begin{align}
-\begin{split}
-\bm{y}^{+} & = \bm{\Lambda_{y}}^{+}\bm{\eta}^{+} \\
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3}
-\end{pmatrix} & = 
-\begin{blockarray}{cccccccc}
- & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
- \begin{block}{c(ccccccc)}
- y_{1} & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
- y_{2} & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
- y_{3} & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\ 
- x_{1} & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
- x_{2} & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
- x_{3} & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
- \end{block}
-\end{blockarray}
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix},
-\end{split} \label{eq:y}
-\end{align} \begin{align}
-\begin{split}
-\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+} \\
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix} & = 
-\begin{blockarray}{cccccccc}
- & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
- \begin{block}{c(ccccccc)}
- y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
- y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
- y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
- x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- \end{block}
-\end{blockarray}
-\begin{pmatrix}
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\alpha
-\end{pmatrix} + 
-\begin{pmatrix}
-\varepsilon_{1} \\
-\varepsilon_{2} \\
-\varepsilon_{3} \\
-x_{1} = \delta_{1} \\
-x_{2} = \delta_{2} \\
-x_{3} = \delta_{3} \\
-\alpha = \xi \\
-\end{pmatrix}. 
-\end{split} \label{eq:eta}
-\end{align} Notice in \(\bm{\zeta}^{+}\), for the independent variables
-we could either write, for example \(x_{t}\) or \(\delta_{t}\). As
-mentioned above, this is due to the model notation treating the
-independent variables like dependent variables with
-variances/covariances to be estimated. For the sake of simplicity, we
-will ignore this subtlety and refer to the observed variable from now
-on, keeping in mind that if the multivariate normality assumption holds,
-the estimated statistics will likely be sufficiently close to the sample
-ones for it to not make much of a difference.
-
-Admittedly, Equations \eqref{eq:y} and \eqref{eq:eta} may not look like
-much yet. We can remedy this by first putting the equation for
-\(\bm{\eta}^{+}\) in reduced form, i.e., by getting rid of the dependent
-variable on the r.h.s.: \begin{align}
-\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+} \\
-\bm{\eta}^{+} - \bm{B}\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
-(\bm{I} - \bm{B})\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
-\bm{\eta}^{+} & = (\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+},
-\end{align} where \(\bm{I}\) is the identity matrix. By substituting
-this back into the equation for the observed variables we get
-\(\bm{y}^{+} = \bm{\Lambda_{y}}^{+}[(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+}]\),
-which works out to: \begin{align}
-\begin{pmatrix} 
-y_{1} \\
-y_{2} \\
-y_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3} \\
-\end{pmatrix} & = 
-\begin{pmatrix}
-\alpha + \beta x_{1} + \varepsilon_{1} \\
-\alpha + \beta x_{2} + \varepsilon_{2} \\
-\alpha + \beta x_{3} + \varepsilon_{3} \\
-x_{1} \\
-x_{2} \\
-x_{3}
-\end{pmatrix},
-\end{align} which is of course exactly what we should expect given
-Equation \eqref{eq:fe}\footnote{We can use the \texttt{sympy} package in
-  \texttt{python} to verify and show the steps for this and other
-  examples, see the
-  \href{https://github.com/henrik-andersen/FE-SEM/blob/master/sympy-doublecheck-matrixnotation.py}{supplementary
-  materials} for the code.}.
-
-\hypertarget{assumptions}{%
-\subsection{Assumptions}\label{assumptions}}
-
-What essentially differentiates an FE from an RE model is our assumption
-concerning the relationship between the unobserved individual effects
-and the model covariates \citep{Bollen2010}. The FE model assumes that
-\(\mathop{\mathrm{\mathbb{E}}}[\alpha x_{t}] \ne 0\). As such, if we
-fail to control for the correlation of the covariate and the
-time-invariant part of the error, then the coefficient of interest, here
-\(\beta\), will be biased. Our assumption regarding whether the
-individual effects are correlated with the model covariates occurs in
-\(\mathop{\mathrm{\mathbb{E}}}[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] = \bm{\Psi}\),
-the covariance matrix of the errors \begin{align}
-\bm{y}^{+}\bm{y}^{+ \intercal} & = \mathop{\mathrm{\mathbb{E}}}[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})^{\intercal}] \\
- & = \mathop{\mathrm{\mathbb{E}}}[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\zeta}^{+ \intercal}(\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal})] \\
- & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \mathop{\mathrm{\mathbb{E}}}[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal} \\
- & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \bm{\Psi} (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal}.
-\end{align} In the case of an FE model, \(\bm{\Psi}\) will reflect our
-belief that the individual effects are correlated with the model
-covariates, here again for demonstration the three-wave model:
-\begin{align}
-\bm{\Psi} & = \mathop{\mathrm{\mathbb{E}}}
-\begin{blockarray}{cccccccc}
- & \varepsilon_{1} & \varepsilon_{2} & \varepsilon_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
- \begin{block}{c(ccccccc)}
- \varepsilon_{1} & \varepsilon_{1}^{2} &                     &                     &              &              &              & \\
- \varepsilon_{2} & 0                   & \varepsilon_{2}^{2} &                     &              &              &              & \\
- \varepsilon_{3} & 0                   & 0                   & \varepsilon_{3}^{2} &              &              &              & \\
- x_{1}           & 0                   & 0                   & 0                   & x_{1}^{2}    &              &              & \\
- x_{2}           & 0                   & 0                   & 0                   & x_{2}x_{1}   & x_{2}^{2}    &              & \\
- x_{3}           & 0                   & 0                   & 0                   & x_{3}x_{1}   & x_{3}x_{2}   & x_{3}^{2}    & \\
- \alpha          & 0                   & 0                   & 0                   & \alpha x_{1} & \alpha x_{2} & \alpha x_{3} & \alpha^{2} \\
- \end{block}
-\end{blockarray}.
-\end{align} Knowing this, we can work out the equation for the
-coefficient of interest, \(\beta\). For the sake of simplicity, assume
-here and throughout mean-centered variables: \begin{align}
-\mathop{\mathrm{\mathrm{Cov}}}(y_{t},x_{t}) & = \mathop{\mathrm{\mathbb{E}}}[y_{t}x_{t}] \\
- & = \mathop{\mathrm{\mathbb{E}}}[(\alpha + \beta x_{t} + \varepsilon_{t})x_{t}] \\
- & = \mathop{\mathrm{\mathbb{E}}}[\alpha x_{t} + \beta x_{t}^{2} + \varepsilon_{t}x_{t}] \\
- & = \mathop{\mathrm{\mathrm{Cov}}}(\alpha, x_{t}) + \beta \mathop{\mathrm{\mathrm{Var}}}(x_{t}) \\
-\hat{\beta}_{FE-SEM} & = \frac{\mathop{\mathrm{\mathrm{Cov}}}(y_{t},x_{t}) - \mathop{\mathrm{\mathrm{Cov}}}(\alpha, x_{t})}{\mathop{\mathrm{\mathrm{Var}}}(x_{t})}. 
-\end{align} This should make intuitive sense. From the observed
-covariance between the dependent and the independent variable, we are
-partialling out the part that is due to the covariance between the
-independent variable and the individual effects per unit, and then
-dividing by the variance of the independent variable, as usual. For the
-RE model, we assume \(\mathop{\mathrm{\mathbb{E}}}[\alpha x_{t}] = 0\)
-and the equation reduces to
-\(\hat{\beta}_{RE-SEM} = \mathop{\mathrm{\mathrm{Cov}}}(y_{t},x_{t})/\mathop{\mathrm{\mathrm{Var}}}(x_{t})\).
-The rest of the model-implied covariance matrix results from
-\(\bm{y}^{+}\bm{y}^{+ \intercal}\).
-
 \hypertarget{fesem}{%
-\section{Fixed effects in lavaan}\label{fesem}}
+\section{Random and fixed effects in lavaan}\label{fesem}}
+
+The basic panel models discussed in this tutorial consist of three main
+parts. First, we want to create a latent variable representing the
+individual effects. Second, we regress the time-varying dependent
+variable on the time-varying (and time-invariant, see Section
+\ref{time-invariant}) covariates. Third, we specify the correlations
+depending on our assumptions. If we believe the individual effects are
+unrelated to the time-varying covariates (we must assume they are
+unrelated to the time-invariant ones, otherwise the model is not
+identified), then we apply the RE assumptions and constrain the
+covariances between the individual effects and the time-varying
+covariates to zero. If we believe the individual effects are indeed
+related to the model covariates (or more realistic assumption in many
+circumstances), then we must specify these correlations between the
+individual effects and the time-varying covariates. Finally, in an
+optional step, we can constrain the residual variances to be equal over
+time, if we want or need to, potentially in order to save degrees of
+freedom. The following section will explain these steps in
+\texttt{lavaan} in detail.
+
+\hypertarget{the-lavaan-package-in-r}{%
+\subsection{\texorpdfstring{The \texttt{lavaan} package in
+\texttt{R}}{The lavaan package in R}}\label{the-lavaan-package-in-r}}
 
 The package \texttt{lavaan} needs to be installed once with
-\texttt{install.packages("lavaan")}. To be able to use it, we need to
-load it for every new \texttt{R} session:
+\texttt{install.packages("lavaan")}. This can be entered directly into
+the console or at the top of the script. To be able to use the package,
+we need to load it for every new \texttt{R} session:
 
@@
 
-Specifying the most basic fixed effects model, like the one shown in
-\citet{Bollen2010} (the same model as Equation \eqref{eq:fe} but with
-just one time-varying predictor) involves four components. First, we
-define the latent individual effects variable using the
-\texttt{=\textasciitilde{}} `measured by' or `manifested by'
-\citep{R-lavaan} operator at the same time constraining the factor
-loadings at each timepoint to one. I will call the latent variable
-\texttt{a} to stand for \(\alpha\). Constraining all of the factor
-loadings to one reflects our implicit assumption that the combined
-effect of the unit-specific unobserved factors is constant over time.
-This is the default behaviour of traditional POLS-based approaches to FE
-that use the stacked long-format data.
+To use \texttt{lavaan}, we create an \texttt{R} object using the
+assignment operator \texttt{\textless{}-}, see the model syntax example
+below. Here, the object has been called \texttt{fe\_sem} because the
+option for the individual effects to covary with the time-varying
+covariates has been chosen (more on that in a moment). The object can be
+named anything that complies with object naming conventions in
+\texttt{R} (e.g., the object name must start with a letter or dot,
+separate using underscores or dots, etc.). The model syntax is enclosed
+in quotes, either single \texttt{\textquotesingle{}\textquotesingle{}}
+or double \texttt{""}. This means that the model syntax is essentially a
+string that the \texttt{lavaan} package interprets in a second step.
+Once the model has been specified, we use the \texttt{sem()} function to
+`fit' the model. Notice a second object is made out of the fitted
+\texttt{lavaan} object. Here the fitted \texttt{lavaan} object has been
+named \texttt{fe\_sem.fit}. That is, we specify the SEM by writing the
+model syntax as a string and saving it as an object. Then, in a second
+step, we run the \texttt{sem()} function on that object. The
+\texttt{sem()} function requires at least two arguments: \texttt{model},
+i.e., the model object (here: \texttt{fe\_sem}), and \texttt{data},
+i.e., the dataframe or covariance matrix (along with the mean vector, if
+desired). That is, at a bare minimum, we must tell \texttt{lavaan} how
+the model is specified and where the data is. There are a number of
+other optional arguments that can be included. If they are not, the
+defaults of the \texttt{sem()} wrapper are used.\footnote{See
+  \href{https://rdrr.io/cran/lavaan/man/sem.html}{further details on the
+  \texttt{sem()} wrapper defaults}, or enter \texttt{lavOptions()} into
+  the console to get a full list of defaults. An explanation of the
+  optional arguments can be found by entering \texttt{?lavOptions} in
+  the console. There are other `wrappers' with slightly different
+  default options, like \texttt{cfa()} for example, see
+  \href{https://lavaan.ugent.be/tutorial/cfa.html}{the \texttt{lavaan}
+  tutorial website}.} Many default options are uninteresting for the
+current tutorial, but it is important to note that the \texttt{sem()}
+wrapper allows all latent exogenous variables to covary,
+
+For this example, even though it is the default estimator,
+\texttt{estimator\ =\ "ML"} has been included as an optional argument to
+emphasize the fact. The
+\href{https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R}{online
+supplementary materials} provide some guidance on using robust
+estimators and full information maximum likelihood (FIML) to deal with
+non-normal data and missing values; both of which can be accessed with
+optional arguments in the \texttt{sem()} function call.
+
+\hypertarget{model-syntax}{%
+\subsection{Model syntax}\label{model-syntax}}
+
+Again, specifying the most basic random or fixed effects model, like the
+one shown in \citet{Bollen2010} (the same model as Equation
+\eqref{eq:fe} but with just one time-varying predictor) involves three
+to four components. First, we define the latent individual effects
+variable using the \texttt{=\textasciitilde{}} `measured by' or
+`manifested by' \citep{R-lavaan} operator at the same time constraining
+the factor loadings at each timepoint to one with \texttt{1*} (see line
+3 in the model code below). I will call the latent variable \texttt{a}
+to stand for \(\alpha\). Constraining all of the factor loadings to one
+reflects our implicit assumption that the combined effect of the
+unit-specific unobserved factors is constant over time. This is the
+default behaviour of traditional POLS-based approaches to RE and FE that
+use the stacked long-format data.
+
+Second, we regress the dependent variable on the independent variable
+using the \texttt{\textasciitilde{}} regression operator (see lines
+5--9). With stacked, long-format data, only one regression coefficient
+is estimated over all observed timepoints. To have our model mimic this
+behaviour, we need to constrain the the estimated coefficient to equal
+over time. We do so by adding the same label to the regression
+coefficient at every time point. We will use the label \texttt{b} (this
+label was chosen arbitrarily, we could have used any letter or string of
+characters) and have it act as an equality constraint for the regression
+coefficient of interest \(\beta\).
+
+The key to a FE model, as opposed to an RE model are our assumptions
+about the relatedness of our time-varying covariates and the individual
+effects, i.e., \(\mathop{\mathrm{\mathbb{E}}}[x_{t}\alpha]\). For an FE
+model, we want to partial out any potential covariance between the
+independent variable and the individual effects. This accounts for any
+linear relationship between \(x_{t}\) and the unit-specific
+characteristics influencing the dependent variable. Further, allowing
+unrestricted covariances between the independent variable itself over
+time will not affect how the coefficient \(\beta\) is estimated, but
+will have an effect on the standard errors (see lines 12--17). To mimic
+the behaviour of a conventional FE model, we allow the independent
+variable to be correlated with the individual effects and itself over
+time. Covariances (including covariances between a variable and itself,
+i.e., variances) are specified using the
+\texttt{\textasciitilde{}\textasciitilde{}} operator. The alternative RE
+model is achieved by replacing line 11 with line 12, which is currently
+commented out. In line 12, the covariances between the individual
+effects and the time-varying covariate are constrained to zero; the RE
+assumption. This is done in the same way as fixing the factor loadings
+to one in line 3. Here, we fix the covariances to zero with \texttt{0*}.
+
+The last component of our code involves the variances of the residuals
+(lines 19--23). This component is optional, but we can constrain the
+residual variances to be equal over time to again mimic the behaviour of
+a conventional RE/FE model using POLS on stacked data. Here, again, we
+use labels to make equality constraints. Because \(y_{t}\) is
+endogenous, the \texttt{\textasciitilde{}\textasciitilde{}} operator
+specifies the variances of \emph{residuals}, i.e., \(\varepsilon_{t}\).
 
@@
 
+\doublespacing
+
+\singlespacing
+
 \begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{a =}\ErrorTok{\textasciitilde{}}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y1 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y2 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y3 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y4 }\OperatorTok{+}\StringTok{ }\DecValTok{1}\OperatorTok{*}\NormalTok{y5}
-\end{Highlighting}
-\end{Shaded}
-
-\doublespacing
-
-Second, we regress the dependent variable on the independent variable
-using the \texttt{\textasciitilde{}} regression operator. With stacked,
-long-format data, only one regression coefficient is estimated over all
-observed timepoints. To have our FE-SEM model mimic this behaviour, we
-need to constrain the the estimated coefficient to equal over time. We
-do so by adding the same label to the regression coefficient at every
-time point. We will use the label \texttt{b} (this label was chosen
-arbitrarily, we could have used any letter or string of characters) and
-have it act as an equality constraint for the regression coefficient of
-interest \(\beta\):
-
-\singlespacing
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{y1 }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x1}
-\NormalTok{y2 }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x2 }
-\NormalTok{y3 }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x3}
-\NormalTok{y4 }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x4}
-\NormalTok{y5 }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{b}\OperatorTok{*}\NormalTok{x5}
-\end{Highlighting}
-\end{Shaded}
-
-\doublespacing
-
-The key to a FE model, as opposed to an RE model are our assumptions
-about the relatedness of our covariate and the individual effects, i.e.,
-\(\mathop{\mathrm{\mathbb{E}}}[x_{t}\alpha]\). For an FE model, we want
-to partial out any potential covariance between the independent variable
-and the individual effects. This accounts for any linear relationship
-between \(x_{t}\) and the unit-specific characteristics influencing the
-dependent variable. Further, allowing unrestricted covariances between
-the independent variable itself over time will not affect how the
-coefficient \(\beta\) is estimated, but will have an effect on the
-standard errors. To mimic the behaviour of a conventional FE model, we
-allow the independent variable to be correlated with the individual
-effects and itself over time. Covariances (including covariances between
-a variable and itself, i.e., variances) are specified using the
-\texttt{\textasciitilde{}\textasciitilde{}} operator:
-
-\singlespacing
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{a }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{x1 }\OperatorTok{+}\StringTok{ }\NormalTok{x2 }\OperatorTok{+}\StringTok{ }\NormalTok{x3 }\OperatorTok{+}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
-\NormalTok{x1 }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{x2 }\OperatorTok{+}\StringTok{ }\NormalTok{x3 }\OperatorTok{+}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
-\NormalTok{x2 }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{x3 }\OperatorTok{+}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
-\NormalTok{x3 }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{x4 }\OperatorTok{+}\StringTok{ }\NormalTok{x5}
-\NormalTok{x4 }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{x5}
-\end{Highlighting}
-\end{Shaded}
-
-\doublespacing
-
-The last component of our code involves the variances of the residuals.
-This component is optional, but we can constrain the residual variances
-to be equal over time to again mimic the behaviour of a conventional FE
-model using POLS on stacked data. Here, again, we use labels to make
-equality constraints. Because \(y_{t}\) is endogenous, the
-\texttt{\textasciitilde{}\textasciitilde{}} operator specifies the
-variances of \emph{residuals}, i.e., \(\varepsilon_{t}\).
-
-\singlespacing
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{y1 }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y1}
-\NormalTok{y2 }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y2}
-\NormalTok{y3 }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y3}
-\NormalTok{y4 }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y4}
-\NormalTok{y5 }\OperatorTok{\textasciitilde{}}\ErrorTok{\textasciitilde{}}\StringTok{ }\NormalTok{e}\OperatorTok{*}\NormalTok{y5}
-\end{Highlighting}
-\end{Shaded}
-
-\doublespacing
-
-\hypertarget{ex1}{%
-\section{A simulated example}\label{ex1}}
-
-\singlespacing
-
-\doublespacing
-
-To demonstrate the application of FE models in SEM, a dataset can be
-simulated that embodies the FE assumptions. Again, the code for data
-simulation can be found in the
-\href{https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R}{online
-supplementary materials}.
-
-To show that the latent individual effects variables represent the
-\emph{combined} effect of all time-invariant characteristics, the
-dependent variable will be influenced by two separate unit-specific
-variables, which we can call \(\alpha_{1}\) and \(\alpha_{2}\). We will
-construct the simulated data such that the independent variable is
-correlated with both of the time-invariant variables. This means that
-approaches that fail to account for this confounding influence, such as
-POLS or RE, will be biased.
-
-The wide-format equations for the data generating process can be
-described as: \begin{align}
-\bm{x_{t}} & = \bm{\alpha_{1}}\beta_{x_{t},\alpha_{1}} + \bm{\alpha_{2}}\beta_{x_{t},\alpha_{2}} + \bm{\delta_{t}}, \\
-\bm{y_{t}} & = \bm{x_{t}}\beta_{y_{t},x_{t}} + \bm{\alpha_{1}}\beta_{y_{t},\alpha_{1}} + \bm{\alpha_{2}} \beta_{y_{t},\alpha_{2}} + \bm{\varepsilon_{t}} 
-\end{align} where, for the sake of simplicity, \(\bm{\alpha_{1}}\),
-\(\bm{\alpha_{2}}\), \(\bm{\delta_{t}}\) and \(\bm{\varepsilon_{t}}\)
-are \(\sim N(0,1)\).
-
-For the following example, a sample size of 1,000, observed over five
-waves, was chosen. The unique variance of \(\bm{x}\), as well as both
-the individual-effect variables is also \(\sim N(0,1)\). The coefficient
-of interest, \(\beta_{y,x}\) is set to be equal to \(0.3\). A
-correlation between \(\bm{x}\) and the individual effects is induced
-through \(\beta_{x,\alpha_{1}} = 0.85\) and
-\(\beta_{x,\alpha_{1}} = 0.50\). With the variances above set to one,
-the covariances will be roughly
-\(\mathop{\mathrm{\mathrm{Cov}}}(x_{t},\alpha_{1}) = 0.85\) and
-\(\mathop{\mathrm{\mathrm{Cov}}}(x_{t},\alpha_{2}) = 0.5\). The
-dependent variable is also influenced by the individual effects
-variables with \(\beta_{y_{t},\alpha_{1}} = 0.75\) and
-\(\beta_{y_{t},\alpha_{2}} = 0.45\) These values were chosen
-arbitrarily.
-
-Now, we run the FE-SEM in \texttt{lavaan}.
-
-\singlespacing
-
-\begin{Shaded}
-\begin{Highlighting}[]
+\begin{Highlighting}[numbers=left,,]
 \NormalTok{fe\_sem \textless{}{-}}\StringTok{ \textquotesingle{}}
@@
 \StringTok{y5 \textasciitilde{} b*x5}
-\StringTok{\# Allow unrestricted correlation between eta and covariates}
-\StringTok{a \textasciitilde{}\textasciitilde{} x1 + x2 + x3 + x4 + x5}
+\StringTok{\# Correlations, individual effects are related to time{-}varying }
+\StringTok{\# covariates depending on RE or FE assumptions }
+\StringTok{a \textasciitilde{}\textasciitilde{} x1 + x2 + x3 + x4 + x5             \# FE Model }
+\StringTok{\# a \textasciitilde{}\textasciitilde{} 0*x1 + 0*x2 + 0*x3 + 0*x4 + 0*x5 \# RE Model }
 \StringTok{x1 \textasciitilde{}\textasciitilde{} x2 + x3 + x4 + x5}
@@
 \doublespacing
+
+\hypertarget{ex1}{%
+\section{A simulated example}\label{ex1}}
+
+To demonstrate the application of FE models in SEM, a dataset can be
+simulated that embodies the FE assumptions. Again, the code for data
+simulation can be found in the
+\href{https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R}{online
+supplementary materials}.
+
+To show that the latent individual effects variables represent the
+\emph{combined} effect of all time-invariant characteristics, the
+dependent variable will be influenced by two separate unit-specific
+variables, which we can call \(\alpha_{1}\) and \(\alpha_{2}\). We will
+construct the simulated data such that the independent variable is
+correlated with both of the time-invariant variables. This means that
+approaches that fail to account for this confounding influence, such as
+POLS or RE, will be biased.
+
+The wide-format equations for the data generating process can be
+described as: \begin{align}
+\bm{x_{t}} & = \bm{\alpha_{1}}\beta_{x_{t},\alpha_{1}} + \bm{\alpha_{2}}\beta_{x_{t},\alpha_{2}} + \bm{\delta_{t}}, \\
+\bm{y_{t}} & = \bm{x_{t}}\beta_{y_{t},x_{t}} + \bm{\alpha_{1}}\beta_{y_{t},\alpha_{1}} + \bm{\alpha_{2}} \beta_{y_{t},\alpha_{2}} + \bm{\varepsilon_{t}} 
+\end{align} where, for the sake of simplicity, \(\bm{\alpha_{1}}\),
+\(\bm{\alpha_{2}}\), \(\bm{\delta_{t}}\) and \(\bm{\varepsilon_{t}}\)
+are \(\sim N(0,1)\).
+
+For the following example, a sample size of 1,000, observed over five
+waves, was chosen. The unique variance of \(\bm{x}\), as well as both
+the individual-effect variables is also \(\sim N(0,1)\). The coefficient
+of interest, \(\beta_{y,x}\) is set to be equal to \(0.3\). A
+correlation between \(\bm{x}\) and the individual effects is induced
+through \(\beta_{x,\alpha_{1}} = 0.85\) and
+\(\beta_{x,\alpha_{1}} = 0.50\). With the variances above set to one,
+the covariances will be roughly
+\(\mathop{\mathrm{\mathrm{Cov}}}(x_{t},\alpha_{1}) = 0.85\) and
+\(\mathop{\mathrm{\mathrm{Cov}}}(x_{t},\alpha_{2}) = 0.5\). The
+dependent variable is also influenced by the individual effects
+variables with \(\beta_{y_{t},\alpha_{1}} = 0.75\) and
+\(\beta_{y_{t},\alpha_{2}} = 0.45\) These values were chosen
+arbitrarily.
 
@@
 
+\hypertarget{extensions}{%
+\section{Extensions}\label{extensions}}
+
+\hypertarget{relax}{%
+\subsection{Relaxing assumptions meant to mimic traditional FE
+models}\label{relax}}
+
+There are a number of implicit assumptions attached to the typical FE
+model that can be relaxed in SEM. Some of these assumptions have been
+discussed already, and a fairly comprehenisve list of assumptions can be
+found in \citet{Bollen2010}. Here, I will go over just a few,
+concentrating on the implementation in \texttt{lavaan} and the
+opportunity to empirically test whether the adjustments are justified or
+not.
+
+The assumptions we will discuss here pertain to the time-invariance of
+the effects of both the latent individual effects and the observed
+covariates, as well as a time-invariant error variance. We can also
+empirically test the correlation between the individual effects and the
+covariates to see whether a RE model is preferable to the FE model.
+
+For example, we can rewrite the original FE equation as \begin{align}
+y_{it} & = \beta_{t}x_{it} + \lambda_{t}\alpha_{i} + \varepsilon_{it}
+\end{align} where \(\beta\) becomes \(\beta_{t}\) and the implicit
+regression weight of one turns to \(\lambda_{t}\) to highlight the fact
+that the effect of \(x\) as well as \(\alpha\) on \(y\) may vary over
+time. We can furthermore easily relax the assumption of time-constant
+error variance, i.e., \(\sigma^{2}_{\varepsilon_{t}}\). As noted in the
+\href{https://github.com/henrik-andersen/FE-SEM/blob/master/article.pdf}{main
+article}, the assumption regarding
+\(\mathop{\mathrm{\mathbb{E}}}[\alpha x_{t}]\) in \(\bm{\Psi}\)
+determines whether we have an FE or RE model. We can set these to zero
+and test whether the RE model would be preferable to the FE model. In
+general, if the individual effects are truly uncorrelated with the model
+covariates, it is advisable to switch to an RE model since because it
+uses up less degrees of freedom, it will have smaller standard errors
+\citep{Bollen2010}.
+
+In the following \texttt{lavaan} code, we simply remove the factor
+loadings of one for the latent individual effect variable which allows
+them to be estimated freely at each timepoint. For the effect of the
+covariate, we can either delete the constraints \texttt{b} in
+\texttt{yt\ \textasciitilde{}\ b*xt} or give each regression a different
+label, e.g., \texttt{b1}, \texttt{b2}, \texttt{b3}, etc. Similarly, to
+allow the error variance to vary over time, we turn the constraints
+\texttt{e} into simple labels, i.e., \texttt{e1}, \texttt{e2},
+\texttt{e3}, etc., or again just delete them. In fact, regarding the
+error variances, they will be estimated necessarily, and do not need to
+be explicitly mentioned in the model syntax at all. Finally, to move
+from an FE to an RE model, we could simply constrain the correlations
+between the individual effects and the covariates to zero, i.e.,
+\texttt{a\ \textasciitilde{}\textasciitilde{}\ 0*x1\ +\ 0*x2\ +\ 0*x3\ +\ 0*x4\ +\ 0*x5}.
+
+\singlespacing
+
+\begin{Shaded}
+\begin{Highlighting}[numbers=left,,]
+\NormalTok{fe\_sem\_fullyrelaxed \textless{}{-}}\StringTok{ \textquotesingle{}}
+\StringTok{\# Define individual effects variable }
+\StringTok{a =\textasciitilde{} y1 + y2 + y3 + y4 + y5}
+\StringTok{\# Regressions, constrain coefficient to be equal over time}
+\StringTok{y1 \textasciitilde{} b1*x1}
+\StringTok{y2 \textasciitilde{} b2*x2 }
+\StringTok{y3 \textasciitilde{} b3*x3}
+\StringTok{y4 \textasciitilde{} b4*x4}
+\StringTok{y5 \textasciitilde{} b5*x5}
+\StringTok{\# Allow unrestricted correlation between eta and covariates}
+\StringTok{a \textasciitilde{}\textasciitilde{} x1 + x2 + x3 + x4 + x5}
+\StringTok{\# Alternatively: constrain all to 0 for RE model, or}
+\StringTok{\# just individual correlations}
+\StringTok{\# a \textasciitilde{}\textasciitilde{} 0*x1 + 0*x2 + 0*x3 + 0*x4 + 0*x5}
+\StringTok{x1 \textasciitilde{}\textasciitilde{} x2 + x3 + x4 + x5}
+\StringTok{x2 \textasciitilde{}\textasciitilde{} x3 + x4 + x5}
+\StringTok{x3 \textasciitilde{}\textasciitilde{} x4 + x5}
+\StringTok{x4 \textasciitilde{}\textasciitilde{} x5}
+\StringTok{\# Constrain residual variances to be equal over time}
+\StringTok{y1 \textasciitilde{}\textasciitilde{} e1*y1}
+\StringTok{y2 \textasciitilde{}\textasciitilde{} e2*y2}
+\StringTok{y3 \textasciitilde{}\textasciitilde{} e3*y3}
+\StringTok{y4 \textasciitilde{}\textasciitilde{} e4*y4}
+\StringTok{y5 \textasciitilde{}\textasciitilde{} e5*y5}
+\StringTok{\textquotesingle{}}
+\NormalTok{fe\_sem\_fullyrelaxed.fit \textless{}{-}}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe\_sem\_fullyrelaxed, }
+                                \DataTypeTok{data =}\NormalTok{ dfw, }
+                                \DataTypeTok{estimator =} \StringTok{"ML"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\doublespacing
+
+\singlespacing
+
+\doublespacing
+
+As outlined in \citet{Bollen2010}, the researcher has the opportunity to
+test each of the assumptions empirically and decide whether a more
+parsimonious, i.e., restrictive model is justifiable. For each
+assumption, a likelihood ratio test can be carried out to determine
+whether the improvement to model fit resulting from the relaxation of
+various assumptions is significant or whether the more parsimonious
+model is preferable after all.
+
+If we use the original model \texttt{fe\_sem.fit} (from the
+\href{https://github.com/henrik-andersen/FE-SEM/blob/master/article.pdf}{main
+article}) as a starting point, the best strategy for testing these
+assumptions is to work in a stepwise fashion, relaxing one assumption at
+a time. We can begin by first constraining the correlation between
+\(\alpha\) and \(x_{t}\) to zero (\texttt{re\_sem}) for an RE model. If
+turning from an FE to an RE model does not significantly worsen model
+fit, we can go forward with the rest of the steps with the RE model. If,
+however, the fit does worsen significantly, it is likely better to stick
+with the FE model; moving forward then with it to see if a less
+restrictive FE model is preferable. We can perform a likelihood ratio
+test in \texttt{R} using the \texttt{anova()} function:
+
+\small
+\singlespacing
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{anova}\NormalTok{( fe\_sem.fit, re\_sem.fit)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Chi-Squared Difference Test
+## 
+##            Df   AIC   BIC   Chisq Chisq diff Df diff Pr(>Chisq)    
+## fe_sem.fit 32 30998 31111  30.137                                  
+## re_sem.fit 37 31809 31897 850.928     820.79       5  < 2.2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{verbatim}
+
+\doublespacing
+\normalsize
+
+The table that is generated shows a comparison of the nested models, in
+decending order according to degrees of freedom. The RE model does not
+estimate the correlations between the individual effects and the
+covariates, so it is more parsimonious and thus listed at the bottom.
+The \texttt{Chisq} column shows the \(\chi^{2}\) statistic for both
+models and the \texttt{Chisq\ diff} column calculates the difference
+between the two. Obviously, according to the DGP, the correlation
+between the individual effects and \(x_{t}\) is not zero, so fixing
+these to zero leads to a substantial amount of misfit. The last column
+puts the \(\chi^{2}\) difference in relation to the difference in
+degrees of freedom and gives a p-value for the probability that the
+difference is solely due to chance. Here, the change in \(\chi^{2}\) is
+highly significant, so the FE model should be retained.
+
+After now having established, based on the likelihood ratio test, that
+FE is our preferred model, we can begin relaxing the rest of the
+assumptions. I show the following merely as a demonstration of the
+procedure, we know already from the DGP that the parsimonious model as
+specified in \texttt{fe\_sem.fit} is appropriate. We can next allow the
+error variances (\texttt{fe\_semb.fit}), the effect of \(x\) on \(y\)
+(\texttt{fe\_semc.fit}) and finally the factor loadings of the
+individual effects (\texttt{fe\_semd.fit}) all to vary over time.
+
+\singlespacing
+
+\doublespacing
+
+\small
+\singlespacing
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{anova}\NormalTok{( fe\_sem.fit, fe\_semb.fit, fe\_semc.fit, fe\_semd.fit)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Chi-Squared Difference Test
+## 
+##             Df   AIC   BIC  Chisq Chisq diff Df diff Pr(>Chisq)
+## fe_semd.fit 20 31017 31189 25.140                              
+## fe_semc.fit 24 31010 31162 25.764     0.6249       4     0.9603
+## fe_semb.fit 28 31003 31135 26.686     0.9215       4     0.9215
+## fe_sem.fit  32 30998 31111 30.137     3.4516       4     0.4853
+\end{verbatim}
+
+\doublespacing
+\normalsize
+
+Keep in mind that a less parsimonious model (fewer degrees of freedom)
+can never fit worse than a more parsimonious one (more degrees of
+freedom). I.e., chance variations due to sampling error mean that adding
+constraints to a model will tend to always worsen fit, at least
+minimally. The question here is whether the improvement to fit by
+loosening constraints is meaningful or not. In the table above, we
+should not expect any meaningful improvements moving from
+\texttt{fe\_sem.fit} to \texttt{fe\_semd.fit}. Here, using simulated
+data, we have the luxury of knowing that any significant differences in
+\(\chi^{2}\) are due to chance. With real data, it is up to the
+researcher to apply their best judgment and decide whether the results
+are plausible or not.
+
+\hypertarget{time-invariant}{%
+\subsection{Time-invariant predictors}\label{time-invariant}}
+
+What if we do not just want to just control for the effects of all
+time-invariant variables, but investigate some of them in detail? Many
+time-invariant variables, like sex, birth cohort, nationality,
+education, etc. can be interesting on their own. And typically, many of
+these variables are readily available in a given dataset. The
+traditional OLS-based FE model does not allow for this, as it wipes out
+the effect of \emph{all} time-invariant variables, whether observed or
+not.
+
+In SEM, we can easily specify a type of \emph{hybrid} FE/RE model
+\citep{Bollen2010} that allows us to control for time-invariant
+unobserved heterogeneity while also investigating the effects of
+specific observed time-invariant predictors.\footnote{These types of
+  models have become well known outside of SEM as well, see for example
+  \citet{Allison2011}; \citet{Schunck2013}; \citet{Bell2018}.}
+
+In the next example, we continue with the most complex model we have
+specified so far, \texttt{fe\_sem4.fit} in which measurement error in
+both the independent and dependent variables is accounted for using
+latent variables. Now, we would like as well to specifically investigate
+the effect of \(\alpha_{2}\) on the dependent variable. The equation for
+this model changes to:
+\(\eta_{t} = \beta \xi_{t} + \alpha + \gamma \alpha_{2} + \zeta_{t}\).
+
+\singlespacing
+
+\begin{Shaded}
+\begin{Highlighting}[numbers=left,,]
+\NormalTok{fe\_sem5 \textless{}{-}}\StringTok{ \textquotesingle{}}
+\StringTok{\# Measurement model for dependent variable, n for eta}
+\StringTok{n1 =\textasciitilde{} 1*y11 + y21 + y31}
+\StringTok{n2 =\textasciitilde{} 1*y12 + y22 + y32}
+\StringTok{n3 =\textasciitilde{} 1*y13 + y23 + y33}
+\StringTok{n4 =\textasciitilde{} 1*y14 + y24 + y34}
+\StringTok{n5 =\textasciitilde{} 1*y15 + y25 + y35}
+\StringTok{\# Define individual effects variable }
+\StringTok{a =\textasciitilde{} 1*n1 + 1*n2 + 1*n3 + 1*n4 + 1*n5}
+\StringTok{\# Measurement model for independent variables, xi }
+\StringTok{xi1 =\textasciitilde{} 1*x11 + x21 + x31 }
+\StringTok{xi2 =\textasciitilde{} 1*x12 + x22 + x32}
+\StringTok{xi3 =\textasciitilde{} 1*x13 + x23 + x33}
+\StringTok{xi4 =\textasciitilde{} 1*x14 + x24 + x34}
+\StringTok{xi5 =\textasciitilde{} 1*x15 + x25 + x35}
+\StringTok{\# Regressions, constrain coefficient to be equal over time}
+\StringTok{n1 \textasciitilde{} b*xi1 + g*a2}
+\StringTok{n2 \textasciitilde{} b*xi2 + g*a2}
+\StringTok{n3 \textasciitilde{} b*xi3 + g*a2}
+\StringTok{n4 \textasciitilde{} b*xi4 + g*a2}
+\StringTok{n5 \textasciitilde{} b*xi5 + g*a2}
+\StringTok{\# Allow unrestricted correlation between eta and covariates}
+\StringTok{a \textasciitilde{}\textasciitilde{} xi1 + xi2 + xi3 + xi4 + xi5 + 0*a2}
+\StringTok{a2 \textasciitilde{}\textasciitilde{} xi1 + xi2 + xi3 + xi4 + xi5}
+\StringTok{xi1 \textasciitilde{}\textasciitilde{} xi2 + xi3 + xi4 + xi5}
+\StringTok{xi2 \textasciitilde{}\textasciitilde{} xi3 + xi4 + xi5}
+\StringTok{xi3 \textasciitilde{}\textasciitilde{} xi4 + xi5}
+\StringTok{xi4 \textasciitilde{}\textasciitilde{} xi5}
+\StringTok{\# Constrain residual variances to be equal over time}
+\StringTok{n1 \textasciitilde{}\textasciitilde{} e*n1}
+\StringTok{n2 \textasciitilde{}\textasciitilde{} e*n2}
+\StringTok{n3 \textasciitilde{}\textasciitilde{} e*n3}
+\StringTok{n4 \textasciitilde{}\textasciitilde{} e*n4}
+\StringTok{n5 \textasciitilde{}\textasciitilde{} e*n5}
+\StringTok{\textquotesingle{}}
+\NormalTok{fe\_sem5.fit \textless{}{-}}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe\_sem5, }
+                    \DataTypeTok{data =}\NormalTok{ dfw, }
+                    \DataTypeTok{estimator =} \StringTok{"ML"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\doublespacing
+
+Keep in mind, based on the DGP, the true parameters are \(\beta = 0.3\)
+and \(\gamma = 0.45\).
+
+\singlespacing
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{summary}\NormalTok{(fe\_sem5.fit)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+...
+## Regressions:
+##                    Estimate  Std.Err  z-value  P(>|z|)
+##   n1 ~                                                
+##     xi1        (b)    0.265    0.023   11.515    0.000
+##     a2         (g)    0.490    0.033   14.999    0.000
+##   n2 ~                                                
+##     xi2        (b)    0.265    0.023   11.515    0.000
+##     a2         (g)    0.490    0.033   14.999    0.000
+##   n3 ~                                                
+##     xi3        (b)    0.265    0.023   11.515    0.000
+##     a2         (g)    0.490    0.033   14.999    0.000
+##   n4 ~                                                
+##     xi4        (b)    0.265    0.023   11.515    0.000
+##     a2         (g)    0.490    0.033   14.999    0.000
+##   n5 ~                                                
+##     xi5        (b)    0.265    0.023   11.515    0.000
+##     a2         (g)    0.490    0.033   14.999    0.000
+...
+\end{verbatim}
+
+\doublespacing
+
+From this we can see that such a hybrid model is does a good job of
+estimating the coefficients of interest, with \(\hat{\beta} =\) 0.265
+(0.023) and \(\hat{\gamma} =\) 0.49 (0.033).
+
+It is important, however, to realize that the unbiasedness of
+\(\hat{\gamma}\) in this model is dependent on the assumption that
+\(\mathop{\mathrm{\mathbb{E}}}[\zeta | \bm{\xi_{t}}, \alpha_{2}] = 0\).
+In other words, the idiosyncratic error is mean independent of
+\(\bm{\xi_{t}} = (\xi_{1}, \xi_{2}, ..., \xi_{T})\) as well as
+\(\alpha_{2}\). The first part is easier to accept because we are
+controlling for all potential time-invariant confounders that could
+induce a relationship between the independent variable and the error.
+The unbiasedness of \(\hat{\gamma}\), on the other hand rests on the
+assumption that the time-invariant predictor is independent of the
+error. If \(\alpha_{2}\) represented the respondent's intelligence and
+\(\eta_{t}\), the dependent variable, represented the respondent's
+income, for example, then \(\hat{\gamma}\) would be biased if both were
+dependent on a third time-invariant variable, say level of schooling, if
+it is not controlled for. For this reason, we need to treat the
+regression on a time-invariant predictor like any other regular
+multivariate regression model and look to include all plausible
+potential confounders as controls in the model, or turn to other
+methods, e.g., instrumental variables.
+
 \hypertarget{conclusion}{%

### 2021-01-14 article_format/extensions_format_revision.Rmd
Started major changes for resubmission

--- 2020-07-28 article_format/extensions_format.Rmd
@@
 Then, I show a type of hybrid FE/RE model that allows us to control for time-invariant unobserved heterogeneity while including time-invariant predictors in the model.
+
+# Model notation
+
+The following is an outline of matrix notation for those that prefer it to the path diagrams and model syntax.
+There are a number of different model notations (see, for example @Bollen1989 for an overview), but the one that will serve us best is one that was proposed by @Graff1979:
+\begin{align}
+\bm{y}^{*} & = \bm{\Lambda_{y}}^{*} \bm{\eta}^{*}, \\
+\bm{\eta}^{*} & = \bm{B}\bm{\eta}^{*} + \bm{\zeta}^{*}, 
+\end{align}
+where $\bm{\eta}^{*} = (\bm{y}, \bm{x}, \bm{\eta}, \bm{\xi})^{\intercal}$, $\bm{\zeta}^{*} = (\bm{\varepsilon}, \bm{\delta}, \bm{\zeta}, \bm{\xi})^{\intercal}$, $\bm{y}^{*} = (\bm{y}, \bm{x})^{\intercal}$.
+$\bm{y}$ is a vector of observed dependent variables and $\bm{x}$ is a vector of observed independent variables.
+$\bm{\eta}$ is a vector of the latent dependent variables and $\bm{\xi}$ is a vector of latent independent variables.
+$\bm{\varepsilon}$ and $\bm{\delta}$ are vectors of the errors of the observed dependent and independent variables, respectively, and $\bm{\zeta}$ is a vector of the errors, or disturbances, of the latent variables.
+Notice the $^{*}$ symbol is just meant to differentiate the vectors with them from those without them.
+That means, $\bm{\eta}^{*}$ is a vector that holds the observed and latent variables, both dependent (in SEM they are referred to as 'endogenous') and independent (i.e., 'exogeneous')^[For our purposes, the terms endogenous and dependent, on the one hand, and exogenous and independent, on the other, can be used interchangeably.], $\bm{\zeta}^{*}$ holds the errors for the observed variables and the disturbances of the latent variables.
+$\bm{y}^{*}$ holds just the observed variables, both dependent and independent, and $\bm{\Lambda_{y}}^{*}$ is a matrix of ones and zeros that selects the observed variables from $\bm{\eta}^{*}$.
+Lastly, $\bm{B}$ is a matrix that holds the regression coefficients.^[If we say that $p$ and $q$ stand for the number of observed dependent and independent variables, respectively, and $m$ and $n$ stand for the number of latent dependent and independent variables, respectively, then $\bm{\eta}^{*}$ and $\bm{\zeta}^{*}$ are $p + q + m + n$, $\bm{y}^{*}$ is $p + q$, $\bm{\Lambda_{y}}^{*}$ is $(p + q) \times (p + q + m + n)$ and $\bm{B}$ is $(p + q + m + n) \times (p + q + m + n)$ [@Bollen1989].]
+
+This notation may be confusing at first, but it has advantages.
+First, it allows us the flexibility we need for the models.
+For example, it allows observed $\bm{x}$ to directly influence observed $\bm{y}$ (more common notation assumes that substantive effects occur only between latent variables, observed ones are only used as indicators, see for example @Bollen1989, @Kline2016).
+It also allows $\bm{\xi}$, i.e., any latent exogenous variables, to influence $\bm{y}$ directly.
+These two scenarios cover the traditional FE model with observed variables, and one in which latent variables are used to account for measurement error in the independent variables.
+It is also consistent with the notation used for these models in `lavaan`.
+In fact, `lavaan` switches automatically between matrix notations depending on the specified model.
+That means the matrix representation of the model one sees if they type in `lavInspect(model, what = "est")` after specifying their model in `lavaan` will match up with the notation used here.^[In `Mplus`, the model matrices can be requested by including `OUTPUT: TECH1` in the input file.] It does have a potential disadvantage however.
+Besides being less intuitive than the typical $\bm{\eta} = \bm{B}\bm{\eta} + \bm{\Gamma}\bm{\xi} + \bm{\zeta}$ notation, it means that by including the observed covariates in the stacked long vector $\bm{y}^{*}$, they are treated as another response variable with variances and covariances to be estimated by the model (instead of just using the sample statistics).
+This means that the assumption of multivariate normality (otherwise just imposed on the dependent variables) also applies to the independent ones [@Skrondal2004, p.
+75].
+This can be problematic for noncontinuous independent variables like sex, nationality dummies, marriage status (married/unmarried), etc.
+See @Skrondal2004 for more on this topic.
+
+Let us, however, make things more concrete and take a look at a simple, three-wave version of the typical FE-SEM using this notation (shown graphically in Figure \ref{fig:fesem}).
+For that, we have the following matrix notation (with labels on the outside of the matrices):
+\begin{align}
+\begin{split}
+\bm{y}^{*} & = \bm{\Lambda_{y}}^{*}\bm{\eta}^{*} \\
+\begin{pmatrix}
+y_{1} \\
+y_{2} \\
+y_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3}
+\end{pmatrix} & = 
+\begin{blockarray}{cccccccc}
+ & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
+ \begin{block}{c(ccccccc)}
+ y_{1} & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
+ y_{2} & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
+ y_{3} & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\ 
+ x_{1} & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
+ x_{2} & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
+ x_{3} & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
+ \end{block}
+\end{blockarray}
+\begin{pmatrix}
+y_{1} \\
+y_{2} \\
+y_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3} \\
+\alpha
+\end{pmatrix},
+\end{split} \label{eq:y}
+\end{align}
+\begin{align}
+\begin{split}
+\bm{\eta}^{*} & = \bm{B}\bm{\eta}^{*} + \bm{\zeta}^{*} \\
+\begin{pmatrix}
+y_{1} \\
+y_{2} \\
+y_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3} \\
+\alpha
+\end{pmatrix} & = 
+\begin{blockarray}{cccccccc}
+ & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
+ \begin{block}{c(ccccccc)}
+ y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
+ y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
+ y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
+ x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+ x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+ x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+ \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+ \end{block}
+\end{blockarray}
+\begin{pmatrix}
+y_{1} \\
+y_{2} \\
+y_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3} \\
+\alpha
+\end{pmatrix} + 
+\begin{pmatrix}
+\varepsilon_{1} \\
+\varepsilon_{2} \\
+\varepsilon_{3} \\
+x_{1} = \delta_{1} \\
+x_{2} = \delta_{2} \\
+x_{3} = \delta_{3} \\
+\alpha = \xi \\
+\end{pmatrix}. 
+\end{split} \label{eq:eta}
+\end{align} 
+Notice in $\bm{\zeta}^{*}$, for the independent variables we could either write, for example $x_{t}$ or $\delta_{t}$.
+As mentioned above, this is due to the model notation treating the independent variables like dependent variables with variances/covariances to be estimated.
+For the sake of simplicity, we will ignore this subtlety and refer to the observed variable from now on, keeping in mind that if the multivariate normality assumption holds, the estimated statistics will likely be sufficiently close to the sample ones for it to not make much of a difference.
+
+Admittedly, Equations \eqref{eq:y} and \eqref{eq:eta} may not look like much yet.
+We can remedy this by first putting the equation for $\bm{\eta}^{*}$ in reduced form, i.e., by getting rid of the dependent variable on the r.h.s.:
+\begin{align}
+\bm{\eta}^{*} & = \bm{B}\bm{\eta}^{*} + \bm{\zeta}^{*} \\
+\bm{\eta}^{*} - \bm{B}\bm{\eta}^{*} & = \bm{\zeta}^{*} \\
+(\bm{I} - \bm{B})\bm{\eta}^{*} & = \bm{\zeta}^{*} \\
+\bm{\eta}^{*} & = (\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*},
+\end{align}
+where $\bm{I}$ is the identity matrix.
+By substituting this back into the equation for the observed variables we get $\bm{y}^{*} = \bm{\Lambda_{y}}^{*}[(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*}]$, which works out to:
+\begin{align}
+\begin{pmatrix} 
+y_{1} \\
+y_{2} \\
+y_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3} \\
+\end{pmatrix} & = 
+\begin{pmatrix}
+\alpha + \beta x_{1} + \varepsilon_{1} \\
+\alpha + \beta x_{2} + \varepsilon_{2} \\
+\alpha + \beta x_{3} + \varepsilon_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3}
+\end{pmatrix},
+\end{align}
+which is of course exactly what we should expect given Equation \eqref{eq:fe}^[We can use the `sympy` package in `python` to verify and show the steps for this and other examples, see the [supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/sympy-doublecheck-matrixnotation.py) for the code.].
+
+## Assumptions
+
+What essentially differentiates an FE from an RE model is our assumption concerning the relationship between the unobserved individual effects and the model covariates [@Bollen2010].
+The FE model assumes that $\E[\alpha x_{t}] \ne 0$.
+As such, if we fail to control for the correlation of the covariate and the time-invariant part of the error, then the coefficient of interest, here $\beta$, will be biased.
+Our assumption regarding whether the individual effects are correlated with the model covariates occurs in $\E[\bm{\zeta}^{*}\bm{\zeta}^{* \intercal}] = \bm{\Psi}$, the covariance matrix of the errors
+\begin{align}
+\bm{y}^{*}\bm{y}^{* \intercal} & = \E[(\bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*})(\bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*})^{\intercal}] \\
+ & = \E[(\bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*})(\bm{\zeta}^{* \intercal}(\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{* \intercal})] \\
+ & = \bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1} \E[\bm{\zeta}^{*}\bm{\zeta}^{* \intercal}] (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{* \intercal} \\
+ & = \bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1} \bm{\Psi} (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{* \intercal}.
+\end{align}
+In the case of an FE model, $\bm{\Psi}$ will reflect our belief that the individual effects are correlated with the model covariates, here again for demonstration the three-wave model:
+\begin{align}
+\bm{\Psi} & = \E
+\begin{blockarray}{cccccccc}
+ & \varepsilon_{1} & \varepsilon_{2} & \varepsilon_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
+ \begin{block}{c(ccccccc)}
+ \varepsilon_{1} & \varepsilon_{1}^{2} &                     &                     &              &              &              & \\
+ \varepsilon_{2} & 0                   & \varepsilon_{2}^{2} &                     &              &              &              & \\
+ \varepsilon_{3} & 0                   & 0                   & \varepsilon_{3}^{2} &              &              &              & \\
+ x_{1}           & 0                   & 0                   & 0                   & x_{1}^{2}    &              &              & \\
+ x_{2}           & 0                   & 0                   & 0                   & x_{2}x_{1}   & x_{2}^{2}    &              & \\
+ x_{3}           & 0                   & 0                   & 0                   & x_{3}x_{1}   & x_{3}x_{2}   & x_{3}^{2}    & \\
+ \alpha          & 0                   & 0                   & 0                   & \alpha x_{1} & \alpha x_{2} & \alpha x_{3} & \alpha^{2} \\
+ \end{block}
+\end{blockarray}.
+\end{align}
+Knowing this, we can work out the equation for the coefficient of interest, $\beta$. For the sake of simplicity, assume here and throughout mean-centered variables: 
+\begin{align}
+\Cov(y_{t},x_{t}) & = \E[y_{t}x_{t}] \\
+ & = \E[(\alpha + \beta x_{t} + \varepsilon_{t})x_{t}] \\
+ & = \E[\alpha x_{t} + \beta x_{t}^{2} + \varepsilon_{t}x_{t}] \\
+ & = \Cov(\alpha, x_{t}) + \beta \Var(x_{t}) \\
+\hat{\beta}_{FE-SEM} & = \frac{\Cov(y_{t},x_{t}) - \Cov(\alpha, x_{t})}{\Var(x_{t})}. 
+\end{align}
+This should make intuitive sense.
+From the observed covariance between the dependent and the independent variable, we are partialling out the part that is due to the covariance between the independent variable and the individual effects per unit, and then dividing by the variance of the independent variable, as usual.
+For the RE model, we assume $\E[\alpha x_{t}] = 0$ and the equation reduces to $\hat{\beta}_{RE-SEM} = \Cov(y_{t},x_{t})/\Var(x_{t})$.
+The rest of the model-implied covariance matrix results from $\bm{y}^{*}\bm{y}^{* \intercal}$.
 
@@
 \end{align}
-where $\theta$ is just the residual of a regression in which the true underlying variable is regressed on all other covariates.
+where $\theta$ is just the residual of a regression in which the underlying theoretical variable is regressed on all other covariates.
 In this case, we need to regress $x_{t}$ on $\alpha_{1}$ and $\alpha_{2}$ for: $x_{t} = \tau + \gamma_{1}\alpha_{1} + \gamma_{2}\alpha_{2} + \theta_{t}$ where $\tau$ is the intercept, and $\gamma_{1}$, $\gamma_{2}$ are the regression coefficients and $\theta_{t}$ is the residual [@Wooldridge2009, p.
@@
 
-Normally it is not possible to reconstruct the bias since in cases where we have to rely on indicators, we would not know the true underlying variable.
+Normally it is not possible to reconstruct the bias since in cases where we have to rely on indicators, we would not have observed the underlying theoretical variable.
 Furthermore, in the case of a fixed-effects model, the covariates are the unobserved time-invariant characteristics.

### 2021-01-14 article_format/extensions_format_revision.tex
Started major changes for resubmission

--- 2020-07-28 article_format/extensions_format.tex
@@
 
+\hypertarget{model-notation}{%
+\section{Model notation}\label{model-notation}}
+
+The following is an outline of matrix notation for those that prefer it
+to the path diagrams and model syntax. There are a number of different
+model notations (see, for example \citet{Bollen1989} for an overview),
+but the one that will serve us best is one that was proposed by
+\citet{Graff1979}: \begin{align}
+\bm{y}^{*} & = \bm{\Lambda_{y}}^{*} \bm{\eta}^{*}, \\
+\bm{\eta}^{*} & = \bm{B}\bm{\eta}^{*} + \bm{\zeta}^{*}, 
+\end{align} where
+\(\bm{\eta}^{*} = (\bm{y}, \bm{x}, \bm{\eta}, \bm{\xi})^{\intercal}\),
+\(\bm{\zeta}^{*} = (\bm{\varepsilon}, \bm{\delta}, \bm{\zeta}, \bm{\xi})^{\intercal}\),
+\(\bm{y}^{*} = (\bm{y}, \bm{x})^{\intercal}\). \(\bm{y}\) is a vector of
+observed dependent variables and \(\bm{x}\) is a vector of observed
+independent variables. \(\bm{\eta}\) is a vector of the latent dependent
+variables and \(\bm{\xi}\) is a vector of latent independent variables.
+\(\bm{\varepsilon}\) and \(\bm{\delta}\) are vectors of the errors of
+the observed dependent and independent variables, respectively, and
+\(\bm{\zeta}\) is a vector of the errors, or disturbances, of the latent
+variables. Notice the \(^{*}\) symbol is just meant to differentiate the
+vectors with them from those without them. That means, \(\bm{\eta}^{*}\)
+is a vector that holds the observed and latent variables, both dependent
+(in SEM they are referred to as `endogenous') and independent (i.e.,
+`exogeneous')\footnote{For our purposes, the terms endogenous and
+  dependent, on the one hand, and exogenous and independent, on the
+  other, can be used interchangeably.}, \(\bm{\zeta}^{*}\) holds the
+errors for the observed variables and the disturbances of the latent
+variables. \(\bm{y}^{*}\) holds just the observed variables, both
+dependent and independent, and \(\bm{\Lambda_{y}}^{*}\) is a matrix of
+ones and zeros that selects the observed variables from
+\(\bm{\eta}^{*}\). Lastly, \(\bm{B}\) is a matrix that holds the
+regression coefficients.\footnote{If we say that \(p\) and \(q\) stand
+  for the number of observed dependent and independent variables,
+  respectively, and \(m\) and \(n\) stand for the number of latent
+  dependent and independent variables, respectively, then
+  \(\bm{\eta}^{*}\) and \(\bm{\zeta}^{*}\) are \(p + q + m + n\),
+  \(\bm{y}^{*}\) is \(p + q\), \(\bm{\Lambda_{y}}^{*}\) is
+  \((p + q) \times (p + q + m + n)\) and \(\bm{B}\) is
+  \((p + q + m + n) \times (p + q + m + n)\) \citep{Bollen1989}.}
+
+This notation may be confusing at first, but it has advantages. First,
+it allows us the flexibility we need for the models. For example, it
+allows observed \(\bm{x}\) to directly influence observed \(\bm{y}\)
+(more common notation assumes that substantive effects occur only
+between latent variables, observed ones are only used as indicators, see
+for example \citet{Bollen1989}, \citet{Kline2016}). It also allows
+\(\bm{\xi}\), i.e., any latent exogenous variables, to influence
+\(\bm{y}\) directly. These two scenarios cover the traditional FE model
+with observed variables, and one in which latent variables are used to
+account for measurement error in the independent variables. It is also
+consistent with the notation used for these models in \texttt{lavaan}.
+In fact, \texttt{lavaan} switches automatically between matrix notations
+depending on the specified model. That means the matrix representation
+of the model one sees if they type in
+\texttt{lavInspect(model,\ what\ =\ "est")} after specifying their model
+in \texttt{lavaan} will match up with the notation used here.\footnote{In
+  \texttt{Mplus}, the model matrices can be requested by including
+  \texttt{OUTPUT:\ TECH1} in the input file.} It does have a potential
+disadvantage however. Besides being less intuitive than the typical
+\(\bm{\eta} = \bm{B}\bm{\eta} + \bm{\Gamma}\bm{\xi} + \bm{\zeta}\)
+notation, it means that by including the observed covariates in the
+stacked long vector \(\bm{y}^{*}\), they are treated as another response
+variable with variances and covariances to be estimated by the model
+(instead of just using the sample statistics). This means that the
+assumption of multivariate normality (otherwise just imposed on the
+dependent variables) also applies to the independent ones
+\citep[p.~75]{Skrondal2004}. This can be problematic for noncontinuous
+independent variables like sex, nationality dummies, marriage status
+(married/unmarried), etc. See \citet{Skrondal2004} for more on this
+topic.
+
+Let us, however, make things more concrete and take a look at a simple,
+three-wave version of the typical FE-SEM using this notation (shown
+graphically in Figure \ref{fig:fesem}). For that, we have the following
+matrix notation (with labels on the outside of the matrices):
+\begin{align}
+\begin{split}
+\bm{y}^{*} & = \bm{\Lambda_{y}}^{*}\bm{\eta}^{*} \\
+\begin{pmatrix}
+y_{1} \\
+y_{2} \\
+y_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3}
+\end{pmatrix} & = 
+\begin{blockarray}{cccccccc}
+ & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
+ \begin{block}{c(ccccccc)}
+ y_{1} & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
+ y_{2} & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
+ y_{3} & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\ 
+ x_{1} & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
+ x_{2} & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
+ x_{3} & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
+ \end{block}
+\end{blockarray}
+\begin{pmatrix}
+y_{1} \\
+y_{2} \\
+y_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3} \\
+\alpha
+\end{pmatrix},
+\end{split} \label{eq:y}
+\end{align} \begin{align}
+\begin{split}
+\bm{\eta}^{*} & = \bm{B}\bm{\eta}^{*} + \bm{\zeta}^{*} \\
+\begin{pmatrix}
+y_{1} \\
+y_{2} \\
+y_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3} \\
+\alpha
+\end{pmatrix} & = 
+\begin{blockarray}{cccccccc}
+ & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
+ \begin{block}{c(ccccccc)}
+ y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
+ y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
+ y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
+ x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+ x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+ x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+ \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+ \end{block}
+\end{blockarray}
+\begin{pmatrix}
+y_{1} \\
+y_{2} \\
+y_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3} \\
+\alpha
+\end{pmatrix} + 
+\begin{pmatrix}
+\varepsilon_{1} \\
+\varepsilon_{2} \\
+\varepsilon_{3} \\
+x_{1} = \delta_{1} \\
+x_{2} = \delta_{2} \\
+x_{3} = \delta_{3} \\
+\alpha = \xi \\
+\end{pmatrix}. 
+\end{split} \label{eq:eta}
+\end{align} Notice in \(\bm{\zeta}^{*}\), for the independent variables
+we could either write, for example \(x_{t}\) or \(\delta_{t}\). As
+mentioned above, this is due to the model notation treating the
+independent variables like dependent variables with
+variances/covariances to be estimated. For the sake of simplicity, we
+will ignore this subtlety and refer to the observed variable from now
+on, keeping in mind that if the multivariate normality assumption holds,
+the estimated statistics will likely be sufficiently close to the sample
+ones for it to not make much of a difference.
+
+Admittedly, Equations \eqref{eq:y} and \eqref{eq:eta} may not look like
+much yet. We can remedy this by first putting the equation for
+\(\bm{\eta}^{*}\) in reduced form, i.e., by getting rid of the dependent
+variable on the r.h.s.: \begin{align}
+\bm{\eta}^{*} & = \bm{B}\bm{\eta}^{*} + \bm{\zeta}^{*} \\
+\bm{\eta}^{*} - \bm{B}\bm{\eta}^{*} & = \bm{\zeta}^{*} \\
+(\bm{I} - \bm{B})\bm{\eta}^{*} & = \bm{\zeta}^{*} \\
+\bm{\eta}^{*} & = (\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*},
+\end{align} where \(\bm{I}\) is the identity matrix. By substituting
+this back into the equation for the observed variables we get
+\(\bm{y}^{*} = \bm{\Lambda_{y}}^{*}[(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*}]\),
+which works out to: \begin{align}
+\begin{pmatrix} 
+y_{1} \\
+y_{2} \\
+y_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3} \\
+\end{pmatrix} & = 
+\begin{pmatrix}
+\alpha + \beta x_{1} + \varepsilon_{1} \\
+\alpha + \beta x_{2} + \varepsilon_{2} \\
+\alpha + \beta x_{3} + \varepsilon_{3} \\
+x_{1} \\
+x_{2} \\
+x_{3}
+\end{pmatrix},
+\end{align} which is of course exactly what we should expect given
+Equation \eqref{eq:fe}\footnote{We can use the \texttt{sympy} package in
+  \texttt{python} to verify and show the steps for this and other
+  examples, see the
+  \href{https://github.com/henrik-andersen/FE-SEM/blob/master/sympy-doublecheck-matrixnotation.py}{supplementary
+  materials} for the code.}.
+
+\hypertarget{assumptions}{%
+\subsection{Assumptions}\label{assumptions}}
+
+What essentially differentiates an FE from an RE model is our assumption
+concerning the relationship between the unobserved individual effects
+and the model covariates \citep{Bollen2010}. The FE model assumes that
+\(\mathop{\mathrm{\mathbb{E}}}[\alpha x_{t}] \ne 0\). As such, if we
+fail to control for the correlation of the covariate and the
+time-invariant part of the error, then the coefficient of interest, here
+\(\beta\), will be biased. Our assumption regarding whether the
+individual effects are correlated with the model covariates occurs in
+\(\mathop{\mathrm{\mathbb{E}}}[\bm{\zeta}^{*}\bm{\zeta}^{* \intercal}] = \bm{\Psi}\),
+the covariance matrix of the errors \begin{align}
+\bm{y}^{*}\bm{y}^{* \intercal} & = \mathop{\mathrm{\mathbb{E}}}[(\bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*})(\bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*})^{\intercal}] \\
+ & = \mathop{\mathrm{\mathbb{E}}}[(\bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{*})(\bm{\zeta}^{* \intercal}(\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{* \intercal})] \\
+ & = \bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1} \mathop{\mathrm{\mathbb{E}}}[\bm{\zeta}^{*}\bm{\zeta}^{* \intercal}] (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{* \intercal} \\
+ & = \bm{\Lambda_{y}}^{*}(\bm{I} - \bm{B})^{-1} \bm{\Psi} (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{* \intercal}.
+\end{align} In the case of an FE model, \(\bm{\Psi}\) will reflect our
+belief that the individual effects are correlated with the model
+covariates, here again for demonstration the three-wave model:
+\begin{align}
+\bm{\Psi} & = \mathop{\mathrm{\mathbb{E}}}
+\begin{blockarray}{cccccccc}
+ & \varepsilon_{1} & \varepsilon_{2} & \varepsilon_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
+ \begin{block}{c(ccccccc)}
+ \varepsilon_{1} & \varepsilon_{1}^{2} &                     &                     &              &              &              & \\
+ \varepsilon_{2} & 0                   & \varepsilon_{2}^{2} &                     &              &              &              & \\
+ \varepsilon_{3} & 0                   & 0                   & \varepsilon_{3}^{2} &              &              &              & \\
+ x_{1}           & 0                   & 0                   & 0                   & x_{1}^{2}    &              &              & \\
+ x_{2}           & 0                   & 0                   & 0                   & x_{2}x_{1}   & x_{2}^{2}    &              & \\
+ x_{3}           & 0                   & 0                   & 0                   & x_{3}x_{1}   & x_{3}x_{2}   & x_{3}^{2}    & \\
+ \alpha          & 0                   & 0                   & 0                   & \alpha x_{1} & \alpha x_{2} & \alpha x_{3} & \alpha^{2} \\
+ \end{block}
+\end{blockarray}.
+\end{align} Knowing this, we can work out the equation for the
+coefficient of interest, \(\beta\). For the sake of simplicity, assume
+here and throughout mean-centered variables: \begin{align}
+\mathop{\mathrm{\mathrm{Cov}}}(y_{t},x_{t}) & = \mathop{\mathrm{\mathbb{E}}}[y_{t}x_{t}] \\
+ & = \mathop{\mathrm{\mathbb{E}}}[(\alpha + \beta x_{t} + \varepsilon_{t})x_{t}] \\
+ & = \mathop{\mathrm{\mathbb{E}}}[\alpha x_{t} + \beta x_{t}^{2} + \varepsilon_{t}x_{t}] \\
+ & = \mathop{\mathrm{\mathrm{Cov}}}(\alpha, x_{t}) + \beta \mathop{\mathrm{\mathrm{Var}}}(x_{t}) \\
+\hat{\beta}_{FE-SEM} & = \frac{\mathop{\mathrm{\mathrm{Cov}}}(y_{t},x_{t}) - \mathop{\mathrm{\mathrm{Cov}}}(\alpha, x_{t})}{\mathop{\mathrm{\mathrm{Var}}}(x_{t})}. 
+\end{align} This should make intuitive sense. From the observed
+covariance between the dependent and the independent variable, we are
+partialling out the part that is due to the covariance between the
+independent variable and the individual effects per unit, and then
+dividing by the variance of the independent variable, as usual. For the
+RE model, we assume \(\mathop{\mathrm{\mathbb{E}}}[\alpha x_{t}] = 0\)
+and the equation reduces to
+\(\hat{\beta}_{RE-SEM} = \mathop{\mathrm{\mathrm{Cov}}}(y_{t},x_{t})/\mathop{\mathrm{\mathrm{Var}}}(x_{t})\).
+The rest of the model-implied covariance matrix results from
+\(\bm{y}^{*}\bm{y}^{* \intercal}\).
+
 \hypertarget{a-comparison-with-non-sem-methods}{%
@@
 
-\CommentTok{# Run the FE model in plm }
-\NormalTok{fe1 <-}\StringTok{ }\KeywordTok{plm}\NormalTok{(y }\OperatorTok{~}\StringTok{ }\NormalTok{x, }
+\CommentTok{\# Run the FE model in plm }
+\NormalTok{fe1 \textless{}{-}}\StringTok{ }\KeywordTok{plm}\NormalTok{(y }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{x, }
            \DataTypeTok{effect =} \StringTok{"individual"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"within"}\NormalTok{, }
@@
 \begin{Highlighting}[]
-\CommentTok{# Generate the cluster means for x per id }
-\NormalTok{clusterMeanx <-}\StringTok{ }\KeywordTok{aggregate}\NormalTok{(df}\OperatorTok{$}\NormalTok{x, }\DataTypeTok{by =} \KeywordTok{list}\NormalTok{(df}\OperatorTok{$}\NormalTok{id), }\DataTypeTok{FUN =}\NormalTok{ mean)}
-\CommentTok{# Rename the columns }
-\KeywordTok{names}\NormalTok{(clusterMeanx) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"id"}\NormalTok{, }\StringTok{"xbar"}\NormalTok{)}
-
-\CommentTok{# Add the cluster means back into df  }
-\NormalTok{df <-}\StringTok{ }\KeywordTok{merge}\NormalTok{(df, clusterMeanx, }\DataTypeTok{by =} \StringTok{"id"}\NormalTok{)}
+\CommentTok{\# Generate the cluster means for x per id }
+\NormalTok{clusterMeanx \textless{}{-}}\StringTok{ }\KeywordTok{aggregate}\NormalTok{(df}\OperatorTok{$}\NormalTok{x, }\DataTypeTok{by =} \KeywordTok{list}\NormalTok{(df}\OperatorTok{$}\NormalTok{id), }\DataTypeTok{FUN =}\NormalTok{ mean)}
+\CommentTok{\# Rename the columns }
+\KeywordTok{names}\NormalTok{(clusterMeanx) \textless{}{-}}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"id"}\NormalTok{, }\StringTok{"xbar"}\NormalTok{)}
+
+\CommentTok{\# Add the cluster means back into df  }
+\NormalTok{df \textless{}{-}}\StringTok{ }\KeywordTok{merge}\NormalTok{(df, clusterMeanx, }\DataTypeTok{by =} \StringTok{"id"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{fe2 <-}\StringTok{ }\KeywordTok{plm}\NormalTok{(y }\OperatorTok{~}\StringTok{ }\NormalTok{x }\OperatorTok{+}\StringTok{ }\NormalTok{xbar, }
+\NormalTok{fe2 \textless{}{-}}\StringTok{ }\KeywordTok{plm}\NormalTok{(y }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{x }\OperatorTok{+}\StringTok{ }\NormalTok{xbar, }
            \DataTypeTok{effect =} \StringTok{"individual"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"random"}\NormalTok{,}
@@
 
-\CommentTok{# Run the mixed model in lmer with the cluster means for x }
-\NormalTok{mixed1 <-}\StringTok{ }\KeywordTok{lmer}\NormalTok{(y }\OperatorTok{~}\StringTok{ }\NormalTok{x }\OperatorTok{+}\StringTok{ }\NormalTok{xbar }\OperatorTok{+}\StringTok{ }\NormalTok{(}\DecValTok{1} \OperatorTok{|}\StringTok{ }\NormalTok{id), }\DataTypeTok{data =}\NormalTok{ df)}
+\CommentTok{\# Run the mixed model in lmer with the cluster means for x }
+\NormalTok{mixed1 \textless{}{-}}\StringTok{ }\KeywordTok{lmer}\NormalTok{(y }\OperatorTok{\textasciitilde{}}\StringTok{ }\NormalTok{x }\OperatorTok{+}\StringTok{ }\NormalTok{xbar }\OperatorTok{+}\StringTok{ }\NormalTok{(}\DecValTok{1} \OperatorTok{|}\StringTok{ }\NormalTok{id), }\DataTypeTok{data =}\NormalTok{ df)}
 \KeywordTok{summary}\NormalTok{(mixed1)}
@@
 \begin{Highlighting}[]
-\CommentTok{# Print the variance components for the plm model  }
+\CommentTok{\# Print the variance components for the plm model  }
 \KeywordTok{print}\NormalTok{(}\KeywordTok{ercomp}\NormalTok{(fe2), }
@@
 \begin{Highlighting}[]
-\CommentTok{# Print the variance components for the lmer model }
+\CommentTok{\# Print the variance components for the lmer model }
 \KeywordTok{print}\NormalTok{(}\KeywordTok{VarCorr}\NormalTok{(mixed1), }
@@
 \begin{Highlighting}[]
-\NormalTok{fe_sem_fullyrelaxed <-}\StringTok{ '}
-\StringTok{# Define individual effects variable }
-\StringTok{a =~ y1 + y2 + y3 + y4 + y5}
-\StringTok{# Regressions, constrain coefficient to be equal over time}
-\StringTok{y1 ~ b1*x1}
-\StringTok{y2 ~ b2*x2 }
-\StringTok{y3 ~ b3*x3}
-\StringTok{y4 ~ b4*x4}
-\StringTok{y5 ~ b5*x5}
-\StringTok{# Allow unrestricted correlation between eta and covariates}
-\StringTok{a ~~ x1 + x2 + x3 + x4 + x5}
-\StringTok{# Alternatively: constrain all to 0 for RE model, or}
-\StringTok{# just individual correlations}
-\StringTok{# a ~~ 0*x1 + 0*x2 + 0*x3 + 0*x4 + 0*x5}
-\StringTok{x1 ~~ x2 + x3 + x4 + x5}
-\StringTok{x2 ~~ x3 + x4 + x5}
-\StringTok{x3 ~~ x4 + x5}
-\StringTok{x4 ~~ x5}
-\StringTok{# Constrain residual variances to be equal over time}
-\StringTok{y1 ~~ e1*y1}
-\StringTok{y2 ~~ e2*y2}
-\StringTok{y3 ~~ e3*y3}
-\StringTok{y4 ~~ e4*y4}
-\StringTok{y5 ~~ e5*y5}
-\StringTok{'}
-\NormalTok{fe_sem_fullyrelaxed.fit <-}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe_sem_fullyrelaxed, }
+\NormalTok{fe\_sem\_fullyrelaxed \textless{}{-}}\StringTok{ \textquotesingle{}}
+\StringTok{\# Define individual effects variable }
+\StringTok{a =\textasciitilde{} y1 + y2 + y3 + y4 + y5}
+\StringTok{\# Regressions, constrain coefficient to be equal over time}
+\StringTok{y1 \textasciitilde{} b1*x1}
+\StringTok{y2 \textasciitilde{} b2*x2 }
+\StringTok{y3 \textasciitilde{} b3*x3}
+\StringTok{y4 \textasciitilde{} b4*x4}
+\StringTok{y5 \textasciitilde{} b5*x5}
+\StringTok{\# Allow unrestricted correlation between eta and covariates}
+\StringTok{a \textasciitilde{}\textasciitilde{} x1 + x2 + x3 + x4 + x5}
+\StringTok{\# Alternatively: constrain all to 0 for RE model, or}
+\StringTok{\# just individual correlations}
+\StringTok{\# a \textasciitilde{}\textasciitilde{} 0*x1 + 0*x2 + 0*x3 + 0*x4 + 0*x5}
+\StringTok{x1 \textasciitilde{}\textasciitilde{} x2 + x3 + x4 + x5}
+\StringTok{x2 \textasciitilde{}\textasciitilde{} x3 + x4 + x5}
+\StringTok{x3 \textasciitilde{}\textasciitilde{} x4 + x5}
+\StringTok{x4 \textasciitilde{}\textasciitilde{} x5}
+\StringTok{\# Constrain residual variances to be equal over time}
+\StringTok{y1 \textasciitilde{}\textasciitilde{} e1*y1}
+\StringTok{y2 \textasciitilde{}\textasciitilde{} e2*y2}
+\StringTok{y3 \textasciitilde{}\textasciitilde{} e3*y3}
+\StringTok{y4 \textasciitilde{}\textasciitilde{} e4*y4}
+\StringTok{y5 \textasciitilde{}\textasciitilde{} e5*y5}
+\StringTok{\textquotesingle{}}
+\NormalTok{fe\_sem\_fullyrelaxed.fit \textless{}{-}}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe\_sem\_fullyrelaxed, }
                                 \DataTypeTok{data =}\NormalTok{ dfw, }
@@
 \begin{Highlighting}[]
-\KeywordTok{anova}\NormalTok{( fe_sem.fit, re_sem.fit)}
+\KeywordTok{anova}\NormalTok{( fe\_sem.fit, re\_sem.fit)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\KeywordTok{anova}\NormalTok{( fe_sem.fit, fe_semb.fit, fe_semc.fit, fe_semd.fit)}
+\KeywordTok{anova}\NormalTok{( fe\_sem.fit, fe\_semb.fit, fe\_semc.fit, fe\_semd.fit)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{fe_sem2 <-}\StringTok{ '}
-\StringTok{# Define individual effects variable }
-\StringTok{a =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5}
-\StringTok{# Regressions, constrain coefficient to be equal over time}
-\StringTok{# Now the imprecisely measured indicator tilde\{x\}_kt}
-\StringTok{# instead of the true variable x_t}
-\StringTok{y1 ~ b*x11 }
-\StringTok{y2 ~ b*x12 }
-\StringTok{y3 ~ b*x13}
-\StringTok{y4 ~ b*x14}
-\StringTok{y5 ~ b*x15}
-\StringTok{# Allow unrestricted correlation between eta and covariates}
-\StringTok{a ~~ x11 + x12 + x13 + x14 + x15}
-\StringTok{x11 ~~ x12 + x13 + x14 + x15}
-\StringTok{x12 ~~ x13 + x14 + x15}
-\StringTok{x13 ~~ x14 + x15}
-\StringTok{x14 ~~ x15}
-\StringTok{# Constrain residual variances to be equal over time}
-\StringTok{y1 ~~ e*y1}
-\StringTok{y2 ~~ e*y2}
-\StringTok{y3 ~~ e*y3}
-\StringTok{y4 ~~ e*y4}
-\StringTok{y5 ~~ e*y5}
-\StringTok{'}
-\NormalTok{fe_sem2.fit <-}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe_sem2, }
+\NormalTok{fe\_sem2 \textless{}{-}}\StringTok{ \textquotesingle{}}
+\StringTok{\# Define individual effects variable }
+\StringTok{a =\textasciitilde{} 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5}
+\StringTok{\# Regressions, constrain coefficient to be equal over time}
+\StringTok{\# Now the imprecisely measured indicator tilde\{x\}\_kt}
+\StringTok{\# instead of the true variable x\_t}
+\StringTok{y1 \textasciitilde{} b*x11 }
+\StringTok{y2 \textasciitilde{} b*x12 }
+\StringTok{y3 \textasciitilde{} b*x13}
+\StringTok{y4 \textasciitilde{} b*x14}
+\StringTok{y5 \textasciitilde{} b*x15}
+\StringTok{\# Allow unrestricted correlation between eta and covariates}
+\StringTok{a \textasciitilde{}\textasciitilde{} x11 + x12 + x13 + x14 + x15}
+\StringTok{x11 \textasciitilde{}\textasciitilde{} x12 + x13 + x14 + x15}
+\StringTok{x12 \textasciitilde{}\textasciitilde{} x13 + x14 + x15}
+\StringTok{x13 \textasciitilde{}\textasciitilde{} x14 + x15}
+\StringTok{x14 \textasciitilde{}\textasciitilde{} x15}
+\StringTok{\# Constrain residual variances to be equal over time}
+\StringTok{y1 \textasciitilde{}\textasciitilde{} e*y1}
+\StringTok{y2 \textasciitilde{}\textasciitilde{} e*y2}
+\StringTok{y3 \textasciitilde{}\textasciitilde{} e*y3}
+\StringTok{y4 \textasciitilde{}\textasciitilde{} e*y4}
+\StringTok{y5 \textasciitilde{}\textasciitilde{} e*y5}
+\StringTok{\textquotesingle{}}
+\NormalTok{fe\_sem2.fit \textless{}{-}}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe\_sem2, }
                     \DataTypeTok{data =}\NormalTok{ dfw, }
@@
 \begin{Highlighting}[]
-\KeywordTok{summary}\NormalTok{( fe_sem2.fit)}
+\KeywordTok{summary}\NormalTok{( fe\_sem2.fit)}
 \end{Highlighting}
@@
 \end{align} where \(\theta\) is just the residual of a regression in
-which the true underlying variable is regressed on all other covariates.
-In this case, we need to regress \(x_{t}\) on \(\alpha_{1}\) and
-\(\alpha_{2}\) for:
+which the underlying theoretical variable is regressed on all other
+covariates. In this case, we need to regress \(x_{t}\) on \(\alpha_{1}\)
+and \(\alpha_{2}\) for:
 \(x_{t} = \tau + \gamma_{1}\alpha_{1} + \gamma_{2}\alpha_{2} + \theta_{t}\)
@@
 Normally it is not possible to reconstruct the bias since in cases where
-we have to rely on indicators, we would not know the true underlying
-variable. Furthermore, in the case of a fixed-effects model, the
-covariates are the unobserved time-invariant characteristics. However,
-because we are working with simulated data, we have everything we need.
-Going back to the results above, we can get the residuals of \(x_{t}\)
-by either running a regression and saving the residuals, or we could
-skip a step and get them directly using the `residual maker' matrix
-\citep{Ruettenauer2020} which is
+we have to rely on indicators, we would not have observed the underlying
+theoretical variable. Furthermore, in the case of a fixed-effects model,
+the covariates are the unobserved time-invariant characteristics.
+However, because we are working with simulated data, we have everything
+we need. Going back to the results above, we can get the residuals of
+\(x_{t}\) by either running a regression and saving the residuals, or we
+could skip a step and get them directly using the `residual maker'
+matrix \citep{Ruettenauer2020} which is
 \(\bm{M} = \bm{I} - \bm{A}(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\)
@@
 \begin{Highlighting}[]
-\CommentTok{# Make the n x n identity matrix}
-\NormalTok{Id <-}\StringTok{ }\KeywordTok{diag}\NormalTok{( n)}
-
-\CommentTok{# n x 2 matrix of covariates a1 and a2}
-\NormalTok{A <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{( }\KeywordTok{c}\NormalTok{( }\KeywordTok{rep}\NormalTok{( }\DecValTok{1}\NormalTok{, n), dfw}\OperatorTok{$}\NormalTok{a1, dfw}\OperatorTok{$}\NormalTok{a2), }
+\CommentTok{\# Make the n x n identity matrix}
+\NormalTok{Id \textless{}{-}}\StringTok{ }\KeywordTok{diag}\NormalTok{( n)}
+
+\CommentTok{\# n x 2 matrix of covariates a1 and a2}
+\NormalTok{A \textless{}{-}}\StringTok{ }\KeywordTok{matrix}\NormalTok{( }\KeywordTok{c}\NormalTok{( }\KeywordTok{rep}\NormalTok{( }\DecValTok{1}\NormalTok{, n), dfw}\OperatorTok{$}\NormalTok{a1, dfw}\OperatorTok{$}\NormalTok{a2), }
              \DataTypeTok{nrow =}\NormalTok{ n, }\DataTypeTok{ncol =} \DecValTok{3}\NormalTok{)}
 
-\CommentTok{# The residual maker matrix M = I - A(A'A)^-1 A'}
-\NormalTok{M <-}\StringTok{ }\NormalTok{Id }\OperatorTok{-}\StringTok{ }\NormalTok{A }\OperatorTok{%*%}\StringTok{ }\KeywordTok{solve}\NormalTok{( }\KeywordTok{t}\NormalTok{( A) }\OperatorTok{%*%}\StringTok{ }\NormalTok{A) }\OperatorTok{%*%}\StringTok{ }\KeywordTok{t}\NormalTok{( A)}
-
-\CommentTok{# Save the residuals, t for 'theta'}
-\NormalTok{t <-}\StringTok{ }\NormalTok{M }\OperatorTok{%*%}\StringTok{ }\NormalTok{dfw}\OperatorTok{$}\NormalTok{x1}
-
-\CommentTok{# Re-run the FE model from above with the 'true' }
-\CommentTok{# independent variable for the correct estimate for beta }
-\NormalTok{fe_sem.fit <-}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe_sem, }\DataTypeTok{data =}\NormalTok{ dfw, }\DataTypeTok{estimator =} \StringTok{"ML"}\NormalTok{)}
-
-\CommentTok{# The equation for the biased beta }
-\KeywordTok{lavInspect}\NormalTok{( fe_sem.fit, }\StringTok{"list"}\NormalTok{)[ }\DecValTok{6}\NormalTok{, }\DecValTok{14}\NormalTok{]}\OperatorTok{*}
-\StringTok{  }\NormalTok{(( }\KeywordTok{var}\NormalTok{( t))}\OperatorTok{/}\NormalTok{( }\KeywordTok{var}\NormalTok{( t) }\OperatorTok{+}\StringTok{ }\KeywordTok{var}\NormalTok{( dfw}\OperatorTok{$}\NormalTok{x11 }\OperatorTok{-}\StringTok{ }\NormalTok{dfw}\OperatorTok{$}\NormalTok{x1)))}
+\CommentTok{\# The residual maker matrix M = I {-} A(A\textquotesingle{}A)\^{}{-}1 A\textquotesingle{}}
+\NormalTok{M \textless{}{-}}\StringTok{ }\NormalTok{Id }\OperatorTok{{-}}\StringTok{ }\NormalTok{A }\OperatorTok{\%*\%}\StringTok{ }\KeywordTok{solve}\NormalTok{( }\KeywordTok{t}\NormalTok{( A) }\OperatorTok{\%*\%}\StringTok{ }\NormalTok{A) }\OperatorTok{\%*\%}\StringTok{ }\KeywordTok{t}\NormalTok{( A)}
+
+\CommentTok{\# Save the residuals, t for \textquotesingle{}theta\textquotesingle{}}
+\NormalTok{t \textless{}{-}}\StringTok{ }\NormalTok{M }\OperatorTok{\%*\%}\StringTok{ }\NormalTok{dfw}\OperatorTok{$}\NormalTok{x1}
+
+\CommentTok{\# Re{-}run the FE model from above with the \textquotesingle{}true\textquotesingle{} }
+\CommentTok{\# independent variable for the correct estimate for beta }
+\NormalTok{fe\_sem.fit \textless{}{-}}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe\_sem, }\DataTypeTok{data =}\NormalTok{ dfw, }\DataTypeTok{estimator =} \StringTok{"ML"}\NormalTok{)}
+
+\CommentTok{\# The equation for the biased beta }
+\KeywordTok{lavInspect}\NormalTok{( fe\_sem.fit, }\StringTok{"list"}\NormalTok{)[ }\DecValTok{6}\NormalTok{, }\DecValTok{14}\NormalTok{]}\OperatorTok{*}
+\StringTok{  }\NormalTok{(( }\KeywordTok{var}\NormalTok{( t))}\OperatorTok{/}\NormalTok{( }\KeywordTok{var}\NormalTok{( t) }\OperatorTok{+}\StringTok{ }\KeywordTok{var}\NormalTok{( dfw}\OperatorTok{$}\NormalTok{x11 }\OperatorTok{{-}}\StringTok{ }\NormalTok{dfw}\OperatorTok{$}\NormalTok{x1)))}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{fe_sem3 <-}\StringTok{ '}
-\StringTok{# Define individual effects variable }
-\StringTok{a =~ 1*y11 + 1*y12 + 1*y13 + 1*y14 + 1*y15}
-\StringTok{# Measurement model for independent variables, xi }
-\StringTok{xi1 =~ 1*x11 + x21 + x31 }
-\StringTok{xi2 =~ 1*x12 + x22 + x32}
-\StringTok{xi3 =~ 1*x13 + x23 + x33}
-\StringTok{xi4 =~ 1*x14 + x24 + x34}
-\StringTok{xi5 =~ 1*x15 + x25 + x35}
-\StringTok{# Regressions, constrain coefficient to be equal over time}
-\StringTok{y11 ~ b*xi1}
-\StringTok{y12 ~ b*xi2 }
-\StringTok{y13 ~ b*xi3}
-\StringTok{y14 ~ b*xi4}
-\StringTok{y15 ~ b*xi5}
-\StringTok{# Allow unrestricted correlation between eta and covariates}
-\StringTok{a ~~ xi1 + xi2 + xi3 + xi4 + xi5}
-\StringTok{xi1 ~~ xi2 + xi3 + xi4 + xi5}
-\StringTok{xi2 ~~ xi3 + xi4 + xi5}
-\StringTok{xi3 ~~ xi4 + xi5}
-\StringTok{xi4 ~~ xi5}
-\StringTok{# Constrain residual variances to be equal over time}
-\StringTok{y11 ~~ e*y11}
-\StringTok{y12 ~~ e*y12}
-\StringTok{y13 ~~ e*y13}
-\StringTok{y14 ~~ e*y14}
-\StringTok{y15 ~~ e*y15}
-\StringTok{'}
-\NormalTok{fe_sem3.fit <-}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe_sem3, }
+\NormalTok{fe\_sem3 \textless{}{-}}\StringTok{ \textquotesingle{}}
+\StringTok{\# Define individual effects variable }
+\StringTok{a =\textasciitilde{} 1*y11 + 1*y12 + 1*y13 + 1*y14 + 1*y15}
+\StringTok{\# Measurement model for independent variables, xi }
+\StringTok{xi1 =\textasciitilde{} 1*x11 + x21 + x31 }
+\StringTok{xi2 =\textasciitilde{} 1*x12 + x22 + x32}
+\StringTok{xi3 =\textasciitilde{} 1*x13 + x23 + x33}
+\StringTok{xi4 =\textasciitilde{} 1*x14 + x24 + x34}
+\StringTok{xi5 =\textasciitilde{} 1*x15 + x25 + x35}
+\StringTok{\# Regressions, constrain coefficient to be equal over time}
+\StringTok{y11 \textasciitilde{} b*xi1}
+\StringTok{y12 \textasciitilde{} b*xi2 }
+\StringTok{y13 \textasciitilde{} b*xi3}
+\StringTok{y14 \textasciitilde{} b*xi4}
+\StringTok{y15 \textasciitilde{} b*xi5}
+\StringTok{\# Allow unrestricted correlation between eta and covariates}
+\StringTok{a \textasciitilde{}\textasciitilde{} xi1 + xi2 + xi3 + xi4 + xi5}
+\StringTok{xi1 \textasciitilde{}\textasciitilde{} xi2 + xi3 + xi4 + xi5}
+\StringTok{xi2 \textasciitilde{}\textasciitilde{} xi3 + xi4 + xi5}
+\StringTok{xi3 \textasciitilde{}\textasciitilde{} xi4 + xi5}
+\StringTok{xi4 \textasciitilde{}\textasciitilde{} xi5}
+\StringTok{\# Constrain residual variances to be equal over time}
+\StringTok{y11 \textasciitilde{}\textasciitilde{} e*y11}
+\StringTok{y12 \textasciitilde{}\textasciitilde{} e*y12}
+\StringTok{y13 \textasciitilde{}\textasciitilde{} e*y13}
+\StringTok{y14 \textasciitilde{}\textasciitilde{} e*y14}
+\StringTok{y15 \textasciitilde{}\textasciitilde{} e*y15}
+\StringTok{\textquotesingle{}}
+\NormalTok{fe\_sem3.fit \textless{}{-}}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe\_sem3, }
                     \DataTypeTok{data =}\NormalTok{ dfw, }
@@
 \begin{Highlighting}[]
-\KeywordTok{summary}\NormalTok{( fe_sem3.fit)}
+\KeywordTok{summary}\NormalTok{( fe\_sem3.fit)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\KeywordTok{lavInspect}\NormalTok{( fe_sem.fit, }\StringTok{"r2"}\NormalTok{)[ }\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{]}
+\KeywordTok{lavInspect}\NormalTok{( fe\_sem.fit, }\StringTok{"r2"}\NormalTok{)[ }\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{]}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\KeywordTok{lavInspect}\NormalTok{( fe_sem3.fit, }\StringTok{"r2"}\NormalTok{)[ }\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{]}
+\KeywordTok{lavInspect}\NormalTok{( fe\_sem3.fit, }\StringTok{"r2"}\NormalTok{)[ }\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{]}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{fe_sem4 <-}\StringTok{ '}
-\StringTok{# Measurement model for dependent variable, n for eta}
-\StringTok{n1 =~ 1*y11 + y21 + y31}
-\StringTok{n2 =~ 1*y12 + y22 + y32}
-\StringTok{n3 =~ 1*y13 + y23 + y33}
-\StringTok{n4 =~ 1*y14 + y24 + y34}
-\StringTok{n5 =~ 1*y15 + y25 + y35}
-\StringTok{# Define individual effects variable }
-\StringTok{a =~ 1*n1 + 1*n2 + 1*n3 + 1*n4 + 1*n5}
-\StringTok{# Measurement model for independent variables, xi }
-\StringTok{xi1 =~ 1*x11 + x21 + x31 }
-\StringTok{xi2 =~ 1*x12 + x22 + x32}
-\StringTok{xi3 =~ 1*x13 + x23 + x33}
-\StringTok{xi4 =~ 1*x14 + x24 + x34}
-\StringTok{xi5 =~ 1*x15 + x25 + x35}
-\StringTok{# Regressions, constrain coefficient to be equal over time}
-\StringTok{n1 ~ b*xi1}
-\StringTok{n2 ~ b*xi2 }
-\StringTok{n3 ~ b*xi3}
-\StringTok{n4 ~ b*xi4}
-\StringTok{n5 ~ b*xi5}
-\StringTok{# Allow unrestricted correlation between eta and covariates}
-\StringTok{a ~~ xi1 + xi2 + xi3 + xi4 + xi5}
-\StringTok{xi1 ~~ xi2 + xi3 + xi4 + xi5}
-\StringTok{xi2 ~~ xi3 + xi4 + xi5}
-\StringTok{xi3 ~~ xi4 + xi5}
-\StringTok{xi4 ~~ xi5}
-\StringTok{# Constrain residual variances to be equal over time}
-\StringTok{n1 ~~ e*n1}
-\StringTok{n2 ~~ e*n2}
-\StringTok{n3 ~~ e*n3}
-\StringTok{n4 ~~ e*n4}
-\StringTok{n5 ~~ e*n5}
-\StringTok{'}
-\NormalTok{fe_sem4.fit <-}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe_sem4, }
+\NormalTok{fe\_sem4 \textless{}{-}}\StringTok{ \textquotesingle{}}
+\StringTok{\# Measurement model for dependent variable, n for eta}
+\StringTok{n1 =\textasciitilde{} 1*y11 + y21 + y31}
+\StringTok{n2 =\textasciitilde{} 1*y12 + y22 + y32}
+\StringTok{n3 =\textasciitilde{} 1*y13 + y23 + y33}
+\StringTok{n4 =\textasciitilde{} 1*y14 + y24 + y34}
+\StringTok{n5 =\textasciitilde{} 1*y15 + y25 + y35}
+\StringTok{\# Define individual effects variable }
+\StringTok{a =\textasciitilde{} 1*n1 + 1*n2 + 1*n3 + 1*n4 + 1*n5}
+\StringTok{\# Measurement model for independent variables, xi }
+\StringTok{xi1 =\textasciitilde{} 1*x11 + x21 + x31 }
+\StringTok{xi2 =\textasciitilde{} 1*x12 + x22 + x32}
+\StringTok{xi3 =\textasciitilde{} 1*x13 + x23 + x33}
+\StringTok{xi4 =\textasciitilde{} 1*x14 + x24 + x34}
+\StringTok{xi5 =\textasciitilde{} 1*x15 + x25 + x35}
+\StringTok{\# Regressions, constrain coefficient to be equal over time}
+\StringTok{n1 \textasciitilde{} b*xi1}
+\StringTok{n2 \textasciitilde{} b*xi2 }
+\StringTok{n3 \textasciitilde{} b*xi3}
+\StringTok{n4 \textasciitilde{} b*xi4}
+\StringTok{n5 \textasciitilde{} b*xi5}
+\StringTok{\# Allow unrestricted correlation between eta and covariates}
+\StringTok{a \textasciitilde{}\textasciitilde{} xi1 + xi2 + xi3 + xi4 + xi5}
+\StringTok{xi1 \textasciitilde{}\textasciitilde{} xi2 + xi3 + xi4 + xi5}
+\StringTok{xi2 \textasciitilde{}\textasciitilde{} xi3 + xi4 + xi5}
+\StringTok{xi3 \textasciitilde{}\textasciitilde{} xi4 + xi5}
+\StringTok{xi4 \textasciitilde{}\textasciitilde{} xi5}
+\StringTok{\# Constrain residual variances to be equal over time}
+\StringTok{n1 \textasciitilde{}\textasciitilde{} e*n1}
+\StringTok{n2 \textasciitilde{}\textasciitilde{} e*n2}
+\StringTok{n3 \textasciitilde{}\textasciitilde{} e*n3}
+\StringTok{n4 \textasciitilde{}\textasciitilde{} e*n4}
+\StringTok{n5 \textasciitilde{}\textasciitilde{} e*n5}
+\StringTok{\textquotesingle{}}
+\NormalTok{fe\_sem4.fit \textless{}{-}}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe\_sem4, }
                     \DataTypeTok{data =}\NormalTok{ dfw, }
@@
 \begin{Highlighting}[]
-\KeywordTok{summary}\NormalTok{( fe_sem4.fit)}
+\KeywordTok{summary}\NormalTok{( fe\_sem4.fit)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{fe_sem5 <-}\StringTok{ '}
-\StringTok{# Measurement model for dependent variable, n for eta}
-\StringTok{n1 =~ 1*y11 + y21 + y31}
-\StringTok{n2 =~ 1*y12 + y22 + y32}
-\StringTok{n3 =~ 1*y13 + y23 + y33}
-\StringTok{n4 =~ 1*y14 + y24 + y34}
-\StringTok{n5 =~ 1*y15 + y25 + y35}
-\StringTok{# Define individual effects variable }
-\StringTok{a =~ 1*n1 + 1*n2 + 1*n3 + 1*n4 + 1*n5}
-\StringTok{# Measurement model for independent variables, xi }
-\StringTok{xi1 =~ 1*x11 + x21 + x31 }
-\StringTok{xi2 =~ 1*x12 + x22 + x32}
-\StringTok{xi3 =~ 1*x13 + x23 + x33}
-\StringTok{xi4 =~ 1*x14 + x24 + x34}
-\StringTok{xi5 =~ 1*x15 + x25 + x35}
-\StringTok{# Regressions, constrain coefficient to be equal over time}
-\StringTok{n1 ~ b*xi1 + g*a2}
-\StringTok{n2 ~ b*xi2 + g*a2}
-\StringTok{n3 ~ b*xi3 + g*a2}
-\StringTok{n4 ~ b*xi4 + g*a2}
-\StringTok{n5 ~ b*xi5 + g*a2}
-\StringTok{# Allow unrestricted correlation between eta and covariates}
-\StringTok{a ~~ xi1 + xi2 + xi3 + xi4 + xi5 + 0*a2}
-\StringTok{a2 ~~ xi1 + xi2 + xi3 + xi4 + xi5}
-\StringTok{xi1 ~~ xi2 + xi3 + xi4 + xi5}
-\StringTok{xi2 ~~ xi3 + xi4 + xi5}
-\StringTok{xi3 ~~ xi4 + xi5}
-\StringTok{xi4 ~~ xi5}
-\StringTok{# Constrain residual variances to be equal over time}
-\StringTok{n1 ~~ e*n1}
-\StringTok{n2 ~~ e*n2}
-\StringTok{n3 ~~ e*n3}
-\StringTok{n4 ~~ e*n4}
-\StringTok{n5 ~~ e*n5}
-\StringTok{'}
-\NormalTok{fe_sem5.fit <-}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe_sem5, }
+\NormalTok{fe\_sem5 \textless{}{-}}\StringTok{ \textquotesingle{}}
+\StringTok{\# Measurement model for dependent variable, n for eta}
+\StringTok{n1 =\textasciitilde{} 1*y11 + y21 + y31}
+\StringTok{n2 =\textasciitilde{} 1*y12 + y22 + y32}
+\StringTok{n3 =\textasciitilde{} 1*y13 + y23 + y33}
+\StringTok{n4 =\textasciitilde{} 1*y14 + y24 + y34}
+\StringTok{n5 =\textasciitilde{} 1*y15 + y25 + y35}
+\StringTok{\# Define individual effects variable }
+\StringTok{a =\textasciitilde{} 1*n1 + 1*n2 + 1*n3 + 1*n4 + 1*n5}
+\StringTok{\# Measurement model for independent variables, xi }
+\StringTok{xi1 =\textasciitilde{} 1*x11 + x21 + x31 }
+\StringTok{xi2 =\textasciitilde{} 1*x12 + x22 + x32}
+\StringTok{xi3 =\textasciitilde{} 1*x13 + x23 + x33}
+\StringTok{xi4 =\textasciitilde{} 1*x14 + x24 + x34}
+\StringTok{xi5 =\textasciitilde{} 1*x15 + x25 + x35}
+\StringTok{\# Regressions, constrain coefficient to be equal over time}
+\StringTok{n1 \textasciitilde{} b*xi1 + g*a2}
+\StringTok{n2 \textasciitilde{} b*xi2 + g*a2}
+\StringTok{n3 \textasciitilde{} b*xi3 + g*a2}
+\StringTok{n4 \textasciitilde{} b*xi4 + g*a2}
+\StringTok{n5 \textasciitilde{} b*xi5 + g*a2}
+\StringTok{\# Allow unrestricted correlation between eta and covariates}
+\StringTok{a \textasciitilde{}\textasciitilde{} xi1 + xi2 + xi3 + xi4 + xi5 + 0*a2}
+\StringTok{a2 \textasciitilde{}\textasciitilde{} xi1 + xi2 + xi3 + xi4 + xi5}
+\StringTok{xi1 \textasciitilde{}\textasciitilde{} xi2 + xi3 + xi4 + xi5}
+\StringTok{xi2 \textasciitilde{}\textasciitilde{} xi3 + xi4 + xi5}
+\StringTok{xi3 \textasciitilde{}\textasciitilde{} xi4 + xi5}
+\StringTok{xi4 \textasciitilde{}\textasciitilde{} xi5}
+\StringTok{\# Constrain residual variances to be equal over time}
+\StringTok{n1 \textasciitilde{}\textasciitilde{} e*n1}
+\StringTok{n2 \textasciitilde{}\textasciitilde{} e*n2}
+\StringTok{n3 \textasciitilde{}\textasciitilde{} e*n3}
+\StringTok{n4 \textasciitilde{}\textasciitilde{} e*n4}
+\StringTok{n5 \textasciitilde{}\textasciitilde{} e*n5}
+\StringTok{\textquotesingle{}}
+\NormalTok{fe\_sem5.fit \textless{}{-}}\StringTok{ }\KeywordTok{sem}\NormalTok{( }\DataTypeTok{model =}\NormalTok{ fe\_sem5, }
                     \DataTypeTok{data =}\NormalTok{ dfw, }
@@
 \begin{Highlighting}[]
-\KeywordTok{summary}\NormalTok{( fe_sem5.fit)}
+\KeywordTok{summary}\NormalTok{( fe\_sem5.fit)}
 \end{Highlighting}

### 2021-01-14 non-normality-simulation.R

@@ -1,0 +1,159 @@
+# ======================================= # 
+# FE-SEM Paper REVISION
+# Simulation: Check normality assumptions
+# Henrik Kenneth Andersen 
+# 04.01.2021
+# ======================================= # 
+
+rm(list = ls())
+
+library(lavaan)
+library(ggplot2)
+
+# Types of error distributions 
+par(mfrow = c(2, 2))
+
+hist(rnorm(n, 0, 1))
+hist(runif(n, min(rnorm(n, 0, 1)), max(rnorm(n, 0, 1))))
+hist(exp(rnorm(n, 0, 1)))
+
+# Function to simulate data and estimate FE-SEM
+sim_func <- function(dist, n) {
+  
+  # Generate the individual effects
+  a <- rnorm(n, 0, 1)  
+  
+  x1 <- 0.5*a + rnorm(n, 0, 1)
+  x2 <- 0.5*a + rnorm(n, 0, 1)
+  x3 <- 0.5*a + rnorm(n, 0, 1)
+  
+  if(dist == "norm") {
+    y1 <- -0.25*x1 + a + rnorm(n, 0, 1)
+    y2 <- -0.25*x2 + a + rnorm(n, 0, 1)
+    y3 <- -0.25*x3 + a + rnorm(n, 0, 1)
+    } else if (dist == "unif") {
+      y1 <- -0.25*x1 + a + runif(n, min(rnorm(n, 0, 1)), max(rnorm(n, 0, 1)))
+      y2 <- -0.25*x2 + a + runif(n, min(rnorm(n, 0, 1)), max(rnorm(n, 0, 1)))
+      y3 <- -0.25*x3 + a + runif(n, min(rnorm(n, 0, 1)), max(rnorm(n, 0, 1)))
+    } else if (dist == "skew") {
+      y1 <- -0.25*x1 + a + exp(rnorm(n, 0, 1))
+      y2 <- -0.25*x2 + a + exp(rnorm(n, 0, 1))
+      y3 <- -0.25*x3 + a + exp(rnorm(n, 0, 1))
+    } else {
+        stop("Choose from 'norm', 'unif' or 'skew' as a distribution for the errors.")
+      }
+  
+  df <- data.frame(x1, x2, x3, y1, y2, y3)
+  
+  mx <- '
+  alpha =~ 1*y1 + 1*y2 + 1*y3
[…]
